---
title: "medicalcoder: A Unified and Longitudinally Aware Framework for ICD-Based Comorbidity Assessment in R"

code-fold: false
echo: true
number-sections: false

format:
  docx:
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
stopifnot(packageVersion("medicalcoder") >= "0.7.0")
options(qwraps2_markup = "markdown")
loadNamespace("data.table")
loadNamespace("medicalcoder")
mimicivDT <- readRDS("mimicivDT.rds")
objs <- readRDS("objs.rds")
benchmarks <- readRDS("benchmarks.rds")
si <- readRDS("si.rds")
```

**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; James A. Feinstein, MD, MPH^2,3,4^; Margaret N. Rebull^1^, MA; Tellen D. Bennett^1,5^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA\
^2^ Adult and Child Center for Outcomes Research and Delivery Science, University of Colorado Anschutz Medical Campus, Aurora.\
^3^ Children's Hospital Colorado, Aurora.\
^4^ Department of Pediatrics, University of Colorado Anschutz Medical Campus, Aurora.\
^5^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD\
Department of Biomedical Informatics\
University of Colorado School of Medicine\
1890 N. Revere Court\
Mailstop F600\
Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} 1983 / 2000 (excluding tables)

[**Tables:**]{.underline} 2 / 2

[**Figures:**]{.underline} 2 / 3

[**Supplementary material:**]{.underline}

[**Short running title**]{.underline}: _medicalcoder_: R package for ICD codes and Comorbidities with longitudinal data

<!-- allowed 5; recomend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline} Comorbidity, International Classification of
Diseases (ICD), Software, Longitudinal Studies, Pediatrics

{{< pagebreak >}}

# Abstract {.unnumbered}

Word count: 150 / 150

**Objectives:**
Comorbidity algorithms derived from ICD codes are central to risk adjustment and
cohort characterization. However, existing software is fragmented, inconsistently
handles mixed ICD-9/10 data, and typically operates at the encounter level. We
developed _medicalcoder_ to address these limitations.

**Materials and Methods:**
_medicalcoder_ is a R package with an internal ICD database and unified
interface for applying multiple variants of the Charlson, Elixhauser, and
Pediatric Complex Chronic Conditions algorithms. It supports full and compact
ICD codes, mixed ICD versions, present-on-admission and primary diagnosis
indicators, and both encounter-level and cumulative longitudinal flagging.

**Results:**
The _medicalcoder_ package requires only R ≥ 3.5.0. No additional packages,
internet access, or external data downloads are required. Outputs are consistent
with published reference implementations. Longitudinal flagging captured comorbidities not
identified through naïve encounter-level aggregation for algorithms
incorporating present-on-admission or technology-dependent logic.

**Conclusions:**
_medicalcoder_ provides a unified, reproducible framework for ICD-based
comorbidity assessment in modern clinical datasets.


{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
Word count: 190 / 200

Researchers often use International Classification of Diseases (ICD) diagnosis
and procedure codes to identify health conditions in large medical datasets.
Comorbidity algorithms group these codes into meaningful categories, such as
heart disease or cancer, and are essential for understanding patient complexity,
comparing outcomes, and adjusting for risk. However, existing tools usually
handle only one or two algorithms, work with either ICD-9 or ICD-10 (but rarely
both together), and do not track health conditions across multiple hospital
visits. This can lead to missed or inconsistent results.

We created _medicalcoder_, an R package that solves these problems. It includes
a complete database of ICD-9 and ICD-10 codes and provides a single, easy-to-use
function to apply many comorbidity algorithms. The package can recognize both
full and compact ICD codes, use present-on-admission flags, identify primary
diagnoses, and follow patients across multiple encounters to avoid missing
conditions that appear earlier in their history.

_medicalcoder_ is fast, works with base R, has extensive automated tests, and
has been validated against reference implementations. It helps researchers apply
comorbidity algorithms more accurately and consistently, improving the quality
and reproducibility of studies using electronic health record or administrative
data.

:::

{{< pagebreak >}}

# Background and Significance

Accurate identification of clinical conditions in large healthcare datasets
depends on International Classification of Diseases (ICD) codes
[@quan2005;@quan2011;@feinstein2024pediatric;@feudtner2014pediatric;@ahrq2025;@glasheen2019].
Comorbidity algorithms built on these codes support risk adjustment, cohort
characterization, and reproducible studies of patient populations
[@quan2005;@yurkovich2015systematic;@guo2022evaluation;@li2008risk].

Existing software for flagging comorbidities typically implements one or two
algorithms, most often variants of Charlson and Elixhauser, supports either
ICD-9 or ICD-10 (but not both simultaneously),
and requires manual reconciliation of mixed-version datasets
[@r-comorbidity;@feudtner2014pediatric;@r-pccc;@pcccv3sas;@r-icdcomorbid;@r-multimorbidity].

Despite the importance of longitudinal data in modern medical research,
comorbidity software typically operates at the encounter level. While the AHRQ
[@ahrq2025] Elixhauser variants account for present-on-admission (POA)
indicators and exemptions, it does not propagate conditions across encounters
within a patient record. Conditions that arise during acute care and should not
be considered comorbidities for the present encounter are handled via POA
indicators. However, unless the relevant ICD codes are flagged as POA on
subsequent encounters, those comorbidities will not be correctly classified.

There is a need for a tool that applies multiple comorbidity algorithms through
a unified API while accounting for algorithm-specific nuances, ICD versions and
types, and longitudinal data.

We developed the R package _medicalcoder_ to fill this need.

# Objective

Develop a fully self-contained R package that includes a comprehensive internal
ICD database capable of applying multiple comorbidity algorithms to clinical
datasets, and requires no external dependencies, internet access, or external
data resources. The package should support simultaneous ICD-9 and ICD-10 coding,
accept full and compact ICD codes, incorporate present-on-admission and primary
diagnosis indicators, and enable longitudinal comorbidity assessment across
patient encounters.

# Materials and Methods

_medicalcoder_ was designed to be fully self-contained and deployable in
restricted computing environments with limited or no internet access
and tightly controled system settings and software. No external namespaces are imported; R &geq;
3.5.0 is the only required dependency, reflecting changes to R's serialization
format introduced in that version[^a]. Several packages are suggested for enhanced
performance (data.table [@r-data.table] and dplyr [@r-dplyr]) and for testing
and documentation, but are not required for installation or use.

[^a]: https://cran.r-project.org/src/base/NEWS.3#:~:text=R%20has%20new,to%20version%203.5.0

```{r}
#| label: icd-src-years
#| include: false
aggregate(cbind(known_start, known_end) ~ src + icdv + dx,
  data = medicalcoder::get_icd_codes(),
  FUN = range
  )

cms9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))
cdc9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cdc10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cms10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))

who10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "who"),
    paste(min(known_start), max(known_end), sep = " - "))
```

The internal ICD database includes ICD-9 diagnostic codes from the Centers
for Disease Control and Prevention (CDC; fiscal years `r cdc9`) and ICD-9-CM and
ICD-9-PCS codes from the Centers for Medicare and Medicaid Services (CMS; fiscal
years `r cms9`). ICD-10 diagnostic codes from the CDC (calendar years `r cdc10`)
and the World Health Organization (WHO; calendar years `r who10`) are included,
along with ICD-10-CM and ICD-10-PCS codes from CMS (fiscal years `r cms10`). The
internal database will be extended as additional code sets are published.

Charlson variants include Deyo (as denoted in @quan2005), Quan (2005; 2011), and
CDMF (2019). Elixhauser variants include the original Elixhauser (1988), AHRQ
Web (ICD-9 based as denoted in @quan2005), Quan (2005), and AHRQ ICD-10
fiscal-year implementations (2022–2026). PCCC variants include versions 2.0,
2.1, 3.0, and 3.1.

A comprehensive test suite runs as a GitHub Action and on the Comprehensive R
Archive Network (CRAN). An additional test suite runs on demand for all released
versions of R &geq; 3.5.0 without suggested packages, and for versions &geq;
4.4.0 with suggested packages[^b].

[^b]: https://github.com/dewittpe/medicalcoder/tree/main/testing

The test suite includes validation between `method = "pccc_v2.0"` and the R
package _pccc_ [@r-pccc;@feinstein2018pccc], and between `method =
elixhauser_ahrq20XX` and the published SAS code from AHRQ[^d] [@ahrq2025]. As
part of the website[^e] build, consistency between `method = charlson_quan2005`
and the SQL code provided by MIT-LCP/mimic-code/mimic-iv[@johnson2018mimic] is
verified.

[^d]: The test directory within the source medicalcoder_<version>.tar.gz
  contains these tests are are checked prior to CRAN submissions.  Detail on
  building the expected output from AHRQ via SAS is found here:
  https://github.com/dewittpe/medicalcoder/blob/main/tests/build-expected-ahrq-results/README.md.

[^e]: https://www.peteredewitt.com/medicalcoder/articles/medicalcoder-vs-mimic.html

We use the MIMIC-IV v3.1 Data
[@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000] as the foundation for
the following examples and the supplement. The MIMIC-IV v3.1 dataset is a large,
publicly available clinical database used for method development, validation,
and reproducible research in clinical informatics and outcomes research.

We compare the results of the comorbidity flagging between _medicalcoder_ and
three alternatives, the MIMIC-IV Code released by the MIT Laboratory for
Computational Physiology (MIT\_LCP) on
GitHub[^f], and the R packages _comorbidity_ [@r-comorbidity] and _pccc_
[@r-pccc].

[^f]: https://github.com/MIT-LCP/mimic-code.git

# Results

The MIMIC-IV v3.1 dataset includes
`r qwraps2::frmt(nrow(mimicivDT))` rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
hospital encounters.
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id)][V1 > 1, qwraps2::frmt(.N)]`
subjects' records contain both ICD-9 and ICD-10 codes, and
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id, hadm_id)][V1 > 1, qwraps2::frmt(.N)]`
encounters contain both ICD-9 and ICD-10 codes.

A single function, `medicalcoder::comorbidities()`, is the interface for
applying any of `r length(medicalcoder:::comorbidities_methods())` comorbidity
algorithms implemented by _medicalcoder_, see Figure 1. The input data are expected
to be in long format, one ICD code per row, with optional columns for
identification variables (e.g., patient and encounter ids), ICD version (integer
value 9 or 10), ICD type (1 for diagnostic code, 0 for procedure code), POA
indicator, and primary diagnosis indicator. The ICD codes can be full, compact,
or a combination thereof. A single patient/encounter record may contain ICD-9,
ICD-10, or both.

Several arguments are provided in paired forms (e.g., `dx` and `dx.var`, `icdv`
and `icdv.var`). These pairs represent the same conceptual information at two
levels of granularity. Arguments without the `.var` suffix specify dataset-level
invariants (e.g., all rows are ICD-10, or all rows are diagnoses), whereas
arguments with the `.var` suffix reference columns in the input data that encode
row-level variation.

This design avoids requiring users to create constant-valued columns when
dataset-level properties are already known, while still supporting heterogeneous
datasets in which code attributes vary across rows. When both forms are
supplied, the column-based form (e.g., `dx.var`) takes precedence and a warning
is issued to indicate the override.

The `id.vars` argument controls how comorbidities are summarized into
encounter-level flags. When flag.method = "current", id.vars is optional: if
omitted, the entire dataset is treated as a single encounter and a single row of
flags is returned.

For longitudinal flagging (`flag.method = "cumulative"`), `id.vars` must include
at least two variables. The leading variables define the patient and
the final variable defines within-patient ordering. Specifically, the data are
sorted by `id.vars`, and relevant ICD codes reported on encounter i are carried
forward to encounters i+1, i+2, ... within the same patient history.

The ordering variable may be a date/time (e.g., admission time) or a numeric
sequence (e.g., encounter number). In large datasets, using an integer sequence
often reduces overhead relative to sorting and joining on POSIXct timestamps.

`medicalcoder::comorbidities()` returns a data.frame with additional metadata
and class consistent with the input.  The returned data frame contains the
`id.vars` columns and an indicator column for each comorbidity. All returned
data.frames include columns `num_cmrb` for the number of comorbidities flagged
for the `id.vars` and `cmrb_flag`, an indicator column for at least one
comorbidity.  _medicalcoder_ provides a S3 `summary()` method for the returns to
generate tables of the counts and percentage of encounters flagged, Table 1.

<!--
using the quarto native tbl- for crossref is causing an error in the word doc.
hardcoding the table number as a result
-->

::: {custom-style="tablecap"}
**Table 1:** Counts and percentages of encounters within the MIMIC-IV dataset
flagged with Charlson comorbidities via `medicalcoder::comorbidities()`.
:::

```{r}
#| label: charlson-summary
#| echo: false
#| results: asis
cc <- summary(objs$medicalcoder_charlson_current)$conditions
data.table::setDT(cc)
cc[, condition := NULL]
cc[, count := qwraps2::frmt(as.integer(count))]
cc[, percent := qwraps2::frmt(percent)]
cc[, rgp := data.table::fifelse(grepl(">=", condition_description), "Total Conditions", "Condition")]
cc[, condition_description := sub(">=", "≥", condition_description)]
data.table::setnames(cc, old = c("count", "percent"), new = c("Encounters", "Percent"))

gt::gt(
  data = cc,
  rowname_col = "condition_description",
  groupname_col = "rgp"
)
```

Flagging of comorbidities at the encounter-level only assumes all relevant ICD
codes are reported on all encounters and correctly marked as POA.  This is
rarely the case. Accumulation of of encounter-level flags is insufficient to
capture comorbidities and/or the severity of disease.
For example,
some PCCC v3 conditions depend on whether technology-dependent codes co-occur (or
eventually co-occur) with non–technology-dependent codes.  Also, Elixhauser from
AHRQ post 2022 account for POA status of ICD codes.
In these cases, naïve
post hoc accumulation of encounter-level flags can miss conditions (Table 2); using
`flag.method = "cumulative"` ensures the algorithm’s longitudinal logic is
applied consistently.

Figure 2 shows the large number of additional encounters in the MIMIC-IV data
set flagged with comorbidities when using `flag.method = "cumulative"` compared
to `flag.method = "current"`.  Severity of disease is also captured in this
example.  The number of encounters with uncomplicated diabetes via Charlson and
Elixhauser decreased from the encounter-level to longitudinal flagging
approaches.  This is because _medicalcoder_ accounts for the severity of disease
and once a subject has diabetes with complications that will take precedence
over diabetes without complications.

<!--
using the quarto native tbl- for crossref is causing an error in the word doc.
hardcoding the table number as a result
-->

::: {custom-style="tablecap"}
**Table 2:** Flagging of comorbidities under PCCC v3 for MIMIC-IV subject
10728333. Cells marked with DxPr denote that the comorbidity was flagged due to
a ICD code that is not technology-dependent.  Tech denotes flagging of a
comorbidity based on the presence of at least one technology-dependent ICD
code and the presence of at least one non-technology-dependent code for
another comorbidity. DxPrTech denotes flagging due to both
technology-dependent and non-technology-dependent ICD codes.  Under the Any
Tech columns we mark the encounters where some tech dependence is identified.
For this patient, on encounter 1, a technology-dependent ICD code for a
respiratory comorbidity was found in the subject's record.  Because no
non-technology-dependent code was observed on encounter 1, under PCCC v3 this
subject has no comorbidities for encounter 1.  This persists for encounters 2
and 3 with respect to respiratory and on encounter 3 for miscellaneous as
well.  On encounter 4 a non-technology-dependent code for a metabolic
comorbidity was reported.  Under a cumulative flagging paradigm, the
miscellaneous and respiratory comorbidities are also flagged as the reported
codes from encounters 3 and 1 have been carried forward respectively.  Under
the 'current' flagging method miscellaneous is never flagged as the
technology-dependent code never occurs on the same encounter as the
non-technology-dependent code.
:::

```{r}
#| label: flagging-table
#| echo: false
#| results: asis
s10728333 <-
  merge(
    x = objs$medicalcoder_pcccv3.1_current[subject_id == 10728333],
    y = objs$medicalcoder_pcccv3.1_cumulative[subject_id == 10728333],
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_current", "_cumulative")
 )
DT0 <- s10728333[subject_id == 10728333, .SD, .SDcols = patterns("enc_seq|^(met|resp|misc|num_|any_tech)")]
DT1 <-
  DT0[
    ,
    .(enc_seq,
      metabolic__current = data.table::fcase(
        metabolic_dxpr_only_current == 1, "DxPr",
        metabolic_tech_only_current == 1, "Tech",
        metabolic_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      metabolic__cumulative = data.table::fcase(
        metabolic_dxpr_only_cumulative == 1, "DxPr",
        metabolic_tech_only_cumulative == 1, "Tech",
        metabolic_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      misc__current = data.table::fcase(
        misc_dxpr_only_current == 1, "DxPr",
        misc_tech_only_current == 1, "Tech",
        misc_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      misc__cumulative = data.table::fcase(
        misc_dxpr_only_cumulative == 1, "DxPr",
        misc_tech_only_cumulative == 1, "Tech",
        misc_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      respiratory__current = data.table::fcase(
        respiratory_dxpr_only_current == 1, "DxPr",
        respiratory_tech_only_current == 1, "Tech",
        respiratory_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      respiratory__cumulative = data.table::fcase(
        respiratory_dxpr_only_cumulative == 1, "DxPr",
        respiratory_tech_only_cumulative == 1, "Tech",
        respiratory_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      any_tech_dep_current = data.table::fifelse(any_tech_dep_current == 1, "1", ""),
      any_tech_dep_cumulative = data.table::fifelse(any_tech_dep_cumulative == 1, "1", ""),
      num_cmrb_current = as.character(num_cmrb_current),
      num_cmrb_cumulative = as.character(num_cmrb_cumulative)
    )
  ]

ftm1 <- "*"
ftm2 <- "†"
DT1[enc_seq == 1, `:=`(respiratory__current = ftm1, respiratory__cumulative = ftm1)]
DT1[enc_seq %in% 2:3,`:=`(respiratory__cumulative = ftm2)]
DT1[enc_seq == 3, `:=`(misc__current = ftm1, misc__cumulative = ftm1)]

DT1 |>
  gt::gt(caption = "") |>
  gt::cols_align(align = "center") |>
  gt::tab_spanner("Metabolic", columns = c(metabolic__current, metabolic__cumulative)) |>
  gt::tab_spanner("Miscellaneous", columns = c(misc__current, misc__cumulative)) |>
  gt::tab_spanner("Respiratory", columns = c(respiratory__current, respiratory__cumulative)) |>
  gt::tab_spanner("Any Tech", columns = c(any_tech_dep_current, any_tech_dep_cumulative)) |>
  gt::tab_spanner("Comorbidity Count", columns = c(num_cmrb_current, num_cmrb_cumulative)) |>
  gt::cols_label(
    enc_seq = "Encounter",
    metabolic__current = "Current", metabolic__cumulative = "Cumulative",
    misc__current = "Current", misc__cumulative = "Cumulative",
    respiratory__current = "Current", respiratory__cumulative = "Cumulative",
    any_tech_dep_current = "Current", any_tech_dep_cumulative = "Cumulative",
    num_cmrb_current = "Current", num_cmrb_cumulative = "Cumulative"
  ) |>
  gt::tab_footnote(footnote = "* A technology-dependent code was reported on this encounter.  Since no non-technology-dependent code was reported no comorbidities are flagged.") |>
  gt::tab_footnote(footnote = "† A technology-dependent code has been carried forward in the record.  Since no non-technology-dependent code was reported on this, or a prior, encounter, no comorbidities are flagged.") |>
  #gt::opt_table_font(size = 19)
  #gt::tab_options(table.font.size = gt::px(10))
  gt::tab_style(
    style = gt::cell_text(size = "8pt"),
    locations = gt::cells_body()
  ) |>
  gt::tab_style(
    style = gt::cell_text(size = "8pt", weight = "bold"),
    locations = gt::cells_column_labels()
  )
```


The supplement includes a detailed exploration of the flags produced by
`medicalcoder::comorbidities()` versus the code from MIMIC-IV and the R
packages _comorbidity_ [@r-comorbidity] and _pccc_ [@r-pccc].  The flagging of
Charlson comorbidities within the MIMIC-IV v3.1 dataset between _medicalcoder_
and MIMIC-IV Code are the same. Between _medicalcoder_ and _comorbidity_ we
found the same for Elixhauser comorbidities.  For Charlson comorbidities we
found that the _comorbidity_ package failed to flag malignancies related to
ICD-10-CM codes C4A.x.  PCCC v2 flagging between _medicalcoder_ and _pccc_
showed several technology-dependent and transplant errors in the mappings within
_pccc_.

In addition to _medicalcoder_ showing consistent, or better, flagging of
comorbidities compared to other utilities, _medicalcoder_ does so with less
computational expense, Figure 3.  The package GitHub has details on benchmarking
by input data class and size.

In addition to the comorbidity interface, _medicalcoder_ provides several ICD
utility functions designed to support data preparation and validation workflows.
The function `medicalcoder::get_icd_codes()` exposes the internal ICD database
and allows users to retrieve diagnostic or procedure codes by version, type, and
code range. The function `medicalcoder::is_icd()` provides fast validation of
candidate ICD codes, supporting both full and compact representations and mixed
ICD-9/ICD-10 datasets. These utilities enable users to inspect and verify ICD
content prior to comorbidity classification and to construct reproducible code
filters directly from the same internal database used by the comorbidity
algorithms.

# Discussion

_medicalcoder_ was initially developed to support implementation of Pediatric
Complex Chronic Conditions (PCCC) versions 3 [@feinstein2024pediatric] and 2
[@feudtner2014pediatric]. During this work, it became clear that accurate and
reproducible PCCC classification required access to a comprehensive,
version-aware ICD database, which was subsequently developed and incorporated
into the package. Designing a general framework to support multiple PCCC
versions naturally enabled extension to additional comorbidity algorithms,
including Charlson and Elixhauser.

A primary design requirement was that the package be fully self-contained and
deployable in restricted computing environments, including systems with limited
internet access and/or tightly controlled software installations. Relying only
on base R ensures that _medicalcoder_ can be installed and used wherever R ≥ 3.5.0
is available.  Additionally, R versions and namespace dependency trees are
common barriers to reproducible research.  _medicalcoder_, with no additional
namespace requirements, avoids dependency version conflicts that often undermine
reproducibility, addressing a common but under-discussed barrier to reproducible
clinical research.

Beyond deployment considerations, _medicalcoder_ addresses a substantive
methodological gap in existing comorbidity software: the lack of support for
longitudinal comorbidity assessment. Most available implementations operate
strictly at the encounter level, implicitly assuming that comorbid conditions
are either repeatedly coded or clinically irrelevant once absent from the
current record. As demonstrated in Table 2, this assumption can lead to
systematic under-ascertainment of comorbidities, particularly for conditions
governed by POA rules or technology-dependent classification criteria. By
explicitly tracking diagnostic history across encounters, _medicalcoder_ enables
comorbidity flagging that more closely reflects clinical reality.

The unified interface provided by _medicalcoder_ further reduces analytic
complexity. Mixed ICD-9 and ICD-10 datasets, full and compact codes, diagnostic
and procedure codes, and algorithm-specific nuances are handled within a single
function call. This contrasts with existing workflows that require dataset
partitioning, multiple method calls, and post hoc reconciliation, each of which
introduces opportunities for error and irreproducibility.

Although the primary focus of _medicalcoder_ is comorbidity classification, the
included ICD utility functions also address a common practical challenge in
working with administrative and electronic health record data: verifying and
manipulating large collections of ICD codes. By exposing the internal ICD
database through functions such as get_icd_codes() and providing validation via
is_icd(), _medicalcoder_ allows analysts to perform exploratory and preprocessing
tasks using the same reference data employed by the comorbidity algorithms. This
reduces reliance on ad hoc code lists and external reference tables, improving
transparency and reproducibility when defining cohorts or filtering diagnosis
and procedure codes in large clinical datasets.

Extensive testing against published SAS, SQL, and R reference implementations
demonstrates that _medicalcoder_ produces results consistent with established
algorithms while offering additional flexibility. Automated testing of the
internal and user-facing functions (97% code lines covered as of _medicalcoder_
v0.8.0) ensures consistent and reliable behavior of the package.

To our knowledge, no currently available R package jointly supports multiple
comorbidity algorithms, mixed ICD versions, present-on-admission logic, and
cumulative longitudinal flagging within a single, reproducible framework.
_medicalcoder_ has these capabilities and enables researchers to accurately
characterize patient complexity and to systematically evaluate how different
comorbidity definitions influence analytic results.

## Conclusions

_medicalcoder_ is a fully self-contained R package with a detailed and robust
internal ICD database that can efficiently apply many comorbidity algorithms to
a dataset while accounting for ICD-version, ICD-type, POA, primary diagnosis,
and full and compact ICD codes. Flagging codes can be done for individual
patient/encounters, or longitudinally within a patient record.

{{< pagebreak >}}

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline}
The _medicalcoder_ R package is available from the Comprehensive R Archive Network
(CRAN) at https://cran.r-project.org/package=medicalcoder. Developmental
versions, additional testing suites, benchmarking, and examples are on GitHub at
https://github.com/dewittpe/medicalcoder/.

The MIMIC-IV v3.1 data is available from PhysioNet at
https://physionet.org/content/mimiciv/3.1/.

This manuscript was written using Quarto version
`r system("quarto --version", intern = TRUE)`
(https://quarto.org/) and `r R.version.string`. All R code, materials,
and dependencies can be found at
https://github.com/dewittpe/medicalcoder-jamiaopen/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
_medicalcoder_ R package and drafted the manuscript. SR provided critical code
review and testing.  SR, MR, and TDB all contributed to the conception, design,
and interpretation and provided important feedback. TDB provided funding. All
authors critically reviewed the manuscript for important intellectual content
and approved of the final version of the manuscript.

[**Acknowledgments:**]{.underline} We thank Vincent Rubinetti for his assistance
in designing the _medicalcoder_ logo.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

**Figure 1:** Interface for the `medicalcoder::comorbidities()` function. The
function requires a data object containing one ICD code per row, the column
containing ICD codes, and the comorbidity algorithm specification (e.g.,
Charlson or Elixhauser variants). Optional arguments allow users to specify
encounter identifiers, ICD version, diagnosis type, present-on-admission
indicators, patient age, and primary diagnosis flags. Defaults control how
comorbidities are identified, including encounter-level (`flag.method =
"current"`) or longitudinal (`flag.method = "cumulative"`) evaluation and how ICD
codes are matched.

**Figure 1 ALT TEXT:**
Diagram showing the argument structure of the R function
`medicalcoder::comorbidities()`. The figure lists three required inputs: a data
frame containing one ICD code per row (data), the column name containing ICD
codes (`icd.codes`), and the comorbidity method specification (`method`, such as
"charlson_quan2005"). Additional optional parameters allow users to specify
encounter identifiers, ICD version (9 or 10), diagnosis versus procedure
indicators, present-on- admission status, patient age (for Charlson algorithms),
and primary diagnosis flags. Default arguments include `flag.method = "current"`
for encounter-level detection, `full.codes = TRUE`, `compact.codes = TRUE`, and
`subconditions = FALSE`.

**Figure 2:**

**Figure 2 ALT TEXT:**

**Figure 3:** The time required to flag comorbidities in the MIMIC-IV v3.1
dataset (`r qwraps2::frmt(nrow(mimicivDT))` rows,
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects,
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
encounters) by utility (panels) and flagging method.  The MIMIC-IV code time is based on
running RSQLite code for an in-memory database.  R packages _comorbidity_ and
_pccc_ require multiple calls and aggregations to handle the mixed use of ICD-9
and ICD-10 within the MIMIC-IV dataset.  Applying each of the
comorbidity algorithms and flagging methods within _medicalcoder_ requires only
modifying the `method` and `flag.method` argument to the
`medicalcoder::comorbidities()` call.  Some panels are sparse, illustrating the
limited comorbidities method provided by the utility.  The times were recorded
when building the supplement in `r si$platform$version` on `r si$platform$os`
with a `r si$platform$CPU` and `r si$platform$RAM` RAM.

**Figure 3 ALT TEXT:**

Four-panel grouped bar chart showing time in seconds to flag comorbidities in
the MIMIC-IV v3.1 dataset. Panels compare four utilities: _comorbidity_,
_medicalcoder_, MIMIC-IV code, and _pccc_. The x-axis lists comorbidity
algorithms: Charlson, Elixhauser, PCCC version 2.0, and PCCC version 3.1. The
y-axis is time in seconds from 0 to about 160. Blue bars indicate cumulative
flagging and maroon bars indicate current flagging. Only _medicalcoder_ has
cumulative bars as it is the only utility natively supporting cumulative
flagging.  _medicalcoder_ is the fastest in every comparison shown, with times
of about 3 to 14 seconds across all four algorithms and flag methods.
_comorbidity_ is slower, at about 54 seconds for current Charlson, 76 seconds
for current Elixhauser. The MIMIC-IV code panel shows a single current Charlson
result at about 106 seconds. The _pccc_ panel shows only PCCC version 2.0, with
about 152 seconds for current flagging.
