---
title: "medicalcoder: A Unified and Longitudinally Aware Framework for ICD-Based Comorbidity Assessment in R"

code-fold: false
echo: true
number-sections: false

format:
  docx:
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
#reference-doc: template.docx
stopifnot(packageVersion("medicalcoder") >= "0.7.0")
options(qwraps2_markup = "markdown")
loadNamespace("data.table")
loadNamespace("medicalcoder")
mimicivDT <- readRDS("mimicivDT.rds")
objs <- readRDS("objs.rds")
```

**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; Margaret N. Rebull^1^, MA; Tellen D. Bennett^1,2^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

^2^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD

Department of Biomedical Informatics

University of Colorado School of Medicine

1890 N. Revere Court

Mailstop F600

Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} XXXXX / 2000 (including R chunks, excluding tables)

[**Tables:**]{.underline} 2 / 2

[**Figures:**]{.underline} 1 / 3

[**Supplementary material:**]{.underline}

[**Short running title**]{.underline}: medicalcoder: R package for ICD codes and Comorbidities with longitudinal data

<!-- allowed 5; recomend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline} Comorbidity, International Classification of
Diseases (ICD), Software, Longitudinal Studies, Pediatrics

{{< pagebreak >}}

# Abstract {.unnumbered}

Word count: 145 / 150

**Objectives:**
Comorbidity algorithms derived from International Classification of Diseases
(ICD) codes are central to risk adjustment and cohort characterization, yet
existing software is fragmented, inconsistently handles mixed ICD-9 and ICD-10
data, and typically operates at the encounter level. We developed _medicalcoder_
to address these limitations.

**Materials and Methods:**
_medicalcoder_ is a self-contained R package with an internal ICD database and
unified interface for applying multiple variants of the Charlson, Elixhauser,
and Pediatric Complex Chronic Conditions algorithms. It supports full and
compact ICD codes, mixed ICD versions, present-on-admission and primary
diagnosis indicators, and both encounter-level and cumulative longitudinal
flagging.

**Results:**
The package requires only R ≥ 3.5.0. Outputs are consistent with published
reference implementations. Outputs are consistent with published reference
implementations. Longitudinal flagging increases sensitivity compared with
aggregation of encounter-level flags.

**Conclusions:**
_medicalcoder_ provides a unified, reproducible framework for ICD-based
comorbidity assessment in modern clinical datasets.


{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
Word count: 187 / 200

Researchers often use International Classification of Diseases (ICD) diagnosis
and procedure codes to identify health conditions in large medical datasets.
Comorbidity algorithms group these codes into meaningful categories, such as
heart disease or cancer, and are essential for understanding patient complexity,
comparing outcomes, and adjusting for risk. However, existing tools usually
handle only one or two algorithms, work with either ICD-9 or ICD-10 (but not
both together), and do not track health conditions across multiple hospital
visits. This can lead to missed or inconsistent results.

We created _medicalcoder_, an R package that solves these problems. It includes
a complete database of ICD-9 and ICD-10 codes and provides a single, easy-to-use
function to apply many comorbidity algorithms. The package can recognize both
full and compact ICD codes, use present-on-admission flags, identify primary
diagnoses, and follow patients across multiple encounters to avoid missing
conditions that appear earlier in their history.

_medicalcoder_ is fast, works with base R, and has been thoroughly tested against
published methods. It helps researchers apply comorbidity algorithms more
accurately and consistently, improving the quality and reproducibility of
studies using electronic health record or administrative data.

:::

{{< pagebreak >}}

# Background and Significance

Accurate identification of clinical conditions in large healthcare datasets
depends on International Classification of Diseases (ICD) codes
[@quan2005;@quan2011;@feinstein2024pediatric;@feudtner2014pediatric;@ahrq2025;@glasheen2019].
Comorbidity algorithms built on these codes support risk adjustment, cohort
characterization, and reproducible study of patient populations
[@quan2005;@yurkovich2015systematic;@guo2022evaluation;@li2008risk].

Existing software for flagging comorbidities typically implement one or two comorbidity
algorithms, most often variants of Charlson and Elixhauser, and support either
ICD-9 or ICD-10, but not both, at least no simultaneously
[@r-comorbidity;@feudtner2014pediatric;@r-pccc;@pcccv3sas;@r-icdcomorbid;@r-multimorbidity].
This leaves users to reconcile mixed ICD-version datasets manually. Input
requirements vary across utilities, making it difficult to apply multiple
algorithms consistently.

Despite the importance of longitudinal data in modern EHR research, comorbidity
software typically operates at the encounter level. While the AHRQ [@ahrq2025]
Elixhauser variants account for present-on-admission (POA) indicators and
exemptions, it does not propagate conditions across encounters within a patient
record. Conditions that arise during acute care and should not be considered
comorbidities for the present encounter are handled via POA indicators. However,
unless the relevant ICD codes are flagged as POA on subsequent encounters, those
comorbidities will not be flagged.

There is a need for a tool that applies multiple comorbidity algorithms through
a unified API while accounting for algorithm-specific nuances, ICD versions and
types, and longitudinal data.

We present the R package _medicalcoder_ to fill this need.

# Objective

Develop a fully self-contained R package with a comprehensive internal ICD
database capable of applying multiple comorbidity algorithms to clinical
datasets. The package should support simultaneous ICD-9 and ICD-10 coding,
accept full and compact ICD codes, incorporate present-on-admission and primary
diagnosis indicators, and enable longitudinal comorbidity assessment across
patient encounters.

# Materials and Methods

_medicalcoder_ was designed to be fully self-contained and deployable in
restricted computing environments. No external namespaces are imported; R &geq;
3.5.0 is the only required dependency, reflecting changes in data serialization
introduced in that version[^a]. Several packages are suggested for enhanced
performance (data.table [@r-data.table] and dplyr [@r-dplyr]) and for testing
and documentation, but are not required for installation or use.

[^a]: https://cran.r-project.org/src/base/NEWS.3#:~:text=R%20has%20new,to%20version%203.5.0

```{r}
#| label: icd-src-years
#| include: false
aggregate(cbind(known_start, known_end) ~ src + icdv + dx,
  data = medicalcoder::get_icd_codes(),
  FUN = range
  )

cms9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))
cdc9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cdc10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cms10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))

who10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "who"),
    paste(min(known_start), max(known_end), sep = " - "))
```

The internal ICD database consists of ICD-9 diagnostic codes from the Centers
for Disease Control and Prevention (CDC; fiscal years `r cdc9`) and ICD-9-CM and
ICD-9-PCS codes from the Centers for Medicare and Medicaid Services (CMS; fiscal
years `r cms9`). ICD-10 diagnostic codes from the CDC (calendar years `r cdc10`)
and the World Health Organization (WHO; calendar years `r who10`) are included,
along with ICD-10-CM and ICD-10-PCS codes from CMS (fiscal years `r cms10`). The
internal database will be extended as additional code sets are published.


Charlson variants include Deyo (1992), Quan (2005; 2011), and CDMF (2019).
Elixhauser variants include the original Elixhauser (1988), Quan (2005), and AHRQ ICD-10 fiscal-year implementations (2022–2026).
PCCC variants include versions 2.0, 2.1, 3.0, and 3.1.

Development used R version 4.5.2. A comprehensive test suite runs as a GitHub
Action on R oldrelease (4.4.3), release (4.5.2), and devel (what will become
4.6.0). An additional test suite runs on demand for all released versions of R
&geq; 3.5.0 without suggested packages, and for versions &geq; 4.4.0 with
suggested packages[^b].

[^b]: https://github.com/dewittpe/medicalcoder/tree/main/testing

Within the test suite there are tests for ensuring consistency between
`method = "pccc_v2.0"` and the R package _pccc_ [@r-pccc;@feinstein2018pccc],
and between `method = elixhauser_ahrq20XX` and the published SAS code from AHRQ[^d] [@ahrq2025].
As part of the website[^e] build, consistency between
`method = charlson_quan2005` and the SQL code provided by
MIT-LCP/mimic-code/mimic-iv[@johnson2018mimic] is verified.

[^d]: The test directory within the source medicalcoder_<version>.tar.gz
  contains these tests are are checked prior to CRAN submissions.  Detail on
  building the expected output from AHRQ via SAS is found here:
  https://github.com/dewittpe/medicalcoder/blob/main/tests/build-expected-ahrq-results/README.md.

[^e]: https://www.peteredewitt.com/medicalcoder/articles/medicalcoder-vs-mimic.html

We use the MIMIC-IV v3.1 Data
[@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000] as the foundation for
these examples. The MIMIC-IV v3.1 dataset is a large, publicly available
clinical database containing de-identified health data from patients admitted to
intensive care units at Beth Israel Deaconess Medical Center. It includes
detailed information on diagnoses, procedures, laboratory results, medications,
and hospital encounters, with ICD-9-CM and ICD-10-CM/PCS codes available for
billing and comorbidity research. MIMIC-IV is widely used for method
development, validation, and reproducible research in clinical informatics and
outcomes research.

We compare the results of the comorbidity flagging between _medicalcoder_ and
three alternatives, the MIMIC-IV Code released by the MIT Laboratory for
Computational Physilogy (MIT\_LCP) on
GitHub[^f], and the R packages _comorbidity_ [@r-comorbidity] and _pccc_
[@r-pccc].

[^f]: https://github.com/MIT-LCP/mimic-code.git

# Results

Within the MIMIC-IV v3.1 dataset there are
`r qwraps2::frmt(nrow(mimicivDT))` total rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
hospital encounters.
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id)][V1 > 1, qwraps2::frmt(.N)]`
subjects records contain both ICD-9 and ICD-10 codes, and
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id, hadm_id)][V1 > 1, qwraps2::frmt(.N)]`
encounters contain both ICD versions.

A single function, `medicalcoder::comorbidities()`, is the interface for
applying any of `r length(medicalcoder:::comorbidities_methods())` comorbidity
algorithms implemented by _medicalcoder_. The input data are expected
to be in long format, one ICD code per row, with optional columns for
identification variables (e.g., patient and encounter ids), ICD version (integer
value 9 or 10), ICD type (1 for diagnostic code, 0 for procedure code), POA
indicator, and primary diagnosis indicator. The ICD codes can be full, compact,
or a combination thereof. A single patient/encounter record may contain ICD-9,
ICD-10, or both.

```{r}
#| label: examle-call
#| eval: false
medicalcoder::comorbidities(
 # Required inputs:
  data      = ,   # object inheriting data.frame; one ICD code per row
  icd.codes = ,   # character; name of column in data containing ICD codes
  method    = ,   # <family>_<variant>, e.g., "charlson_quan2005"

  # Optional inputs controlling grouping and code attributes:
  id.vars   = NULL,  # character; encounter identifiers (see text below)

  icdv.var  = NULL,  # character; column name for ICD version (9 or 10)
  icdv      = NULL,  # integer; constant ICD version for all rows

  dx.var    = NULL,  # character; column name indicating diagnosis (1), procedure (0)
  dx        = NULL,  # integer; constant dx indicator for all rows

  poa.var   = NULL,  # character; column name for present-on-admission indicator
  poa       = NULL,  # integer; constant POA indicator for all rows

  age.var   = NULL,  # character; column name for age in years (Charlson only)
  age       = NULL,  # numeric; constant age in years (Charlson only)

  primarydx.var = NULL, # character; column name indicating primary diagnosis
  primarydx     = NULL, # integer; constant primary dx indicator

  # Defaults:
  flag.method   = "current", # "current" (encounter-level) or
                             # "cumulative" (longitudinal)
  full.codes    = TRUE, # reference data[[icd.codes]] against full ICD codes, e.g., C4A.111
  compact.codes = TRUE, # reference data[[icd.codes]] against compact ICD codes, e.g., C4A111
  subconditions = FALSE # PCCC only: return subcondition flags when available
)
```

Several arguments are provided in paired forms (e.g., `dx` and `dx.var`, `icdv`
and `icdv.var`). These pairs represent the same conceptual information at two
levels of granularity. Arguments without the `.var` suffix specify dataset-level
invariants (e.g., all rows are ICD-10, or all rows are diagnoses), whereas
arguments with the `.var` suffix reference columns in the input data that encode
row-level variation.

This design avoids requiring users to create constant-valued columns when
dataset-level properties are already known, while still supporting heterogeneous
datasets in which code attributes vary across rows. When both forms are
supplied, the column-based form (e.g., `dx.var`) takes precedence and a warning
is issued to indicate the override.

The `id.vars` argument controls how comorbidities are summarized into encounter-
level flags. When flag.method = "current", id.vars is optional: if omitted,
the entire dataset is treated as a single encounter and a single row of flags is
returned.

For longitudinal flagging (`flag.method = "cumulative"`), `id.vars` must include
at least two variables. The leading variables define the patient (or unit) and
the final variable defines within-patient ordering. Specifically, the data are
sorted by `id.vars`, and relevent ICD codes reported on encounter i are carried
forward to encounters i+1, i+2, ... within the same patient history.

The ordering variable may be a date/time (e.g., admission time) or a numeric
sequence (e.g., encounter number). In large datasets, using an integer sequence
often reduces overhead relative to sorting and joining on POSIXct timestamps.

`medicalcoder::comorbidities()` returns a data.frame with addition metadata and
class consistent with the input.  The data.frame has column(s) for the `id.vars`
and an indicator column each of the relevant comorbidities.  All returned
data.frames include columns `num_cmrb` for the number of comorbidities flagged
for the `id.vars` and `cmrb_flag`, an indicator column for at least one
comorbidity.  _medicalcoder_ provides a S3 `summary` method for the returns to
generate tables of the counts and percentage of encounters flagged, Table 1.

<!--
using the quarto native tbl- for crossref is causing an error in the word doc.
hardcoding the table number as a result
-->

::: {custom-style="tablecap"}
**Table 1:** Counts and percentages of encounters within the MIMIC-IV dataset
flagged with Charlson comorbidities via `medicalcoder::comorbidities()`.
:::

```{r}
#| label: charlson-summary
#| echo: false
#| results: asis
cc <- summary(objs$medicalcoder_charlson_current)$conditions
data.table::setDT(cc)
cc[, condition := NULL]
cc[, count := qwraps2::frmt(as.integer(count))]
cc[, percent := qwraps2::frmt(percent)]
cc[, rgp := data.table::fifelse(grepl(">=", condition_description), "Total Conditions", "Condition")]
cc[, condition_description := sub(">=", "≥", condition_description)]
data.table::setnames(cc, old = c("count", "percent"), new = c("Encounters", "Percent"))

gt::gt(
  data = cc,
  rowname_col = "condition_description",
  groupname_col = "rgp"
)
```

Finally, note that longitudinal flagging is algorithm-dependent. For example,
some PCCC v3 conditions depend on whether technology-dependent codes co-occur (or
eventually co-occur) with non–technology-dependent codes.  Also, Elixhauser from
AHRQ post 2022 account for POA status of ICD codes.
In these cases, naïve
post hoc accumulation of encounter-level flags can miss conditions (Table 2); using
`flag.method = "cumulative"` ensures the algorithm’s longitudinal logic is
applied consistently.

<!--
using the quarto native tbl- for crossref is causing an error in the word doc.
hardcoding the table number as a result
-->

::: {custom-style="tablecap"}
**Table 2:** Flagging of comorbidities under PCCC v3 for MIMIC-IV subject
10728333. Cells marked with DxPr denote that the comorbidity was flagged due to
a ICD code that is not technology-dependent.  Tech denotes flagging of a
comorbidity base on the presence of at least one technology-dependent ICD
code and the presence of at least one non-technology-dependent code for
another comorbidity. DxPrTech denotes flagging due to both
technology-dependent and non-technology-dependent ICD codes.  Under the Any
Tech columns we mark the encounters where some tech dependence is identified.
For this patient, on encounter 1, a technology-dependent ICD code for a
respiratory comorbidity was found in the subject's record.  Because no
non-technology-dependent code was observed on encounter 1, under PCCC v3 this
subject has no comorbidities for encounter 1.  This persists for encounters 2
and 3 with respect to respiratory and on encounter 3 for miscellaneous as
well.  On encounter 4 a non-technology-dependent code for a metabolic
comorbidity was reported.  Under a cumulative flagging paradigm the
miscellaneous and respiratory comorbidities are also flagged as the reported
codes from encounters 3 and 1 have been carried forward respectively.  Under
the 'current' flagging method miscellaneous is never flagged as the
technology-dependent code never occurs on the same encounter as the
non-technology-dependent code.
:::

```{r}
#| label: flagging-table
#| echo: false
#| results: asis
s10728333 <-
  merge(
    x = objs$medicalcoder_pcccv3.1_current[subject_id == 10728333],
    y = objs$medicalcoder_pcccv3.1_cumulative[subject_id == 10728333],
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_current", "_cumulative")
 )
DT0 <- s10728333[subject_id == 10728333, .SD, .SDcols = patterns("enc_seq|^(met|resp|misc|num_|any_tech)")]
DT1 <-
  DT0[
    ,
    .(enc_seq,
      metabolic__current = data.table::fcase(
        metabolic_dxpr_only_current == 1, "DxPr",
        metabolic_tech_only_current == 1, "Tech",
        metabolic_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      metabolic__cumulative = data.table::fcase(
        metabolic_dxpr_only_cumulative == 1, "DxPr",
        metabolic_tech_only_cumulative == 1, "Tech",
        metabolic_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      misc__current = data.table::fcase(
        misc_dxpr_only_current == 1, "DxPr",
        misc_tech_only_current == 1, "Tech",
        misc_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      misc__cumulative = data.table::fcase(
        misc_dxpr_only_cumulative == 1, "DxPr",
        misc_tech_only_cumulative == 1, "Tech",
        misc_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      respiratory__current = data.table::fcase(
        respiratory_dxpr_only_current == 1, "DxPr",
        respiratory_tech_only_current == 1, "Tech",
        respiratory_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      respiratory__cumulative = data.table::fcase(
        respiratory_dxpr_only_cumulative == 1, "DxPr",
        respiratory_tech_only_cumulative == 1, "Tech",
        respiratory_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      any_tech_dep_current = data.table::fifelse(any_tech_dep_current == 1, "1", ""),
      any_tech_dep_cumulative = data.table::fifelse(any_tech_dep_cumulative == 1, "1", ""),
      num_cmrb_current = as.character(num_cmrb_current),
      num_cmrb_cumulative = as.character(num_cmrb_cumulative)
    )
  ]

ftm1 <- "*"
ftm2 <- "†"
DT1[enc_seq == 1, `:=`(respiratory__current = ftm1, respiratory__cumulative = ftm1)]
DT1[enc_seq %in% 2:3,`:=`(respiratory__cumulative = ftm2)]
DT1[enc_seq == 3, `:=`(misc__current = ftm1, misc__cumulative = ftm1)]

DT1 |>
  gt::gt(caption = "") |>
  gt::cols_align(align = "center") |>
  gt::tab_spanner("Metabolic", columns = c(metabolic__current, metabolic__cumulative)) |>
  gt::tab_spanner("Miscellaneous", columns = c(misc__current, misc__cumulative)) |>
  gt::tab_spanner("Respiratory", columns = c(respiratory__current, respiratory__cumulative)) |>
  gt::tab_spanner("Any Tech", columns = c(any_tech_dep_current, any_tech_dep_cumulative)) |>
  gt::tab_spanner("Comorbidity Count", columns = c(num_cmrb_current, num_cmrb_cumulative)) |>
  gt::cols_label(
    enc_seq = "Encounter",
    metabolic__current = "Current", metabolic__cumulative = "Cumulative",
    misc__current = "Current", misc__cumulative = "Cumulative",
    respiratory__current = "Current", respiratory__cumulative = "Cumulative",
    any_tech_dep_current = "Current", any_tech_dep_cumulative = "Cumulative",
    num_cmrb_current = "Current", num_cmrb_cumulative = "Cumulative"
  ) |>
  gt::tab_footnote(footnote = "* A technology-dependent code was reported on this encounter.  Since no non-technology-dependent was reported no comorbidities are flagged.") |>
  gt::tab_footnote(footnote = "† A technology-dependent code has been carried forward in the record.  Since no non-technology-dependent was reported on this, or a prior, enconter, no comorbidities are flagged.")
```

In the supplement is a detailed exploration of the flags produced by
`medicalcoder::comorbidities()` versus the code from MIMIV-IV, and the R
packages _comorbidity_ [@r-comorbidity] and _pccc_ [@r-pccc].  The flagging of
Charlson comorbidities within the MIMIC-IV v3.1 dataset between _medicalcoder_
and MIMIV-IV Code are the same. Between _medicalcoder_ and _comorbidity_ we
found the same for Elixhauser comorbidities.  For Charlson comorbidities we
found that the _comorbidity_ package failed to flag malignancies related to
ICD-10-CM codes C4A.x.  PCCC v2 flagging between _medicalcoder_ and _pccc_
showed several technology-dependent and transplant errors in the mappings within
_pccc_.

In addition to _medicalcoder_ showing greater sensitivity, _medicalcoder_
demonstrated substantial performance advantages in large-scale testing (Figure 1).

```{r}
#| label: figure1
#| include: false
ttc <-
  lapply(objs, attr, "tictoc") |>
  lapply(as.numeric) |>
  data.table::as.data.table() |>
  data.table::melt(measure.vars = names(objs), variable.factor = FALSE)
ttc[, c("Tool", "Algorithm", "Flag Method") := data.table::tstrsplit(variable, "_")]
ttc[, variable := NULL]
g <-
  ggplot2::ggplot(ttc) +
  ggplot2::aes(x = Algorithm, y = value, fill = `Flag Method`) +
  ggplot2::geom_col(
    position = ggplot2::position_dodge2(width = 0.8, preserve = "single"),
    width = 0.7,
    color = "white",
    linewidth = 0.3
    ) +
  ggplot2::facet_wrap(~ Tool, nrow = 1, strip.position = "top") +
  ggplot2::scale_fill_manual(
    name = "Flag Method",
    values = c("current" = "#6F263D", "cumulative" = "#4F8FCB")
  ) +
  ggplot2::scale_y_continuous(breaks = seq(0, max(ttc$value), by = 15)) +
  ggplot2::labs(x = "Comorbidity Algorithm", y = "Time (seconds)") +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    legend.position = "bottom",
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
  )
ggplot2::ggsave(plot = g, filename = "figure1.png", width = 6, height = 4)
ggplot2::ggsave(plot = g, filename = "figure1.svg", width = 6, height = 4)
```

# Discussion

_medicalcoder_ was initially developed to support implementation of Pediatric
Complex Chronic Conditions (PCCC) versions 3 [@feinstein2024pediatric] and 2
[@feudtner2014pediatric]. During this work, it became clear that accurate and
reproducible PCCC classification required access to a comprehensive,
version-aware ICD database, which was subsequently developed and incorporated
into the package. Designing a general framework to support multiple PCCC
versions naturally enabled extension to additional comorbidity algorithms,
including Charlson and Elixhauser.

A primary design requirement was that the package be fully self-contained and
deployable in restricted computing environments, including systems with limited
internet access or tightly controlled software installations. Relying only on
base R ensures that _medicalcoder_ can be installed and used wherever R &geq;
3.5.0 is available, addressing a common but under-discussed barrier to
reproducible clinical research.

Beyond deployment considerations, _medicalcoder_ addresses a substantive
methodological gap in existing comorbidity software: the lack of support for
longitudinal comorbidity assessment. Most available implementations operate
strictly at the encounter level, implicitly assuming that comorbid conditions
are either repeatedly coded or clinically irrelevant once absent from the
current record. As demonstrated in Figures 1 and 2, this assumption can lead to
systematic under-ascertainment of comorbidities, particularly for conditions
governed by POA rules or technology-dependent classification criteria. By
explicitly tracking diagnostic history across encounters, _medicalcoder_ enables
comorbidity flagging that more closely reflects clinical reality.

The unified interface provided by _medicalcoder_ further reduces analytic
complexity. Mixed ICD-9 and ICD-10 datasets, full and compact codes, diagnostic
and procedure codes, and algorithm-specific nuances are handled within a single
function call. This contrasts with existing workflows that require dataset
partitioning, multiple method calls, and post hoc reconciliation, each of which
introduces opportunities for error and irreproducibility.

Extensive testing against published SAS, SQL, and R reference implementations
demonstrates that _medicalcoder_ produces results consistent with established
algorithms while offering additional flexibility.

To our knowledge, _medicalcoder_ is the only available R package that jointly
supports multiple comorbidity algorithms, mixed ICD versions,
present-on-admission logic, and cumulative longitudinal flagging within a
single, reproducible framework. These capabilities enable researchers to
accurately characterize patient complexity and to systematically evaluate how
different comorbidity definitions influence analytic results.

## Conclusions

_medicalcoder_ is a fully self-contained R package with a detailed and robust
internal ICD database that can efficiently apply many comorbidity algorithms to
a dataset while accounting for ICD-version, ICD-type, POA, primary diagnosis,
and full and compact ICD codes. Flagging codes can be done for individual
patient/encounters, or longitudinally within a patient record.

{{< pagebreak >}}

[**Funding Statement:**]{.underline}

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline}
The _medicalcoder_ R package is available from the Comprehensive R Archive Network
(CRAN) at https://cran.r-project.org/package=medicalcoder. Developmental
versions, additional testing suites, benchmarking, and examples are on GitHub at
https://github.com/dewittpe/medicalcoder/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
_medicalcoder_ R package and drafted the manuscript. SR provided critical code
review and testing.  SR, MR, and TDB all contributed to the conception, design,
and interpretation and provided important feedback. TDB provided funding. All
authors critically reviewed the manuscript for important intellectual content
and approved of the final version of the manuscript.

[**Acknowledgments:**]{.underline} We thank Vincent Rubinetti for his assistance
in designing the _medicalcoder_ logo.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

**Figure 1:** The time required to flag comorbidities in the MIMIC-IV v3.1
dataset (
`r qwraps2::frmt(nrow(mimicivDT))` rows,
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects,
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
encounters) by utility and flagging method.  The MIMIIC-IV code time is based on
running RSQLite code for a in-memory database.  R packages _comorbidity_ and
_pccc_ require multiple calls and aggregations to handle the mixed use of ICD-9
and ICD-10 within the MIMIIC-IV dataset.  Applying the each of of the
comorbidity algorithms and flagging methods within _medicalcoder_ requires only
modifying the `method` and `flag.method` argument to the
`medicalcoder::comorbidities()` call.

