---
title: "medicalcoder: An R Package for Working with ICD Codes and Comorbidity Algorithms with Considerations for Longitudinal Data"

code-fold: false
echo: true
number-sections: false

format:
  docx:
    fig-cap-location: top
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
stopifnot(packageVersion("medicalcoder") >= "0.7.0")
```

**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; Margaret N. Rebull^1^, MA; Tellen D. Bennett^1,2^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

^2^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD

Department of Biomedical Informatics

University of Colorado School of Medicine

1890 N. Revere Court

Mailstop F600

Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} 1764 / 2000 (excluding tables and R code chunks),
2088 (including R chuncks, excluding tables)

[**Tables:**]{.underline} 1 / 2

[**Figures:**]{.underline} 3 / 3

[**Supplementary material:**]{.underline}

[**Short running title**]{.underline}: medicalcoder: R package for ICD codes and Comorbidities with longitudinal data

<!-- allowed 5; recomend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline} Comorbidity, International Classification of
Diseases (ICD), Software, Longitudinal Studies, Pediatrics

{{< pagebreak >}}

# Abstract {.unnumbered}

words: /150 words

**Objectives**

**Materials and Methods**

**Results**

**Conclusions**

{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
<!-- word count:  / 200 -->

:::

{{< pagebreak >}}

# Background and Significance

Accurate identification of clinical conditions in large healthcare datasets
depends  on International Classification of Diseases (ICD) codes. Comorbidity
algorithms built on these codes are central to risk adjustment, cohort
characterization, and the reproducible study of patient populations.

Existing software packages and utilities typically
implement only one or two comorbidity algorithms, usually a
variants of the Charlson and Elixhauser, and support applying either ICD-9 or
ICD-10
[@r-comorbidity;@feudtner2014pediatric;@r-pccc;@pcccv3sas;@r-icdcomorbid;@r-multimorbidity]
leaving the
end user to reconcile mixed-version datasets manually.
Input requirements vary across utilities making it difficult to apply multipe
comorbidity algorithms consistency.

To our knowledge, no current implimentation accounts for longitudinal
data.  While the AHRQ [@ahrq2025] Elixhauser variant accounts for
present-on-admission (POA) indicators and exemptions, it does nto propagate
conditions across encounters within a patient record.

For example, a patient who develops heart failure (HF)
during an encounter should have HF flagged on subsequent encounters regardless
of POA indicators.  Encounter-level flagging alone cannot capture this.

There is a need for a tool which can apply multiple comorbidity algorithms to a
data set through a unified API while accounting for the nuances of each
comorbidity algorithm, ICD versions and types, and can account for longitudinal
data.

We present the R package medicalcoder to fill this need.

# Objective

Develop a self-contained R package with a complete ICD database and multiple
comorbidity algorithm applied via a single function while accounting for POA,
primary diagnoses status, ICD version (9 or 10), ICD type (diagnosis or
procedure), full and compact ICD code strings, and longitudinal patient
histories.
<!-- word count: 46 -->

# Materials and Methods

medicalcoder has minimal dependencies. No namespaces are imported; R &geq; 3.5.0
is the only dependency due to a change in data serialization[^a].  Suggested
packages support building vignettes, testing, and building website articles.
<!-- word count: 32 -->

[^a]: https://cran.r-project.org/src/base/NEWS.3#:~:text=R%20has%20new,to%20version%203.5.0

```{r}
#| label: icd-src-years
#| include: false
aggregate(cbind(known_start, known_end) ~ src + icdv + dx,
  data = medicalcoder::get_icd_codes(),
  FUN = range
  )

cms9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))
cdc9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cdc10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cms10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))

who10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "who"),
    paste(min(known_start), max(known_end), sep = " - "))
```

The internal ICD database consists of

* ICD-9 diagnosis and procedure codes
  * Centers for Diseases Control (CDC) (fiscal years `r cdc9`)
  * Centers for Medicare and Medicaid Services (CMS) (fiscal years `r cms9`)
* ICD-10 diagnosis codes
  * CDC (calendar years `r cdc10`)
  * World Health Organization (WHO) (calendar years `r who10`).
* ICD-10 diagnosis and procedure codes
  * CMS (fiscal years `r cms10`).

Three comorbidity algorithms, Charlson
[@charlson1987;@deyo1992adapting;@ahrq2017;@quan2005;@quan2011;@glasheen2019],
Elixhauser [@elixhauser1998;@quan2005;@ahrq2025], and Pediatrics Complex Chronic
Conditions (PCCC) [@feudtner2014pediatric;@feinstein2024pediatric], each with
multiple variants, have been implemented in medicalcoder and are accessible via
the `method` argument of `medicalcoder::comorbidities()`, @tbl-comorbidities-methods.

::: {#tbl-comorbidities-methods}

+------------------------------------------+-----------------------------------+
| Comorbidity Method                       | Notes                             |
+:=========================================+:==================================+
| **Charlson**                                                                 |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `charlson_deyo1992`         | ICD-9 codes defined in Table 1 of |
|                                          | @quan2005                         |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `charlson_quan2005`         | ICD-9 and ICD-10 defined in       |
|                                          | Table 1 of @quan2005;             |
|                                          | index scoring as                  |
|                                          | reported in Table 2 of @quan2011  |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `charlson_quan2011`         | ICD-9 and ICD-10 defined in       |
|                                          | Table 1 of @quan2005;             |
|                                          | index scoring as                  |
|                                          | reported in Table 2 of @quan2011  |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `charlson_cdmf2019`         | ICD-9 and ICD-10 defined in       |
|                                          | @glasheen2019                     |
+------------------------------------------+-----------------------------------+
| **Elixhauser**                                                               |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_elixhauser1988` | ICD-9 codes defined in Table 2 of |
|                                          | @quan2005                         |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq_web`       | ICD-9 codes defined in Table 2 of |
|                                          | @quan2005                         |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_quan2005`       | ICD-9 and ICD-10 defined in       |
|                                          | Table 2 of @quan2005;             |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq2022`       | ICD-10 codes from @ahrq2025       |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq2023`       |ICD-10 codes from @ahrq2025        |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq2024`       | ICD-10 codes from @ahrq2025       |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq2025`       | ICD-10 codes from @ahrq2025       |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq_icd10`     | Any ICD-10 code from all          |
|                                          | `elixhauser_ahrqYYYY`             |
+------------------------------------------+-----------------------------------+
| **PCCC**                                                                     |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `pccc_v2.0`                 | ICD-9 and ICD-10 diagnostic and   |
|                                          | procedure codes consistent with   |
|                                          | @dewitt2025pccc                   |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `pccc_v2.1`                 | Extended set of `pccc_v2.0`       |
|                                          | codes based on                    |
|                                          | @dewitt2025pccc and supplements   |
|                                          | @feudtner2014pediatric            |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `pccc_v3.0`                 | ICD-9 and ICD-10 diagnostic and   |
|                                          | procedure codes consistent with   |
|                                          | @pcccv3sas                        |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `pccc_v3.1`                 | Extended set of `pccc_v3.0` codes |
|                                          | based on the supplements to       |
|                                          | @feinstein2024pediatric           |
+------------------------------------------+-----------------------------------+

Comorbidity methods implemented in medicalcoder.

:::

Development used R version 4.5.1.  A comprehensive test suite runs as a
GitHub action on R oldrelease, release, and devel.  The test suite also runs on
demand for all released versions of R &geq; 3.5.0 without suggested packages,
and all version &geq; 4.4.0 with suggested packages[^b].

[^b]: https://github.com/dewittpe/medicalcoder/tree/main/testing

# Results {#sec-results}

There are two conceptual groups of functions within medicalcoder: ICD utilities
and comorbidity flagging tools.

## ICD Utilities

The internal ICD database is accessible via `medicalcoder::get_icd_codes()`,
with optional descriptions and hierarchy. `medicalcoder::is_icd()` tests if a
character string is a known ICD code while considering ICD version, type,
source, year, and whether header codes can be consider valid. Both full and
compact codes can be considered simultaneously, and warnings for ambiguous codes
are optionally thrown.

The vignette `vignettes(topic = "icd", package = "medicalcoder")` provides
detailed usage examples for `medicalcoder::get_icd_codes()`,
`medicalcoder::is_icd()` and two more ICD utilities,
`medicalcoder::lookup_icd_codes()` and `medicalcoder::icd_compact_to_full()`.

```{r}
#| label: get-icd-codes
#| include: false
# omit from manuscript to save space and keep the focus on the comorbidities
str(medicalcoder::get_icd_codes())
# icdv:
#   integer 9 or 10 for ICD-9 or ICD-10
# dx:
#   diagnostic code (1); procedure code (0)
# full_code:
#   ICD code string with applicable dot, eg., D71.1
# code:
#   ICD code string without applicable dot, eg., D711
# src:
#   character source information from the cdc, cms, or who
# known_start, known_end:
#   first and last year that the code was known to be in
#   the source
# assignable_start, assignable_end:
#   first and last year that the code was known to be the
#   most granular code from the source
```


## Comorbidities

A single function call, `medicalcoder::comorbidities()`, is the interface for
applying any of `r length(medicalcoder:::comorbidities_methods())` comorbidity
algorithms, @tbl-comorbidities-methods. The input data is
expected to be in a long format, one ICD code per row, with optional columns for
identification variables (e.g., patient and encounter ids), ICD version (integer
value 9 or 10), ICD type (1 for diagnostic code, 0 for procedure code),
present-on-admission (POA) indicator, and primary diagnostic indicator.  The ICD
codes can be full, compact, or a combination thereof.  A singel patient/encouner
record can consist of ICD-9, ICD-10, or both.

A verbose example of calling `medicalcoder::comorbidities()`, including default
values, follows.  The example uses one of two example data sets provided by the
package. `mdcr` is a data.frame with 4 columns: patient identifier (patid), ICD
version (icdv), compact ICD code string (code), and an indicator for diagnostic
or procedure code (dx).

```{r}
#| label: comorbidities-example-1
data(mdcr, package = "medicalcoder") # load data

elixhauser_quan2005_comorbidities <-
  medicalcoder::comorbidities(
    data          = mdcr, # a data.frame
    icd.codes     = "code", # name of column in data with ICD code strings
    method        = "elixhauser_quan2005", # Comorbidity method_variant
    id.vars       = "patid", # (optional) id variables within data
    icdv.var      = "icdv",  # (optional) variable denoting ICD version
    icdv          = NULL, # (default)
    dx.var        = "dx", # (default is NULL)
                          # varaibale in data marking each data[[icd.codes]]
                          # as a diagnostic (1) or a procedure (0) code
    dx            = NULL, # (default)
    poa.var       = NULL, # (default)
    poa           = 1L,   # (default is NULL)
                          # Consider all codes to be present-on-admission
    age.var       = NULL, # (default) only relevent to charlson_<variant>
    primarydx.var = NULL, # (default)
    primarydx     = 0L,   # (default is NULL)
                          # Condiser all codes to be secondary diagnoses
    flag.method   = "current",  # (default), other option "cumulative"
    full.codes    = TRUE, # (default) Reference data[[icd.codes]] against
                          # full ICD codes in lookup tables
    compact.codes = TRUE, # (default) Reference data[[icd.codes]] against
                          # compact ICD codes in lookup tables
    subconditions = FALSE # (default) only relevent to pccc_<variant>
  )
```

Note that several arguements come in pairs, e.g., `icdv.var` and `icdv`.  The
version with the `.var` suffix will take precedence over the other if both are
defined.  The version without the `.var` suffix is provided as a simple way to
say the whole input data set has a particular property.  If both the versions
are left as the defaul NULL value, then that property will be ignored in the
lookup tables.  That is, if `icdv.var` and `icdv` are both NULL, then
`data[[icd.codes]]` will be compared to both ICD-9 and ICD-10 codes.  If
`icdv.var` is NULL and `icdv = 9L` then `data[[icd.codes]]` will only be
compared to ICD-9 codes in the lookup tables.

```{r}
str(elixhauser_quan2005_comorbidities, max.level = 0L)
```

The return from `medicalcoder::comorbidities()` is a data.frame with additional
metadata.  The data.frame has column(s) for the `id.vars` and an indicator
column each of the relevent comorbidities.  All returned data.frames include
columns `num_cmrb` for the number of comorbidities flaged for the `id.vars` and
`cmrb_flag`, an indicator column for at least one comorbidity.

The combination of `poa.var` or `poa` and `flag.method` impact when a
condition/comorbidity is flag, Figure 1. For the
example, below is the record one patient with six encouners.  ICD-10 diagnostic
codes C78.4 (mapping to cancer in Charlson, Elixhauser, and PCCC), and I50.40
(mapping to cardiovascular disease in Charlson, Elixhauser, and PCCC), appear on
encounters 2, 3, and 5.  The POA status of the codes are dyanmic as well.

```{r}
#| label: flag-method-example
#| include: false
# this will load the object record and args into the .Globalenv
load("figure1.Rdata")
```

```{r}
#| label: example-record
record
```

The following call is an example of for the `r nrow(args)`  calls needed to
generate the flags shown in Figure 1. The most important detail to take away is
that the `id.vars` have the name of the sortable encounter id listed last.

```{r}
#| label: example-call-for-flag-example
#| eval: false
medicalcoder::comorbidities(
  data = record,
  id.vars = c("patid", "encid"),
  icdv = 10L,
  dx = 1L,
  poa =     # ... three options: <0, 1, NULL>
  poa.var = # ... NULL when poa = 0 or 1, else "poa"
  method =  # ... charlson_quan2005, elixhauser_ahrq2025, pccc_v3.1
  primarydx = 0L,
  flag.method = # ... "current", "cumulative"
)
```

The utility of the longitudinal methods in medicalcoder are critical in the
assessment of PCCC v3 [@feinstein2024pediatric] conditions.  For these
conditions, some ICD codes are classified technology dependent codes.  If a
technology dependent code for a condition is in the patient record, it only
flags the condition if a non-technology code is also present. For example,
consider the following three ICD-10 codes: H49.811 (non-technology-dependent
metabolic), J84.111 (non-technology-dependent respiratory), and Z96.41
(technology-dependent metabolic).  For a patient with seven encounters, and one
code presented on ecnounter 2, one on encounter 4, and one on encounter 6, we
will apply `method = "pccc_v3.1"` with `flag.method = "cumulative"` and codes
considered to be present-on-admission.  Figure 2 illustrates when the conditions
are flagged.  For permutations 1, 2, and 3, where the non-technology-dependent
code for a metabolic condition appears in the record before the
technology-dependent metabolic code, the flags appear as intuitively expected,
metabolic flags as soon as the non-technology-dependent code appears.
Permutations 4, 5, and 6 are more interesting as the technology-dependent
metabolic code appears before the non-technology-dependent code.  Because the
longitudinal methods in `medicalcoder::comorbidities()` track all ICD codes from
all prior encounters, the metabolic technology-dependent-only flag appears in
permutations 4 and 6 due to the respiratory non-technology-dependent code
existing.  Flagging conditions on a ecnounter only level and aggregating the
results would result in missing metabolic on encounters 4 and 5 in permutations
4 and 6.

## Testing and Benchmarking
Currently, medicalcoder has more than 95% code coverage via testing.

Within the test suite within medicalcoder, there are tests for ensuring
consistency between `method = "pccc_v2.0"` and the R package pccc[@r-pccc],
between `method = elixhauser_ahrq20XX` and the published SAS code from AHRQ[@ahrq2025].
As part of the website[^d] build, consistency between
`method = charlson_quan2005` and the SQL code provided by
MIT-LCP/mimic-code/mimic-iv[@johnson2018mimic] is verified.

[^d]: https://www.peteredewitt.com/medicalcoder/articles/medicalcoder-vs-mimic.html

medicalcoder has been designed to work with just base R.  That said, if the
input to the `data` argument of `medicalcoder::comorbidities()` is a data.table
[@r-data.table] and the data.table namespace is available, guarded S3 method
dispatching will exploit the data.table efficiencies for merge, setting, and
subsetting.
`medicalcoder::comorbidities()` can process
100,000 encounters in less than 10 seconds when `flag.method = "current"`. Using
a data.table can take less than 50% the time compared to a data.frame, Figure 3. Extensive
details on the benchmarking are on GitHub[^c].

[^c]: https://github.com/dewittpe/medicalcoder/tree/main/benchmarking

Comparing medicalcoder to other R packages has been done (see supplements).
medicalcoder has more features, more comorbidity methods, and flexability than
other packages.  Notable alternatives at the time this manuscript was written
are: multimorbidity [@r-multimorbidity], icdcomorbid [@r-icdcomorbid], and
comorbidity [@r-comorbidity].

```{r}
#| label: vs
#| include: false
vs <- new.env()
load(file = "medicalcoder-vs.Rdata", envir = vs)
```

`icdcomorbid::icd9_to_comorbid()` required
`r round(attr(vs$icdcomorbid_icd9_results, "tictoc"), 1)` seconds to process
only `r nrow(vs$icdcomorbid_icd9_results)` patids using only ICD-9 codes.
`medicalcoder::comorbitities()` took
`r round(attr(vs$medicalcoder_charlson_quan2005, "tictoc"), 2)` seconds to
process the whole dataset against ICD-9 and ICD-10 codes.  Due to the several
orders of magnitude different in computation time, we did not compare results
between `medicalcoder::comorbitities()` and `icdcomorbid::icd9_to_comorbid()`.

`multimorbidity::charlson()` required
`r round(attr(vs$multimorbidity_charlson_results, "tictoc"), 1)` seconds to apply the
Charlson comorbidities to the `mdcr` data set. We noted several false negatives
resulting from the `multimorbidity::charlson()` call.

The R package comorbidity is a widely referenced and downloaded package that can
apply the Charlson and Elixhauser (Quan 2005 and Quan 2011) a couple other
variants of other comorbidity algorithms.  The difference in time required to
apply the methods between `medicalcoder::comorbidities()` and
`comorbidity::comorbidity()` is largely due to `comorbidity::comorbidity()`
requiring two calls, one to apply ICD-9 codes, one for ICD-10 codes, aggregating
the results, and then scoring them whereas `medicalcoder::comorbidities()`
required only one call. `medicalcoder::comorbidities()` took
`r round(attr(vs$medicalcoder_charlson_quan2005, "tictoc"), 2)` seconds to apply
`charlson_quan2005`,
`r round(attr(vs$medicalcoder_charlson_quan2011, "tictoc"), 2)`
seconds to apply `charlson_quan2011`, and
`r round(attr(vs$medicalcoder_elixhauser_quan2005, "tictoc"), 2)` seconds to apply
`elixhauser_quan2005`. Compared to `comorbidity:comorbidity()` which took
`r round(attr(vs$comorbidity_charlson_results, "tictoc"), 1)` seconds to apply Charlson, and
`r round(attr(vs$comorbidity_elixhauser_results, "tictoc"), 1)` seconds to apply
Elixhauser.
We found the flags and index scores were consistent between the two packages.

# Discussion

medicalcoder was initially built to implement PCCC v3
[@feinstein2024pediatric].  A requirement was to have a self-contained package
that could be easily installed on a machine with extremely limited or no
internet access.  Using only base R was needed due to a history of insufficient
system dependencies resulting in difficulties compiling other R packages.  The
development of the ICD utilities within medicalcoder were built to inform the
code sets needed for the comorbidity algorithms.

Extensive testing ensures reproducibility and compatibility with published
algorithms. The workflow for updating ICD code sets or adding new algorithm
variants is streamlined; recent CMS ICD‑10 updates were integrated within hours.

medicalcoder provides computationally efficient implementations of
`r length(medicalcoder:::comorbidities_methods())` comorbidity methods.
This opens research opportunities for investigation of different comorbidity
classifications within the same data sets.

Guarded S3 methods allow medicalcoder to exploit data.table performance without
requiring it.

Longitudinal support distinguishes medicalcoder from existing tools. Figures 1
and 2 illustrate how encounter‑level aggregation can miss clinically important
conditions.

## Conclusions

medicalcoder is a fully-self contained R package with a detailed and robust
internal ICD database and is able to efficiently apply many comorbidity
algorithms to a dataset while accounting for ICD-version, ICD-type,
present-on-admission, primary diagnostic, full and compact ICD codes.  Flagging
codes can be done for individual patient/encounters, or longitudinally within a
patient record.

{{< pagebreak >}}

[**Funding Statement:**]{.underline}

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline}
The medicalcoder R packaage is available from the Comprehensive R Archive
Network (CRAN) at https://cran.r-project.org/package=medicalcoder.
Developmental versions, additional testing suites, and examples are on GitHub at
https://github.com/dewittpe/medicalcoder/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
medicalcoder R package and drafted the manuscript. SR provided critical code
review and testing.  SR, MR, nd TDB all contributed to the conception, design,
and interpretation and provided important feedback. TDB provided funding. All
authors critically reviewed the manuscript for important intellectual content
and approved of the final version of the manuscript.

[**Acknowledgments:**]{.underline} We thank Vincent Rubinetti for his assistance
in designing the hex-sticker logo for medicalcoder.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

Figure 1: Flagging cancer and cardiovascular disease by Comorbidity method,
present on admission (POA) flagging, and flag method.  Consider a patient record
with six encounters.  ICD-10 diagnostic code C78.4 (mapping to cancer in
Charlson, Elixhauser, and PCCC) is reported on encounter 2 as not present on
admission (POA), and on encounter 5 as POA.  Additionally, ICD-10 diagnostic
code I50.40 (mapping to cardiovascular disease (CVD) or congenstie heart
failure) is reported as POA on encounter 3 and not-POA on encounter 5.  This
figure illustrates when cancer and when CVD are flagged by the different
comorbidity algorithms and by how POA is handeled.  `poa = 0` considers all
codes to be not-POA, `poa = 1` considers all codes to be POA, and `poa.var =
'poa'` uses a dynamic, code-specific, flag in the input data to determine the
POA status.  The flag method `current` only considers the codes and POA status
on the specific encounter, where as `cumulative` considers if the code has been
reported previously, hence the flagging of both conditons on encounter 4 under
all comorbidity agorithems under `flag.method = 'cumulative'` as the codes from
encounters 2 and 3 are considered in the history for encounter 4.

Figure 2: Flagging of metabolic and respiratory conditiions via the Pediatrics
Complex Chronic Conditions v3 [@feinstein2024pediatric] for an example patient
with ICD-10 codes: H49.811 (non-technology-dependent metabolic), J84.111
(non-technology-dependent respiratory), and Z96.41 (technology-dependennt
metabolic).  One code appears on encounter 2, 4, and 6.  PCCC v3 only flags a
condition based on a technology-dependennt code if at least one non-technology
dependennt code also exits.  Permutations 1, 2, 3 easily flag the metabolic
condition because the H49.811 appears before Z96.41 in the record.  Permutations
4, 5, 6 do not flag metabolic until a non-technology dependennt code is in the
record, e.g., in permutation 6, Z96.41 appears on encounter 2. Since no
non-technology-dependennt code is in the record no condition is flagged on
encounter 2 or 3.  On encounter 4 when J84.111 is reported, then both metabolic
and respiratory conditions are flagged as a non-technology-dependennt code
exists in the record.

Figure 3: Benchmarking statistics for the medicalcoder package.  Comorbidity
algorithms Charlson [@quan2005], Elixhauser [@quan2005], and Pediatric Complex
Chronic Conditions v2 [@feinstein2024pediatric] were applied to simulated data
sets.  The benchmarking was performed on a computational server.  All processes,
including the data.table ones, were limited to a single thread.  General
performance between base-R data.frames and tidyverse tibbles are the same
whereas notable time and memory use reductions are seen for data.tables.  While
medicalcoder does not import any namespaces, it does use guarded S3 method
dispatching when data.tables are present to achieve the performance gains.  The
benchmarking was done on a computation server running Ubuntu (24.04.3 LTS), R
4.5.1, with Intel(R) Xeon(R) Gold 6416H CPUs and 2 TBs RAM.

