---
title: "medicalcoder: An R Package for Working with ICD Codes and Comorbidity Algorithms with Considerations for Longitudinal Data"

code-fold: false
echo: true
number-sections: false

format:
  docx:
    fig-cap-location: top
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
stopifnot(packageVersion("medicalcoder") >= "0.7.0")
```

**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; Margaret N. Rebull^1^, MA; Tellen D. Bennett^1,2^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

^2^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD

Department of Biomedical Informatics

University of Colorado School of Medicine

1890 N. Revere Court

Mailstop F600

Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} XXXX / 2000 (excluding tables and R code chunks),
XXXX (including R chunks, excluding tables)

[**Tables:**]{.underline} 1 / 2

[**Figures:**]{.underline} 3 / 3

[**Supplementary material:**]{.underline}

[**Short running title**]{.underline}: medicalcoder: R package for ICD codes and Comorbidities with longitudinal data

<!-- allowed 5; recomend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline} Comorbidity, International Classification of
Diseases (ICD), Software, Longitudinal Studies, Pediatrics

{{< pagebreak >}}

# Abstract {.unnumbered}

Word count: 158 / 150
<!--
  word count:
    objectives:            38
    materials and methods: 68
    results:               35
    conclusions:           17
-->

**Objectives**

Develop a self-contained R package with an internal ICD database and utilities
and applies multiple comorbidity algorithms to ICD-coded data while accounting
for ICD version, type, full and compact codes, present-on-admission status,
primary diagnoses, and longitudinal patient histories.

**Materials and Methods**

```{r}
#| label: import-benchmark-results
#| include: false
mimicivbenchmark <- readRDS("benchmark.rds")
data.table::setDT(mimicivbenchmark)
best_case_benchmark <-
  mimicivbenchmark[
    subjects == 1e6 &
      encounters == max(encounters) &
      data_class == "data.table",
    min(time_smooth),
    keyby = .(data_class, subjects, encounters)
    ]
```

We created medicalcoder, an R package containing a comprehensive ICD-9 and
ICD-10 diagnostic and procedure code database and implementations of
`r length(medicalcoder:::comorbidities_methods())` variants of the Charlson,
Elixhauser, and Pediatric Complex Chronic Conditions (PCCC) algorithms.
A single function, `medicalcoder::comorbidities()`, is used to apply the
comorbidities methods with suppport for both encounter-level and cumulative
longitudinal flagging of comorbidities. Extensive testing ensures consistency
with published SAS, SQL, and R reference implementations.

**Results**

medicalcoder accurately applies all included algorithms across mixed ICD-version
datasets and efficiently processes large inputs.  1 million subjects with a
total of more than 2.7 million encounters, can be processed in less than 30
seconds.

**Conclusions**

medicalcoder provides a reproducible, efficient, and comprehensive framework for
applying comorbidity algorithms to modern longitudinal clinical datasets.

{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
Word count: 187 / 200

Researchers often use International Classification of Diseases (ICD) diagnosis
and procedure codes to identify health conditions in large medical datasets.
Comorbidity algorithms group these codes into meaningful categories, such as
heart disease or cancer, and are essential for understanding patient complexity,
comparing outcomes, and adjusting for risk. However, existing tools usually
handle only one or two algorithms, work with either ICD-9 or ICD-10 (but not
both together), and do not track health conditions across multiple hospital
visits. This can lead to missed or inconsistent results.

We created medicalcoder, an R package that solves these problems. It includes a
complete database of ICD-9 and ICD-10 codes and provides a single, easy-to-use
function to apply many comorbidity algorithms. The package can recognize both
full and compact ICD codes, use present-on-admission flags, identify primary
diagnoses, and follow patients across multiple encounters to avoid
missing conditions that appear earlier in their history.

medicalcoder is fast, works with base R, and has been thoroughly tested against
published methods. It helps researchers apply comorbidity algorithms more
accurately and consistently, improving the quality and reproducibility of
studies using electronic health record or administrative data.

:::

{{< pagebreak >}}

# Background and Significance

Accurate identification of clinical conditions in large healthcare datasets
depends on International Classification of Diseases (ICD) codes
[@quan2005;@quan2011;@feinstein2024pediatric;@feudtner2014pediatric;@ahrq2025].
Comorbidity algorithms built on these codes support risk adjustment, cohort
characterization, and reproducible study of patient populations
[@quan2005;@yurkovich2015systematic;@guo2022evaluation;@li2008risk].

Existing software packages typically implement one or two comorbidity
algorithms—most often variants of Charlson and Elixhauser—and support either
ICD-9 or ICD-10, but not both
[@r-comorbidity;@feudtner2014pediatric;@r-pccc;@pcccv3sas;@r-icdcomorbid;@r-multimorbidity].
This leaves users to reconcile mixed ICD-version datasets manually, and input
requirements vary across utilities, making it difficult to apply multiple
algorithms consistently.

To our knowledge, no current implementation accounts for longitudinal data.
While the AHRQ [@ahrq2025] Elixhauser variant accounts for present-on-admission
(POA) indicators and exemptions, it does not propagate conditions across
encounters within a patient record. Conditions that arise during acute care and
should not be considered comorbidities for the present encounter are handled via
POA indicators. However, unless the relevant ICD codes are flagged as POA on
subsequent encounters, those comorbidities will not be flagged.

There is a need for a tool that applies multiple comorbidity algorithms through
a unified API while accounting for algorithm-specific nuances, ICD versions and
types, and longitudinal data.

We present the R package medicalcoder to fill this need.

# Objective

Develop an R package with a comprehensive internal ICD database and the ability
to apply multiple comorbidity algorithms to a dataset. The assessment of
comorbidities should handle ICD-9 and ICD-10 simultaneously, accept full and
compact codes, incorporate POA flags and primary diagnosis indicators, and
support longitudinal assessment across patient encounters.

The package must have no required dependencies beyond base R. When available, it
should dynamically use tidyverse or data.table backends to improve performance
without requiring them for installation or use.

# Materials and Methods

medicalcoder has minimal dependencies. No namespaces are imported; R &geq; 3.5.0
is the only dependency due to a change in data serialization[^a]. Several
packages are suggested for enhanced performance (data.table[@r-data.table] and
dplyr[@r-dplyr]) and for testing and documentation, but suggested packages are
not required for installation or use.

[^a]: https://cran.r-project.org/src/base/NEWS.3#:~:text=R%20has%20new,to%20version%203.5.0

```{r}
#| label: icd-src-years
#| include: false
aggregate(cbind(known_start, known_end) ~ src + icdv + dx,
  data = medicalcoder::get_icd_codes(),
  FUN = range
  )

cms9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))
cdc9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cdc10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cms10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))

who10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "who"),
    paste(min(known_start), max(known_end), sep = " - "))
```

The internal ICD database consists of ICD-9 diagnostic codes from the Centers
for Disease Control and Prevention (CDC; fiscal years `r cdc9`) and ICD-9-CM and
ICD-9-PCS codes from the Centers for Medicare and Medicaid Services (CMS; fiscal
years `r cms9`). ICD-10 diagnostic codes from the CDC (calendar years `cdc10`)
and the World Health Organization (WHO; calendar years `r who10`) are included,
along with ICD-10-CM and ICD-10-PCS codes from CMS (fiscal years `r cms10`). The
internal database will be extended as additional code sets are published.

Three comorbidity algorithms, Charlson
[@charlson1987;@deyo1992adapting;@ahrq2017;@quan2005;@quan2011;@glasheen2019],
Elixhauser [@elixhauser1998;@quan2005;@ahrq2025], and Pediatrics Complex Chronic
Conditions (PCCC) [@feudtner2014pediatric;@feinstein2024pediatric], each with
multiple variants, have been implemented in medicalcoder and are accessible via
the `method` argument of `medicalcoder::comorbidities()`, Table 1.

<!-- for reasons I can't explain, I cannot get the left justfication of the
table to persist in a table div in the docx file.  By keeping it out of a table
div the text justification persists -->

**Table 1:** Comorbidity methods implemented in medicalcoder version `r packageVersion("medicalcoder")`.

| Comorbidity Method                       | Notes                                                                                                   |
| :---                                     | :----------                                                                                             |
| **Charlson**                             |                                                                                                         |
| &nbsp;&nbsp; `charlson_deyo1992`         | ICD-9 codes defined in Table 1 of [@quan2005]                                                           |
| &nbsp;&nbsp; `charlson_quan2005`         | ICD-9 and ICD-10 defined in Table 1 of [@quan2005]; index scoring as reported in Table 2 of [@quan2011] |
| &nbsp;&nbsp; `charlson_quan2011`         | ICD-9 and ICD-10 defined in Table 1 of [@quan2005]; index scoring as reported in Table 2 of [@quan2011] |
| &nbsp;&nbsp; `charlson_cdmf2019`         | ICD-9 and ICD-10 defined in [@glasheen2019]                                                             |
| **Elixhauser**                           |                                                                                                         |
| &nbsp;&nbsp; `elixhauser_elixhauser1988` | ICD-9 codes defined in Table 2 of [@quan2005]                                                           |
| &nbsp;&nbsp; `elixhauser_ahrq_web`       | ICD-9 codes defined in Table 2 of [@quan2005]                                                           |
| &nbsp;&nbsp; `elixhauser_quan2005`       | ICD-9 and ICD-10 defined in Table 2 of [@quan2005]                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2022`       | ICD-10 codes from [@ahrq2025] for fiscal year 2022                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2023`       | ICD-10 codes from [@ahrq2025] for fiscal year 2023                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2024`       | ICD-10 codes from [@ahrq2025] for fiscal year 2024                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2025`       | ICD-10 codes from [@ahrq2025] for fiscal year 2025                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2026`       | ICD-10 codes from [@ahrq2025] for fiscal year 2026                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq_icd10`     | Any ICD-10 code from all `elixhauser_ahrqYYYY`                                                          |
| **PCCC**                                 |                                                                                                         |
| &nbsp;&nbsp; `pccc_v2.0`                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with [@dewitt2026pccc]                       |
| &nbsp;&nbsp; `pccc_v2.1`                 | Extended set of `pccc_v2.0` pccc and supplements [@feudtner2014pediatric]                               |
| &nbsp;&nbsp; `pccc_v3.0`                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with [@pcccv3sas]                            |
| &nbsp;&nbsp; `pccc_v3.1`                 | Extended set of `pccc_v3.0` codes based on the supplements to [@feinstein2024pediatric]                 |

Development used R version 4.5.2. A comprehensive test suite runs as a GitHub
Action on R oldrelease (4.4.3), release (4.5.2), and devel (what will become
4.6.0). An additional test suite runs on demand for all released versions of R
&geq; 3.5.0 without suggested packages, and for versions &geq; 4.4.0 with
suggested packages[^b].

[^b]: https://github.com/dewittpe/medicalcoder/tree/main/testing

# Results {#sec-results}

There are two conceptual groups of functions within medicalcoder: ICD utilities
and comorbidity flagging tools.

## ICD Utilities

The internal ICD database is accessible via `medicalcoder::get_icd_codes()`,
with optional descriptions and hierarchy. `medicalcoder::is_icd()` tests if a
character string is a known ICD code while considering ICD version, type,
source, year, and whether header codes can be considered valid. Both full and
compact codes can be considered simultaneously, and warnings for ambiguous codes
are optionally thrown.

The `"icd"` vignette provides
detailed usage examples for `medicalcoder::get_icd_codes()`,
`medicalcoder::is_icd()`, and two more ICD utilities,
`medicalcoder::lookup_icd_codes()` and `medicalcoder::icd_compact_to_full()`.

```{r}
#| label: get-icd-codes
#| include: false
# omit from manuscript to save space and keep the focus on the comorbidities
str(medicalcoder::get_icd_codes())
# icdv:
#   integer 9 or 10 for ICD-9 or ICD-10
# dx:
#   diagnostic code (1); procedure code (0)
# full_code:
#   ICD code string with applicable dot, eg., D71.1
# code:
#   ICD code string without applicable dot, eg., D711
# src:
#   character source information from the cdc, cms, or who
# known_start, known_end:
#   first and last year that the code was known to be in
#   the source
# assignable_start, assignable_end:
#   first and last year that the code was known to be the
#   most granular code from the source
```


## Comorbidities

A single function, `medicalcoder::comorbidities()`, is the interface for
applying any of `r length(medicalcoder:::comorbidities_methods())` comorbidity
algorithms implemented by medicalcoder (Table 1). The input data are expected
to be in long format, one ICD code per row, with optional columns for
identification variables (e.g., patient and encounter ids), ICD version (integer
value 9 or 10), ICD type (1 for diagnostic code, 0 for procedure code), POA
indicator, and primary diagnosis indicator. The ICD codes can be full, compact,
or a combination thereof. A single patient/encounter record can contain ICD-9,
ICD-10, or both.

A verbose example of calling `medicalcoder::comorbidities()`, including default
values, follows.  The example uses one of two example datasets provided by the
package. `mdcr` is a data.frame with 4 columns: patient identifier (patid), ICD
version (icdv), compact ICD code string (code), and an indicator for diagnostic
or procedure code (dx).

```{r}
#| label: comorbidities-example-1
data(mdcr, package = "medicalcoder") # load data

elixhauser_quan2005_comorbidities <-
  medicalcoder::comorbidities(
    data          = mdcr,  # a data.frame
    icd.codes     = "code", # name of column in data with ICD code strings
    method        = "elixhauser_quan2005", # Comorbidity method_variant
    id.vars       = "patid", # (optional) id variables within data
    icdv.var      = "icdv",  # (optional) variable denoting ICD version
    icdv          = NULL, # (default)
    dx.var        = "dx", # (default is NULL)
                          # variable in data marking each data[[icd.codes]]
                          # as a diagnostic (1) or a procedure (0) code
    dx            = NULL, # (default)
    poa.var       = NULL, # (default)
    poa           = 1L,   # (default is NULL)
                          # Consider all codes to be present-on-admission
    age.var       = NULL, # (default) only relevant to charlson_<variant>
    primarydx.var = NULL, # (default)
    primarydx     = 0L,   # (default is NULL)
                          # Consider all codes to be secondary diagnoses
    flag.method   = "current",  # (default), other option "cumulative"
    full.codes    = TRUE, # (default) Reference data[[icd.codes]] against
                          # full ICD codes in lookup tables
    compact.codes = TRUE, # (default) Reference data[[icd.codes]] against
                          # compact ICD codes in lookup tables
    subconditions = FALSE # (default) only relevant to pccc_<variant>
  )
```

Several arguments come in pairs, e.g., `icdv.var` and `icdv`. The version with
the `.var` suffix takes precedence if both are defined. The version without the
`.var` suffix is provided as a simple way to indicate a dataset-wide property.
If both are left as the default NULL value, that property is ignored in the
lookup tables. That is, if `icdv.var` and `icdv` are both NULL, then
`data[[icd.codes]]` will be compared to both ICD-9 and ICD-10 codes. If
`icdv.var` is NULL and `icdv = 9L` then `data[[icd.codes]]` will only be
compared to ICD-9 codes in the lookup tables.

```{r}
str(elixhauser_quan2005_comorbidities, max.level = 0L)
```

The return from `medicalcoder::comorbidities()` is a data.frame with additional
metadata.  The data.frame has column(s) for the `id.vars` and an indicator
column each of the relevant comorbidities.  All returned data.frames include
columns `num_cmrb` for the number of comorbidities flagged for the `id.vars` and
`cmrb_flag`, an indicator column for at least one comorbidity.

The combination of `poa.var` or `poa` and `flag.method` affects when a
condition/comorbidity is flagged (Figure 1). For example, below is the record
for one patient with six encounters.  ICD-10 diagnostic codes C78.4 (mapping to
cancer in Charlson, Elixhauser, and PCCC), and I50.40 (mapping to cardiovascular
disease in Charlson, Elixhauser, and PCCC), appear on encounters 2, 3, and 5.
The POA status of the codes are dynamic as well.

```{r}
#| label: flag-method-example
#| include: false
# this will load the object record and args into the .Globalenv
load("figure1.Rdata")
```

```{r}
#| label: example-record
record
```

The following call is an example of the `r nrow(args)`  calls needed to
generate the flags shown in Figure 1. The most important detail to take away is
that the `id.vars` have the name of the sortable encounter id listed last.  The
assumption, when `flag.method = "cumulative"`, is that the
`data[[tail(id.vars, 1)]]` is a sortable variable which puts encounters in
chronological order within the other `id.vars`.

```{r}
#| label: example-call-for-flag-example
#| eval: false
medicalcoder::comorbidities(
  data = record,
  id.vars = c("patid", "encid"),
  icdv = 10L,
  dx = 1L,
  poa =     # ... three options: <0, 1, NULL>
  poa.var = # ... NULL when poa = 0 or 1, else "poa"
  method =  # ... charlson_quan2005, elixhauser_ahrq2025, pccc_v3.1
  primarydx = 0L,
  flag.method = # ... "current", "cumulative"
)
```

```{r}
#| label: load-figure2permutations
#| include: false
figure2permutations <- readRDS("figure2permutations.rds")
```

The longitudinal methods in medicalcoder are critical for PCCC v3
[@feinstein2024pediatric]. For these conditions, some ICD codes are classified
as technology dependent. A technology-dependent code flags a condition only if
at least one non-technology-dependent code is also present. For example,
consider three ICD-10 codes: H49.811 (non-technology-dependent metabolic),
J84.111 (non-technology-dependent respiratory), and Z96.41
(technology-dependent metabolic). For a patient with seven encounters and one
code reported on encounters 2, 4, and 6, we apply `method = "pccc_v3.1"` with
`flag.method = "cumulative"` and codes treated as POA for all six permutations 
`r sprintf("(%s).", paste(sort(unique(figure2permutations$plabel)), collapse = "; "))`
Figure 2 illustrates when conditions are flagged for each permutation. For
permutations 1, 2, and 3, the non-technology-dependent metabolic code appears
before the technology-dependent metabolic code, so the metabolic condition flags
immediately after H49.811 appears. Permutations 4, 5, and 6 are more interesting
because Z96.41 appears first. Because `medicalcoder::comorbidities()` tracks all
ICD codes from prior encounters, the metabolic technology-dependent-only flag
appears in permutations 4 and 6 due to the earlier respiratory
non-technology-dependent code. Encounter-level flagging with post hoc
aggregation would miss metabolic flags on encounters 4 and 5 in permutations 4
and 6.

## Testing

Within the testing suite there are tests for ensuring consistency between
`method = "pccc_v2.0"` and the R package pccc[@r-pccc], and between
`method = elixhauser_ahrq20XX` and the published SAS code from AHRQ[@ahrq2025].
As part of the website[^d] build, consistency between
`method = charlson_quan2005` and the SQL code provided by
MIT-LCP/mimic-code/mimic-iv[@johnson2018mimic] is verified.

[^d]: https://www.peteredewitt.com/medicalcoder/articles/medicalcoder-vs-mimic.html

## Benchmarking
medicalcoder has been designed to work with just base R.  Any system with R >=
3.5.0 can install and use medicalcoder.

That said, medicalcoder suggests the use of data.frame extensions tibbles or
data.tables. When the input data class to `medicalcoder::comorbidities()` is a
tibble and the dplyr namespace is available, or when the input is a data.table
and the data.table namespace is available, then guarded S3 methods apply the
appropriate data-manipulation tools. The result is a 4- to 8-fold decrease in
expected computation time and memory use (Figure 3). Extensive details on the
benchmarking are on GitHub[^c].

[^c]: https://github.com/dewittpe/medicalcoder/tree/main/benchmarking

## Comparing to Other R packages

Comparing medicalcoder to other R packages has been done (see supplements).
medicalcoder has more features, more comorbidity methods, and greater
flexibility than other packages. Notable alternatives at the time this
manuscript was written are multimorbidity [@r-multimorbidity],
icdcomorbid [@r-icdcomorbid], and comorbidity [@r-comorbidity].

```{r}
#| label: vs
#| include: false
vs <- new.env()
load(file = "medicalcoder-vs.Rdata", envir = vs)
```

The R package comorbidity[@r-comorbidity] is a widely referenced and downloaded
package that can apply Charlson and Elixhauser (Quan 2005 and Quan 2011), along
with a few other variants of comorbidity algorithms.
We compared `medicalcoder::comorbidities()` to `comorbidity::comorbidity()`
using the `medicalcoder::mdcr` dataset.
We found the flags and index scores were consistent between the two packages.

The computation time required was different.  `comorbidity::comorbidity()`
required
`r round(attr(vs$comorbidity_charlson_results, "tictoc"), 1)` seconds to apply Charlson, and
`r round(attr(vs$comorbidity_elixhauser_results, "tictoc"), 1)` seconds to apply
Elixhauser.  Compared to `medicalcoder::comorbidities()` taking
`r round(attr(vs$medicalcoder_charlson_quan2005, "tictoc"), 2)` seconds to apply
`charlson_quan2005`,
`r round(attr(vs$medicalcoder_charlson_quan2011, "tictoc"), 2)`
seconds to apply `charlson_quan2011`, and
`r round(attr(vs$medicalcoder_elixhauser_quan2005, "tictoc"), 2)` seconds to apply
`elixhauser_quan2005`.
The difference in time required to
apply the methods between `medicalcoder::comorbidities()` and
`comorbidity::comorbidity()` is largely due to `comorbidity::comorbidity()`
requiring two calls, one to apply ICD-9 codes, one for ICD-10 codes, aggregating
the results, and then scoring them whereas `medicalcoder::comorbidities()`
required only one call.

The comparison between `medicalcoder::comorbidities()` and
`icdcomorbid::icd9_to_comorbid()` [@r-icdcomorbid] was stopped due to computation
time.
`icdcomorbid::icd9_to_comorbid()` took
`r round(attr(vs$icdcomorbid_icd9_results, "tictoc"), 1)` seconds to process
`r nrow(vs$icdcomorbid_icd9_results)` patients from the `medicalcoder::mdcr` data
sets and only using only ICD-9 codes.
`medicalcoder::comorbidities()` took
`r round(attr(vs$medicalcoder_charlson_quan2005, "tictoc"), 2)` seconds to
process the `r length(unique(medicalcoder::mdcr[["patid"]]))`-patient dataset
against ICD-9 and ICD-10 codes. Due to the several orders-of-magnitude
difference in computation time, we did not compare results between
`medicalcoder::comorbidities()` and `icdcomorbid::icd9_to_comorbid()`.

`multimorbidity::charlson()` [@r-multimorbidity] required
`r round(attr(vs$multimorbidity_charlson_results, "tictoc"), 1)` seconds to apply
the Charlson comorbidities to the `medicalcoder::mdcr` dataset. We noted several
false negatives relative to the published definitions (details in the
supplement).

# Discussion

medicalcoder was initially built to implement PCCC v3 [@feinstein2024pediatric].
A requirement was a self-contained package that could be installed on machines
with extremely limited or no internet access. Using only base R was necessary
given prior difficulties acquiring and compiling other R packages due to system
dependencies. The ICD utilities in medicalcoder informed the code sets needed
for the comorbidity algorithms. Guarded S3 methods allow medicalcoder to exploit
data.table and dplyr performance gains without requiring either namespace for
installation and basic use.

Extensive testing ensures reproducibility and compatibility with published
algorithms. The workflow for updating ICD code sets or adding new algorithm
variants is streamlined; recent CMS ICD‑10 updates were integrated within hours.

medicalcoder provides computationally efficient implementations of
`r length(medicalcoder:::comorbidities_methods())` comorbidity methods. This
opens research opportunities for investigation of different comorbidity
classifications within the same datasets.

To our knowledge, medicalcoder is the first and only package to support
comorbidity flagging in longitudinal datasets.  Additionally, medicalcoder
provides extended levels of detail for PCCC, namely subconditions can be flagged
in addition to the primary conditions.


## Conclusions

medicalcoder is a fully self-contained R package with a detailed and robust
internal ICD database that can efficiently apply many comorbidity
algorithms to a dataset while accounting for ICD-version, ICD-type,
present-on-admission, primary diagnosis, and full and compact ICD codes. Flagging
codes can be done for individual patient/encounters, or longitudinally within a
patient record.

{{< pagebreak >}}

[**Funding Statement:**]{.underline}

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline}
The medicalcoder R package is available from the Comprehensive R Archive Network
(CRAN) at https://cran.r-project.org/package=medicalcoder. Developmental
versions, additional testing suites, benchmarking, and examples are on GitHub at
https://github.com/dewittpe/medicalcoder/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
medicalcoder R package and drafted the manuscript. SR provided critical code
review and testing.  SR, MR, and TDB all contributed to the conception, design,
and interpretation and provided important feedback. TDB provided funding. All
authors critically reviewed the manuscript for important intellectual content
and approved of the final version of the manuscript.

[**Acknowledgments:**]{.underline} We thank Vincent Rubinetti for his assistance
in designing the hex-sticker logo for medicalcoder.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

**Figure 1:** Flagging cancer and cardiovascular disease by comorbidity method,
present on admission (POA) flagging, and flag method.  Consider a patient record
with six encounters.  ICD-10 diagnostic code C78.4 (mapping to cancer in
Charlson, Elixhauser, and PCCC) is reported on encounter 2 as not present on
admission (POA), and on encounter 5 as POA.  Additionally, ICD-10 diagnostic
code I50.40 (mapping to cardiovascular disease (CVD) or congestive heart
failure) is reported as POA on encounter 3 and not-POA on encounter 5.  This
figure illustrates when cancer and when CVD are flagged by the different
comorbidity algorithms and by how POA is handled.  `poa = 0` considers all
codes to be not-POA, `poa = 1` considers all codes to be POA, and `poa.var =
'poa'` uses a dynamic, code-specific flag in the input data to determine the
POA status.  The flag method `current` only considers the codes and POA status
on the specific encounter, whereas `cumulative` considers if the code has been
reported previously, hence the flagging of both conditions on encounter 4 under
all comorbidity algorithms under `flag.method = 'cumulative'` as the codes from
encounters 2 and 3 are considered in the history for encounter 4.

**Figure 2:** Flagging of metabolic and respiratory conditions via the Pediatric
Complex Chronic Conditions v3 [@feinstein2024pediatric] for an example patient
with ICD-10 codes: H49.811 (non-technology-dependent metabolic), J84.111
(non-technology-dependent respiratory), and Z96.41 (technology-dependent
metabolic).  One code appears on encounter 2, 4, and 6.  PCCC v3 only flags a
condition based on a technology-dependent code if at least one non-technology
dependent code also exists.  Permutations 1, 2, 3 easily flag the metabolic
condition because the H49.811 appears before Z96.41 in the record.  Permutations
4, 5, 6 do not flag metabolic until a non-technology dependent code is in the
record, e.g., in permutation 6, Z96.41 appears on encounter 2. Since no
non-technology-dependent code is in the record, no condition is flagged on
encounter 2 or 3.  On encounter 4, when J84.111 is reported, both metabolic and
respiratory conditions are flagged because a non-technology-dependent code now
exists in the record.

**Figure 3:** Benchmarking statistics for the medicalcoder package. Charlson
[@quan2005], Elixhauser [@quan2005], and Pediatric Complex Chronic Conditions
v3.1 [@feinstein2024pediatric] were applied to simulated datasets using both
`current` and `cumulative` flag methods and three input data classes
(data.frame, tibble, data.table). The benchmarking was performed on a
computational server runing Ubuntu 24.04.3 LTS, R 4.5.1, Intel Xeon Gold 6416H
CPUs, and 2 TB RAM. All processes limited to a single thread. While medicalcoder
does not import any namespaces, guarded S3 method dispatching will use
data.table [@r-data.table] or dplyr [@r-dplyr] native methods for performance
gains when the respective namespaces are available and the input data is of the
appropriate class.
