---
title: "medicalcoder: A Unified and Longitudinally Aware Framework for ICD-Based Comorbidity Assessment in R"

code-fold: false
echo: true
number-sections: false

format:
  docx:
    fig-cap-location: top
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
stopifnot(packageVersion("medicalcoder") >= "0.7.0")
options(qwraps2_markup = "markdown")
loadNamespace("data.table")
loadNamespace("medicalcoder")
mimicivDT <- readRDS("mimicivDT.rds")
objs <- readRDS("objs.rds")
```

**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; Margaret N. Rebull^1^, MA; Tellen D. Bennett^1,2^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

^2^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD

Department of Biomedical Informatics

University of Colorado School of Medicine

1890 N. Revere Court

Mailstop F600

Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} XXXXX / 2000 (including R chunks, excluding tables)

[**Tables:**]{.underline} X / 2

[**Figures:**]{.underline} X / 3

[**Supplementary material:**]{.underline}

[**Short running title**]{.underline}: medicalcoder: R package for ICD codes and Comorbidities with longitudinal data

<!-- allowed 5; recomend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline} Comorbidity, International Classification of
Diseases (ICD), Software, Longitudinal Studies, Pediatrics

{{< pagebreak >}}

# Abstract {.unnumbered}

Word count: XXX / 150

**Objectives:**
Comorbidity algorithms derived from International Classification of Diseases
(ICD) codes are central to risk adjustment and cohort characterization, yet
existing software implementations are fragmented, inconsistently handle mixed
ICD-9 and ICD-10 data, and typically ignore longitudinal patient histories.
We present _medicalcoder_ to address these limitations.

**Materials and Methods:**
We developed _medicalcoder_, a self-contained R package with an internal ICD
database and a unified interface for applying
`r length(medicalcoder:::comorbidities_methods())` variants of the Charlson,
Elixhauser, and Pediatric Complex Chronic Conditions algorithms. The package
supports full and compact ICD codes, mixed ICD versions, present-on-admission
and primary diagnosis indicators, and both encounter-level and cumulative
longitudinal flagging.

**Results:**
Outputs are consistent with published SAS, SQL, and R reference implementations.
_medicalcoder_ only depends on R &geq; 3.5.0 while dynamically leveraging
_data.table_ or _dplyr_ when available, yielding substantial performance
improvements.  Flagging comorbidities in longitudinal data results in greater
sensitivity, especially for PCCC version 3, than ad hoc aggregations after
ecounter level flagging.

**Conclusions:**
_medicalcoder_ is a fully self-contained R package that provides a unified,
reproducible framework for applying multiple comorbidity algorithms to modern
clinical datasets. By supporting mixed ICD-9 and ICD-10 data, full and compact
codes, present-on-admission and primary diagnosis indicators, and cumulative
longitudinal flagging, _medicalcoder_ addresses both methodological and
practical limitations of existing implementations.

The ability to propagate clinically relevant conditions across patient histories
improves comorbidity ascertainment in longitudinal data and reduces reliance on
ad hoc preprocessing or repeated method calls. Combined with extensive
validation against published reference implementations and efficient performance
at scale, _medicalcoder_ enables accurate, transparent, and reproducible
comorbidity assessment across a wide range of research and deployment
environments.

{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
Word count: 187 / 200

Researchers often use International Classification of Diseases (ICD) diagnosis
and procedure codes to identify health conditions in large medical datasets.
Comorbidity algorithms group these codes into meaningful categories, such as
heart disease or cancer, and are essential for understanding patient complexity,
comparing outcomes, and adjusting for risk. However, existing tools usually
handle only one or two algorithms, work with either ICD-9 or ICD-10 (but not
both together), and do not track health conditions across multiple hospital
visits. This can lead to missed or inconsistent results.

We created _medicalcoder_, an R package that solves these problems. It includes
a complete database of ICD-9 and ICD-10 codes and provides a single, easy-to-use
function to apply many comorbidity algorithms. The package can recognize both
full and compact ICD codes, use present-on-admission flags, identify primary
diagnoses, and follow patients across multiple encounters to avoid missing
conditions that appear earlier in their history.

_medicalcoder_ is fast, works with base R, and has been thoroughly tested against
published methods. It helps researchers apply comorbidity algorithms more
accurately and consistently, improving the quality and reproducibility of
studies using electronic health record or administrative data.

:::

{{< pagebreak >}}

# Background and Significance

Accurate identification of clinical conditions in large healthcare datasets
depends on International Classification of Diseases (ICD) codes
[@quan2005;@quan2011;@feinstein2024pediatric;@feudtner2014pediatric;@ahrq2025;@glasheen2019].
Comorbidity algorithms built on these codes support risk adjustment, cohort
characterization, and reproducible study of patient populations
[@quan2005;@yurkovich2015systematic;@guo2022evaluation;@li2008risk].

Existing software for flagging comorbidities typically implement one or two comorbidity
algorithms, most often variants of Charlson and Elixhauser, and support either
ICD-9 or ICD-10, but not both, at least no simultaneously
[@r-comorbidity;@feudtner2014pediatric;@r-pccc;@pcccv3sas;@r-icdcomorbid;@r-multimorbidity].
This leaves users to reconcile mixed ICD-version datasets manually. Input
requirements vary across utilities, making it difficult to apply multiple
algorithms consistently.

To our knowledge, no current implementation accounts for longitudinal data.
While the AHRQ [@ahrq2025] Elixhauser variant accounts for present-on-admission
(POA) indicators and exemptions, it does not propagate conditions across
encounters within a patient record. Conditions that arise during acute care and
should not be considered comorbidities for the present encounter are handled via
POA indicators. However, unless the relevant ICD codes are flagged as POA on
subsequent encounters, those comorbidities will not be flagged.

There is a need for a tool that applies multiple comorbidity algorithms through
a unified API while accounting for algorithm-specific nuances, ICD versions and
types, and longitudinal data.

We present the R package _medicalcoder_ to fill this need.

# Objective

Develop a fully self-contained R package with a comprehensive internal ICD
database capable of applying multiple comorbidity algorithms to clinical
datasets. The package should support simultaneous ICD-9 and ICD-10 coding,
accept full and compact ICD codes, incorporate present-on-admission and primary
diagnosis indicators, and enable longitudinal comorbidity assessment across
patient encounters.

# Materials and Methods

_medicalcoder_ was designed to be fully self-contained and deployable in
restricted computing environments. No external namespaces are imported; R &geq;
3.5.0 is the only required dependency, reflecting changes in data serialization
introduced in that version[^a]. Several packages are suggested for enhanced
performance (data.table [@r-data.table] and dplyr [@r-dplyr]) and for testing
and documentation, but are not required for installation or use.

[^a]: https://cran.r-project.org/src/base/NEWS.3#:~:text=R%20has%20new,to%20version%203.5.0

```{r}
#| label: icd-src-years
#| include: false
aggregate(cbind(known_start, known_end) ~ src + icdv + dx,
  data = medicalcoder::get_icd_codes(),
  FUN = range
  )

cms9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))
cdc9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cdc10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cms10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))

who10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "who"),
    paste(min(known_start), max(known_end), sep = " - "))
```

The internal ICD database consists of ICD-9 diagnostic codes from the Centers
for Disease Control and Prevention (CDC; fiscal years `r cdc9`) and ICD-9-CM and
ICD-9-PCS codes from the Centers for Medicare and Medicaid Services (CMS; fiscal
years `r cms9`). ICD-10 diagnostic codes from the CDC (calendar years `r cdc10`)
and the World Health Organization (WHO; calendar years `r who10`) are included,
along with ICD-10-CM and ICD-10-PCS codes from CMS (fiscal years `r cms10`). The
internal database will be extended as additional code sets are published.

Three comorbidity algorithms, Charlson
[@charlson1987;@deyo1992adapting;@ahrq2017;@quan2005;@quan2011;@glasheen2019],
Elixhauser [@elixhauser1998;@quan2005;@ahrq2025], and Pediatrics Complex Chronic
Conditions (PCCC) [@feudtner2014pediatric;@feinstein2024pediatric], each with
multiple variants, have been implemented in _medicalcoder_ and are accessible via
the `method` argument of `medicalcoder::comorbidities()`, Table 1.

<!-- for reasons I can't explain, I cannot get the left justfication of the
table to persist in a table div in the docx file.  By keeping it out of a table
div the text justification persists -->

**Table 1:** Comorbidity methods implemented in _medicalcoder_ version `r packageVersion("medicalcoder")`.

| Comorbidity Method                       | Notes                                                                                                   |
| :-----                                   | :---------                                                                                              |
| **Charlson**                             |                                                                                                         |
| &nbsp;&nbsp; `charlson_deyo1992`         | ICD-9 codes defined in Table 1 of [@quan2005]                                                           |
| &nbsp;&nbsp; `charlson_quan2005`         | ICD-9 and ICD-10 defined in Table 1 of [@quan2005]; index scoring as reported in Table 2 of [@quan2011] |
| &nbsp;&nbsp; `charlson_quan2011`         | ICD-9 and ICD-10 defined in Table 1 of [@quan2005]; index scoring as reported in Table 2 of [@quan2011] |
| &nbsp;&nbsp; `charlson_cdmf2019`         | ICD-9 and ICD-10 defined in [@glasheen2019]                                                             |
| **Elixhauser**                           |                                                                                                         |
| &nbsp;&nbsp; `elixhauser_elixhauser1988` | ICD-9 codes defined in Table 2 of [@quan2005]                                                           |
| &nbsp;&nbsp; `elixhauser_ahrq_web`       | ICD-9 codes defined in Table 2 of [@quan2005]                                                           |
| &nbsp;&nbsp; `elixhauser_quan2005`       | ICD-9 and ICD-10 defined in Table 2 of [@quan2005]                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2022`       | ICD-10 codes from [@ahrq2025] for fiscal year 2022                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2023`       | ICD-10 codes from [@ahrq2025] for fiscal year 2023                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2024`       | ICD-10 codes from [@ahrq2025] for fiscal year 2024                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2025`       | ICD-10 codes from [@ahrq2025] for fiscal year 2025                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2026`       | ICD-10 codes from [@ahrq2025] for fiscal year 2026                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq_icd10`     | Any ICD-10 code from all `elixhauser_ahrqYYYY`                                                          |
| **PCCC**                                 |                                                                                                         |
| &nbsp;&nbsp; `pccc_v2.0`                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with [@dewitt2026pccc]                       |
| &nbsp;&nbsp; `pccc_v2.1`                 | Extended set of `pccc_v2.0` pccc and supplements [@feudtner2014pediatric]                               |
| &nbsp;&nbsp; `pccc_v3.0`                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with [@pcccv3sas]                            |
| &nbsp;&nbsp; `pccc_v3.1`                 | Extended set of `pccc_v3.0` codes based on the supplements to [@feinstein2024pediatric]                 |

Development used R version 4.5.2. A comprehensive test suite runs as a GitHub
Action on R oldrelease (4.4.3), release (4.5.2), and devel (what will become
4.6.0). An additional test suite runs on demand for all released versions of R
&geq; 3.5.0 without suggested packages, and for versions &geq; 4.4.0 with
suggested packages[^b].

[^b]: https://github.com/dewittpe/medicalcoder/tree/main/testing

Within the test suite there are tests for ensuring consistency between
`method = "pccc_v2.0"` and the R package _pccc_ [@r-pccc;@feinstein2018pccc],
and between `method = elixhauser_ahrq20XX` and the published SAS code from AHRQ[^d] [@ahrq2025].
As part of the website[^e] build, consistency between
`method = charlson_quan2005` and the SQL code provided by
MIT-LCP/mimic-code/mimic-iv[@johnson2018mimic] is verified.

[^d]: The test directory within the source medicalcoder_<version>.tar.gz
  contains these tests are are checked prior to CRAN submissions.  Detail on
  building the expected output from AHRQ via SAS is found here:
  https://github.com/dewittpe/medicalcoder/blob/main/tests/build-expected-ahrq-results/README.md.

[^e]: https://www.peteredewitt.com/medicalcoder/articles/medicalcoder-vs-mimic.html

We use the MIMIC-IV v3.1 Data
[@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000] as the foundation for
these examples. The MIMIC-IV v3.1 dataset is a large, publicly available
clinical database containing de-identified health data from patients admitted to
intensive care units at Beth Israel Deaconess Medical Center. It includes
detailed information on diagnoses, procedures, laboratory results, medications,
and hospital encounters, with ICD-9-CM and ICD-10-CM/PCS codes available for
billing and comorbidity research. MIMIC-IV is widely used for method
development, validation, and reproducible research in clinical informatics and
outcomes research.

We compare the results of the comorbidity flagging between _medicalcoder_ and
three alternatives, the MIMIC-IV Code released by the MIT Laboratory for
Computational Physilogy (MIT\_LCP) on
GitHub[^f], and the R packages _comorbidity_ [@r-comorbidity] and _pccc_
[@r-pccc].

[^f]: https://github.com/MIT-LCP/mimic-code.git

# Results

Within the MIMIC-IV v3.1 dataset there are
`r qwraps2::frmt(nrow(mimicivDT))` total rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
hospital encounters.
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id)][V1 > 1, qwraps2::frmt(.N)]`
subjects records contain both ICD-9 and ICD-10 codes, and
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id, hadm_id)][V1 > 1, qwraps2::frmt(.N)]`
encounters contain both ICD versions.

A single function, `medicalcoder::comorbidities()`, is the interface for
applying any of `r length(medicalcoder:::comorbidities_methods())` comorbidity
algorithms implemented by _medicalcoder_ (Table 1). The input data are expected
to be in long format, one ICD code per row, with optional columns for
identification variables (e.g., patient and encounter ids), ICD version (integer
value 9 or 10), ICD type (1 for diagnostic code, 0 for procedure code), POA
indicator, and primary diagnosis indicator. The ICD codes can be full, compact,
or a combination thereof. A single patient/encounter record may contain ICD-9,
ICD-10, or both.

A verbose example of calling `medicalcoder::comorbidities()`, including default
values, follows.  Detailed data preperation and additional examples are in the
supplement.

```{r}
#| label: example-call
#| eval: false
medicalcoder::comorbidities(
  data        = mimicivDT,  # object inheriting data.frame
  icd.codes   = "icd_code", # character string name of icd codes in data
  id.vars     = c("subject_id", "enc_seq"),
  icdv.var    = "icd_version",
  dx.var      = "dx",
  poa         = 1L,    # consider all codes present on admission
  age.var     = "age", # in years, only relevant to charlson
  primarydx   = 0L,    # consider all diagnosis codes secondary diagnoses.
  method      = "charlson_quan2005",
  flag.method = "current" # (default) will flag comorbidities within each
                          # unique data[id.vars].  "cumulative", will flag
                          # comorbidities within each unique
                          # data[id.vars[-length(id.vars)]]
                          # sorted by data[id.vars[length(id.vars)]]
)
```

Several function arguments in `medicalcoder::comorbidities()` are provided in
paired forms (e.g., `dx` and `dx.var`, `icdv` and `icdv.var`). These pairs
represent the same conceptual information at different levels of granularity.
Arguments without the `.var` suffix specify dataset-level invariants, such as
when all ICD codes in a dataset are diagnostic or belong to a single ICD
version, whereas arguments with the `.var` suffix reference columns in the input
data that encode row-level variation.

This design avoids requiring users to create constant-valued columns when
dataset-level properties are already known, while still supporting heterogeneous
datasets in which code attributes vary across rows.  When both forms of an
argument are supplied, the `.var` version takes precedence and a warning is
issued to alert the user to the override.

The order the `id.vars` is optional when `flag.method = 'current'`.  Omision
would treat the whole of `data` as a single encounter.  When flagging
comorbidities in longitudinal data use `flag.method =
"cumulative"` and the length of `id.vars` is expected to be at least 2.  The
last entry in `id.vars` is used to sort the data chronologically within the
other `id.vars`.

For the MIMIC-IV data, we use a
generated variable, `enc_seq`, which counts hospital admissions (`hadm_id`) by admission time,
within a `subject_id`.  The `admittime` variable from the MIMIC-IV data could be
used directly for this purpose, but the sorting and joining by a POSIXct
variable requires aditional computational overhead which is eleviated by using
the simple integer value `enc_seq`.

`medicalcoder::comorbidities()` returns a data.frame with addition metadata and
class consistent with the input.  The data.frame has column(s) for the `id.vars`
and an indicator column each of the relevant comorbidities.  All returned
data.frames include columns `num_cmrb` for the number of comorbidities flagged
for the `id.vars` and `cmrb_flag`, an indicator column for at least one
comorbidity.  _medicalcoder_ provides a S3 `summary` method for the returns to
generate tables of the counts and percentage of encounters flagged, Table 2.

**Table 2:** Counts and percentages of encounters flag with Charlson
comorbidities via `medicalcoder::comorbidities()`.
```{r}
#| label: charlson-summary
#| echo: false
#| results: asis
cc <- summary(objs$medicalcoder_charlson_current)$conditions
data.table::setDT(cc)
cc[, condition := NULL]
cc[, count := qwraps2::frmt(as.integer(count))]
cc[, percent := qwraps2::frmt(percent)]
cc[, condition_description := sub(">=", "≥", condition_description)]

qwraps2::qable(cc[, 2:3],
  rgroup = c("Conditions" = 17, "Total Conditions" = 9),
  rnames = cc[["condition_description"]],
  kable_args = list(
    format = "markdown",
    rows.names = FALSE
  )
)
```

When working with longitudinal data, flagging comorbidities by cumulative
sumation of flags over encounters can fail, Table 2. This table focuses on PCCC
v3, but the issue can occur with Elixhauser from the AHRQ post 2022 where
present-on-admission requirements are considered in the flagging of the codes.
In the example shown in Table 2, the subject had ICD codes mapping to
respiratory and miscellaneous comorbidities on encounters 1 and 3 of 13 total
encounters.  Due to the nauces of the PCCC v3 method, those do not flag as
comorbidities since no non-technology-dependent code is also present in the
encounter.  It is not until encounter 4 when a non-technology-dependent code for
a metabolic condition is reported that the respiratory and miscellaneous
conditions are also flagged under the `flag.method = 'cumulative'` paradigm but
not under `flag.method = 'current'`.

```{r}
#| label: flagging-table
#| echo: false
#| results: asis
s10728333 <-
  merge(
    x = objs$medicalcoder_pcccv3.1_current[subject_id == 10728333],
    y = objs$medicalcoder_pcccv3.1_cumulative[subject_id == 10728333],
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_current", "_cumulative")
 )
DT0 <- s10728333[subject_id == 10728333, .SD, .SDcols = patterns("enc_seq|^(met|resp|misc|num_|any_tech)")]
DT1 <-
  DT0[
    ,
    .(enc_seq,
      metabolic__current = data.table::fcase(
        metabolic_dxpr_only_current == 1, "DxPr",
        metabolic_tech_only_current == 1, "Tech",
        metabolic_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      metabolic__cumulative = data.table::fcase(
        metabolic_dxpr_only_cumulative == 1, "DxPr",
        metabolic_tech_only_cumulative == 1, "Tech",
        metabolic_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      misc__current = data.table::fcase(
        misc_dxpr_only_current == 1, "DxPr",
        misc_tech_only_current == 1, "Tech",
        misc_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      misc__cumulative = data.table::fcase(
        misc_dxpr_only_cumulative == 1, "DxPr",
        misc_tech_only_cumulative == 1, "Tech",
        misc_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      respiratory__current = data.table::fcase(
        respiratory_dxpr_only_current == 1, "DxPr",
        respiratory_tech_only_current == 1, "Tech",
        respiratory_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      respiratory__cumulative = data.table::fcase(
        respiratory_dxpr_only_cumulative == 1, "DxPr",
        respiratory_tech_only_cumulative == 1, "Tech",
        respiratory_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      any_tech_dep_current = data.table::fifelse(any_tech_dep_current == 1, "1", ""),
      any_tech_dep_cumulative = data.table::fifelse(any_tech_dep_cumulative == 1, "1", ""),
      num_cmrb_current = as.character(num_cmrb_current),
      num_cmrb_cumulative = as.character(num_cmrb_cumulative)
    )
  ]

ftm1 <- "*"
ftm2 <- "†"
DT1[enc_seq == 1, `:=`(respiratory__current = ftm1, respiratory__cumulative = ftm1)]
DT1[enc_seq %in% 2:3,`:=`(respiratory__cumulative = ftm2)]
DT1[enc_seq == 3, `:=`(misc__current = ftm1, misc__cumulative = ftm1)]

cap <- "Table 2: Flagging of comorbidities under PCCC v3 for MIMIC-IV subject 10728333. Cells marked with DxPr denote that the comorbidity was flagged due to a ICD code that is not technology-dependent.  Tech denotes flagging of a comorbidity base on the presence of at least one technology-dependent ICD code and the presence of at least one non-technology-dependent code for another comorbidity. DxPrTech denotes flagging due to both technology-dependent and non-technology-dependent ICD codes.  Under the Any Tech columns we mark the encounters where some tech dependence is identified. For this patient, on encounter 1, a technology-dependent ICD code for a respiratory comorbidity was found in the subject's record.  Because no non-technology-dependent code was observed on encounter 1, under PCCC v3 this subject has no comorbidities for encounter 1.  This persists for encounters 2 and 3 with respect to respiratory and on encounter 3 for miscellaneous as well.  On encounter 4 a non-technology-dependent code for a metabolic comorbidity was reported.  Under a cumulative flagging paradigm the miscellaneous and respiratory comorbidities are also flagged as the reported codes from encounters 3 and 1 have been carried forward respectively.  Under the 'current' flagging method miscellaneous is never flagged as the technology-dependent code never occurs on the same encounter as the non-technology-dependent code."
DT1 |>
  gt::gt(caption = cap) |>
  gt::cols_align(align = "center") |>
  gt::tab_spanner("Metabolic", columns = c(metabolic__current, metabolic__cumulative)) |>
  gt::tab_spanner("Miscellaneous", columns = c(misc__current, misc__cumulative)) |>
  gt::tab_spanner("Respiratory", columns = c(respiratory__current, respiratory__cumulative)) |>
  gt::tab_spanner("Any Tech", columns = c(any_tech_dep_current, any_tech_dep_cumulative)) |>
  gt::tab_spanner("Comorbidity Count", columns = c(num_cmrb_current, num_cmrb_cumulative)) |>
  gt::cols_label(
    enc_seq = "Encounter",
    metabolic__current = "Current", metabolic__cumulative = "Cumulative",
    misc__current = "Current", misc__cumulative = "Cumulative",
    respiratory__current = "Current", respiratory__cumulative = "Cumulative",
    any_tech_dep_current = "Current", any_tech_dep_cumulative = "Cumulative",
    num_cmrb_current = "Current", num_cmrb_cumulative = "Cumulative"
  ) |>
  gt::tab_footnote(footnote = "* A technology-dependent code was reported on this encounter.  Since no non-technology-dependent was reported no comorbidities are flagged.") |>
  gt::tab_footnote(footnote = "† A technology-dependent code has been carried forward in the record.  Since no non-technology-dependent was reported on this, or a prior, enconter, no comorbidities are flagged.")
```

In the supplement is a detailed exploration of the flags produced by
`medicalcoder::comorbidities()` versus the code from MIMIV-IV, and the R
packages _comorbidity_ [@r-comorbidity] and _pccc_ [@r-pccc].  The flagging of
Charlson comorbidities within the MIMIC-IV v3.1 dataset between _medicalcoder_
and MIMIV-IV Code are the same. Between _medicalcoder_ and _comorbidity_ we
found the same for Elixhauser comorbidities.  For Charlson comorbidities we
found that the _comorbidity_ package failed to flag malignancies related to
ICD-10-CM codes C4A.x.  PCCC v2 flagging between _medicalcoder_ and _pccc_
showed several technology-dependent and transplant errors in the mappings within
_pccc_.

In addition to _medicalcoder_ showing greater sensitivity, _medicalcoder_ 
demonstrated substantial performance advantages in large-scale testing (Figure 1).

```{r}
#| label: figure1
#| include: false
ttc <-
  lapply(objs, attr, "tictoc") |>
  lapply(as.numeric) |>
  data.table::as.data.table() |>
  data.table::melt(measure.vars = names(objs), variable.factor = FALSE)
ttc[, c("Tool", "Algorithm", "Flag Method") := data.table::tstrsplit(variable, "_")]
ttc[, variable := NULL]
g <-
  ggplot2::ggplot(ttc) +
  ggplot2::aes(x = Algorithm, y = value, fill = `Flag Method`) +
  ggplot2::geom_col(
    position = ggplot2::position_dodge2(width = 0.8, preserve = "single"),
    width = 0.7,
    color = "white",
    linewidth = 0.3
    ) +
  ggplot2::facet_wrap(~ Tool, nrow = 1, strip.position = "top") +
  ggplot2::scale_fill_manual(
    name = "Flag Method",
    values = c("current" = "#6F263D", "cumulative" = "#4F8FCB")
  ) +
  ggplot2::scale_y_continuous(breaks = seq(0, max(ttc$value), by = 15)) +
  ggplot2::labs(x = "Comorbidity Algorithm", y = "Time (seconds)") +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    legend.position = "bottom",
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
  )
ggplot2::ggsave(plot = g, filename = "figure1.png", width = 6, height = 4)
ggplot2::ggsave(plot = g, filename = "figure1.svg", width = 6, height = 4)
```

# Discussion

_medicalcoder_ was initially developed to support implementation of Pediatric
Complex Chronic Conditions (PCCC) versions 3 [@feinstein2024pediatric] and 2
[@feudtner2014pediatric]. During this work, it became clear that accurate and
reproducible PCCC classification required access to a comprehensive,
version-aware ICD database, which was subsequently developed and incorporated
into the package. Designing a general framework to support multiple PCCC
versions naturally enabled extension to additional comorbidity algorithms,
including Charlson and Elixhauser.

A primary design requirement was that the package be fully self-contained and
deployable in restricted computing environments, including systems with limited
internet access or tightly controlled software installations. Relying only on
base R ensures that _medicalcoder_ can be installed and used wherever R &geq;
3.5.0 is available, addressing a common but under-discussed barrier to
reproducible clinical research.

Beyond deployment considerations, _medicalcoder_ addresses a substantive
methodological gap in existing comorbidity software: the lack of support for
longitudinal comorbidity assessment. Most available implementations operate
strictly at the encounter level, implicitly assuming that comorbid conditions
are either repeatedly coded or clinically irrelevant once absent from the
current record. As demonstrated in Figures 1 and 2, this assumption can lead to
systematic under-ascertainment of comorbidities, particularly for conditions
governed by POA rules or technology-dependent classification criteria. By
explicitly tracking diagnostic history across encounters, _medicalcoder_ enables
comorbidity flagging that more closely reflects clinical reality.

The unified interface provided by _medicalcoder_ further reduces analytic
complexity. Mixed ICD-9 and ICD-10 datasets, full and compact codes, diagnostic
and procedure codes, and algorithm-specific nuances are handled within a single
function call. This contrasts with existing workflows that require dataset
partitioning, multiple method calls, and post hoc reconciliation, each of which
introduces opportunities for error and irreproducibility.

Extensive testing against published SAS, SQL, and R reference implementations
demonstrates that _medicalcoder_ produces results consistent with established
algorithms while offering additional flexibility.

To our knowledge, _medicalcoder_ is the only available R package that jointly
supports multiple comorbidity algorithms, mixed ICD versions,
present-on-admission logic, and cumulative longitudinal flagging within a
single, reproducible framework. These capabilities enable researchers to
accurately characterize patient complexity and to systematically evaluate how
different comorbidity definitions influence analytic results.

## Conclusions

_medicalcoder_ is a fully self-contained R package with a detailed and robust
internal ICD database that can efficiently apply many comorbidity algorithms to
a dataset while accounting for ICD-version, ICD-type, POA, primary diagnosis,
and full and compact ICD codes. Flagging codes can be done for individual
patient/encounters, or longitudinally within a patient record.

{{< pagebreak >}}

[**Funding Statement:**]{.underline}

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline}
The _medicalcoder_ R package is available from the Comprehensive R Archive Network
(CRAN) at https://cran.r-project.org/package=medicalcoder. Developmental
versions, additional testing suites, benchmarking, and examples are on GitHub at
https://github.com/dewittpe/medicalcoder/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
_medicalcoder_ R package and drafted the manuscript. SR provided critical code
review and testing.  SR, MR, and TDB all contributed to the conception, design,
and interpretation and provided important feedback. TDB provided funding. All
authors critically reviewed the manuscript for important intellectual content
and approved of the final version of the manuscript.

[**Acknowledgments:**]{.underline} We thank Vincent Rubinetti for his assistance
in designing the _medicalcoder_ logo.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

**Figure 1:** The time required to flag comorbidities in the MIMIC-IV v3.1
dataset (
`r qwraps2::frmt(nrow(mimicivDT))` rows,
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects,
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
encounters) by utility and flagging method.  The MIMIIC-IV code time is based on
running RSQLite code for a in-memory database.  R packages _comorbidity_ and
_pccc_ require multiple calls and aggregations to handle the mixed use of ICD-9
and ICD-10 within the MIMIIC-IV dataset.  Applying the each of of the
comorbidity algorithms and flagging methods within _medicalcoder_ requires only
modifying the `method` and `flag.method` argument to the
`medicalcoder::comorbidities()` call.

