---
title: "medicalcoder vs other R packages"


format:
  pdf:
    toc: true

bibliography: references.bib
---

# Objective and Setup

Compare results and time to compute between medicalcoder::comorbidities() and
other implementations of comorbidity algorithms.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE
)
```

```{r}
#| label: namespaces
#| echo: true
#| results: "markup"
# required namespaces
library(medicalcoder)
library(data.table)
library(icdcomorbid)
library(multimorbidity)
library(comorbidity)
library(pccc)
```

For this exercise we will use the `medicalcoder::mdcr` dataset.

```{r}
#| label: data-setup
#| echo: true
#| results: "markup"
mdcrDT <- data.table::copy(medicalcoder::mdcr)
data.table::setDT(mdcrDT)
data.table::setkey(mdcrDT, patid)
str(mdcrDT)
```

# Apply Comorbidity Algorithms

## medicalcoder

Here we apply several comorbidity algorithms to the `mdcrDT` data set via
`medicalcoder::comorbidities()`.  The results are stored, along with a timing
for the work which will be used to compare to other implementations.

We will make several calls to `medicalcoder::comorbidities()` with only the
`method` argument changing.  As such we define a common set of arguments to
simplify the calls.

```{r}
#| label: medicalcoder-setup
call_medicalcoder <- function(method, ...) {
  tic <- Sys.time()
  rtn <-
    medicalcoder::comorbidities(
      data      = mdcrDT,
      id.vars   = "patid",
      icd.codes = "code",
      dx.var    = "dx",
      icdv.var  = "icdv",
      poa       = 1L, # assume all codes are present on admission
      primarydx = 0L, # assume all codes are secondary diagnoses
      method    = method,
      ...
  )
  toc <- Sys.time()
  attr(rtn, "tictoc") <- as.numeric(difftime(toc, tic, units = "secs"))
  rtn
}
```

```{r}
#| label: medicalcoder-calls
medicalcoder_charlson_quan2005 <-
  call_medicalcoder(method = "charlson_quan2005")

medicalcoder_charlson_quan2011 <-
  call_medicalcoder(method = "charlson_quan2011")

medicalcoder_elixhauser_quan2005 <-
  call_medicalcoder(method = "elixhauser_quan2005")

medicalcoder_pccc_v2.0 <-
  call_medicalcoder(method = "pccc_v2.0")
```

## comorbidity

To use the R package comorbidity [@r-comorbidity] we need to split the data by
ICD version, apply `comorbidity::comorbidity()` to each split, and reconcile the
two sets. We restrict the input data to only diagnostic codes for Charlson and
Elixhauser comorbidities.

```{r}
#| label: comorbidity-setup
mdcr_icd9  <- subset(mdcrDT, icdv ==  9L)
mdcr_icd10 <- subset(mdcrDT, icdv == 10L)

# we will add rows without a code for each patid not in the subsets.
mdcr_icd9 <-
  rbind(
    mdcr_icd9,
    data.table(
      patid = setdiff(mdcr$patid, mdcr_icd9$patid),
      icdv = 9L,
      code = "",
      dx = 1L
    )
  )

mdcr_icd10 <-
  rbind(
    mdcr_icd10,
    data.table(
      patid = setdiff(mdcr$patid, mdcr_icd10$patid),
      icdv = 10L,
      code = "",
      dx = 1L
    )
  )
```

Now that the data has been split and prepared, we can apply the comorbidity
algorithms and aggregate the results.  In the calls that follow we include the
subsetting of the data to just diagnostic codes.  While that could (should) have
been done when building the datasets it was left to the following calls because
`medicalcoder::comorbidities()` handles diagnostic and procedure codes by
function argument.

```{r}
#| label: comorbidity-charlson
tic <- Sys.time()
comorbidity_charlson_icd9_results <-
  comorbidity::comorbidity(
    x = subset(mdcr_icd9, dx == 1L),
    id = "patid",
    code = "code",
    map = "charlson_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe comorbidities is also flagged
  )

comorbidity_charlson_icd10_results <-
  comorbidity::comorbidity(
    x = subset(mdcr_icd10, dx == 1L),
    id = "patid",
    code = "code",
    map = "charlson_icd10_quan",
    assign0 = TRUE
  )

# combine the ICD-9 and ICD-10 results into one set
comorbidity_charlson_results <-
  rbind(comorbidity_charlson_icd9_results, comorbidity_charlson_icd10_results)

comorbidity_charlson_results <-
  aggregate(. ~ patid, data = comorbidity_charlson_results, FUN = max)

# add the attributes to the combine set
attributes(comorbidity_charlson_results)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_charlson_icd9_results)[c("class", "variable.labels", "map")]

comorbidity_charlson_results[["score"]] <-
  comorbidity::score(
    x = comorbidity_charlson_results,
    weights = "quan",
    assign0 = TRUE
  )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
comorbidity_charlson_results[["score"]] <-
  as.integer(comorbidity_charlson_results[["score"]])
toc <- Sys.time()

attr(comorbidity_charlson_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))

str(comorbidity_charlson_results)
```

The same work is done for applying the Elixhauser comorbidities save no scoring
is required here.

```{r}
#| label: comorbidity-elixhauser
tic <- Sys.time()
comorbidity_elixhauser_icd9_results <-
  comorbidity::comorbidity(
    x = subset(mdcr_icd9, dx == 1L),
    id = "patid",
    code = "code",
    map = "elixhauser_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe comorbidities is also flagged
  )

comorbidity_elixhauser_icd10_results <-
  comorbidity::comorbidity(
    x = subset(mdcr_icd10, dx == 1L),
    id = "patid",
    code = "code",
    map = "elixhauser_icd10_quan",
    assign0 = TRUE
  )

# combine the ICD-9 and ICD-10 results into one set
comorbidity_elixhauser_results <-
  rbind(comorbidity_elixhauser_icd9_results, comorbidity_elixhauser_icd10_results)

comorbidity_elixhauser_results <-
  aggregate(. ~ patid, data = comorbidity_elixhauser_results, FUN = max)

# add the attributes to the combine set
attributes(comorbidity_elixhauser_results)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_elixhauser_icd9_results)[c("class", "variable.labels", "map")]

toc <- Sys.time()

attr(comorbidity_elixhauser_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))

str(comorbidity_elixhauser_results)
```

## icdcomorbid

icdcomorbid [@r-icdcomorbid] requires the input data to be in a "wide" format.
The package provides a tool for this but it takes considerably more time to run
than alternatives.

```{r}
#| label: icdcomorbid-data-format
#  mdcr_icd9dx_wide <-
#    icdcomorbid::long_to_wide(
#      df = mdcr_icd9dx,
#      idx = "patid",
#      icd_cols = "code"
#    )
#
# killed process after 15 seconds

tic <- Sys.time()

mdcr_icd9dx  <- mdcr_icd9[ dx == 1L][, DX := sprintf("DX%02d", seq_along(code)), by = .(patid)]
mdcr_icd10dx <- mdcr_icd10[dx == 1L][, DX := sprintf("DX%02d", seq_along(code)), by = .(patid)]

mdcr_icd9dx_wide <-
  data.table::dcast(mdcr_icd9dx, patid ~ DX, value.var = "code")

mdcr_icd10dx_wide <-
  data.table::dcast(mdcr_icd10dx, patid ~ DX, value.var = "code")

toc <- Sys.time()

difftime(toc, tic)
```

We now apply the Charlson comorbidity algorithm to the data

```{r}
#| label: icdcomorbid
#| cache: true
tic <- Sys.time()

icdcomorbid_charlson_icd9_results <-
  icdcomorbid::icd9_to_comorbid(
    df = mdcr_icd9dx_wide,
    idx = "patid",
    icd_cols = grep("^DX", names(mdcr_icd9dx_wide), value = TRUE),
    mapping = "charlson9"
  )

icdcomorbid_charlson_icd10_results <-
  icdcomorbid::icd10_to_comorbid(
    df = mdcr_icd10dx_wide,
    idx = "patid",
    icd_cols = grep("^DX", names(mdcr_icd10dx_wide), value = TRUE),
    mapping = "charlson10"
  )

data.table::setDT(icdcomorbid_charlson_icd9_results)
data.table::setDT(icdcomorbid_charlson_icd10_results)

icdcomorbid_charlson_results <-
  rbind(
    icdcomorbid_charlson_icd9_results,
    icdcomorbid_charlson_icd10_results,
    use.names = TRUE,
    fill = TRUE
  )

icdcomorbid_charlson_results <-
  aggregate(
    . ~ patid,
    data = comorbidity_charlson_results,
    FUN = any,
    na.rm = TRUE
  )

toc <- Sys.time()

attr(icdcomorbid_charlson_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))

str(icdcomorbid_charlson_results)
```

## multimorbidity

```{r}
#| label: multimorbidity-charlson
tic <- Sys.time()

multimorbidity_charlson_results <-
  multimorbidity::charlson(
    dat = mdcrDT,
    id  = patid,
    version = 19, # default - both ICD-9 and ICD-10 data
    version_var = icdv
  )
toc <- Sys.time()

attr(multimorbidity_charlson_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))

str(multimorbidity_charlson_results)
```

## pccc

The pccc package [@r-pccc] required the data to be in a wide format and applied
the scoring by ICD version.

```{r}
#| label: oldpccc
#| cache: true
tic <- Sys.time()

mdcr_wide <-
  mdcrDT[][,
    variable := data.table::fcase(
      dx == 0, sprintf("PR%02d", seq_along(code)),
      dx == 1, sprintf("DX%02d", seq_along(code))
    ),
    by = .(patid, icdv)
  ]
mdcr_wide <-
  data.table::dcast(mdcr_wide, patid +icdv ~ variable, value.var = "code")

pccc_results <-
  rbind(
    pccc::ccc(
      data = mdcr_wide[icdv == 9L],
      id = patid,
      dx_cols = grep("^DX\\d+", names(mdcr_wide), value = TRUE),
      pc_cols = grep("^PR\\d+", names(mdcr_wide), value = TRUE),
      icdv = 9
    )
    ,
    pccc::ccc(
      data = mdcr_wide[icdv == 10L],
      id = patid,
      dx_cols = grep("^DX\\d+", names(mdcr_wide), value = TRUE),
      pc_cols = grep("^PR\\d+", names(mdcr_wide), value = TRUE),
      icdv = 10
    )
  )

pccc_results <-
  aggregate(
    . ~ patid,
    data = pccc_results,
    FUN = max,
    na.rm = TRUE
  )

toc <- Sys.time()

attr(pccc_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))
```

# medicalcoder vs...

## comorbidity

## icdcomorbid

## multimorbidity

## pccc

The time required to apply PCCC version 2 [@feudtner2014pediatric] to the data
set is considerably different between `pccc::ccc()` and `medicalcoder::comorbidities()`.
medicalcoder was
`r qwraps2::frmt(attr(pccc_results, "tictoc") / attr(medicalcoder_pccc_v2.0, "tictoc"))`
times faster.

```{r}
#| label: medicalcoder-vs-pccc-timing
attr(medicalcoder_pccc_v2.0, "tictoc")
attr(pccc_results, "tictoc")
```

The speed difference is impressive, but it only matters if the results are the
same or better. We start by checking the flagging for the conditions.

```{r}
#| label: medicalcoder-vs-pccc-results
delta <-
  merge(
    x = medicalcoder_pccc_v2.0,
    y = pccc_results,
    all = TRUE,
    by = "patid",
    suffixes = c("_mdcr", "_pccc")
  )

for (j in setdiff(intersect(names(medicalcoder_pccc_v2.0), names(pccc_results)), "patid")) {
  xj <-sprintf("%s_mdcr", j)
  yj <-sprintf("%s_pccc", j)
  if (identical(delta[[xj]], delta[[yj]])) {
    print(sprintf("medicalcoder and pccc flagged %s the same.", j))
    data.table::set(delta, j = xj, value = NULL)
    data.table::set(delta, j = yj, value = NULL)
  } else {
    print(sprintf("medicalcoder and pccc did not flagged %s the same.", j))
  }
}
```

What remains are columns with differnt names or which were unique to
medicalcoder.  Let's check the technology dependence flag.

```{r}
identical(
  delta[["any_tech_dep"]], # medicalcoder::comorbidities()
  delta[["tech_dep"]]      # pccc::ccc()
)

all(delta[["any_tech_dep"]] >= delta[["tech_dep"]])
```

There are differences between the two methods.  Since the
`medicalcoder::comorbidities()` flag is
always greater or equal to the `pccc::ccc()` flag it means that there are
technology dependece flags that `medicalcoder::comorbidities()` is flagging
conditions when `pccc::ccc()` is not.



```{r}
mdcrDT[, code2 := paste(icdv, dx, code, sep = "_")]
medicalcoder::comorbidities(
  data = mdcrDT[patid %in% delta[any_tech_dep != tech_dep, patid]],
  id.vars = c("patid", "code2"),
  dx.var = "dx",
  icd.code = "code",
  icdv.var = "icdv",
  method = "pccc_v2.0"
)[any_tech_dep == 1, unique(code2)]
```
There are four codes which do not match:

```{r}
#| include: false
subset(
  medicalcoder::get_pccc_codes(),
  full_code %in% c("V45.85", "V56.2", "86.06", "Z94.2")
)

subset(
  medicalcoder::get_icd_codes(with.descriptions = TRUE),
  full_code %in% c("V45.85", "V56.2", "86.06", "Z94.2")
)
```

1. ICD-9-CM V45.85
   * Documented for version 2 as metabolic (devices) but is only listed under metabolic in the software.  Becuase of how `medicalcoder::comorbidities()` is implemented the subconditon (devices) had to be provided in the tables for `method = "pccc_v2.0"`.  This results in the (improved) mapping to technology dependence.  Perhaps more concerning is that the mapping for V45.85 in `pccc::ccc()` maps to a transplant, which is incorrect.  GitHub issue and pull request have been submitted to resolve this issue[^1].
2. ICD-9-CM V56.2
   * Not in the v2 document, but in the software for neuromusc only. This raises a problem that the subcondition is not defined.  To allow for the methods in medicalcoder to work, a non-missing subcondition is needed. device and technology use seems appropriate[^2].
3. ICD-9-PCS 86.06
   * Same issue as ICD-9-CM V48.85[^3]
4. ICD-10-CM Z49.01
   * Same issue as ICD-9-CM V56.2[^4]


[^1]: https://github.com/CUD2V/pccc/issues/49
[^2]: https://github.com/CUD2V/pccc/issues/46
[^3]: https://github.com/CUD2V/pccc/issues/48
[^4]: https://github.com/CUD2V/pccc/issues/55

For transplant we have a similar missmatch between
`medicalcoder::comorbidities()` and `pccc::ccc()`.
```{r}
identical(
  delta[["any_transplant"]],  # medicalcoder::comorbidities()
  delta[["transplant"]]       # pccc::ccc()
)
  
all(delta[["any_transplant"]] >= delta[["transplant"]])

medicalcoder::comorbidities(
  data = mdcrDT[patid %in% delta[any_transplant != transplant, patid]],
  id.vars = c("patid", "code2"),
  dx.var = "dx",
  icd.code = "code",
  icdv.var = "icdv",
  method = "pccc_v2.0"
)[any_transplant == 1, unique(code2)]
```

* ICD-9-CM 
* ICD-10-CM Z94.x

# Nice-to-Haves in medicalcoder

As shown above, medicalcoder applies these common comorbidity algorithms
consistent with other implementations, or with increased precision and
considerably faster as well.  Additionally, medicalcoder has some nice-to-haves
for working with the results from calling `medicalcoder::comorbidities()`.

## summary()

An S3 `summary()` method is available for the `medicalcoder_comorbidities` class
objects returned from `medicalcoder::comorbidities()`.

Below is the structure of the summary for the Charlson, Elixhauser, and PCCC
returns.  Each is slightly different based on the method.

```{r}
#| label: medicalcoder-summary
str( summary(medicalcoder_charlson_quan2005) )
str( summary(medicalcoder_elixhauser_quan2005) )
str( summary(medicalcoder_pccc_v2.0) )
```

We highlight the `conditions` element from `medicalcoder_charlson_quan2005`,
`medicalcoder_elixhauser_quan2005` and the summary from the
`medicalcoder_pccc_v2.0` object.  
These data.frames can be the basis for useful summary tables which can be nicely
formatted via packages like kableExtra [@r-kableExtra].


```{r}
#| label: summary-table-charlson
#| echo: true
#| results: "asis"
x <- summary(medicalcoder_charlson_quan2005)[["conditions"]]
kableExtra::kbl(
  x = x[, c("condition_description", "count", "percent")],
  format = "latex",
  booktabs = TRUE
) |>
kableExtra::pack_rows("Conditions", 1, 17) |>
kableExtra::pack_rows("Total Number of Conditions", 18, nrow(x))
```

```{r}
#| label: old-stuff
#| eval: false


################################################################################
# Charlson 2011 and Elixhauser Quan 2005 via comorbidity

tic <- Sys.time()

# check that medicalcoder::comorbidities() and comorbidity::comorbidity()
# results are the same
charlson_delta <-
  merge(
    x = comorbidity_charlson_results,
    y = medicalcoder_charlson_quan2011,
    all = TRUE,
    by = "patid"
  )

charlson_columns <- data.table::fread(text = "
Condition                          | comorbidity | medicalcoder
AIDS                               | aids        | aidshiv
Cancer (Any malignancy)            | canc        | mal
Cancer (Metastatic solid tumor)    | metacanc    | mst
Cerebrovascular disease            | cevd        | cebvd
Chronic pulmonary disease          | cpd         | copd
Congestive heart failure           | chf         | chf
Dementia                           | dementia    | dem
Hemiplegia or paraplegia           | hp          | hp
Myocardial infarction              | mi          | mi
Peripheral vascular disease        | pvd         | pvd
Peptic ulcer disease               | pud         | pud
Rheumatic disease                  | rheumd      | rhd
Diabetes (uncomplicated)           | diab        | dm
Diabetes (complicated)             | diabwc      | dmc
Renal disease                      | rend        | rnd
Liver disease (mild)               | mld         | mld
Liver disease (moderate or severe) | msld        | msld
Score (based on Quan 2011)         | score       | cci
  ")
data.table::setDF(charlson_columns)

for (i in seq_len(nrow(charlson_columns))) {
  x <- charlson_columns[["comorbidity"]][i]
  y <- charlson_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(identical(charlson_delta[[X]], charlson_delta[[Y]]), list(X = x, Y = y))
  print(e)
  r <- eval(e)
  print(r)
  stopifnot(r)
  charlson_delta[[x]] <- NULL
  charlson_delta[[y]] <- NULL
}

stopifnot(
  identical(
    names(charlson_delta),
    c("patid", "num_cmrb", "cmrb_flag", "age_score")
  )
)

elixhauser_delta <-
  merge(
    x = comorbidity_elixhauser_results,
    y = medicalcoder_elixhauser_quan2005,
    all = TRUE,
    by = "patid"
  )

elixhauser_columns <- data.table::fread(text = "
Condition                                     | comorbidity | medicalcoder
AIDS/HIV                                      | aids        | AIDS
Alcohol abuse                                 | alcohol     | ALCOHOL
Anemias (Blood loss anaemia)                  | blane       | BLDLOSS
Anemias (Deficiency anaemia)                  | dane        | ANEMDEF
Cardiac Arrhythmias                           | carit       | CARDIAC_ARRHYTHMIAS
Cancer (Solid tummor without metastasis)      | solidtum    | TUMOR
Cancer (Metastatic)                           | metacanc    | METS
Chronic pulmonary disease                     | cpd         | CHRNLUNG
Coagulopathy                                  | coag        | COAG
Congestive Heart Failure                      | chf         | CHF
Depression                                    | depre       | DEPRESS
Diabetes (Complicated)                        | diabc       | DMCX
Diabetes (Uncomplicated)                      | diabunc     | DM
Drug abuse                                    | drug        | DRUG
Fluid and electrolyte disorders               | fed         | LYTES
Hypertension (Complicated)                    | hypc        | HTN_CX
Hypertension (Uncomplicated)                  | hypunc      | HTN_UNCX
Hypothyroidism                                | hypothy     | HYPOTHY
Liver disease                                 | ld          | LIVER
Lymphoma                                      | lymph       | LYMPH
Obesity                                       | obes        | OBESE
Other neurological disease                    | ond         | NEURO
Paralysis                                     | para        | PARA
Peptic ulcer disease excluding bleeding       | pud         | ULCER
Peripheral Vascular Disorders                 | pvd         | PERIVASC
Psychoses                                     | psycho      | PSYCH
Pulmonary Circulation Disorders               | pcd         | PULMCIRC
Renal Failure                                 | rf          | RENLFAIL
Rheumatoid artritis/collaged vascular disease | rheumd      | ARTH
Valvular disease                              | valv        | VALVE
Weight loss                                   | wloss       | WGHTLOSS
  ")
data.table::setDF(elixhauser_columns)

# the HTN_C is only in the medicalcoder results
stopifnot(
  identical(
    medicalcoder_elixhauser_quan2005[["HTN_C"]],
    as.integer(
      medicalcoder_elixhauser_quan2005[["HTN_UNCX"]] |
        medicalcoder_elixhauser_quan2005[["HTN_CX"]]
    )
  )
)

for (i in seq_len(nrow(elixhauser_columns))) {
  x <- elixhauser_columns[["comorbidity"]][i]
  y <- elixhauser_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(identical(elixhauser_delta[[X]], elixhauser_delta[[Y]]), list(X = x, Y = y))
  print(e)
  r <- eval(e)
  print(r)
  stopifnot(r)
  elixhauser_delta[[x]] <- NULL
  elixhauser_delta[[y]] <- NULL
}

stopifnot(identical(names(elixhauser_delta), c("patid", "HTN_C", "num_cmrb", "cmrb_flag", "mortality_index", "readmission_index")))


################################################################################
# multimorbidity


# false negatives:
deltas <-
  merge(
    x = medicalcoder_charlson_quan2005,
    y = multimorbidity_charlson_results,
    all = TRUE,
    by.x = "patid",
    by.y = "id"
  )
setDT(deltas)

missing_rheum_codes <-
  merge(
    x = subset(mdcr, patid %in% deltas[rhd != charlson_rheum, patid]),
    y = subset(medicalcoder::get_charlson_codes(), charlson_quan2005 == 1L),
    by = c("code", "icdv", "dx")
  )
missing_rheum_codes <- subset(missing_rheum_codes, condition == "rhd")
missing_rheum_codes <- unique(missing_rheum_codes[, c("icdv", "dx", "code", "full_code")])
setDT(missing_rheum_codes)
missing_rheum_codes[, codeid := 1:.N]

stopifnot(all(
  medicalcoder::comorbidities(
    data = missing_rheum_codes,
    id.vars = "codeid",
    icd.codes = "code",
    icdv.var = "icdv",
    dx.var = "dx",
    method = "charlson_quan2005",
    poa = 1L,
    primarydx = 0L
  )$rhd == 1
))

stopifnot(all(
  multimorbidity::charlson(
    dat = missing_rheum_codes,
    id = codeid,
    version = 19,
    version_var = icdv
  )$charlson_rheum == 0
))

################################################################################
# save
save(
  medicalcoder_charlson_quan2005,
  medicalcoder_charlson_quan2011,
  medicalcoder_elixhauser_quan2005,
  comorbidity_charlson_results,
  comorbidity_elixhauser_results,
  icdcomorbid_icd9_results,
  multimorbidity_charlson_results,
  missing_rheum_codes,
  file = "medicalcoder-vs.Rdata"
)

################################################################################
#                                 End of File                                  #
################################################################################
```

# Session Info

```{r}
#| label: sessioninfo
sessionInfo()
```

# References
