---
title: "medicalcoder: A Unified and Longitudinally Aware Framework for ICD-Based Comorbidity Assessment in R"

code-fold: false
echo: true
number-sections: false

format:
  docx:
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
#| cache: false
stopifnot(packageVersion("medicalcoder") >= "0.8.1")
options(qwraps2_markup = "markdown")
loadNamespace("data.table")
loadNamespace("medicalcoder")
knitr::read_chunk("mimiciv-data-analysis.R")
qwraps2::lazyload_cache_labels(
  labels = c("benchmarking"),
  path = "mimiciv-data-analysis_cache/pdf/"
)
```
```{r}
#| label: import-mimiciv-data
#| include: false
```
```{r}
#| label: build-mimicivdata-ages
#| include: false
```
```{r}
#| label: build-mimicivdata-icd
#| include: false
```
```{r}
#| label: one-data-set
#| include: false
```
```{r}
#| label: build-enc-seq
#| include: false
```


**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; James A. Feinstein, MD, MPH^2,3,4^; Margaret N. Rebull^1^, MA; Tellen D. Bennett^1,5^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA\
^2^ Adult and Child Center for Outcomes Research and Delivery Science, University of Colorado Anschutz Medical Campus, Aurora.\
^3^ Children's Hospital Colorado, Aurora.\
^4^ Department of Pediatrics, University of Colorado Anschutz Medical Campus, Aurora.\
^5^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD\
Department of Biomedical Informatics\
University of Colorado School of Medicine\
1890 N. Revere Court\
Mailstop F600\
Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} 1880 / 2000 (excluding tables)

[**Tables:**]{.underline} 2 / 2

[**Figures:**]{.underline} 3 / 3

[**Supplementary material:**]{.underline}

[**Short running title**]{.underline}: _medicalcoder_: R package for ICD codes and Comorbidities with longitudinal data

<!-- allowed 5; recomend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline} Comorbidity, International Classification of
Diseases (ICD), Software, Longitudinal Studies, Pediatrics

{{< pagebreak >}}

# Abstract {.unnumbered}

Word count: 148 / 150

**Objectives:**
Comorbidity algorithms derived from ICD codes are central to risk adjustment and
cohort characterization. However, existing software inconsistently handles mixed
ICD-9/10 data, and operates at the encounter level, failing to reflect the
longitudinal nature of disease. We developed _medicalcoder_ to address these
limitations.

**Materials and Methods:**
_medicalcoder_ is a R package with an internal ICD database and unified
interface for applying variants of the Charlson, Elixhauser, and Pediatric
Complex Chronic Conditions algorithms. It supports full and compact ICD codes,
mixed ICD versions, present-on-admission and primary diagnosis indicators, and
encounter-level and cumulative longitudinal flagging.

**Results:**
The _medicalcoder_ package requires only R ≥ 3.5.0; no additional packages,
internet access, or external data. Outputs are consistent with published
reference implementations. Longitudinal flagging identifies additional
comorbidity flags under a carry-forward assumption while accounting for
present-on-admission or technology-dependent logic.

**Conclusions:**
_medicalcoder_ provides a unified, reproducible framework for ICD-based
comorbidity assessment in modern clinical datasets.


{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
Word count: 190 / 200

Researchers often use International Classification of Diseases (ICD) diagnosis
and procedure codes to identify health conditions in large medical datasets.
Comorbidity algorithms group these codes into meaningful categories, such as
heart disease or cancer, and are essential for understanding patient complexity,
comparing outcomes, and adjusting for risk. However, existing tools usually
handle only one or two algorithms, work with either ICD-9 or ICD-10 (but rarely
both together), and do not track health conditions across multiple hospital
visits. This can lead to missed or inconsistent results.

We created _medicalcoder_, an R package that solves these problems. It includes
a complete database of ICD-9 and ICD-10 codes and provides a single, easy-to-use
function to apply many comorbidity algorithms. The package can recognize both
full and compact ICD codes, use present-on-admission flags, identify primary
diagnoses, and follow patients across multiple encounters to avoid missing
conditions that appear earlier in their history.

_medicalcoder_ is fast, works with base R, has extensive automated tests, and
has been validated against reference implementations. It helps researchers apply
comorbidity algorithms with consistency, improving the quality and
reproducibility of studies using electronic health records or administrative
data.

:::

{{< pagebreak >}}

# Background and Significance

Accurate identification of clinical conditions in large healthcare datasets
depends on International Classification of Diseases (ICD) codes
[@quan2005;@quan2011;@feinstein2024pediatric;@feudtner2014pediatric;@ahrq2025;@glasheen2019;@ludvigsson2021adaptation;@ludvigsson2023adaptation].
Comorbidity algorithms built on these codes support risk adjustment, cohort
characterization, and reproducible studies of patient populations
[@quan2005;@yurkovich2015systematic;@guo2022evaluation;@li2008risk].

Existing software for flagging comorbidities typically implements one or two
algorithms, most often variants of Charlson and Elixhauser, supports either
ICD-9 or ICD-10 (but not both simultaneously),
and requires manual reconciliation of mixed-version datasets
[@r-comorbidity;@feudtner2014pediatric;@dewitt2026pccc;@pcccv3sas;@r-icdcomorbid;@r-multimorbidity].

Despite the importance of longitudinal data in modern medical
research[@hripcsak2013next;@casey2016using;@jensen2014temporal], comorbidity
software typically operates at the encounter
level[@pcccv3sas;@ahrq2017;@dewitt2026pccc;@ahrq2025;@feinstein2018pccc;@r-comorbidity;@r-icdcomorbid;@r-multimorbidity;@johnson2018mimic;@johnson2022mimiccode].
While the AHRQ [@ahrq2025] Elixhauser variants account for present-on-admission
(POA) indicators and exemptions, they do not propagate conditions across
encounters within a patient record. As a result, conditions that arise during
acute care and are appropriately excluded from the index encounter may be missed
in subsequent encounters unless the relevant ICD codes are explicitly
re-reported and flagged as POA.

This limitation is particularly important given the high prevalence of chronic
disease in the United States, where 76.4% of adults report at least one chronic
condition and 59.5% report two or more[@watson2025trends]. Failing to account
for longitudinal patient history can therefore lead to systematic
under-ascertainment of comorbidities and misclassification of patient risk.

There is a need for a tool that applies multiple comorbidity algorithms through
a unified API while accounting for algorithm-specific nuances, ICD versions and
types, and longitudinal data.

We developed the R package _medicalcoder_ to fill this need.

# Objective

Develop a fully self-contained R package that includes a comprehensive internal
ICD database capable of applying multiple comorbidity algorithms to clinical
datasets, and requires no external dependencies, internet access, or external
data resources. The package should support simultaneous ICD-9 and ICD-10 coding,
accept full and compact ICD codes, incorporate present-on-admission and primary
diagnosis indicators, and enable longitudinal comorbidity assessment across
patient encounters.

# Materials and Methods

_medicalcoder_ was designed to be fully self-contained and deployable in
restricted computing environments with limited or no internet access and tightly
controlled system settings and software. No external namespaces are imported; R
&geq; 3.5.0 is the only required dependency, ensuring compatibility with modern
R serialization formats. Several packages are suggested for enhanced performance
(data.table [@r-data.table] and dplyr [@r-dplyr]) and for testing and
documentation, but are not required for installation or use.

```{r}
#| label: icd-src-years
#| include: false
```

As of _medicalcoder_ version `r packageVersion("medicalcoder")`, the internal
ICD database includes ICD-9 diagnostic codes from the Centers for Disease
Control and Prevention (CDC; fiscal years `r cdc9`) and ICD-9-CM and ICD-9-PCS
codes from the Centers for Medicare and Medicaid Services (CMS; fiscal years `r
cms9`). ICD-10 diagnostic codes from the CDC (calendar years `r cdc10`) and the
World Health Organization (WHO; calendar years `r who10`) are included, along
with ICD-10-CM and ICD-10-PCS codes from CMS (fiscal years `r cms10`)
(TODO: ADD SE AND AM NOTES). The
internal database will be extended as additional code sets are published.

Charlson variants include Deyo (as denoted in [@quan2005]), Quan (2005; 2011),
CDMF (2019) [@glasheen2019], Ludvigsson (2021)
[@ludvigsson2021adaptation;@ludvigsson2023adaptation], Sundararajan (2004)
[@sundararajan2004], and Beyrer (2021) [@beyrer2021validation]. Elixhauser
variants include the original Elixhauser (1988), AHRQ Web (ICD-9 based as
denoted in @quan2005), Quan (2005), and AHRQ ICD-10 fiscal-year implementations
(2022–2026). PCCC variants include versions 2.0, 2.1, 3.0, and 3.1. Additional
variants will be added as they are released.

A comprehensive test suite runs as a GitHub Action and on the Comprehensive R
Archive Network (CRAN). An additional test suite runs on demand for all released
versions of R &geq; 3.5.0 without suggested packages, and for versions &geq;
4.4.0 with suggested packages[^b].

[^b]: https://github.com/dewittpe/medicalcoder/tree/main/testing

The test suite includes validation between `method = "pccc_v2.0"` and the R
package _pccc_ [@dewitt2026pccc;@feinstein2018pccc], and between `method =
elixhauser_ahrq20XX` and the published SAS code from AHRQ[^d] [@ahrq2025]. As
part of the website[^e] build, consistency between `method = charlson_quan2005`
and the SQL code provided by
MIT-LCP/mimic-code/mimic-iv[@johnson2018mimic;@johnson2022mimiccode] is
verified.

[^d]: The test directory within the source medicalcoder_<version>.tar.gz
  contains these tests, which are checked prior to CRAN submissions. Detail on
  building the expected output from AHRQ via SAS is found here:
  https://github.com/dewittpe/medicalcoder/blob/main/tests/build-expected-ahrq-results/README.md.

[^e]: https://www.peteredewitt.com/medicalcoder/articles/medicalcoder-vs-mimic.html

We use the MIMIC-IV v3.1 Data
[@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000] as the foundation for
the following examples and the supplement. The MIMIC-IV v3.1 dataset is a large,
publicly available clinical database used for method development, validation,
and reproducible research in clinical informatics and outcomes research.

We compare the results of the comorbidity flagging between _medicalcoder_ and
three alternatives, the MIMIC-IV Code released by the MIT Laboratory for
Computational Physiology (MIT\_LCP) on GitHub[^f] and
Zenodo[@johnson2018mimic;@johnson2022mimiccode], and the R packages
_comorbidity_ [@r-comorbidity] and _pccc_ [@dewitt2026pccc].

[^f]: https://github.com/MIT-LCP/mimic-code.git

# Results

The MIMIC-IV v3.1 dataset includes
`r qwraps2::frmt(nrow(mimicivDT))` rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
hospital encounters.
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id)][V1 > 1, qwraps2::frmt(.N)]`
subjects' records contain both ICD-9 and ICD-10 codes, and
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id, hadm_id)][V1 > 1, qwraps2::frmt(.N)]`
encounters contain both ICD-9 and ICD-10 codes.

A single function, `medicalcoder::comorbidities()`, is the interface for
applying any of `r length(medicalcoder:::comorbidities_methods())` comorbidity
algorithms implemented by _medicalcoder_, see Figure 1. The input data are expected
to be in long format, one ICD code per row, with optional columns for
identification variables (e.g., patient and encounter ids), ICD version (integer
value 9 or 10), ICD type (1 for diagnostic code, 0 for procedure code), POA
indicator, and primary diagnosis indicator. The ICD codes can be full, compact,
or a combination thereof. A single patient/encounter record may contain ICD-9,
ICD-10, or both.

Several arguments are provided in paired forms (e.g., `dx` and `dx.var`, `icdv`
and `icdv.var`): arguments without the `.var` suffix specify dataset-level
invariants, whereas `.var` arguments reference columns encoding row-level
variation. This avoids creating constant-valued columns when dataset-level
properties are already known while still supporting heterogeneous datasets. When
both forms are supplied, the `.var` form takes precedence and a warning is
issued.

The `id.vars` argument controls how comorbidities are summarized into
encounter-level flags. When flag.method = "current", `id.vars` is optional; if
omitted, the entire dataset is treated as a single encounter. For longitudinal
flagging (`flag.method = "cumulative"`), `id.vars` must include at least two
variables: the leading variables define the patient and the final variable
defines within-patient order for carrying ICD codes forward across encounters.

The ordering variable may be a date/time (e.g., admission time) or a numeric
sequence (e.g., encounter number). In large datasets, using an integer sequence
often reduces overhead relative to sorting and joining on POSIXct timestamps.

`medicalcoder::comorbidities()` returns an object of the same class as the input
data.frame, containing the `id.vars`, one indicator column per comorbidity, and
summary columns `num_cmrb` and `cmrb_flag`. _medicalcoder_ also provides an S3
`summary()` method to generate count and percentage tables (Table 1).

<!--
using the quarto native tbl- for crossref is causing an error in the word doc.
hardcoding the table number as a result
-->

::: {custom-style="tablecap"}
**Table 1:** Counts and percentages of encounters within the MIMIC-IV dataset
flagged with Charlson comorbidities via `medicalcoder::comorbidities()`.
:::

```{r}
#| label: medicalcoder-charlson-current
#| include: false
```
```{r}
#| label: medicalcoder-charlson-cumulative
#| include: false
```
```{r}
#| label: medicalcoder-elixhauser-current
#| include: false
```
```{r}
#| label: medicalcoder-elixhauser-cumulative
#| include: false
```
```{r}
#| label: medicalcoder-pcccv3.1-current
#| include: false
```
```{r}
#| label: medicalcoder-pcccv3.1-cumulative
#| include: false
```
```{r}
#| label: charlson-summary
#| echo: false
#| results: asis
```

Flagging of comorbidities at the encounter-level only assumes all relevant ICD
codes are reported on all encounters and correctly marked as POA.  This is
rarely the case.
Naïve accumulation of encounter-level flags may fail to preserve condition history or severity.

```{r}
#| label: figure2
#| include: false
```
```{r}
#| label: figure2-pros
#| include: false
```
Figure 2 shows the impact on risk profiles for subjects and encounters by
flagging method and algorithm.  For example,
`r figure2pros$N` subjects with at least two encounters and at least one Elixhauser
comorbidity flagged on encounter 2 via `flag.method = "cumulative"`,
`r figure2pros$cs[1]` subjects have at least one encounter with a different risk profile
(different classification of comorbidities and/or count of comorbidities) under
`flag.method = "cumulative"` compared to `flag.method = "current"`.
`r figure2pros$cs[2]` subjects had at least one encounter with zero comorbidities per
`flag.method = "current"` and at least one comorbidity with `flag.method =
'cumulative'`.

The PCCC v3 and post-2022 AHRQ Elixhauser algorithms are more complex. PCCC
v3 conditions depend on whether technology-dependent codes co-occur (or
eventually co-occur) with non–technology-dependent codes (Table 2). Elixhauser
from AHRQ post 2022 account for POA status of ICD codes (example in Supplement.)
Using `flag.method = "cumulative"` ensures the algorithm’s longitudinal logic is
applied consistently.

<!--
using the quarto native tbl- for crossref is causing an error in the word doc.
hardcoding the table number as a result
-->

::: {custom-style="tablecap"}
**Table 2:** Flagging of comorbidities under PCCC v3 for MIMIC-IV subject
10728333. Cells marked with DxPr denote that the comorbidity was flagged due to
an ICD code that is not technology-dependent.  Tech denotes flagging of a
comorbidity based on the presence of at least one technology-dependent ICD
code and the presence of at least one non-technology-dependent code for
another comorbidity. DxPrTech denotes flagging due to both
technology-dependent and non-technology-dependent ICD codes.  Under the Any
Tech columns we mark the encounters where some tech dependence is identified.
For this patient, on encounter 1, a technology-dependent ICD code for a
respiratory comorbidity was found in the subject's record.  Because no
non-technology-dependent code was observed on encounter 1, under PCCC v3 this
subject has no comorbidities for encounter 1.  This persists for encounters 2
and 3 with respect to respiratory and on encounter 3 for miscellaneous as
well.  On encounter 4 a non-technology-dependent code for a metabolic
comorbidity was reported.  Under a cumulative flagging paradigm, the
miscellaneous and respiratory comorbidities are also flagged as the reported
codes from encounters 3 and 1 have been carried forward respectively.  Under
the 'current' flagging method miscellaneous is never flagged as the
technology-dependent code never occurs on the same encounter as the
non-technology-dependent code.
:::

```{r}
#| label: s10728333-manuscript
#| echo: false
#| results: asis
```

The supplement details comparisons with MIMIC-IV code and the R packages
_comorbidity_ [@r-comorbidity] and _pccc_ [@dewitt2026pccc]. Charlson flags matched
MIMIC-IV code, and Elixhauser flags matched _comorbidity_. For Charlson,
_comorbidity_ missed malignancies related to ICD-10-CM codes C4A.x; for PCCC
v2, _pccc_ showed several technology-dependent and transplant mapping errors.

In addition to _medicalcoder_ showing consistent, or better, flagging of
comorbidities compared to other utilities, _medicalcoder_ does so with less
computational expense, Figure 3.  The package GitHub repository has details on
benchmarking by input data class and size.

_medicalcoder_ also provides ICD utility functions for data preparation and
validation. `medicalcoder::get_icd_codes()` exposes the internal ICD database,
and `medicalcoder::is_icd()` validates candidate codes across full and compact
representations and mixed ICD-9/ICD-10 datasets.

# Discussion

_medicalcoder_ was initially developed to support the implementation of Pediatric
Complex Chronic Conditions (PCCC) versions 3 [@feinstein2024pediatric] and 2
[@feudtner2014pediatric]. During this work, it became clear that accurate and
reproducible PCCC classification required access to a comprehensive,
version-aware ICD database, which was subsequently developed and incorporated
into the package. Designing a general framework to support multiple PCCC
versions naturally enabled extension to additional comorbidity algorithms,
including Charlson and Elixhauser.

A primary design requirement was that the package be fully self-contained and
deployable in restricted computing environments, including systems with limited
internet access and/or tightly controlled software installations. Relying only
on base R ensures that _medicalcoder_ can be installed and used wherever R ≥ 3.5.0
is available.  Additionally, R versions and namespace dependency trees are
common barriers to reproducible research.  _medicalcoder_, with no additional
namespace requirements, avoids dependency version conflicts that often undermine
reproducibility, addressing a common but under-discussed barrier to reproducible
clinical research.

Beyond deployment considerations, _medicalcoder_ addresses a substantive
methodological gap in existing comorbidity software: the lack of support for
longitudinal comorbidity assessment. Most available implementations operate
strictly at the encounter level, implicitly assuming that comorbid conditions
are either repeatedly coded or clinically irrelevant once absent from the
current record. As demonstrated in Table 2, this assumption can lead to
systematic under-ascertainment of comorbidities, particularly for conditions
governed by POA rules or technology-dependent classification criteria. By
explicitly tracking diagnostic history across encounters, _medicalcoder_ enables
comorbidity flagging that supports analyses incorporating prior observed
diagnoses under an explicit carry-forward assumption.

The unified interface provided by _medicalcoder_ further reduces analytic
complexity. Mixed ICD-9 and ICD-10 datasets, full and compact codes, diagnostic
and procedure codes, and algorithm-specific nuances are handled within a single
function call. This contrasts with existing workflows that require dataset
partitioning, multiple method calls, and post hoc reconciliation, each of which
introduces opportunities for error and irreproducibility.

The included ICD utility functions also support exploratory and preprocessing
tasks using the same reference data employed by the comorbidity algorithms. This
reduces reliance on ad hoc code lists and external reference tables when
defining cohorts or filtering diagnosis and procedure codes in large clinical
datasets.

Extensive testing against published SAS, SQL, and R reference implementations
demonstrates that _medicalcoder_ produces results consistent with established
algorithms while offering additional flexibility. Automated testing of the
internal and user-facing functions (97% code lines covered as of _medicalcoder_
v0.8.0) ensures consistent and reliable behavior of the package.

To our knowledge, no currently available R package jointly supports multiple
comorbidity algorithms, mixed ICD versions, present-on-admission logic, and
cumulative longitudinal flagging within a single, reproducible framework.
_medicalcoder_ has these capabilities and enables researchers to accurately
characterize patient complexity and to systematically evaluate how different
comorbidity definitions influence analytic results.

## Conclusions

_medicalcoder_ is a fully self-contained R package with a detailed and robust
internal ICD database that can efficiently apply many comorbidity algorithms to
a dataset while accounting for ICD-version, ICD-type, POA, primary diagnosis,
and full and compact ICD codes. Flagging codes can be done for individual
patient encounters, or longitudinally within a patient record.

{{< pagebreak >}}

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline}
The _medicalcoder_ R package is available from the Comprehensive R Archive Network
(CRAN) at https://cran.r-project.org/package=medicalcoder. Developmental
versions, additional testing suites, benchmarking, and examples are on GitHub at
https://github.com/dewittpe/medicalcoder/.

The MIMIC-IV v3.1 data is available from PhysioNet at
https://physionet.org/content/mimiciv/3.1/.

This manuscript was written using Quarto version
`r system("quarto --version", intern = TRUE)`
(https://quarto.org/) and `r R.version.string`. All R code, materials,
and dependencies can be found at
https://github.com/dewittpe/medicalcoder-jamia-open/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
_medicalcoder_ R package and drafted the manuscript. SR provided critical code
review and testing.  JAF, SR, MR, and TDB all contributed to the conception,
design, and interpretation and provided important feedback. TDB provided
funding. All authors critically reviewed the manuscript for important
intellectual content and approved of the final version of the manuscript.

[**Acknowledgments:**]{.underline} We thank Vincent Rubinetti for his assistance
in designing the _medicalcoder_ logo.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

**Figure 1:** Interface for the `medicalcoder::comorbidities()` function. The
function requires a data object containing one ICD code per row, the column
containing ICD codes, and the comorbidity algorithm specification (e.g.,
Charlson or Elixhauser variants). Optional arguments allow users to specify
encounter identifiers, ICD version, diagnosis type, present-on-admission
indicators, patient age, and primary diagnosis flags. Defaults control how
comorbidities are identified, including encounter-level (`flag.method =
"current"`) or longitudinal (`flag.method = "cumulative"`) evaluation and how ICD
codes are matched.

**Figure 1 ALT TEXT:** Diagram showing the argument structure of the R function
`medicalcoder::comorbidities()`. The figure lists three required inputs: a data
frame containing one ICD code per row (data), the column name containing ICD
codes (`icd.codes`), and the comorbidity method specification (`method`, such as
"charlson_quan2005"). Additional optional parameters allow users to specify
encounter identifiers, ICD version (9 or 10), diagnosis versus procedure
indicators, present-on-admission status, patient age (for Charlson algorithms),
and primary diagnosis flags. Default arguments include `flag.method = "current"`
for encounter-level detection, `full.codes = TRUE`, `compact.codes = TRUE`, and
`subconditions = FALSE`.

**Figure 2:** Percent of subjects and encounters in the MIMIC-IV dataset with
different comorbidity profiles under `flag.method = "cumulative"` versus
`flag.method = "current"`. Panel (A) uses subjects with at least two
encounters and at least one comorbidity flagged on encounter 2 as the
denominator. Panel (B) uses non-initial encounters with at least one
comorbidity flagged under `flag.method = "cumulative"` as the denominator.

**Figure 2 ALT TEXT:** Two-panel grouped bar chart comparing how cumulative
longitudinal flagging changes comorbidity results relative to encounter-level
current flagging for three algorithms: Charlson, Elixhauser, and PCCC v3.1.
Panel A shows the percentage of subjects and Panel B shows the percentage of
encounters in four categories: different risk profile, zero encounter-level
comorbidities but at least one longitudinal comorbidity, only differences in
classification, and differences in classification and/or additional
comorbidities. Across both panels, the largest percentages are for different
risk profile and for differences in classification and/or additional
comorbidities, with Elixhauser generally highest, PCCC v3.1 intermediate, and
Charlson lowest; the other two categories are much smaller.

```{r}
#| label: mem-total-helper
#| include: false
```
```{r}
#| label: get-cpu-brand
#| include: false
```
```{r}
#| label: sessioninfo-with-memory-cpu
#| include: false
```

**Figure 3:** The median time (in seconds; bar height), and IQR (error bars),
required to flag comorbidities in the MIMIC-IV v3.1 dataset
(`r qwraps2::frmt(nrow(mimicivDT))` rows,
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects,
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
encounters) by utility (panels) and flagging method for N = 10 iterations.
The MIMIC-IV code time is based on running RSQLite code for an in-memory
database.  R packages _comorbidity_ and _pccc_ require multiple calls and
aggregations to handle the mixed use of ICD-9 and ICD-10 within the MIMIC-IV
dataset.  Applying each of the comorbidity algorithms and flagging methods
within _medicalcoder_ requires only modifying the `method` and `flag.method`
argument to the `medicalcoder::comorbidities()` call.  Some panels are sparse,
illustrating the limited comorbidities method provided by the utility.  The
times were recorded when building the supplement in `r si$platform$version` on
`r si$platform$os` with a `r si$platform$CPU` and `r si$platform$RAM` RAM.

```{r}
#| label: for-figure3-alt-text
#| include: false
```

**Figure 3 ALT TEXT:** Four-panel grouped bar chart showing median time in
seconds to flag comorbidities in the MIMIC-IV v3.1 dataset (N = 10 iterations).
Panels compare four utilities: _comorbidity_, _medicalcoder_, MIMIC-IV code, and
_pccc_. The x-axis lists comorbidity algorithms: Charlson, Elixhauser, PCCC
version 2.0, and PCCC version 3.1. The y-axis is time in seconds from 0 to about
160. Blue bars indicate cumulative flagging and maroon bars indicate current
flagging. Only _medicalcoder_ has cumulative bars as it is the only utility
natively supporting cumulative flagging.  _medicalcoder_ is the fastest in
every comparison shown, with times of about `r mdcr_f3_range[1]` to
`r mdcr_f3_range[2]` seconds across all four algorithms and flag methods.
_comorbidity_ is slower, at about `r cmrb_f3_char_median` seconds for Charlson,
`r cmrb_f3_elix_median` seconds for Elixhauser. The MIMIC-IV code panel shows a
single Charlson result at about `r mmic_f3_median` seconds. The _pccc_ panel
shows only PCCC version 2.0, with about `r pccc_f3_median` seconds for current
flagging.
