---
title: "medicalcoder: An R Package for Working with ICD Codes and Comorbidity Algorithms with Considerations for Longitudinal Data"

code-fold: false
echo: true
number-sections: false

format:
  docx:
    fig-cap-location: top
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
stopifnot(packageVersion("medicalcoder") >= "0.7.0")
```

**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; Margaret N. Rebull^1^, MA; Tellen D. Bennett^1,2^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

^2^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD

Department of Biomedical Informatics

University of Colorado School of Medicine

1890 N. Revere Court

Mailstop F600

Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} 1,992 / 2000 (including R chunks, excluding tables)

[**Tables:**]{.underline} 2 / 2

[**Figures:**]{.underline} 3 / 3

[**Supplementary material:**]{.underline}
* figure1-flag-method-example.R: R script for building Figure 1
* figure2-pcccv3-cumulative-flagging.R: R script for building Figure 2
* medicalcoder-vs-others-r-packages.pdf: detailed comparisons between
  _medicalcoder_ and alternative R packages.

[**Short running title**]{.underline}: medicalcoder: R package for ICD codes and Comorbidities with longitudinal data

<!-- allowed 5; recomend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline} Comorbidity, International Classification of
Diseases (ICD), Software, Longitudinal Studies, Pediatrics

{{< pagebreak >}}

# Abstract {.unnumbered}

Word count: 142 / 150 <!-- Count from MS Word; omit section titles -->

**Objectives**

Comorbidity algorithms derived from International Classification of Diseases
(ICD) codes are central to risk adjustment and cohort characterization, yet
existing software implementations are fragmented, inconsistently handle mixed
ICD-9 and ICD-10 data, and typically ignore longitudinal patient histories.
We present _medicalcoder_ to address these limitations.

**Materials and Methods**
We developed _medicalcoder_, a self-contained R package with an internal ICD
database and a unified interface for applying
`r length(medicalcoder:::comorbidities_methods())` variants of the Charlson,
Elixhauser, and Pediatric Complex Chronic Conditions algorithms. The package
supports full and compact ICD codes, mixed ICD versions, present-on-admission
and primary diagnosis indicators, and both encounter-level and cumulative
longitudinal flagging.

**Results**
Outputs are consistent with published SAS, SQL, and R reference implementations.
_medicalcoder_ only depends on R &geq; 3.5.0 while dynamically leveraging
_data.table_ or _dplyr_ when available, yielding substantial performance
improvements.

**Conclusions**

_medicalcoder_ is a fully self-contained R package that provides a unified,
reproducible framework for applying multiple comorbidity algorithms to modern
clinical datasets. By supporting mixed ICD-9 and ICD-10 data, full and compact
codes, present-on-admission and primary diagnosis indicators, and cumulative
longitudinal flagging, _medicalcoder_ addresses both methodological and
practical limitations of existing implementations.

The ability to propagate clinically relevant conditions across patient histories
improves comorbidity ascertainment in longitudinal data and reduces reliance on
ad hoc preprocessing or repeated method calls. Combined with extensive
validation against published reference implementations and efficient performance
at scale, _medicalcoder_ enables accurate, transparent, and reproducible
comorbidity assessment across a wide range of research and deployment
environments.

{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
Word count: 187 / 200

Researchers often use International Classification of Diseases (ICD) diagnosis
and procedure codes to identify health conditions in large medical datasets.
Comorbidity algorithms group these codes into meaningful categories, such as
heart disease or cancer, and are essential for understanding patient complexity,
comparing outcomes, and adjusting for risk. However, existing tools usually
handle only one or two algorithms, work with either ICD-9 or ICD-10 (but not
both together), and do not track health conditions across multiple hospital
visits. This can lead to missed or inconsistent results.

We created _medicalcoder_, an R package that solves these problems. It includes
a complete database of ICD-9 and ICD-10 codes and provides a single, easy-to-use
function to apply many comorbidity algorithms. The package can recognize both
full and compact ICD codes, use present-on-admission flags, identify primary
diagnoses, and follow patients across multiple encounters to avoid missing
conditions that appear earlier in their history.

_medicalcoder_ is fast, works with base R, and has been thoroughly tested against
published methods. It helps researchers apply comorbidity algorithms more
accurately and consistently, improving the quality and reproducibility of
studies using electronic health record or administrative data.

:::

{{< pagebreak >}}

# Background and Significance

Accurate identification of clinical conditions in large healthcare datasets
depends on International Classification of Diseases (ICD) codes
[@quan2005;@quan2011;@feinstein2024pediatric;@feudtner2014pediatric;@ahrq2025].
Comorbidity algorithms built on these codes support risk adjustment, cohort
characterization, and reproducible study of patient populations
[@quan2005;@yurkovich2015systematic;@guo2022evaluation;@li2008risk].

Existing software packages typically implement one or two comorbidity
algorithms, most often variants of Charlson and Elixhauser, and support either
ICD-9 or ICD-10, but not both
[@r-comorbidity;@feudtner2014pediatric;@r-pccc;@pcccv3sas;@r-icdcomorbid;@r-multimorbidity].
This leaves users to reconcile mixed ICD-version datasets manually, and input
requirements vary across utilities, making it difficult to apply multiple
algorithms consistently.

To our knowledge, no current implementation accounts for longitudinal data.
While the AHRQ [@ahrq2025] Elixhauser variant accounts for present-on-admission
(POA) indicators and exemptions, it does not propagate conditions across
encounters within a patient record. Conditions that arise during acute care and
should not be considered comorbidities for the present encounter are handled via
POA indicators. However, unless the relevant ICD codes are flagged as POA on
subsequent encounters, those comorbidities will not be flagged.

There is a need for a tool that applies multiple comorbidity algorithms through
a unified API while accounting for algorithm-specific nuances, ICD versions and
types, and longitudinal data.

We present the R package _medicalcoder_ to fill this need.

# Objective

Develop a fully self-contained R package with a comprehensive internal ICD
database capable of applying multiple comorbidity algorithms to clinical
datasets. The package should support simultaneous ICD-9 and ICD-10 coding,
accept full and compact ICD codes, incorporate present-on-admission and primary
diagnosis indicators, and enable longitudinal comorbidity assessment across
patient encounters.

# Materials and Methods

_medicalcoder_ was designed to be fully self-contained and deployable in
restricted computing environments. No external namespaces are imported; R &geq;
3.5.0 is the only required dependency, reflecting changes in data serialization
introduced in that version[^a]. Several packages are suggested for enhanced
performance (data.table [@r-data.table] and dplyr [@r-dplyr]) and for testing
and documentation, but are not required for installation or use.

[^a]: https://cran.r-project.org/src/base/NEWS.3#:~:text=R%20has%20new,to%20version%203.5.0

```{r}
#| label: icd-src-years
#| include: false
aggregate(cbind(known_start, known_end) ~ src + icdv + dx,
  data = medicalcoder::get_icd_codes(),
  FUN = range
  )

cms9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))
cdc9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cdc10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cms10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))

who10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "who"),
    paste(min(known_start), max(known_end), sep = " - "))
```

The internal ICD database consists of ICD-9 diagnostic codes from the Centers
for Disease Control and Prevention (CDC; fiscal years `r cdc9`) and ICD-9-CM and
ICD-9-PCS codes from the Centers for Medicare and Medicaid Services (CMS; fiscal
years `r cms9`). ICD-10 diagnostic codes from the CDC (calendar years `r cdc10`)
and the World Health Organization (WHO; calendar years `r who10`) are included,
along with ICD-10-CM and ICD-10-PCS codes from CMS (fiscal years `r cms10`). The
internal database will be extended as additional code sets are published.

Three comorbidity algorithms, Charlson
[@charlson1987;@deyo1992adapting;@ahrq2017;@quan2005;@quan2011;@glasheen2019],
Elixhauser [@elixhauser1998;@quan2005;@ahrq2025], and Pediatrics Complex Chronic
Conditions (PCCC) [@feudtner2014pediatric;@feinstein2024pediatric], each with
multiple variants, have been implemented in _medicalcoder_ and are accessible via
the `method` argument of `medicalcoder::comorbidities()`, Table 1.

<!-- for reasons I can't explain, I cannot get the left justfication of the
table to persist in a table div in the docx file.  By keeping it out of a table
div the text justification persists -->

**Table 1:** Comorbidity methods implemented in _medicalcoder_ version `r packageVersion("medicalcoder")`.

| Comorbidity Method                       | Notes                                                                                                   |
| :-----                                   | :---------                                                                                              |
| **Charlson**                             |                                                                                                         |
| &nbsp;&nbsp; `charlson_deyo1992`         | ICD-9 codes defined in Table 1 of [@quan2005]                                                           |
| &nbsp;&nbsp; `charlson_quan2005`         | ICD-9 and ICD-10 defined in Table 1 of [@quan2005]; index scoring as reported in Table 2 of [@quan2011] |
| &nbsp;&nbsp; `charlson_quan2011`         | ICD-9 and ICD-10 defined in Table 1 of [@quan2005]; index scoring as reported in Table 2 of [@quan2011] |
| &nbsp;&nbsp; `charlson_cdmf2019`         | ICD-9 and ICD-10 defined in [@glasheen2019]                                                             |
| **Elixhauser**                           |                                                                                                         |
| &nbsp;&nbsp; `elixhauser_elixhauser1988` | ICD-9 codes defined in Table 2 of [@quan2005]                                                           |
| &nbsp;&nbsp; `elixhauser_ahrq_web`       | ICD-9 codes defined in Table 2 of [@quan2005]                                                           |
| &nbsp;&nbsp; `elixhauser_quan2005`       | ICD-9 and ICD-10 defined in Table 2 of [@quan2005]                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2022`       | ICD-10 codes from [@ahrq2025] for fiscal year 2022                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2023`       | ICD-10 codes from [@ahrq2025] for fiscal year 2023                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2024`       | ICD-10 codes from [@ahrq2025] for fiscal year 2024                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2025`       | ICD-10 codes from [@ahrq2025] for fiscal year 2025                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq2026`       | ICD-10 codes from [@ahrq2025] for fiscal year 2026                                                      |
| &nbsp;&nbsp; `elixhauser_ahrq_icd10`     | Any ICD-10 code from all `elixhauser_ahrqYYYY`                                                          |
| **PCCC**                                 |                                                                                                         |
| &nbsp;&nbsp; `pccc_v2.0`                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with [@dewitt2026pccc]                       |
| &nbsp;&nbsp; `pccc_v2.1`                 | Extended set of `pccc_v2.0` pccc and supplements [@feudtner2014pediatric]                               |
| &nbsp;&nbsp; `pccc_v3.0`                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with [@pcccv3sas]                            |
| &nbsp;&nbsp; `pccc_v3.1`                 | Extended set of `pccc_v3.0` codes based on the supplements to [@feinstein2024pediatric]                 |

Development used R version 4.5.2. A comprehensive test suite runs as a GitHub
Action on R oldrelease (4.4.3), release (4.5.2), and devel (what will become
4.6.0). An additional test suite runs on demand for all released versions of R
&geq; 3.5.0 without suggested packages, and for versions &geq; 4.4.0 with
suggested packages[^b].

[^b]: https://github.com/dewittpe/medicalcoder/tree/main/testing

# Results {#sec-results}

There are two conceptual groups of functions within medicalcoder: ICD utilities
and comorbidity flagging tools.

The internal ICD database is accessible via `medicalcoder::get_icd_codes()`,
with optional descriptions and hierarchy. `medicalcoder::is_icd()` tests if a
character string is a known ICD code while considering ICD version, type,
source, year, and whether header codes can be considered valid. Both full and
compact codes can be considered simultaneously, and warnings for ambiguous codes
are optionally thrown.  The `"icd"` vignette provides
detailed usage and examples (see supplement).

A single function, `medicalcoder::comorbidities()`, is the interface for
applying any of `r length(medicalcoder:::comorbidities_methods())` comorbidity
algorithms implemented by _medicalcoder_ (Table 1). The input data are expected
to be in long format, one ICD code per row, with optional columns for
identification variables (e.g., patient and encounter ids), ICD version (integer
value 9 or 10), ICD type (1 for diagnostic code, 0 for procedure code), POA
indicator, and primary diagnosis indicator. The ICD codes can be full, compact,
or a combination thereof. A single patient/encounter record can contain ICD-9,
ICD-10, or both.

A verbose example of calling `medicalcoder::comorbidities()`, including default
values, follows.  The example uses one of two example datasets provided by the
package. `mdcr` is a data.frame with 4 columns: patient identifier (patid;
`r length(unique(medicalcoder::mdcr$patid))` unique ids), ICD
version (icdv; ICD-9 and ICD-10, some patid have both),
compact ICD code string (code), and an indicator for diagnostic
or procedure code (dx).

```{r}
#| label: comorbidities-example-1
data(mdcr, package = "medicalcoder") # load data

elixhauser_quan2005_comorbidities <-
  medicalcoder::comorbidities(
    data          = mdcr,  # a data.frame
    icd.codes     = "code", # name of column in data with ICD code strings
    method        = "elixhauser_quan2005", # Comorbidity method_variant
    id.vars       = "patid", # (optional) id variables within data
    icdv.var      = "icdv",  # (optional) variable denoting ICD version
    icdv          = NULL, # (default)
    dx.var        = "dx", # (default is NULL)
                          # variable in data marking each data[[icd.codes]]
                          # as a diagnostic (1) or a procedure (0) code
    dx            = NULL, # (default)
    poa.var       = NULL, # (default)
    poa           = 1L,   # (default is NULL)
                          # Consider all codes to be present-on-admission
    age.var       = NULL, # (default) only relevant to charlson_<variant>
    primarydx.var = NULL, # (default)
    primarydx     = 0L,   # (default is NULL)
                          # Consider all codes to be secondary diagnoses
    flag.method   = "current",  # (default), other option "cumulative"
    full.codes    = TRUE, # (default) Reference data[[icd.codes]] against
                          # full ICD codes in lookup tables
    compact.codes = TRUE, # (default) Reference data[[icd.codes]] against
                          # compact ICD codes in lookup tables
    subconditions = FALSE # (default) only relevant to pccc_<variant>
  )
```

Several function arguments in `medicalcoder::comorbidities()` are provided in
paired forms (e.g., `dx` and `dx.var`, `icdv` and `icdv.var`). These pairs
represent the same conceptual information at different levels of granularity.
Arguments without the `.var` suffix specify dataset-level invariants, such as
when all ICD codes in a dataset are diagnostic or belong to a single ICD
version, whereas arguments with the `.var` suffix reference columns in the input
data that encode row-level variation.

This design avoids requiring users to create constant-valued columns when
dataset-level properties are already known, while still supporting heterogeneous
datasets in which code attributes vary across rows.  When both forms of an
argument are supplied, the `.var` version takes precedence and a warning is
issued to alert the user to the override. The same pattern is applied
consistently across diagnostic and procedure indicators, ICD version,
present-on-admission status, and primary diagnosis flags.

`medicalcoder::comorbidities()` returns a data.frame with addition metadata and
class consistent with the input.  The data.frame has column(s) for the `id.vars`
and an indicator column each of the relevant comorbidities.  All returned
data.frames include columns `num_cmrb` for the number of comorbidities flagged
for the `id.vars` and `cmrb_flag`, an indicator column for at least one
comorbidity.

Flagging comorbidities depends on the combination of `flag.method` and `poa.var`
or `poa`. An example patient record with six encounters and varying POA status
is used to illustrate encounter and cumulative flagging (Figure 1).

```{r}
#| label: load-figure2permutations
#| include: false
figure2permutations <- readRDS("figure2permutations.rds")
```

The longitudinal methods in _medicalcoder_ are critical for PCCC v3
[@feinstein2024pediatric] were some some ICD codes are classified
as technology dependent. A technology-dependent code flags a condition only if
at least one non-technology-dependent code is also present. For example,
consider three ICD-10 codes: H49.811 (non-technology-dependent metabolic),
J84.111 (non-technology-dependent respiratory), and Z96.41
(technology-dependent metabolic). For a patient with seven encounters and one
code reported on encounters 2, 4, and 6, we apply `method = "pccc_v3.1"` with
`flag.method = "cumulative"` and codes treated as POA for all six permutations
`r sprintf("(%s).", paste(sort(unique(figure2permutations$plabel)), collapse = "; "))`
Figure 2 illustrates when conditions are flagged for each permutation. For
permutations 1, 2, and 3, the non-technology-dependent metabolic code appears
before the technology-dependent metabolic code, so the metabolic condition flags
immediately after H49.811 appears. Permutations 4, 5, and 6 are more interesting
because Z96.41 appears first. Because `medicalcoder::comorbidities()` tracks all
ICD codes from prior encounters, the metabolic technology-dependent-only flag
appears in permutations 4 and 6 due to the earlier respiratory
non-technology-dependent code. Encounter-level flagging with post hoc
aggregation would miss metabolic flags on encounters 4 and 5 in permutations 4
and 6.

### Testing

Within the testing suite there are tests for ensuring consistency between
`method = "pccc_v2.0"` and the R package _pccc_ [@r-pccc;@feinstein2018pccc],
and between `method = elixhauser_ahrq20XX` and the published SAS code from AHRQ [@ahrq2025].
As part of the website[^d] build, consistency between
`method = charlson_quan2005` and the SQL code provided by
MIT-LCP/mimic-code/mimic-iv[@johnson2018mimic] is verified.

[^d]: https://www.peteredewitt.com/medicalcoder/articles/medicalcoder-vs-mimic.html

### Benchmarking

```{r}
#| label: benchmarking-results
#| include: false
mimicivbenchmark <- readRDS("benchmark.rds")
data.table::setDT(mimicivbenchmark)

best_case_benchmark <-
  mimicivbenchmark[
    subjects == 1e6 &
      encounters == max(encounters) &
      data_class == "data.table",
    .(relative_time = sprintf("%s%%", qwraps2::frmt(100 * min(relative_time), digits = 1))),
    keyby = .(data_class, subjects, encounters)
    ]

bnch <-
  mimicivbenchmark[
    encounters == max(encounters)
    ,
    unique(.SD)
    ,
    .SDcols = c("data_class", "method", "subconditions", "encounters", "subjects", "flag.method", "time_smooth", "relative_time", "max_rss_kib_smooth")
    ]

bnch_encounters <- unique(bnch$encounters)
bnch_subjects   <- unique(bnch$subjects)
stopifnot(length(bnch_encounters) == 1L)
stopifnot(length(bnch_subjects) == 1L)
bnch[, encounters := NULL]
bnch[, subjects := NULL]

bnch <-
  data.table::melt(
    data = bnch,
    measure.vars = c("time_smooth", "relative_time", "max_rss_kib_smooth")
  )

bnch[, flag.method := factor(flag.method, levels = c("current", "cumulative"))]
bnch[, data_class := factor(data_class, levels = c("data.frame", "tibble", "data.table"))]
bnch[, method := data.table::fcase(method == "charlson_quan2005", "Charlson (Quan 2005)",
  method == "elixhauser_quan2005", "Elixhauser (Quan 2005)",
  method == "pccc_v3.1" & !subconditions, "PCCC 3.1 (without subconditions)",
  method == "pccc_v3.1" & subconditions, "PCCC 3.1 (with subconditions)")
  ]
bnch[, subconditions := NULL]

data.table::setkey(bnch, method, flag.method, variable)

bnch[,
  value := data.table::fcase(
    variable == "time_smooth", formatC(      value,      digits = 1, format = "f"),
    variable == "relative_time" & data_class == "data.frame", "",
    variable == "relative_time", paste0(formatC(100 * value,      digits = 1, format = "f"), "%"),
    variable == "max_rss_kib_smooth", formatC(value / (1024^2), digits = 1, format = "f")
  )]
bnch[, variable := factor(variable,
  levels = c("time_smooth", "relative_time", "max_rss_kib_smooth"),
  labels = c("Average Time (seconds)", "Relative time (to data.frame)", "Max Memory (GiB)")
  )]

bnch <-
  data.table::dcast(
    data = bnch,
    formula = method + flag.method + variable ~ data_class,
    value.var = "value"
  )

bnchmat <- as.matrix(bnch[, .(data.frame, tibble, data.table)])
```

Benchmarking `medicalcoder::comorbidities()` was done by resampling the MIMIC-IV
demo dataset [@PhysioNet-mimic-iv-demo-2.2] and applying the Charlson (Quan
2005), Elixhauser (Quan 2005), and PCCC v3.1 with and without subconditions, to
the data on a computational server running Ubuntu 24.04.3 LTS, R 4.5.1, Intel
Xeon Gold 6416H CPUs, and 2 TB RAM. All processes limited to a single thread.

**Table 2:** Benchmarking `medicalcoder::comorbidities()` by data class and
flagging method based on a resampling the MIMIC-IV demonstration dataset
[@PhysioNet-mimic-iv-demo-2.2] to generate a data set of
`r qwraps2::frmt(as.integer(bnch_subjects))` subjects, and a total of
`r qwraps2::frmt(as.integer(bnch_encounters))` encounters.  The benchmarking was performed on a
computational server running Ubuntu 24.04.3 LTS, R 4.5.1, Intel Xeon Gold 6416H
CPUs, and 2 TB RAM. All processes limited to a single thread.  Figure 3
illustrates the expected time, relative time, and memory use for other
encounter sizes. Extensive details on the benchmarking are on GitHub[^c].

[^c]: https://github.com/dewittpe/medicalcoder/tree/main/benchmarking


|                                                        | data.frame         | tibble             | data.table         |
| :----------                                            | ----:              | ----:              | ----:              |
| __**Charlson**__                                       |                    |                    |                    |
| &nbsp;&nbsp; **flag.method = "current"**               |                    |                    |                    |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (Seconds)                | `r bnchmat[ 1, 1]` | `r bnchmat[ 1, 2]` | `r bnchmat[ 1, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (relative to data.frame) | `r bnchmat[ 2, 1]` | `r bnchmat[ 2, 2]` | `r bnchmat[ 2, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Memory (GiB)                  | `r bnchmat[ 3, 1]` | `r bnchmat[ 3, 2]` | `r bnchmat[ 3, 3]` |
| &nbsp;&nbsp; **flag.method = "cumulative"**            |                    |                    |                    |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (Seconds)                | `r bnchmat[ 4, 1]` | `r bnchmat[ 4, 2]` | `r bnchmat[ 4, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (relative to data.frame) | `r bnchmat[ 5, 1]` | `r bnchmat[ 5, 2]` | `r bnchmat[ 5, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Memory (GiB)                  | `r bnchmat[ 6, 1]` | `r bnchmat[ 6, 2]` | `r bnchmat[ 6, 3]` |
| __**Elixhauser**__                                     |                    |                    |                    |
| &nbsp;&nbsp; **flag.method = "current"**               |                    |                    |                    |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (Seconds)                | `r bnchmat[ 7, 1]` | `r bnchmat[ 7, 2]` | `r bnchmat[ 7, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (relative to data.frame) | `r bnchmat[ 8, 1]` | `r bnchmat[ 8, 2]` | `r bnchmat[ 8, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Memory (GiB)                  | `r bnchmat[ 9, 1]` | `r bnchmat[ 9, 2]` | `r bnchmat[ 9, 3]` |
| &nbsp;&nbsp; **flag.method = "cumulative"**            |                    |                    |                    |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (Seconds)                | `r bnchmat[10, 1]` | `r bnchmat[10, 2]` | `r bnchmat[10, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (relative to data.frame) | `r bnchmat[11, 1]` | `r bnchmat[11, 2]` | `r bnchmat[11, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Memory (GiB)                  | `r bnchmat[12, 1]` | `r bnchmat[12, 2]` | `r bnchmat[12, 3]` |
| __**PCCC v3.1 (with subconditions)**__                 |                    |                    |                    |
| &nbsp;&nbsp; **flag.method = "current"**               |                    |                    |                    |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (Seconds)                | `r bnchmat[13, 1]` | `r bnchmat[13, 2]` | `r bnchmat[13, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (relative to data.frame) | `r bnchmat[14, 1]` | `r bnchmat[14, 2]` | `r bnchmat[14, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Memory (GiB)                  | `r bnchmat[15, 1]` | `r bnchmat[15, 2]` | `r bnchmat[15, 3]` |
| &nbsp;&nbsp; **flag.method = "cumulative"**            |                    |                    |                    |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (Seconds)                | `r bnchmat[16, 1]` | `r bnchmat[16, 2]` | `r bnchmat[16, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (relative to data.frame) | `r bnchmat[17, 1]` | `r bnchmat[17, 2]` | `r bnchmat[17, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Memory (GiB)                  | `r bnchmat[18, 1]` | `r bnchmat[18, 2]` | `r bnchmat[18, 3]` |
| __**PCCC v3.1 (without subconditions)**__              |                    |                    |                    |
| &nbsp;&nbsp; **flag.method = "current"**               |                    |                    |                    |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (Seconds)                | `r bnchmat[19, 1]` | `r bnchmat[19, 2]` | `r bnchmat[19, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (relative to data.frame) | `r bnchmat[20, 1]` | `r bnchmat[20, 2]` | `r bnchmat[20, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Memory (GiB)                  | `r bnchmat[21, 1]` | `r bnchmat[21, 2]` | `r bnchmat[21, 3]` |
| &nbsp;&nbsp; **flag.method = "cumulative"**            |                    |                    |                    |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (Seconds)                | `r bnchmat[22, 1]` | `r bnchmat[22, 2]` | `r bnchmat[22, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Time (relative to data.frame) | `r bnchmat[23, 1]` | `r bnchmat[23, 2]` | `r bnchmat[23, 3]` |
| &nbsp;&nbsp;&nbsp;&nbsp; Memory (GiB)                  | `r bnchmat[24, 1]` | `r bnchmat[24, 2]` | `r bnchmat[24, 3]` |

### Comparing to Other R packages

Comparisons with existing R packages, including
_multimorbidity_ [@r-multimorbidity],
_icdcomorbid_ [@r-icdcomorbid], _comorbidity_ [@r-comorbidity], and 
_pccc_ [@dewitt2026pccc;@feinstein2018pccc], showed that _medicalcoder_ produces
results consistent with published reference implementations while requiring
fewer preprocessing steps and method calls.  In several cases, alternative
packages exhibited false negative or false positive flags due to differences in
ICD handling.  Detailed comparisons, including ICD-code-level examples and
performance measurements, are provided in the supplements.

<!--
Comparing _medicalcoder_ to other R packages has been done (see supplements).
_medicalcoder_ has more features, more comorbidity methods, and greater
flexibility than other packages. Notable alternatives at the time this
manuscript was written are _multimorbidity_ [@r-multimorbidity],
_icdcomorbid_ [@r-icdcomorbid], and _comorbidity_ [@r-comorbidity] for Charlson and
Elixhauser comorbidities and the _pccc_ [@dewitt2026pccc;@feinstein2018pccc]
package for PCCC (version 2).
-->

```{r}
#| label: vs
#| include: false
vs <- new.env()
load(file = "medicalcoder-vs.Rdata", envir = vs)
```

<!--
The R package _comorbidity_[@r-comorbidity] is a widely referenced and downloaded
package that can apply Charlson and Elixhauser (Quan 2005 and Quan 2011), and
others.
We compared `medicalcoder::comorbidities()` to `comorbidity::comorbidity()`
using the `medicalcoder::mdcr` dataset.
We found the flags and index scores were consistent between the two packages
(see supplement).

For these two methods, the only notable differences between the two packages is
that only one call to `medicalcoder::comorbidities()` for each comorbidity
algorithm was needed.  For `comorbidity::comorbidity()`, two calls, one for each
ICD version, was needed per comorbidity algorithm.  The additional call and
aggregation of the results required more coding and execution time.

`comorbidity::comorbidity()` required
`r round(attr(vs$comorbidity_charlson_results, "tictoc"), 1)` seconds to apply Charlson, and
`r round(attr(vs$comorbidity_elixhauser_results, "tictoc"), 1)` seconds to apply
Elixhauser.  Compared to `medicalcoder::comorbidities()` taking
`r round(attr(vs$medicalcoder_charlson_quan2005, "tictoc"), 2)` seconds to apply
`charlson_quan2005`, and
`r round(attr(vs$medicalcoder_elixhauser_quan2005, "tictoc"), 2)` seconds to apply
`elixhauser_quan2005`.

The supplement provides additional detail on the comparisons between
`medicalcoder::comorbidities()` and the packages _icdcomorbid_, and
_multimorbidity._  We found both _icdcomorbid_ and _multimorbidity_ had several false
negative flags for Charlson comorbidities, examples: ICD-10 P29.0 should flag for
congestive heart failure, but fails to do so in _icdcomorbid_ and ICD-10 I27.81
and I27.82 failed to flag for chronic pulmonary disease.  Additionally,
_icdcomorbid_ had false positives (P29.30), and _multimorbidity_ returned lower and
higher severity levels of conditions such as diabetes. (_medicalcoder_ only
returns the highest level, `comorbidity::comorbidity()` provides an option for
end users to control the behavior).

In addition to the greater sensitivity and specificity,
`medicalcoder::comorbidities()` is considerably faster that _icdcomorbid_,
`r attr(vs$icdcomorbid_charlson_results, "tictoc")` seconds to apply Charlson,
and _multimorbidity_,
`r attr(vs$multimorbidity_charlson_results, "tictoc")` seconds.

For the PCCC (version 2) comparison between `medicalcoder::comorbidities()`
and `pccc::ccc()` showed a handful of false negatives for `pccc::ccc()` with
respect to technology dependence and transplantation.  For the primary
conditions, the results between `medicalcoder::comorbidities()` with `method =
"pccc_v2.0"` and `pccc::ccc()` are consistent.  Again,
`medicalcoder::comorbidities()` is considerably faster than the competitor;
`r round(attr(vs$medicalcoder_pccc_v2.0, "tictoc"), 2)` seconds vs
`r round(attr(vs$pccc_results, "tictoc"), 1)` seconds, respectively.
-->

# Discussion

_medicalcoder_ was initially developed to support implementation of Pediatric
Complex Chronic Conditions (PCCC) versions 3 [@feinstein2024pediatric] and 2
[@feudtner2014pediatric]. During this work, it became clear that accurate and
reproducible PCCC classification required access to a comprehensive,
version-aware ICD database, which was subsequently developed and incorporated
into the package. Designing a general framework to support multiple PCCC
versions naturally enabled extension to additional comorbidity algorithms,
including Charlson and Elixhauser.

A primary design requirement was that the package be fully self-contained and
deployable in restricted computing environments, including systems with limited
internet access or tightly controlled software installations. Relying only on
base R ensures that _medicalcoder_ can be installed and used wherever R &geq;
3.5.0 is available, addressing a common but under-discussed barrier to
reproducible clinical research.

Beyond deployment considerations, _medicalcoder_ addresses a substantive
methodological gap in existing comorbidity software: the lack of support for
longitudinal comorbidity assessment. Most available implementations operate
strictly at the encounter level, implicitly assuming that comorbid conditions
are either repeatedly coded or clinically irrelevant once absent from the
current record. As demonstrated in Figures 1 and 2, this assumption can lead to
systematic under-ascertainment of comorbidities, particularly for conditions
governed by POA rules or technology-dependent classification criteria. By
explicitly tracking diagnostic history across encounters, _medicalcoder_ enables
comorbidity flagging that more closely reflects clinical reality.

The unified interface provided by _medicalcoder_ further reduces analytic
complexity. Mixed ICD-9 and ICD-10 datasets, full and compact codes, diagnostic
and procedure codes, and algorithm-specific nuances are handled within a single
function call. This contrasts with existing workflows that require dataset
partitioning, multiple method calls, and post hoc reconciliation, each of which
introduces opportunities for error and irreproducibility.

Extensive testing against published SAS, SQL, and R reference implementations
demonstrates that _medicalcoder_ produces results consistent with established
algorithms while offering additional flexibility. In several comparisons,
alternative R packages exhibited false negative or false positive flags,
underscoring that implementation details materially affect comorbidity
classification. Performance benchmarking further shows that _medicalcoder_
scales efficiently to large datasets, with substantial reductions in computation
time and memory usage when optional backends such as _data.table_ or _dplyr_ are
available, despite not being required dependencies.

To our knowledge, _medicalcoder_ is the only available R package that jointly
supports multiple comorbidity algorithms, mixed ICD versions,
present-on-admission logic, and cumulative longitudinal flagging within a
single, reproducible framework. These capabilities enable researchers to
accurately characterize patient complexity and to systematically evaluate how
different comorbidity definitions influence analytic results.

## Conclusions

_medicalcoder_ is a fully self-contained R package with a detailed and robust
internal ICD database that can efficiently apply many comorbidity algorithms to
a dataset while accounting for ICD-version, ICD-type, POA, primary diagnosis,
and full and compact ICD codes. Flagging codes can be done for individual
patient/encounters, or longitudinally within a patient record.

{{< pagebreak >}}

[**Funding Statement:**]{.underline}

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline}
The _medicalcoder_ R package is available from the Comprehensive R Archive Network
(CRAN) at https://cran.r-project.org/package=medicalcoder. Developmental
versions, additional testing suites, benchmarking, and examples are on GitHub at
https://github.com/dewittpe/medicalcoder/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
_medicalcoder_ R package and drafted the manuscript. SR provided critical code
review and testing.  SR, MR, and TDB all contributed to the conception, design,
and interpretation and provided important feedback. TDB provided funding. All
authors critically reviewed the manuscript for important intellectual content
and approved of the final version of the manuscript.

[**Acknowledgments:**]{.underline} We thank Vincent Rubinetti for his assistance
in designing the hex-sticker logo for medicalcoder.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

**Figure 1:**
Flagging cancer and cardiovascular disease by comorbidity method,
present-on-admission (POA) flagging, and flag method. Consider a patient record
with six encounters. ICD-10 diagnostic code C78.4 (mapping to cancer in
Charlson, Elixhauser, and Pediatric Complex Chronic Conditions (PCCC)) is
reported on encounter 2 as not present on admission (POA), and on encounter 5 as
POA. Additionally, ICD-10 diagnostic code I50.40 (mapping to cardiovascular
disease (CVD) or congestive heart failure) is reported as POA on encounter 3 and
not-POA on encounter 5.

This example illustrates how encounter-level flagging can fail to identify
clinically relevant comorbidities when diagnoses are not repeatedly coded or
when POA status varies across encounters. `flag.method = "current"` considers
only codes and POA status on the specific encounter, whereas `flag.method =
"cumulative"` propagates conditions identified in prior encounters.

**Figure 2:**
Flagging of metabolic and respiratory conditions via Pediatric Complex Chronic
Conditions, version 3, for an example patient with ICD-10 codes: H49.811
(non-technology-dependent metabolic), J84.111 (non-technology-dependent
respiratory), and Z96.41 (technology-dependent metabolic). One code appears on
encounters 2, 4, and 6. Encounter-level flagging with post hoc aggregation would
fail to identify metabolic conditions in several permutations shown here. PCCC
v3 only flags a condition based on a technology-dependent code if at least one
non-technology-dependent code exists in the patient history. _medicalcoder_
tracks prior encounters, allowing conditions to be flagged once the necessary
clinical context is present.

**Figure 3:** Benchmarking statistics for the _medicalcoder_ package. Charlson
[@quan2005], Elixhauser [@quan2005], and Pediatric Complex Chronic Conditions
v3.1 [@feinstein2024pediatric] were applied to simulated datasets using both
`current` and `cumulative` flag methods and three input data classes
(data.frame, tibble, data.table). The benchmarking was performed on a
computational server running Ubuntu 24.04.3 LTS, R 4.5.1, Intel Xeon Gold 6416H
CPUs, and 2 TB RAM. All processes limited to a single thread. While _medicalcoder_
does not import any namespaces, guarded S3 method dispatching will use
_data.table_ [@r-data.table] or _dplyr_ [@r-dplyr] native methods which reduce
computation time and memory use scales more efficiently with increasing
encounter counts.

