---
title: "medicalcoder: An R Package for Working with ICD Codes and Comorbidity Algorithms with Considerations for Longitudinal Data"

code-fold: false
echo: true
number-sections: false

format:
  docx:
    fig-cap-location: top
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
stopifnot(packageVersion("medicalcoder") >= "0.7.0")
```

**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; Margaret N. Rebull^1^, MA; Tellen D. Bennett^1,2^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

^2^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD

Department of Biomedical Informatics

University of Colorado School of Medicine

1890 N. Revere Court

Mailstop F600

Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} / 2000 (excluding tables)

[**Tables:**]{.underline} / 2

[**Figures:**]{.underline} 1 / 3

[**Supplementary material:**]{.underline}

[**Short running title**]{.underline}: R package

<!-- allowed 5; recomend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline}

{{< pagebreak >}}

# Abstract {.unnumbered}

words: /150 words

**Objectives**

**Materials and Methods**

**Results**

**Conclusions**

{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
<!-- word count:  / 200 -->

:::

{{< pagebreak >}}

# Background and Significance

Accurate identification of clinical conditions in large healthcare datasets
depends  on International Classification of Diseases (ICD) codes, which
serve as a primary structured representation of diagnoses and procedures.
Comorbidity algorithms built on these codes are central to risk adjustment,
cohort characterization, and the reproducible study of patient populations.
Software to apply comorbidity algorithms to

# Objective

Build a self-contained R package with a robust and complete ICD database and
multiple comorbidity algorithms.  Applying comorbidity algorithms to data via a
single function call while accounting for present-on-admission (POA), primary
diagnoses status, ICD version (9 or 10), ICD type (diagnostic or procedural),
full and compact ICD code strings, and longitudinal data.
<!-- word count 53 -->

* ICD utilities
  * `lookup_icd_codes()`
  * `is_icd()`
  * `get_icd_codes()`

* Comorbidity Algorithms
  * One function `comorbidities()`
  * One data structure: long data - one ICD per row
  * Handle ICD-9 and ICD-10 codes at the same time
  * Handle full and compact ICD codes at the same time
  * Consider present on admission
  * Consider primary diagnoses
  * Apply the comorbidity algorithms to
    * current encounter
    * cumulative (longitudinally)

# Materials and Methods

The medicalcoder R package has been developed to have minimal dependencies.
No namespaces are imported; R &geq; 3.5.0 is the only dependency due to a change
in data serialization[^a].  There are seveal namespaces listed as suggested
packages needed for building vignettes, testing, and building web-site only
articles.

[^a]: https://cran.r-project.org/src/base/NEWS.3#:~:text=R%20has%20new,to%20version%203.5.0

```{r}
#| label: icd-src-years
#| include: false
aggregate(cbind(known_start, known_end) ~ src + icdv + dx,
  data = medicalcoder::get_icd_codes(),
  FUN = range
  )

cms9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))
cdc9 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 9 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cdc10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cdc"),
    paste(min(known_start), max(known_end), sep = " - "))

cms10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "cms"),
    paste(min(known_start), max(known_end), sep = " - "))

who10 <-
  with(subset(medicalcoder::get_icd_codes(), icdv == 10 & src == "who"),
    paste(min(known_start), max(known_end), sep = " - "))
```

The ICD database contained within medicalcoder is built on the publicly
available ICD-9 and ICD-10 code sets from several sources.  ICD-9 diagnostic and
procedure codes from the
Centers for Medicare and Medicaid Services (CMS) (fiscal years `r cms9`) and
Centers for Diseases Control (fiscal years `r cdc9`).  ICD-10 diagnostic codes
from the CDC (calendar years `r cdc10`), World Health Organization (WHO)
(calendar years `r who10`).  ICD-10 diagnostic and procedure codes from CMS
(fiscal years `r cms10`).

Three comorbidity algorithms, Charlson
[@charlson1987;@deyo1992adapting;@ahrq2017;@quan2005;@quan2011;@glasheen2019],
Elixhauser [@elixhauser1998;@quan2005;@ahrq2025], and Pediatrics Complex Chronic
Conditions (PCCC) [@feudtner2014pediatric;@feinstein2024pediatric], each with
multiple variants, have been implemented in medicalcoder and are accessable via
the `method` argument of `medicalcoder::comorbidities()`.

medicalcoder was developed in R version 4.5.1.  The package has been tested
with, and without, suggested packages, for all version of R &geq; 3.5.0[^b].

[^b]: https://github.com/dewittpe/medicalcoder/tree/main/testing



`medicalcoder::comorbidities()` can apply the comorbidity algorithms over
longitudinal data to flag conditions as they arise over multiple encounters
within a patient record.


# Results {#sec-results}

There are two conceptual groups of functions within medicalcoder: those for
working with ICD codes, and those for applying and summarizing comorbidity
algorithms to a dataset.

## ICD Codes

End users can access the full internal ICD data base as a data.frame via
`medicalcoder::get_icd_codes()`.  Additional details, such as year and source
specific descriptions and the hierarchical structure of the ICD
codes, are returned when the arguments `with.descriptions = TRUE` and/or
`with.hierarchy = TRUE` are passed to `medicalcoder::get_icd_codes()`.

```{r}
#| label: get-icd-codes
str(medicalcoder::get_icd_codes())
# icdv:
#   integer 9 or 10 for ICD-9 or ICD-10
# dx:
#   diagnostic code (1); procedure code (0)
# full_code:
#   ICD code string with applicable dot, eg., D71.1
# code:
#   ICD code string without applicable dot, eg., D711
# src:
#   character indicating if the source infromation was from the cdc, cms, or who
# known_start,known_end:
#   first and last year that the code was known to be in the source
# assignable_start,assignable_end:
#   first and last year that the code was known to be the most granular code
#   from the source
```

To test if a ICD code string is a known ICD code
`medicalcoder::is_icd()` provides a

```{r}
medicalcoder::is_icd(c("D71", "D71.1"))

medicalcoder::is_icd(c("D71", "D71.1"), src = "cms")
medicalcoder::is_icd(c("D71", "D71.1"), src = "cms", year = 2026)
medicalcoder::is_icd(c("D71", "D71.1"), src = "cms", ever.assignable = TRUE)
medicalcoder::is_icd(c("D71", "D71.1"), src = "cms", headerok = TRUE, ever.assignable = TRUE)

medicalcoder::is_icd(c("D71", "D71.1"), src = "who")
medicalcoder::is_icd(c("D71", "D71.1"), src = "who", ever.assignable = TRUE)

medicalcoder::lookup_icd_codes(c("D71", "D71.1"))
```

A bridge between the ICD utilities and comorbidity methods are the functions
`medicalcoder::get_charlson_codes()`,
`medicalcoder::get_elixhause_codes()`, and
`medicalcoder::get_pccc_codes()`.  All of these return data.frames with the ICD
codes and the conditions/comorbidities the ICD code maps to under each variant
of the primary comorbidity algorithm.

## Comorbidities

A single function call, `medicalcoder::comorbidities()` is the interface for
applying any of the comorbidity algorithms to a dataset. The input data is
expected to be in a long format, one ICD code per row, with optional columns for
identification variables (e.g., patient and encounter ids), ICD version (integer
value 9 or 10), ICD type (1 for diagnostic code, 0 for procedure code),
present-on-admission (POA) indicator, and primary diagnostic indicator.  The ICD
codes can be full, compact, or a combination thereof.  A singel patient/encouner
record can consist of ICD-9, ICD-10, or both.

medicalcoder provides two example data sets.  `mdcr` is a data.frame with 4
columns: patient identifier (patid), ICD version (icdv), compact ICD code string
(code), and an indicator for diagnostic or procedure code (dx).

```{r}
library(medicalcoder)
str(mdcr)
head(mdcr)
```

A second dataset, `mdcr_longitudinal` provides an example dataset for exploring
comorbidities over time within a patient record.
```{r}
head(mdcr_longitudinal)
str(mdcr_longitudinal)
```

A verbose example of calling `medicalcoder::comorbidities()`, including default values is:
```{r}
#| label: comorbidities-example-1
elixhauser_quan2005_comorbidities <-
  medicalcoder::comorbidities(
    data          = mdcr, # a data.frame
    icd.codes     = "code", # name of column in data with ICD code strings
    method        = "elixhauser_quan2005", # name of <method>_<variant> to apply
    id.vars       = "patid", # (optional) id variables within data
    icdv.var      = "icdv",  # (optional) variable denoting ICD version
    icdv          = NULL, # (default)
    dx.var        = "dx", # (optional) variable denoting diagnostic (1) or procedure (0) code
    dx            = NULL, # (default)
    poa.var       = NULL, # (default)
    poa           = 1L,   # Consider all codes to be present-on-admission
    age.var       = NULL, # (default) only relevent to charlson_<variant>
    primarydx.var = NULL, # (default)
    primarydx     = 0L,   # Condiser all codes to be secondary diagnoses
    flag.method   = "current",  # (default), other option "cumulative"
    full.codes    = TRUE, # (default) Reference data[[icd.codes]] against full ICD codes in lookup tables
    compact.codes = TRUE, # (default) Reference data[[icd.codes]] against compact ICD codes in lookup tables
    subconditions = FALSE # (default) only relevent to pccc_<variant>
  )
```

Note that several arguements come in pairs, e.g., `icdv.var` and `icdv`.  The
version with the `.var` suffix will take precedence over the other if both are
defined.  The version without the `.var` suffix is provided as a simple way to
say the whole input data set has a particular property.  If both the versions
are left as the defaul NULL value, then that property will be ignored in the
lookup tables.  That is, if `icdv.var` and `icdv` are both NULL, then
`data[[icd.codes]]` will be compared to both ICD-9 and ICD-10 codes.  If
`icdv.var` is NULL and `icdv = 9L` then `data[[icd.codes]]` will only be
compared to ICD-9 codes in the lookup tables.

The return from `medicalcoder::comorbidities()` is a data.frame with some
additional metadata.  The data.frame has column(s) for the `id.vars` and an
indicator column each of the relevent
conditions/comorbidities.  All comorbidity methods will return a data.frame with
columns `num_cmrb` for the number of conditions/comorbidities
flaged for the `id.vars` and `cmrb_flag`, an indicator column for the
presence/absence of at least one condition/comorbidity.

```{r}
str(elixhauser_quan2005_comorbidities, max.level = 0L)
#names(elixhauser_quan2005_comorbidities)
```

A summary method is provided to provide count and percentages of each condition.

```{r}
head(summary(elixhauser_quan2005_comorbidities)$conditions)
```


<!--

```{r}
#| label: comorbitities-args
#| include: false
args(medicalcoder::comorbidities)
```

::: {#tbl-comorbidities-args}

+-----------------+----------+-------------------------------------------------+
| Argument        | Default  | Definition                                      |
+:================+:========:+:================================================+
| `data`          |          | A data.frame in long format - one row per ICD   |
|                 |          | code.                                           |
+-----------------+----------+-------------------------------------------------+
| `icd.codes`     |          | quoted name of character vector in `data` with  |
|                 |          | ICD codes                                       |
+-----------------+----------+-------------------------------------------------+
| `method`        |          |                                                 |
+-----------------+----------+-------------------------------------------------+
| `id.vars`       |          |                                                 |
+-----------------+----------+-------------------------------------------------+
| `icdv.var`      | NULL     |                                                 |
+-----------------+----------+-------------------------------------------------+
| `icdv`          | NULL     |                                                 |
+-----------------+----------+-------------------------------------------------+
| `dx.var`        | NULL     |                                                 |
+-----------------+----------+-------------------------------------------------+
| `dx`            | NULL     |                                                 |
+-----------------+----------+-------------------------------------------------+
| `poa.var`       | NULL     |                                                 |
+-----------------+----------+-------------------------------------------------+
| `poa`           | NULL     |                                                 |
+-----------------+----------+-------------------------------------------------+
| `age.var`       | NULL     |                                                 |
+-----------------+----------+-------------------------------------------------+
| `primarydx.var` | NULL     |                                                 |
+-----------------+----------+-------------------------------------------------+
| `primarydx`     | NULL     |                                                 |
+-----------------+----------+-------------------------------------------------+
| `flag.method`   | "current"|                                                 |
+-----------------+----------+-------------------------------------------------+
| `full.codes`    | `TRUE`   |                                                 |
+-----------------+----------+-------------------------------------------------+
| `compact.codes` | `TRUE`   |                                                 |
+-----------------+----------+-------------------------------------------------+
| `subconditions` | `FALSE`  |                                                 |
+-----------------+----------+-------------------------------------------------+

Arguments and default values for `medicalcoder::comorbidities()`.

:::

-->

::: {#tbl-comorbidities-methods}

+------------------------------------------+-----------------------------------+
| Comorbidity Method                       | Notes                             |
+:=========================================+:==================================+
| Charlson                                                                     |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `charlson_deyo1992`         | ICD-9 codes defined in Table 1 of |
|                                          | @quan2005                         |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `charlson_quan2005`         | ICD-9 and ICD-10 defined in       |
|                                          | Table 1 of @quan2005;             |
|                                          | index scoring as                  |
|                                          | reported in Table 2 of @quan2011  |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `charlson_quan2011`         | ICD-9 and ICD-10 defined in       |
|                                          | Table 1 of @quan2005;             |
|                                          | index scoring as                  |
|                                          | reported in Table 2 of @quan2011  |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `charlson_cdmf2019`         | ICD-9 and ICD-10 defined in       |
|                                          | @glasheen2019                     |
+------------------------------------------+-----------------------------------+
| Elixhauser                                                                   |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_elixhauser1988` | ICD-9 codes defined in Table 2 of |
|                                          | @quan2005                         |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq_web`       | ICD-9 codes defined in Table 2 of |
|                                          | @quan2005                         |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_quan2005`       | ICD-9 and ICD-10 defined in       |
|                                          | Table 2 of @quan2005;             |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq2022`       | ICD-10 codes from @ahrqicd10      |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq2023`       |ICD-10 codes from @ahrqicd10       |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq2024`       | ICD-10 codes from @ahrqicd10      |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq2025`       | ICD-10 codes from @ahrqicd10      |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `elixhauser_ahrq_icd10`     | Any ICD-10 code from any specific |
|                                          | `elixhauser_ahrqYYYY`             |
+------------------------------------------+-----------------------------------+
| PCCC                                                                         |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `pccc_v2.0`                 | ICD-9 and ICD-10 diagnostic and   |
|                                          | procedure codes consistent with   |
|                                          | @dewitt2025pccc                   |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `pccc_v2.1`                 | Extended set of `pccc_v2.0`       |
|                                          | codes based on                    |
|                                          | @dewitt2025pccc and supplements   |
|                                          | @feudtner2014pediatric            |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `pccc_v3.0`                 | ICD-9 and ICD-10 diagnostic and   |
|                                          | procedure codes consistent with   |
|                                          | @pcccv3sas                        |
+------------------------------------------+-----------------------------------+
| &nbsp;&nbsp; `pccc_v3.1`                 | Extended set of `pccc_v3.0` codes |
|                                          | based on the supplements to       |
|                                          | @feinstein2024pediatric           |
+------------------------------------------+-----------------------------------+

Comorbidity methods implemented in medicalcoder.

:::

```{r}
#| label: flag-method-example
#| include: false
source("flag-method-example.R")
```

The the combination of `poa.var` or `poa` and `flag.method` impact when a
condition/comorbidity is flag.  Figure 1 illustrates this impact. For the
example record below we have one patient with six encouners.  ICD-10 diagnostic
codes C78.4 (mapping to cancer in Charlson, Elixhauser, and PCCC), and I50.40
(mapping to cardiovascular disease in Charlson, Elixhauser, and PCCC), appear on
encounters 2, 3, adn 5.  The POA status of the codes are dyanmic as well.
```{r}
#| label: example-record
record
```

The following call is an example of for the `r nrow(args)`  calls needed to
generate the flags shown in Figure 1.  See Supplementary material for the
specific code to reproduce the graphic.  The most important detail to take away
is that the `id.vars` have the name of the sortable encounter id.
```{r}
#| label: example-call-for-flag-example
#| eval: false
medicalcoder::comorbidities(
  data = record,
  id.vars = c("patid", "encid"),
  icdv = 10L,
  dx = 1L,
  poa =     # ... three options: <0, 1, NULL>
  poa.var = # ... NULL when poa = 0 or 1, else "poa"
  method =  # ... charlson_quan2005, elixhauser_ahrq2025, pccc_v3.1
  primarydx = 0L,
  flag.method = # ... "current", "cumulative"
)
```

The utility of the longitudinal methods in medicalcoder are critical in the
assessment of PCCC v3 [@feinstein2024pediatric] conditions.  For these
conditions, some ICD codes are classified technology
dependent codes.  If a technology dependent code for a condition is in the
patient record, it only flags the condition if a non-technology code is also
present.
For example, consider
* H49.811: metabolic (other metabolic disorders),
* J84.111: respiratory (chronic respiratory diseases), and
* Z96.41:  metabolic (device and technology use).

We'll apply the `pccc_v3.1` to this code set with `flag.method = "cumulative"`
and all codes considered to be present-on-admission.  Figure 2 illustrates when
the conditions are flagged.  For permutations 1, 2, and 3, where the
non-technology-dependent code for a metabolic condition appears in the record
before the technology-dependent metabolic code, the flags appear as intuitively
expected, metabolic flags as soon as the non-technology-dependent code appears.
Permutations 4, 5, and 6 are more interesting as the technology-dependent
metabolic code appears before the non-technology-dependent code.  Because the
longitudinal methods in `medicalcoder::comorbidities()` track all ICD codes from
all prior encounters, the metabolic technology-dependent-only flag appears in
permutations 4 and 6 due to the respiratory non-technology-dependent code
existing.  Flagging conditions on a ecnounter only level and aggregating the
results would result in missing metabolic on encounters 4 and 5 in permutations
4 and 6.

## Benchmarking

There are two benchmarks we consider. First, medicalcoder has been designed to
work with base R.  That said, if the input to the `data` arguement of
`medicalcoder::comorbidities()` is a data.table and the data.table namespace is
available, the gaurded S3 method dispacting will exploit the data.table
efficientcys for merge, setting, and subsetting compaired to base R data.frame.
Figure 2 shows the average time, relative time (versus data.frame), and the
amount of RAM used for datasets of different sizes and several comorbidity
algorithms.  In general, `medicalcoder::comorbidities()` will return the condition/comorbidity
flags for 100,000 encounters in less than 10 seconds when `flag.method =
"current"`.  Using a data.table can take less than 50% the time compared to a
data.frame.  Extensive details on the the benmarking and code to reproduce the
graphic are on GitHub[^c].

[^c]: https://github.com/dewittpe/medicalcoder/tree/main/benchmarking

Comparing medicalcoder to other R packages has been done (see supplements).
medicalcoder has more features, more comorbidity methods, and flexability than
other packages.  Notable alternatives at the time this manuscript was written
are: multimorbidity [@r-multimorbidity], icdcomorbid [@r-icdcomorbid], and
comorbidity [@r-comorbidity].

```{r}
#| label: vs
#| include: false
vs <- new.env()
load(file = "medicalcoder-vs.Rdata", envir = vs)
```

It took `icdcomorbid::icd9_to_comorbid()`
`r attr(vs$icdcomorbid_icd9_results, "tictoc")` seconds to
process only `r nrow(vs$icdcomorbid_icd9_results)` patids from the `mdcr`
dataset using only ICD-9 codes.  `medicalcoder::comorbitities()` took
`r attr(vs$medicalcoder_charlson_quan2005, "tictoc")` seconds to process the
whole dataset.  Due to the several orders of magnitude differene in computation
time, we did not compare results between `medicalcoder::comorbitities()` and
`icdcomorbid::icd9_to_comorbid()`.

`multimorbidity::charlson()` took
`r attr(vs$multimorbidity_charlson_results, "tictoc")` seconds to apply the
Charlson comorbidities to the `mdcr` data set. We noted several flase negatives
resulting from the `multimorbidity::charlson()` call.

The comorbidity is a widely referenced and downloaded R package and applys the
Charlson and Elixhauser (Quan 2005 and Quan 2011) a couple other variants of
other comorbidity algorithms.  The difference in time required to apply the
methods between `medicalcoder::comorbidities()` and `comorbidity::comorbidity()`
is largely due to `comorbidity::comorbidity()`
requiring two calls, one to apply ICD-9 codes, one for ICD-10 codes, aggregating
the results, and then scoring them whereas `medicalcoder::comorbidities()`
required only one call. `medicalcoder::comorbidities()` took
`r attr(vs$medicalcoder_charlson_quan2005, "tictoc")` seconds to apply
`charlson_quan2005`, `r attr(vs$medicalcoder_charlson_quan2011, "tictoc")`
seconds to apply `charlson_quan2011`, and
`r attr(vs$medicalcoder_elixhauser_quan2005, "tictoc")` seconds to apply
`elixhauser_quan2005`. Compared to `comorbidity:comorbidity()` which took
`r attr(vs$comorbidity_charlson_results, "tictoc")` seconds to apply
Charlson, and `r attr(vs$comorbidity_elixhauser_results, "tictoc")` seconds to
apply Elixhauser.

# Discussion

medicalcoder was initially built to implement PCCC v3
[@feinstein2024pediatric].  A requirement was to have a self contained package
that could be easily installed on a machine with extremely limited or no
internet access.  Using only base R was needed due to a history of insufficient
system dependencies resulting in difficulting compiling other R packages.  The
development of the ICD utilities within medicalcoder were built to inform the
code sets needed for the comorbidity algorithms.

An extensive testing suite has been developed for medicalcoder.  Comparisons to
prior package results, AHRQ Elixhauser SAS code, PCCC SAS code, and MIMIC-IV code for
Charlson comorbidities have all been used to verify consistency of results from
`medicalcoder::comorbidities()`.  These tests run as a GitHub action for the
oldrelease, release, and developmental versions of R.
Additionally, the test suite runs on demand in all versions of R &geq; 3.5.0
without suggested packages via docker.  Many versions also run the test suite
with suggested packages.

The workflow for extending medicalcoder with additional ICD codes, or
comorbidity algorithms is inplace.  Anecdotaly, the release of the fiscal year
2026 CMS ICD-10 codes took about three hours of work to fully integrate into the
package, examples, and tests.

Careful use of gaurded S3 method dispacting allows medicalcoder to be installed
without any imported namespaces while being able to exploit the computational
efficientcies of data.table when the namespace is available on the system.

Overall, medicalcoder provides computationally efficient implementations of
`r length(medicalcoder:::comorbidities_methods())` comorbidity methods.  This is
far more methods than any other known utility.  This opens research
oppurtunities for investigation of different comorbidity classifications within
the same data sets.

Crucially, medicalcoder provides support of longitudinal data, another feature
we have not seen in other software packages.  Figure 1 and Figure SXX
illustrates the importance of this feature as aggregating conditions after
flagging can result in false negatives as some codes will not flag a conditions
without the presence of other codes (PCCC v3).

## Conclusions

{medicalcoder} is a fully-self contained R package able to efficienty apply many
comorbidity algorithms to a dataset while accounting for ICD-version, ICD-type,
present-on-admission, primary diagnostic, full and compact ICD codes.  Flagging
codes can be done for individual patient/encounters, or longitudinally within a
patient record.

{{< pagebreak >}}

[**Funding Statement:**]{.underline}

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline}
The {medicalcoder} R packaage is available from the Comprehensive R Archive
Network (CRAN) at https://cran.r-project.org/package=medicalcoder.
Developmental versions, additional testing suites, and examples are on GitHub at
https://github.com/dewittpe/medicalcoder/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
medicalcoder R package and drafted the manuscript. SR provided critical code
review and testing.  SR, MR, nd TDB all contributed to the conception, design,
and interpretation and provided important feedback. TDB provided funding. All
authors critically reviewed the manuscript for important intellectual content
and approved of the final version of the manuscript.

<!--
[**Acknowledgments:**]{.underline}
-->

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

Figure 1: Flagging cancer and cardiovascular disease by Comorbidity method,
present on admission (POA) flagging, and flag method.  Consider a patient record
with six encounters.  ICD-10 diagnostic code C78.4 (mapping to cancer in
Charlson, Elixhauser, and PCCC) is reported on encounter 2 as not present on
admission (POA), and on encounter 5 as POA.  Additionally, ICD-10 diagnostic
code I50.40 (mapping to cardiovascular disease (CVD) or congenstie heart
failure) is reported as POA on encounter 3 and not-POA on encounter 5.  This
figure illustrates when cancer and when CVD are flagged by the different
comorbidity algorithms and by how POA is handeled.  `poa = 0` considers all
codes to be not-POA, `poa = 1` considers all codes to be POA, and `poa.var =
'poa'` uses a dynamic, code-specific, flag in the input data to determine the
POA status.  The flag method `current` only considers the codes and POA status
on the specific encounter, where as `cumulative` considers if the code has been
reported previously, hence the flagging of both conditons on encounter 4 under
all comorbidity agorithems under `flag.method = 'cumulative'` as the codes from
encounters 2 and 3 are considered in the history for encounter 4.

Figure 2: Flagging of metabolic and respiratory conditiions via the Pediatrics
Complex Chronic Conditions v3 [@feinstein2024pediatric] for an example patient
with ICD-10 codes: H49.811 (non-technology-dependent metabolic), J84.111
(non-technology-dependent respiratory), and Z96.41 (technology-dependennt
metabolic).  One code appears on encounter 2, 4, and 6.  PCCC v3 only flags a
condition based on a technology-dependennt code if at least one non-technology
dependennt code also exits.  Permutations 1, 2, 3 easily flag the metabolic
condition because the H49.811 appears before Z96.41 in the record.  Permutations
4, 5, 6 do not flag metabolic until a non-technology dependennt code is in the
record, e.g., in permutation 6, Z96.41 appears on encounter 2. Since no
non-technology-dependennt code is in the record no condition is flagged on
encounter 2 or 3.  On encounter 4 when J84.111 is reported, then both metabolic
and respiratory conditions are flagged as a non-technology-dependennt code
exists in the record.
