---
title: "medicalcoder: A Unified and Longitudinally Aware Framework for ICD-Based Comorbidity Assessment in R"

code-fold: false
echo: true
number-sections: false

format:
  docx:
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
#| cache: false
stopifnot(packageVersion("medicalcoder") >= "0.8.1")
options(qwraps2_markup = "markdown")
loadNamespace("data.table")
loadNamespace("medicalcoder")
knitr::read_chunk("mimiciv-data-analysis.R")
qwraps2::lazyload_cache_labels(
  labels = c("benchmarking"),
  path = "mimiciv-data-analysis_cache/pdf/"
)
```
```{r}
#| label: import-mimiciv-data
#| include: false
```
```{r}
#| label: build-mimicivdata-ages
#| include: false
```
```{r}
#| label: build-mimicivdata-icd
#| include: false
```
```{r}
#| label: one-data-set
#| include: false
```
```{r}
#| label: build-enc-seq
#| include: false
```
```{r}
#| label: medicalcoder-charlson-current
#| include: false
```
```{r}
#| label: medicalcoder-charlson-cumulative
#| include: false
```
```{r}
#| label: medicalcoder-elixhauser-current
#| include: false
```
```{r}
#| label: medicalcoder-elixhauser-cumulative
#| include: false
```
```{r}
#| label: medicalcoder-pcccv3.1-current
#| include: false
```
```{r}
#| label: medicalcoder-pcccv3.1-cumulative
#| include: false
```


**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; James A. Feinstein, MD, MPH^2,3,4^; Margaret N. Rebull^1^, MA; Tellen D. Bennett^1,5^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA\
^2^ Adult and Child Center for Outcomes Research and Delivery Science, University of Colorado Anschutz Medical Campus, Aurora.\
^3^ Children's Hospital Colorado, Aurora.\
^4^ Department of Pediatrics, University of Colorado Anschutz Medical Campus, Aurora.\
^5^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD\
Department of Biomedical Informatics\
University of Colorado School of Medicine\
1890 N. Revere Court\
Mailstop F600\
Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} 1957 / 2000 (excluding tables)

[**Tables:**]{.underline} 2 / 2

[**Figures:**]{.underline} 3 / 3

[**Supplementary material:**]{.underline} mimiciv-data-analysis.pdf

[**Short running title**]{.underline}: _medicalcoder_: R package for ICD codes and comorbidities with longitudinal data

<!-- allowed 5; recommend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline} Comorbidity, International Classification of
Diseases (ICD), Software, Longitudinal Studies, Pediatrics

{{< pagebreak >}}

# Abstract {.unnumbered}

Word count: 145 / 150

**Objectives:**
Comorbidity algorithms derived from ICD codes are central to risk adjustment and
cohort characterization, but existing software inconsistently handles mixed
ICD-9/10 data and typically operates at the encounter level. We developed
_medicalcoder_ to address these limitations.

**Materials and Methods:**
_medicalcoder_ is an R package with an internal ICD database and unified
interface for applying variants of the Charlson, Elixhauser, and Pediatric
Complex Chronic Conditions algorithms. It supports full and compact ICD codes,
mixed ICD versions, present-on-admission and primary diagnosis indicators, and
encounter-level and cumulative longitudinal flagging.

**Results:**
The _medicalcoder_ package requires only R &geq; 3.5.0; no additional packages,
internet access, or external data are required. Outputs are consistent with published
reference implementations. Longitudinal flagging identifies additional
comorbidity flags under a carry-forward assumption while accounting for
present-on-admission or technology-dependent logic.

**Conclusions:**
_medicalcoder_ provides a unified, reproducible framework for ICD-based
comorbidity assessment in modern clinical datasets.


{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
Word count: 189 / 200

Researchers often use International Classification of Diseases (ICD) diagnosis
and procedure codes to identify health conditions in large medical datasets.
Comorbidity algorithms group these codes into meaningful categories, such as
heart disease or cancer, and are essential for understanding patient complexity,
comparing outcomes, and adjusting for risk. However, existing tools usually
handle only one or two algorithms, work with either ICD-9 or ICD-10 (but rarely
both together), and do not track health conditions across multiple hospital
visits. This can lead to missed or inconsistent results.

We created _medicalcoder_, an R package that solves these problems. It includes
an internal database of ICD-9 and ICD-10 codes and provides a single, easy-to-use
function to apply many comorbidity algorithms. The package can recognize both
full and compact ICD codes, use present-on-admission flags, identify primary
diagnoses, and follow patients across multiple encounters to avoid missing
conditions that appear earlier in their history.

_medicalcoder_ is fast, works with base R, has extensive automated tests, and
produces encounter-level outputs consistent with reference implementations. It
helps researchers apply comorbidity algorithms with consistency, improving the quality and
reproducibility of studies using electronic health records or administrative
data.

:::

{{< pagebreak >}}

# Background and Significance

Accurate identification of clinical conditions in large healthcare datasets
depends on International Classification of Diseases (ICD) codes
[@quan2005;@quan2011;@feinstein2024pediatric;@feudtner2014pediatric;@ahrqicd10;@glasheen2019;@ludvigsson2021adaptation;@ludvigsson2023adaptation].
Comorbidity algorithms built on these codes support risk adjustment, cohort
characterization, and reproducible studies of patient populations
[@quan2005;@yurkovich2015systematic;@guo2022evaluation;@li2008risk].

Existing software for flagging comorbidities typically implements one or two
algorithms, most often variants of Charlson and Elixhauser, supports either
ICD-9 or ICD-10 (but not both simultaneously),
and requires manual reconciliation of mixed-version datasets
[@r-comorbidity;@feudtner2014pediatric;@dewitt2026pccc;@pcccv3sas;@r-icdcomorbid;@r-multimorbidity].

Other R packages for working with ICD codes and comorbidities include the
archived _icd_ package, which provides ICD validation, manipulation, and
comorbidity mapping [@r-icd], and the _comorbidity_ and _pccc_ packages which
implement specific algorithms [@r-comorbidity;@dewitt2026pccc].

Despite the importance of longitudinal data in medical
research [@hripcsak2013next;@casey2016using;@jensen2014temporal], comorbidity
software typically operates at the encounter
level [@pcccv3sas;@ahrq2017;@dewitt2026pccc;@ahrqicd10;@feinstein2018pccc;@r-comorbidity;@r-icdcomorbid;@r-multimorbidity;@johnson2018mimic;@johnson2022mimiccode].
While the AHRQ [@ahrqicd10] Elixhauser variants account for present-on-admission
(POA) indicators and exemptions, they do not propagate conditions across
encounters within a patient record. As a result, conditions that arise during
acute care and are excluded from the index encounter may be missed
in subsequent encounters unless the relevant ICD codes are
re-reported and flagged as POA.

This limitation is particularly important given the high prevalence of chronic
disease in the United States, where 76.4% of adults report at least one chronic
condition and 59.5% report two or more [@watson2025trends]. Failing to account
for longitudinal patient history can therefore lead to systematic
under-ascertainment of comorbidities and misclassification of patient risk.

We developed _medicalcoder_ as a unified, POA-aware API supporting multiple
algorithms, algorithm-specific requirements, mixed ICD versions and types, and
longitudinal data.

# Objective

Develop a self-contained R package with an internal ICD database capable of
applying multiple comorbidity algorithms to clinical datasets without external
dependencies, internet access, or data resources. The package should support
simultaneous ICD-9 and ICD-10 coding,
accept full and compact ICD codes, incorporate present-on-admission and primary
diagnosis indicators, and enable longitudinal comorbidity assessment across
patient encounters.

# Materials and Methods

_medicalcoder_ was designed to run within restricted computing environments. No
namespaces are imported, and no network access is required. R
&geq; 3.5.0 is the only required installation and runtime dependency.

```{r}
#| label: icd-src-years
#| include: false
```

As of _medicalcoder_ version `r packageVersion("medicalcoder")`, the internal
ICD database includes ICD-9 diagnostic codes from the Centers for Disease
Control and Prevention (CDC) and ICD-9-CM and ICD-9-PCS codes from the Centers
for Medicare and Medicaid Services (CMS). ICD-10 diagnostic codes from the CDC
and the World Health Organization (WHO) are included, along with ICD-10-CM and
ICD-10-PCS codes from CMS, ICD-10-AM (Australian Modification) data from the
Independent Health and Aged Care Pricing Authority, and ICD-10-SE (Swedish
modification) data from Socialstyrelsen. Future releases will incorporate
published updates to supported code sets and additional national modifications.

Charlson variants include Deyo (as denoted in [@quan2005]), Quan (2005; 2011)
[@quan2005;@quan2011], CDMF (2019) [@glasheen2019], Ludvigsson (2021)
[@ludvigsson2021adaptation;@ludvigsson2023adaptation], Sundararajan (2004)
[@sundararajan2004], and Beyrer (2021) [@beyrer2021validation]. Elixhauser
variants include the original Elixhauser (1998) [@elixhauser1998], AHRQ Web
(ICD-9 based as denoted in [@quan2005]), Quan (2005) [@quan2005], and AHRQ
ICD-10 fiscal-year implementations (2022–2026) [@ahrqicd10]. PCCC variants
include versions 2.0 [@feinstein2018pccc], 2.1, 3.0 [@feinstein2024pediatric],
and 3.1. The reference data and comorbidity mappings are versioned within
each _medicalcoder_ release.

_medicalcoder_ maps ICD codes to comorbidity conditions using method-specific
lookup tables derived from, but distinct from, its broader internal ICD
database. Each table defines the ICD versions (9 or 10), code types (diagnosis
or procedure), full and compact code forms, and condition mappings supported by
the selected algorithm. Unlike implementations based primarily on partial string
matching or regular expressions
[@johnson2022mimiccode;@r-comorbidity;@dewitt2026pccc], _medicalcoder_ defaults
to matching input against precomputed valid ICD codes through database-style
joins. For supported methods, users may select method-specific regular-expression mappings. Dataset-
or row-level metadata can identify ICD version and code type, allowing mixed
ICD-9 and ICD-10 diagnosis and procedure codes within a dataset or encounter.
Matches are reduced to unique encounter-condition combinations before binary
indicators are assigned.

Because code text alone cannot reliably distinguish jurisdictional variants such
as WHO ICD-10 and ICD-10-CM, users must select a method appropriate to their
source coding system and provide accurate code-system metadata.

A comprehensive test suite runs through GitHub Actions and the Comprehensive R
Archive Network (CRAN), with additional compatibility testing across supported R
versions. Encounter-level outputs are compared with published R, SAS, and SQL
implementations: PCCC v2.0 with pccc [@dewitt2026pccc], AHRQ Elixhauser variants
with the published AHRQ SAS programs [@ahrqicd10], and Charlson variants with
_comorbidity_ [@r-comorbidity] and MIMIC-IV Code [@johnson2022mimiccode].
Testing workflows and comparisons are provided in the package repository and
Supplement.

We use the MIMIC-IV v3.1 dataset [@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000]
for examples, validation, and benchmarking, not clinical inference. Pediatric
comorbidity algorithms are demonstrated using adult data. We compare
_medicalcoder_ with MIMIC-IV Code
[@johnson2018mimic;@johnson2022mimiccode], _comorbidity_ [@r-comorbidity], and
_pccc_ [@dewitt2026pccc].

# Results

The MIMIC-IV v3.1 dataset includes
`r qwraps2::frmt(nrow(mimicivDT))` rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
hospital encounters.
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id)][V1 > 1, qwraps2::frmt(.N)]`
subjects' records contain both ICD-9 and ICD-10 codes, and
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id, hadm_id)][V1 > 1, qwraps2::frmt(.N)]`
encounters contain both versions.

`medicalcoder::comorbidities()` can apply
`r length(medicalcoder:::comorbidities_methods())` comorbidity
algorithms to long-format, one ICD code per row, data. Optional columns identify
patients and/or encounters, ICD version and type, POA, and primary diagnosis
(Figure 1). A single patient record may contain mixed ICD code versions and
types in full or compact forms.

Paired arguments (e.g., `dx` and `dx.var`) specify dataset-level values or row-level
columns, respectively, avoiding constant-valued columns while supporting
row-level variation. If both are supplied, the `.var` argument takes precedence
and a warning is issued.

`id.vars` defines encounter-level summaries:
with `flag.method = "current"`, omission treats the dataset as one encounter;
with `flag.method = "cumulative"`, `id.vars` must include at least two
variables defining the patient and within-patient encounter order (Supplement
Section 5.1).

Output preserves the input `data.frame` class and includes
`id.vars`, one indicator column per comorbidity, and
summary columns `num_cmrb` and `cmrb_flag`. An S3
`summary()` method generates count and percentage tables (Table 1).

<!--
using the quarto native tbl- for crossref is causing an error in the word doc.
hardcoding the table number as a result
-->

::: {custom-style="tablecap"}
**Table 1:** Counts and percentages of encounters within the MIMIC-IV dataset
flagged with Charlson comorbidities via `medicalcoder::comorbidities()` with
`method = "charlson_quan2011"`.
:::

```{r}
#| label: charlson-summary
#| echo: false
#| results: asis
```
Encounter-level flagging assumes relevant ICD codes are reported, and when
required, marked POA at each encounter. Omissions may under-ascertain conditions
or severity.
```{r}
#| label: figure2
#| include: false
```
```{r}
#| label: figure2-pros
#| include: false
```
Figure 2 compares cumulative and current profiles using descriptive,
nonexclusive measures. "Different risk profile" denotes any difference in
condition classification or count. Denominators are subjects with at least two
encounters and a cumulative flag after the index encounter, and non-index
encounters with at least one cumulative flag. Among `r figure2pros$N` eligible
Elixhauser subjects, `r figure2pros$cs[1]` had a different profile on at least
one encounter, and `r figure2pros$cs[2]` had no current but at least one
cumulative comorbidity.

PCCC v3 and the 2022-2026 AHRQ Elixhauser algorithms require additional logic. PCCC
v3 conditions depend on whether technology-dependent codes co-occur (or
eventually co-occur) with non–technology-dependent codes (Table 2).
For AHRQ Elixhauser variants, POA-required conditions are flagged when POA is
present, or the ICD code is POA-exempt (Supplement Section 8.4).

<!--
using the quarto native tbl- for crossref is causing an error in the word doc.
hardcoding the table number as a result
-->

::: {custom-style="tablecap"}
**Table 2:** Flagging of comorbidities under PCCC v3 for MIMIC-IV subject
10728333. Cells marked with DxPr denote that the comorbidity was flagged due to
an ICD code that is not technology-dependent.  Tech denotes flagging of a
comorbidity based on the presence of at least one technology-dependent ICD
code and the presence of at least one non-technology-dependent code for
another comorbidity. DxPrTech denotes flagging due to both
technology-dependent and non-technology-dependent ICD codes.  Under the Any
Tech columns we mark the encounters where some tech dependence is identified.
For this patient, on encounter 1, a technology-dependent ICD code for a
respiratory comorbidity was found in the subject's record.  Because no
non-technology-dependent code was observed on encounter 1, under PCCC v3 this
subject has no comorbidities for encounter 1.  This persists for encounters 2
and 3 with respect to respiratory and on encounter 3 for miscellaneous as
well.  On encounter 4 a non-technology-dependent code for a metabolic
comorbidity was reported.  Under a cumulative flagging paradigm, the
miscellaneous and respiratory comorbidities are also flagged as the reported
codes from encounters 3 and 1 have been carried forward respectively.  Under
the 'current' flagging method miscellaneous is never flagged as the
technology-dependent code never occurs on the same encounter as the
non-technology-dependent code.
:::

```{r}
#| label: s10728333
#| include: false
```
```{r}
#| label: s10728333-manuscript
#| echo: false
#| results: asis
```

Across encounter-level comparisons with MIMIC-IV Code and _comorbidity_,
_medicalcoder_ showed 100% agreement for shared condition flags after
harmonizing output conventions; PCCC v2.0 discordance with _pccc_ was limited to
documented technology-dependence and transplant-category differences
(Supplementary Table 3).

The comparators all use some form of partial string matching for mapping ICD
codes to conditions. _medicalcoder_ uses a set of precomputed valid ICD codes to
map between input data and conditions using database-style joins. As a result,
comparator implementations that rely on prefix or regular-expression matching
can return false-positive flags when an invalid input string starts with a valid
ICD code substring. For example, the ICD-10 code C189 maps to "any malignancy"
under Charlson; _comorbidity_ maps C189NOTACODE to that condition because the
string starts with C18. _medicalcoder_, by default, does not map the invalid
string to a condition.

Precomputed ICD-to-condition maps reduce false-positive flags compared with
partial string matching and reduce computational expense. Benchmarking in the
supplement and Figure 3 shows that _medicalcoder_ requires less time to compute
comorbidity flags in the MIMIC-IV examples.

_medicalcoder_ provides ICD utility functions for data preparation and
validation. `medicalcoder::get_icd_codes()` exposes the internal ICD database,
and `medicalcoder::is_icd()` validates candidate codes across full and compact
representations and mixed ICD-9/ICD-10 datasets.

# Discussion

_medicalcoder_ originated as a PCCC implementation
[@feinstein2024pediatric; @feudtner2014pediatric],
for which reproducible classification required a version-aware ICD database that
subsequently supported Charlson and Elixhauser algorithms. The package imports
no namespaces and can be installed wherever R &geq; 3.5.0 is available,
including restricted environments. Pinning the package version locks its ICD
database and comorbidity mappings, supporting reproducibility. “Dependency-free”
applies to installation and runtime; tests, documentation, benchmarks, and
manuscript workflows require the external tools and data described here and in
the Supplement.

_medicalcoder_ addresses a methodological gap in existing comorbidity software:
the lack of support for longitudinal comorbidity assessment. Most
implementations operate strictly at the encounter level, implicitly assuming
that comorbid conditions are either repeatedly coded or clinically irrelevant
once absent from the current record. As demonstrated in Table 2, this assumption
can lead to systematic under-ascertainment of comorbidities, particularly for
chronic conditions, or conditions governed by POA rules or technology-dependent
classification criteria. By tracking condition history across encounters,
_medicalcoder_ supports comorbidity classification based on prior observed
diagnoses under an explicit carry-forward assumption.

The utility of cumulative flags depends on whether every ICD code mapped to a
condition represents persistent disease. For example, carry-forward
of uncomplicated diabetes is defensible for the Quan et al. Charlson and
Elixhauser mappings [@quan2005], which identify non-gestational diabetes. In the
AHRQ fiscal year 2022 and beyond mappings, however, uncomplicated diabetes also
includes ICD-10 codes under the O24 category for gestational diabetes, a
pregnancy-related and potentially transient condition. Because _medicalcoder_
carries forward the comorbidity flag rather than the originating ICD code,
subsequent output cannot distinguish gestational from non-gestational diabetes.
Thus, cumulative uncomplicated-diabetes flags may be reasonable with Quan-style
mappings but require additional qualification with AHRQ-style mappings.

Malignancy illustrates a related limitation. Suppose C18.2 "malignant neoplasm
of the colon" is reported. Upon resolution of the neoplasm, the C18.2 code
should no longer be reported and the ICD-10 code Z85.038, "personal history of
malignant neoplasm of the large intestine" should be reported. The cumulative
flag remains positive on and after the encounter with the initial C18.2 report
and therefore represents current or previously observed malignancy rather than
current malignancy alone. However, because Z85.038 is not included in any of the
comorbidity mappings, a history-only encounter preceding any mapped
active-malignancy code remains unflagged. Cumulative malignancy flags should
therefore be interpreted as indicating current or previously observed
malignancy, not a comprehensive current-or-history phenotype.

The magnitude and interpretation of cumulative flagging may differ in pediatric,
outpatient, claims, non-ICU, and non-U.S. datasets compared to the results shown
using the MIMIC-IV v3.1 dataset because coding density, encounter spacing, POA
availability, and condition persistence vary across settings.

The unified interface provided by _medicalcoder_ further reduces analytic
complexity. Mixed ICD-9 and ICD-10 datasets, full and compact codes, diagnostic
and procedure codes, and algorithm-specific nuances are handled within a single
function call. This contrasts with existing workflows that require dataset
partitioning, multiple method calls, and post hoc reconciliation, each of which
can introduce error and irreproducibility.

The included ICD utility functions also support exploratory and preprocessing
tasks using the same reference data employed by the comorbidity algorithms. This
reduces reliance on ad hoc code lists and external reference tables when
defining cohorts or filtering diagnosis and procedure codes in large clinical
datasets.

Testing against published SAS, SQL, and R reference implementations
demonstrates that _medicalcoder_ produces results consistent with established
encounter-level algorithms while offering additional flexibility.

To our knowledge, no available R package jointly supports multiple
comorbidity algorithms, mixed ICD versions, present-on-admission logic, and
cumulative longitudinal flagging within a single, reproducible framework.
_medicalcoder_ has these capabilities and enables researchers to consistently
characterize patient complexity and evaluate how different
comorbidity definitions influence analytic results.

## Conclusions

_medicalcoder_ is a self-contained R package with an internal ICD database that
can apply many comorbidity algorithms while accounting for ICD-version,
ICD-type, POA, primary diagnosis, and full and compact ICD codes. Codes can be
flagged for individual encounters or longitudinally within a patient record.

{{< pagebreak >}}

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline}
The _medicalcoder_ R package is available from the Comprehensive R Archive Network
(CRAN) at https://cran.r-project.org/package=medicalcoder. Developmental
versions, additional testing suites, benchmarking, and examples are on GitHub at
https://github.com/dewittpe/medicalcoder/.

The MIMIC-IV v3.1 data is available from PhysioNet at
https://physionet.org/content/mimiciv/3.1/.

This manuscript was written using Quarto version
`r system("quarto --version", intern = TRUE)`
(https://quarto.org/) and `r R.version.string`.
All R code, materials, and dependencies can be found at
https://github.com/dewittpe/medicalcoder-jamia-open/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
_medicalcoder_ R package and drafted the manuscript. SR provided critical code
review and testing.  JAF, SR, MR, and TDB all contributed to the conception,
design, and interpretation and provided important feedback. TDB provided
funding. All authors critically reviewed the manuscript for important
intellectual content and approved of the final version of the manuscript.

[**Acknowledgments:**]{.underline} We thank Vincent Rubinetti for his assistance
in designing the _medicalcoder_ logo.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

**Figure 1:** Interface for the `medicalcoder::comorbidities()` function. The
function requires a data object containing one ICD code per row, the column
containing ICD codes, and the comorbidity algorithm specification (e.g.,
Charlson or Elixhauser variants). Optional arguments allow users to specify
encounter identifiers, ICD version, diagnosis type, present-on-admission
indicators, patient age, and primary diagnosis flags. Defaults control how
comorbidities are identified, including encounter-level (`flag.method =
"current"`) or longitudinal (`flag.method = "cumulative"`) evaluation and how ICD
codes are matched.

**Figure 1 ALT TEXT:** Diagram showing the argument structure of the R function
`medicalcoder::comorbidities()`. The figure lists three required inputs: a data
frame containing one ICD code per row (data), the column name containing ICD
codes (`icd.codes`), and the comorbidity method specification (`method`, such as
"charlson_quan2005"). Additional optional parameters allow users to specify
encounter identifiers, ICD version (9 or 10), diagnosis versus procedure
indicators, present-on-admission status, patient age (for Charlson algorithms),
and primary diagnosis flags. Default arguments include `flag.method = "current"`
for encounter-level detection, `full.codes = TRUE`, `compact.codes = TRUE`, and
`subconditions = FALSE`.

**Figure 2:** Percent of subjects and encounters in the MIMIC-IV dataset with
different comorbidity profiles under `flag.method = "cumulative"` versus
`flag.method = "current"`. The bars are not mutually exclusive.  "Different risk
profile" is an aggregation measure indicating any difference in the comorbidity
classification or count between cumulative and current flagging.
The remaining bars describe selected components of that aggregate, including
encounters with no encounter-level comorbidity but at least one cumulative
comorbidity, classification or severity differences with the same comorbidity
count, and differences involving additional comorbidities. Panel A uses subjects
with at least two encounters and at least one non-initial encounter with a
comorbidity flagged under flag.method = "cumulative" as the denominator. Panel B
uses non-initial encounters with at least one comorbidity flagged under
flag.method = "cumulative" as the denominator.

**Figure 2 ALT TEXT:** Two-panel grouped bar chart comparing how cumulative
longitudinal flagging changes comorbidity results relative to encounter-level
current flagging for three algorithms: Charlson, Elixhauser, and PCCC v3.1.
Panel A shows the percentage of subjects and Panel B shows the percentage of
encounters in four categories: different risk profile, zero encounter-level
comorbidities but at least one longitudinal comorbidity, only differences in
classification, and differences in classification and/or additional
comorbidities. Across both panels, the largest percentages are for different
risk profile and for differences in classification and/or additional
comorbidities, with Elixhauser generally highest, PCCC v3.1 intermediate, and
Charlson lowest; the other two categories are much smaller.

```{r}
#| label: mem-total-helper
#| include: false
```
```{r}
#| label: get-cpu-brand
#| include: false
```
```{r}
#| label: sessioninfo-with-memory-cpu
#| include: false
```

**Figure 3:** The median time (in seconds; bar height), and IQR (error bars),
required to flag comorbidities in the MIMIC-IV v3.1 dataset
(`r qwraps2::frmt(nrow(mimicivDT))` rows,
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects,
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
encounters) by utility (panels) and flagging method for N = 10 iterations.
The MIMIC-IV code time is based on running RSQLite code for an in-memory
database.  R packages _comorbidity_ and _pccc_ require multiple calls and
aggregations to handle the mixed use of ICD-9 and ICD-10 within the MIMIC-IV
dataset.  Applying each of the comorbidity algorithms and flagging methods
within _medicalcoder_ requires only modifying the `method` and `flag.method`
argument to the `medicalcoder::comorbidities()` call.  Some panels are sparse,
because each comparator supports fewer methods.  The
times were recorded when building the supplement in `r si$platform$version` on
`r si$platform$os` with an `r si$platform$CPU` and `r si$platform$RAM` RAM.

```{r}
#| label: for-figure3-alt-text
#| include: false
```

**Figure 3 ALT TEXT:** Four-panel grouped bar chart showing median time in
seconds to flag comorbidities in the MIMIC-IV v3.1 dataset (N = 10 iterations).
Panels compare four utilities: _comorbidity_, _medicalcoder_, MIMIC-IV code, and
_pccc_. The x-axis lists comorbidity algorithms: Charlson, Elixhauser, PCCC
version 2.0, and PCCC version 3.1. The y-axis is time in seconds from 0 to about
`r floor(max(f3_yaxis_range))`. Blue bars indicate cumulative flagging and maroon bars indicate current
flagging. Only _medicalcoder_ has cumulative bars as it is the only utility
natively supporting cumulative flagging.  _medicalcoder_ is the fastest in
every comparison shown, with times of about `r mdcr_f3_range[1]` to
`r mdcr_f3_range[2]` seconds across all four algorithms and flag methods.
_comorbidity_ is slower, at about `r cmrb_f3_char_median` seconds for Charlson,
`r cmrb_f3_elix_median` seconds for Elixhauser. The MIMIC-IV code panel shows a
single Charlson result at about `r mmic_f3_median` seconds. The _pccc_ panel
shows only PCCC version 2.0, with about `r pccc_f3_median` seconds for current
flagging.  The times were recorded when building the supplement in
`r si$platform$version` on `r si$platform$os` with an `r si$platform$CPU` and
`r si$platform$RAM` RAM.
