---
title: "medicalcoder vs other R packages"

format:
  pdf:
    toc: true
    number-sections: true

bibliography: references.bib
---

# Objective and Setup

Compare results and time to compute between `medicalcoder::comorbidities()` and
other implementations of comorbidity algorithms.

```{r}
#| label: setup
#| include: false
#| cache: false
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  cache = TRUE
)

pkg_with_version <- function(pkg) {
  sprintf("%s v%s", pkg, packageVersion(pkg))
}

tictoc_ratio <- function(numerator, denominator) {
  formatC(
    attr(numerator, "tictoc") / attr(denominator, "tictoc"),
    format = "d"
  )
}

loadNamespace('medicalcoder')
loadNamespace('data.table')
```

## Data Prep
For this exercise we will use the `medicalcoder::mdcr` dataset.

```{r}
#| label: data-setup
#| echo: true
#| results: "markup"
mdcrDT <- data.table::copy(medicalcoder::mdcr)
data.table::setDT(mdcrDT)
data.table::setkey(mdcrDT, patid)
str(mdcrDT)
```

## Flagging Comorbidities via ``medicalcoder::comorbidities()`
Here we apply several comorbidity algorithms to the `mdcrDT` dataset via
`medicalcoder::comorbidities()`.  The results are stored, along with a timing
for the work which will be used to compare to other implementations.
We will make several calls to `medicalcoder::comorbidities()` with only the
`method` argument changing.  We also record the system time before and after the
calls so we can report the time required to apply the methods.
As such we define a function to simplify the calls.

```{r}
#| label: medicalcoder-setup
call_medicalcoder <- function(method, ...) {
  tic <- Sys.time()
  rtn <-
    medicalcoder::comorbidities(
      data      = mdcrDT,
      id.vars   = "patid",
      icd.codes = "code",
      dx.var    = "dx",
      icdv.var  = "icdv",
      poa       = 1L, # assume all codes are present on admission
      primarydx = 0L, # assume all codes are secondary diagnoses
      method    = method,
      ...
  )
  toc <- Sys.time()
  attr(rtn, "tictoc") <- as.numeric(difftime(toc, tic, units = "secs"))
  rtn
}
```

```{r}
#| label: medicalcoder-warm-up
#| include: false
# to eliminate the time on the first call resulting from namespace loading run
# the code once here and then again for the timing, the full set need not be
# run, just enough to get all the namespaces loaded and data sets built.
call_medicalcoder(method = "charlson_quan2005")
```

```{r}
#| label: medicalcoder-calls
medicalcoder_charlson_quan2005 <-
  call_medicalcoder(method = "charlson_quan2005")

medicalcoder_charlson_quan2011 <-
  call_medicalcoder(method = "charlson_quan2011")

medicalcoder_elixhauser_quan2005 <-
  call_medicalcoder(method = "elixhauser_quan2005")

medicalcoder_pccc_v2.0 <-
  call_medicalcoder(method = "pccc_v2.0")
```

# `r pkg_with_version("medicalcoder")` vs \ldots

## `r pkg_with_version("comorbidity")`

### Data Prep

To use the R package comorbidity [@r-comorbidity] we need to split the data by
ICD version, apply `comorbidity::comorbidity()` to each split, and reconcile the
two sets.  This need to reconcile the results is critcally important as there
are many patients with both ICD-9 and ICD-10 codes.

```{r}
#| label: comorbidity-setup
# how many patid have both ICD-9 and ICD-10 codes?
mdcrDT[
  ,
  .(both_icdv = any(icdv == 9) & any(icdv == 10)),
  by = .(patid)][(both_icdv)
  ][
  ,
  data.table::uniqueN(patid)
  ]

# split by icvd
mdcr_icd9  <- subset(mdcrDT, icdv ==  9L)
mdcr_icd10 <- subset(mdcrDT, icdv == 10L)

# we will add rows without a code for each patid not in the subsets.
mdcr_icd9 <-
  rbind(
    mdcr_icd9,
    data.table::data.table(
      patid = setdiff(mdcrDT$patid, mdcr_icd9$patid),
      icdv = 9L,
      code = "",
      dx = 1L
    )
  )

mdcr_icd10 <-
  rbind(
    mdcr_icd10,
    data.table::data.table(
      patid = setdiff(mdcrDT$patid, mdcr_icd10$patid),
      icdv = 10L,
      code = "",
      dx = 1L
    )
  )
```

Charlson and Elixhauser comorbidities only require the use of diagnosic codes.
The `mdcr_icd9` and `mdcr_icd10` objects contain both diagnosic and procedure
codes.  Since `medicalcoder::comorbidities()` is able to handle diagnosic and
procedure codes via function argument we leave the subsetting of the data to
diagnosic codes to the timed calls to `comorbidity::comorbidity()` forth coming.

### Charlson

Now that the data has been split and prepared, we can apply the
`comorbidity::comorbidity()` algorithms and aggregate the results.

```{r}
#| label: comorbidity-charlson
tic <- Sys.time()

comorbidity_charlson_icd9_results <-
  comorbidity::comorbidity(
    x = subset(mdcr_icd9, dx == 1L),
    id = "patid",
    code = "code",
    map = "charlson_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe
                   # comorbidities is also flagged
  )

comorbidity_charlson_icd10_results <-
  comorbidity::comorbidity(
    x = subset(mdcr_icd10, dx == 1L),
    id = "patid",
    code = "code",
    map = "charlson_icd10_quan",
    assign0 = TRUE
  )

# combine the ICD-9 and ICD-10 results into one set
comorbidity_charlson_results <-
  rbind(comorbidity_charlson_icd9_results, comorbidity_charlson_icd10_results)

# aggregate to find the flags across each patient record.
comorbidity_charlson_results <-
  aggregate(. ~ patid, data = comorbidity_charlson_results, FUN = max)

# add the attributes to the combine set
attributes(comorbidity_charlson_results)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_charlson_icd9_results)[c("class", "variable.labels", "map")]

comorbidity_charlson_results[["score"]] <-
  comorbidity::score(
    x = comorbidity_charlson_results,
    weights = "quan",
    assign0 = TRUE
  )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
comorbidity_charlson_results[["score"]] <-
  as.integer(comorbidity_charlson_results[["score"]])

# stop timming
toc <- Sys.time()

# record time needed to build the charlson results
attr(comorbidity_charlson_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))
```

### Elixhauser


```{r}
#| label: comorbidity-elixhauser
tic <- Sys.time()

comorbidity_elixhauser_icd9_results <-
  comorbidity::comorbidity(
    x = subset(mdcr_icd9, dx == 1L),
    id = "patid",
    code = "code",
    map = "elixhauser_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe
                   # comorbidities is also flagged
  )

comorbidity_elixhauser_icd10_results <-
  comorbidity::comorbidity(
    x = subset(mdcr_icd10, dx == 1L),
    id = "patid",
    code = "code",
    map = "elixhauser_icd10_quan",
    assign0 = TRUE
  )

# combine the ICD-9 and ICD-10 results into one set
comorbidity_elixhauser_results <-
  rbind(comorbidity_elixhauser_icd9_results, comorbidity_elixhauser_icd10_results)

comorbidity_elixhauser_results <-
  aggregate(. ~ patid, data = comorbidity_elixhauser_results, FUN = max)

# add the attributes to the combine set
attributes(comorbidity_elixhauser_results)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_elixhauser_icd9_results)[c("class", "variable.labels", "map")]

toc <- Sys.time()

attr(comorbidity_elixhauser_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))
```

### Compare Results

Expected computation time for `comorbidity::comorbidity()` is greater than the
expected computation time for `medicalcoder::comorbidities()`.  This is mostly due to the need for two
calls to `comorbidity::comorbidity()`, one for each ICD version, and
aggregation.
`comorbidity::comorbidity()` does uses `data.table` for performance, and so does
`medicalcoder::comorbidities()` in this example.  Still,
`medicalcoder::comorbidities()` is
`r tictoc_ratio(comorbidity_charlson_results, medicalcoder_charlson_quan2005)`
times faster for Charlson comorbidities, and
`r tictoc_ratio(comorbidity_elixhauser_results, medicalcoder_elixhauser_quan2005)`
times faster for Elixhauser comorbidities.

```{r}
#| label: medicalcoder-vs-comorbidity-timing
attr(medicalcoder_charlson_quan2005, "tictoc")   # seconds
attr(medicalcoder_charlson_quan2011, "tictoc")   # seconds
attr(medicalcoder_elixhauser_quan2005, "tictoc") # seconds
attr(comorbidity_charlson_results, "tictoc")     # seconds
attr(comorbidity_elixhauser_results, "tictoc")   # seconds
```

The results are consistent as well.

```{r}
#| label: define-column-matches-between-medicalcoder-and-comorbidity
charlson_delta <-
  merge(
    x = comorbidity_charlson_results,
    y = medicalcoder_charlson_quan2011,
    all = TRUE,
    by = "patid"
  )

charlson_columns <- data.table::fread(text = "
Condition                          | comorbidity | medicalcoder
AIDS                               | aids        | aidshiv
Cancer (Any malignancy)            | canc        | mal
Cancer (Metastatic solid tumor)    | metacanc    | mst
Cerebrovascular disease            | cevd        | cebvd
Chronic pulmonary disease          | cpd         | copd
Congestive heart failure           | chf         | chf
Dementia                           | dementia    | dem
Hemiplegia or paraplegia           | hp          | hp
Myocardial infarction              | mi          | mi
Peripheral vascular disease        | pvd         | pvd
Peptic ulcer disease               | pud         | pud
Rheumatic disease                  | rheumd      | rhd
Diabetes (uncomplicated)           | diab        | dm
Diabetes (complicated)             | diabwc      | dmc
Renal disease                      | rend        | rnd
Liver disease (mild)               | mld         | mld
Liver disease (moderate or severe) | msld        | msld
Score (based on Quan 2011)         | score       | cci
  ")

for (i in seq_len(nrow(charlson_columns))) {
  x <- charlson_columns[["comorbidity"]][i]
  y <- charlson_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(
    identical(charlson_delta[[X]], charlson_delta[[Y]]),
    list(X = x, Y = y)
  )
  print(e)
  r <- eval(e)
  print(r)
  stopifnot(r)
  # if you are here then the results are the same, remove the columns from the
  # charlson_delta dataset
  charlson_delta[[x]] <- NULL
  charlson_delta[[y]] <- NULL
}
```

The only remaining non-id columns are all from `medicalcoder::comorbidities()`.
```{r}
#| label: charlson-delta-summary
stopifnot(
  identical(
    names(charlson_delta),
    c("patid", "num_cmrb", "cmrb_flag", "age_score")
  )
)
summary(medicalcoder_charlson_quan2005$num_cmrb)
summary(medicalcoder_charlson_quan2005$cmrb_flag)
summary(medicalcoder_charlson_quan2005$age_score)
```

* `num_cmrb`: number of comorbidities flagged for the patid
* `cmrb_flag`: indicator, 0 if no comorbidities are flagged, 1 if at least one
  comorbidity is flagged.
* `age_score`: NA in this example as age was not part of the scoring.  Age
  scores are reported when age is available.

When comparing the results between `medicalcoder::comorbidities()` and
`comorbidity::comorbidity()` for Elixhauser we again find consistent results.

```{r}
#| label: elixhauser-delta-compare
elixhauser_delta <-
  merge(
    x = comorbidity_elixhauser_results,
    y = medicalcoder_elixhauser_quan2005,
    all = TRUE,
    by = "patid"
  )

elixhauser_columns <- data.table::fread(text = "
Condition                                     | comorbidity | medicalcoder
AIDS/HIV                                      | aids        | AIDS
Alcohol abuse                                 | alcohol     | ALCOHOL
Anemias (Blood loss anaemia)                  | blane       | BLDLOSS
Anemias (Deficiency anaemia)                  | dane        | ANEMDEF
Cardiac Arrhythmias                           | carit       | CARDIAC_ARRHYTHMIAS
Cancer (Solid tummor without metastasis)      | solidtum    | TUMOR
Cancer (Metastatic)                           | metacanc    | METS
Chronic pulmonary disease                     | cpd         | CHRNLUNG
Coagulopathy                                  | coag        | COAG
Congestive Heart Failure                      | chf         | CHF
Depression                                    | depre       | DEPRESS
Diabetes (Complicated)                        | diabc       | DMCX
Diabetes (Uncomplicated)                      | diabunc     | DM
Drug abuse                                    | drug        | DRUG
Fluid and electrolyte disorders               | fed         | LYTES
Hypertension (Complicated)                    | hypc        | HTN_CX
Hypertension (Uncomplicated)                  | hypunc      | HTN_UNCX
Hypothyroidism                                | hypothy     | HYPOTHY
Liver disease                                 | ld          | LIVER
Lymphoma                                      | lymph       | LYMPH
Obesity                                       | obes        | OBESE
Other neurological disease                    | ond         | NEURO
Paralysis                                     | para        | PARA
Peptic ulcer disease excluding bleeding       | pud         | ULCER
Peripheral Vascular Disorders                 | pvd         | PERIVASC
Psychoses                                     | psycho      | PSYCH
Pulmonary Circulation Disorders               | pcd         | PULMCIRC
Renal Failure                                 | rf          | RENLFAIL
Rheumatoid artritis/collaged vascular disease | rheumd      | ARTH
Valvular disease                              | valv        | VALVE
Weight loss                                   | wloss       | WGHTLOSS
  ")
data.table::setDF(elixhauser_columns)

for (i in seq_len(nrow(elixhauser_columns))) {
  x <- elixhauser_columns[["comorbidity"]][i]
  y <- elixhauser_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(
    identical(elixhauser_delta[[X]], elixhauser_delta[[Y]]),
    list(X = x, Y = y)
  )
  print(e)
  r <- eval(e)
  print(r)
  stopifnot(r)
  elixhauser_delta[[x]] <- NULL
  elixhauser_delta[[y]] <- NULL
}
```
The remaining non-id columns are unique to `medicalcoder::comorbidities()`.
```{r}
#| label: medicalcoder-vs-comorbidity-elixhauser-remainder
stopifnot(
  identical(
    names(elixhauser_delta),
    c("patid", "HTN_C",
      "num_cmrb", "cmrb_flag",
      "mortality_index", "readmission_index")
  )
)
str(elixhauser_delta)
```

`HTN_C` is any hypertension and is the same as an OR statement between
uncomplicated and complicated hypertension.

```{r}
#| label: medicalcoder-vs-comorbidity-htnc
# the HTN_C is only in the medicalcoder results
stopifnot(
  identical(
    medicalcoder_elixhauser_quan2005[["HTN_C"]],
    as.integer(
      medicalcoder_elixhauser_quan2005[["HTN_UNCX"]] |
        medicalcoder_elixhauser_quan2005[["HTN_CX"]]
    )
  )
)
```
The `mortality_index` and `readmission_index`

```{r}
#| label: medicalcoder-vs-comorbidity-index
summary(medicalcoder_elixhauser_quan2005$mortality_index)
summary(medicalcoder_elixhauser_quan2005$readmission_index)
```

## `r pkg_with_version("icdcomorbid")`

### Data Prep

icdcomorbid [@r-icdcomorbid] requires the input data to be in a "wide" format.
The package provides a tool for this but it takes considerably more time to run
than alternatives.

```{r}
#| label: icdcomorbid-data-format

#  mdcr_icd9dx_wide <-
#    icdcomorbid::long_to_wide(
#      df = mdcr_icd9dx,
#      idx = "patid",
#      icd_cols = "code"
#    )
#
# killed process after 15 seconds, data.table is much faster

tic <- Sys.time()

mdcr_wide <-
  split(subset(mdcrDT, dx == 1L), by = "icdv") |>
  lapply(function(DT) {
    DT[, DX := sprintf("DX%02d", seq_along(code)), by = .(patid)]
  }) |>
  lapply(data.table::dcast, patid ~ DX, value.var = "code")

toc <- Sys.time()

difftime(toc, tic)
```

### Charlson

We now apply the Charlson comorbidity algorithm to the data

```{r}
#| label: icdcomorbid
tic <- Sys.time()

icdcomorbid_charlson_icd9_results <-
  icdcomorbid::icd9_to_comorbid(
    df = mdcr_wide[["9"]],
    idx = "patid",
    icd_cols = grep("^DX", names(mdcr_wide[["9"]]), value = TRUE),
    mapping = "charlson9"
  )

icdcomorbid_charlson_icd10_results <-
  icdcomorbid::icd10_to_comorbid(
    df = mdcr_wide[["10"]],
    idx = "patid",
    icd_cols = grep("^DX", names(mdcr_wide[["10"]]), value = TRUE),
    mapping = "charlson10"
  )

icdcomorbid_charlson_results <-
  rbind(
    data.table::setDT(icdcomorbid_charlson_icd9_results),
    data.table::setDT(icdcomorbid_charlson_icd10_results),
    use.names = TRUE,
    fill = TRUE
  )

icdcomorbid_charlson_results <-
  aggregate(
    . ~ patid,
    data = icdcomorbid_charlson_results,
    FUN = any,
    na.rm = TRUE
  )

toc <- Sys.time()

attr(icdcomorbid_charlson_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))
```

### Compare Results

To process the `r length(unique(mdcrDT$patid))` patient records in the `mdcrDT`
data set, icdcomorbid took
`r tictoc_ratio(icdcomorbid_charlson_results, medicalcoder_charlson_quan2005)`
times longer to process the data than medicalcoder did.


```{r}
#| label: medicalcoder-vs-icdcomorbid-timing
attr(medicalcoder_charlson_quan2005, "tictoc") # seconds
attr(icdcomorbid_charlson_results, "tictoc")   # seconds
```

The return from icdcomorbid are booleans. We will coerce to integers before
comparing to medicalcoder.

```{r}
#| label: icdcomorbid-to-int
data.table::setDT(icdcomorbid_charlson_results)
for (j in setdiff(names(icdcomorbid_charlson_results), "patid")) {
  data.table::set(
    icdcomorbid_charlson_results,
    j = j,
    value = as.integer(icdcomorbid_charlson_results[[j]])
  )
}
mdcr_v_icdcomorbid <-
  merge(
    x = medicalcoder_charlson_quan2011,
    y = icdcomorbid_charlson_results,
    all = TRUE,
    by = c("patid")
  )
```

Let's compare the results starting the AIDS/HIV.
```{r}
#| label: icdcomorbid-aidshiv-compare
identical(
  mdcr_v_icdcomorbid[["aidshiv"]],  # medicalcoder::comorbidities()
  mdcr_v_icdcomorbid[["aids_hiv"]]  # icdcomorbid
)

mdcr_v_icdcomorbid[aidshiv != aids_hiv, .(patid, aidshiv, aids_hiv)]
```

There are differences in the flags for AIDS/HIV, all false-negatives from
icdcomorbid.  The issue, as shown below, is that ICD-9-CM code 042 is not
flagged as AIDS/HIV by icdcomorbid.  This appears to be due to the string
"042.x" using within the json files within icdcomorbid for mapping codes to
comorbidities.  Table 1 witin Quan (2005) [@quan2005] reports "042.x-044.x"
which should match any ICD-9 code under the three digit code 042, 043, or 044.
(Note: 043 and 044 were removed from the ICD-9-CM standard starting October 1,
1994 (fiscal year 1995) which is before the earliest set of codes medicalcoder
has in its internal database.)

```{r}
#| label: icdcomorbid-aidshiv-codes
merge(
  x = mdcrDT[patid %in% mdcr_v_icdcomorbid[aidshiv != aids_hiv, patid]],
  y = subset(
        x      = medicalcoder::get_charlson_codes(),
        subset = condition == "aidshiv",
        select = c("code", "icdv", "dx", "condition", "charlson_quan2005")
        ),
  by = c("code", "icdv", "dx")
)
```

```{r}
#| label: icdcomorbid-drop-aidshiv
mdcr_v_icdcomorbid[, aidshiv := NULL]
mdcr_v_icdcomorbid[, aids_hiv := NULL]
```

Next, congestive heart failure.
```{r}
#| label: icdcomorbid-chf-compare
identical(
  mdcr_v_icdcomorbid[["chf"]],  # medicalcoder::comorbidities()
  mdcr_v_icdcomorbid[["congestive_heart_failure"]]  # icdcomorbid
)
mdcr_v_icdcomorbid[
  chf != congestive_heart_failure,
  .(patid, chf, congestive_heart_failure)
]

mdcr_v_icdcomorbid[, .N, keyby = .(chf, congestive_heart_failure)]
```

Focusing on the differences:
```{r}
#| label: icdcomorbid-chf-diff-count
mdcr_v_icdcomorbid[
  !Vectorize(identical)(chf, congestive_heart_failure),
  .N,
  keyby = .(chf, congestive_heart_failure)]
```

There are records which are missing form the results from icdcomorbid.  In these
cases, it appears that the missing rows are from records with no comorbidities
flagged. It should be noted that not all cases of `num_cmrb == 0` have missing
values for `congestive_heart_failure`.

```{r}
#| label: icdcomorbid-chf-missing
mdcr_v_icdcomorbid[num_cmrb == 0, .N]
mdcr_v_icdcomorbid[is.na(congestive_heart_failure), all(num_cmrb == 0)]
```

What about the cases where icdcomorbid flagged congestive_heart_failure and
medicalcoder did not
```{r}
#| label: icdcomorbid-chf-false-positives
# where does icdcomorbid flag that medicalcoder does not?
DT <-
  mdcrDT[patid %in% mdcr_v_icdcomorbid[congestive_heart_failure > chf, patid]]

all(DT$icdv == 10L)

DT <- DT[, unique(.SD), .SDcols = c("code")]
DT[, code_id := 1:.N]
DT[, DX := "DX"]
DT <- data.table::dcast(DT, code_id ~ DX, value.var = "code")

false_positives <-
  icdcomorbid::icd10_to_comorbid(
    df = DT,
    idx = "code_id",
    icd_cols = "DX", mapping = "charlson10"
  )
data.table::setDT(false_positives)

false_positives[(congestive_heart_failure), .N]

DT[
  false_positives[(congestive_heart_failure)], on = "code_id"
  ][
  , .(DX, congestive_heart_failure)
  ]
```

* P2930 (full code P29.30) should not be part of the Charlson comorbidities for
  CHF, I'm not sure why this is flagging at the moment, the JSON files within
  icdcomorbid report the code P29.0 only, which is correct and consistent with
  Table 1 of [@quan2005].

* I1310 (full code I13.10) should not flag based on what I see in the
  icdcomorbid JSON files.

```{r}
#| label: icdcomorbid-json-snippet
cat(
  scan(
    system.file(package = "icdcomorbid", "comorbidity_mappings", "charlson10.json"),
    what = "character", sep = "\n", quiet = TRUE
    )[3]
)
```

And what about the cases where medicalcoder flags and icdcomorbid does not?
```{r}
#| label: icdcomorbid_false_negatives
icdcomorbid_false_negatives <-
  merge(
    x = mdcrDT[patid %in% mdcr_v_icdcomorbid[congestive_heart_failure < chf, patid]],
    y = subset(
          x      = medicalcoder::get_charlson_codes(),
          subset = condition == "chf",
          select = c("code", "icdv", "dx", "condition", "charlson_quan2005")
          ),
    by = c("code", "icdv", "dx")
  )
str(icdcomorbid_false_negatives)

unique(icdcomorbid_false_negatives[icdv == 9 & dx == 1, .(code)][["code"]])
unique(icdcomorbid_false_negatives[icdv == 10 & dx == 1, .(code)][["code"]])
```

These are all false negatives from icdcomorbid.

Given the false positive and false negative results, along with several orders
of magnitude longer computation time, we conclude here that medicalcoder is
superior to icdcomorbid.

There are many more conditions to explore, but the above should be sufficent to
support using medicalcoder over icdcomorbid.

## `r pkg_with_version("multimorbidity")`

### Data Prep

The R package multimorbidity [@r-multimorbidity] can apply Charlson and
Elixhauser comorbidities.  The call to `multimorbidity::charlson()` requires
that the input data "has been properly prepared using
`multimorbidity::prepare_data()`."  This is not a suggestion.  Failure to do so
properly will result in errors.  For example, not using the column name `dx` for
the diagnostic code can result in an error.

```{r}
#| label: multimorbidity-data-prep-issue-without-error
DF <-
  data.frame(
    code_id = c("ICD-9-CM 042", "ICD-10-CM B20"),
    dx      = c("042", "B20"),
    icdv = c(9, 10)
  )

multimorbidity::charlson(
  dat = DF,
  id = code_id,
  dx = dx,
  version_var = icdv
) |>
dplyr::select(id, charlson_hiv)
```

Now if we rename the dx to code and call the method again we will get an error.
```{r}
#| label: multimorbidity-data-prep-issue-with-error
#| error: true
DF <- dplyr::rename(DF, code = dx)
DF
multimorbidity::charlson(
  dat = DF,
  id = code_id,
  dx = code,
  version_var = icdv
)
```

To prepare the data for calling `multimorbidity::charlson()` we do the
following:

```{r}
#| label: multimorbidity-data-prep
mdcrDT2 <- data.table::copy(mdcrDT)
mdcrDT2 <- mdcrDT2[dx == 1L]
mdcrDT2[, dx := NULL]
data.table::setnames(mdcrDT2, old = c("code"), new = c("dx"))
mdcrDT2 <-
  rbind(mdcrDT2,
    data.table::data.table(
      patid = setdiff(mdcrDT$patid, mdcrDT2$patid),
      icdv = 9L,
      dx = ""
    )
  )
```

### Charlson

The following should flag Charlson comorbidities via
`multimorbidity::charlson()`.

```{r}
#| label: multimorbidity-charlson
tic <- Sys.time()

multimorbidity_charlson_results <-
  multimorbidity::charlson(
    dat = mdcrDT2,
    id  = patid,
    version_var = icdv
  )

toc <- Sys.time()

attr(multimorbidity_charlson_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))
```

### Compare Results

`multimorbidity::charlson` took
`r tictoc_ratio(multimorbidity_charlson_results, medicalcoder_charlson_quan2005)`
times longer to process the `r length(unique(mdcrDT$patid))` patient records in
`mdcrDT`.

```{r}
#| label: medicalcoder-vs-multimorbidity-timing
attr(medicalcoder_charlson_quan2005, "tictoc")  # seconds
attr(medicalcoder_charlson_quan2011, "tictoc")  # seconds
attr(multimorbidity_charlson_results, "tictoc") # seconds
```

```{r}
#| label: multimorbidity-coerce-merge
# the return form multimorbidity have numeric columns for the indicators not
# integers
for (j in names(multimorbidity_charlson_results)) {
  multimorbidity_charlson_results[[j]] <-
    as.integer(multimorbidity_charlson_results[[j]])
}
mdcr_v_multimorbidity <-
  merge(
    x = medicalcoder_charlson_quan2011,
    y = multimorbidity_charlson_results,
    all = TRUE,
    by.x = "patid",
    by.y = "id"
  )
```

Some of the condition flags are the same:
```{r}
#| label: multimorbidity-matching-flags
stopifnot(
  identical(
    mdcr_v_multimorbidity[["aidshiv"]],     # medicalcoder
    mdcr_v_multimorbidity[["charlson_hiv"]] # multimorbidity
  )
  ,
  identical(
    mdcr_v_multimorbidity[["cebvd"]],           # medicalcoder
    mdcr_v_multimorbidity[["charlson_cerebro"]] # multimorbidity
  )
  ,
  identical(
    mdcr_v_multimorbidity[["hp"]],                # medicalcoder
    mdcr_v_multimorbidity[["charlson_hemi_para"]] # multimorbidity
  )
  ,
  identical(
    mdcr_v_multimorbidity[["pud"]],                 # medicalcoder
    mdcr_v_multimorbidity[["charlson_peptic_ulcer"]] # multimorbidity
  )
  ,
  identical(
    mdcr_v_multimorbidity[["rhd"]],                 # medicalcoder
    mdcr_v_multimorbidity[["charlson_rheum"]] # multimorbidity
  )
)
```

Other conditions have differences in the flags between
`medicalcoder::comorbidities()` and `multimorbidity::charlson()`.

For example, dementia was not flagged by `multimorbidity::charlson().
```{r}
#| label: multimorbidity-dementia-diff
mdcr_v_multimorbidity[, .N, keyby = .(dem, charlson_dementia)]

dem_fn <-
  merge(
    x = mdcrDT[patid %in% mdcr_v_multimorbidity[dem != charlson_dementia, patid]],
    y = subset(
          x      = medicalcoder::get_charlson_codes(),
          subset = condition == "dem",
          select = c("code", "icdv", "dx", "condition", "charlson_quan2005")
          ),
    by = c("code", "icdv", "dx")
  )
unique(dem_fn[, .(code, icdv, dx)])
medicalcoder::lookup_icd_codes("2941")
medicalcoder::lookup_icd_codes("29410")
```

ICD-9-CM 294.10 is a valid code that should map to Dementia (Table 1 of
[@quan2005].  Likely issue is that Table 1 of [@quan2005] reports the four digit
"249.1" and not the five digit "249.10". The more granular 249.10 should still
flag, this is a flag negative from `multimorbidity::charlson()`.

Chronic pulmonary disease (COPD), has differences too.
```{r}
#| label: multimorbidity-copd-diff
mdcr_v_multimorbidity[, .N, keyby = .(copd, charlson_chronic_pulm)]
copd_fn <-
  merge(
    x = mdcrDT[patid %in% mdcr_v_multimorbidity[copd != charlson_chronic_pulm, patid]],
    y = subset(
          x      = medicalcoder::get_charlson_codes(),
          subset = condition == "copd",
          select = c("code", "icdv", "dx", "condition", "charlson_quan2005")
          ),
    by = c("code", "icdv", "dx")
  )
unique(copd_fn[, .(code, icdv, dx)])
medicalcoder::lookup_icd_codes("490")
subset(
  medicalcoder::lookup_icd_codes("^490\\d*", regex = TRUE),
  dx == 1
)

medicalcoder::lookup_icd_codes("I278[12]", regex = TRUE)
```

ICD-9 490 is a valid diagnostic code.  The false negative from the
`multimorbidity::charlson()` is likely do to the  use of "490.x" in Table 1 of
[@quan2005].

The ICD-10 I27.81 and I27.82 codes are likely false negatives form
`medicalcoder::charlson()` as [@quan2005] only reports the four digit I27.8 code
in Table 1.


Diabetes has two levels, complicated and uncomplicated.  The results between
`medicalcoder::comorbidities()` and `multimorbidity::charlson()` are consistent
for complicated diabetes, but not the uncomplicated.
```{r}
#| label: multimorbidity-diabetes-compare
identical(
  mdcr_v_multimorbidity[["dmc"]],            # medicalcoder
  mdcr_v_multimorbidity[["charlson_diab_c"]] # multimorbidity
)
identical(
  mdcr_v_multimorbidity[["dm"]],              # medicalcoder
  mdcr_v_multimorbidity[["charlson_diab_uc"]] # multimorbidity
)

mdcr_v_multimorbidity[, .N, keyby = .(dm, charlson_diab_uc)]
mdcr_v_multimorbidity[, .N, keyby = .(dmc, charlson_diab_c)]
mdcr_v_multimorbidity[, .N, keyby = .(dm, dmc, charlson_diab_uc, charlson_diab_c)]
```
The above indicates that records can be flagged as both uncomplicated and
complicated diabetes by `multimorbidity::charlson()` whereas
`medicalcoder::comorbidities()` returns only the more sever level.

```{r}
#| label: multimorbidity-diabetes-false-negatives
dm_fn <-
  merge(
    x = mdcrDT[patid %in% mdcr_v_multimorbidity[dm > charlson_diab_uc, patid]],
    y = subset(
          x      = medicalcoder::get_charlson_codes(),
          subset = condition == "dm",
          select = c("code", "icdv", "dx", "condition", "charlson_quan2005")
          ),
    by = c("code", "icdv", "dx")
  )
unique(dm_fn[, .(code, icdv, dx)])
medicalcoder::lookup_icd_codes(c("E13.6", "E13.69"))
```
A quick spot check of the codes indicate more false negatives from
`multimorbidity::charlson()`.  For example ICD-10-CM E13.6 is listed in Table 1
of [@quan2005] as diabetes without chronic complication.  It appears the five
digit E13.6x codes are missed by `multimorbidity::charlson()`.

Additonal mismatches:
```{r}
#| label: multimorbidity-other-mismatches
mdcr_v_multimorbidity[, .N, keyby = .(mal, charlson_malig)]
mdcr_v_multimorbidity[, .N, keyby = .(mst, charlson_met_solid)]
mdcr_v_multimorbidity[, .N, keyby = .(chf, charlson_chf)]
mdcr_v_multimorbidity[, .N, keyby = .(mld, charlson_mild_liv)]
mdcr_v_multimorbidity[, .N, keyby = .(msld, charlson_mod_sev_liv)]
mdcr_v_multimorbidity[, .N, keyby = .(mi, charlson_myocar)]
mdcr_v_multimorbidity[, .N, keyby = .(pvd, charlson_periph_vasc)]
mdcr_v_multimorbidity[, .N, keyby = .(rnd, charlson_renal)]
```

Given that `medicalcoder::comorbidities()` is considerably faster than
`multimorbidity::charlson()` and `multimorbidity::charlson()` has demonstribable
false negavtives, we conclude `medicalcoder` is supperior to `multimorbidity`
for flagging Charlson comorbidities.

## `r pkg_with_version("pccc")`

### Data Prep

The pccc package [@r-pccc] required the data to be in a wide format and applied
the scoring by ICD version.

```{r}
#| label: oldpccc
mdcr_wide <-
  mdcrDT[][,
    variable := data.table::fcase(
      dx == 0, sprintf("PR%02d", seq_along(code)),
      dx == 1, sprintf("DX%02d", seq_along(code))
    ),
    by = .(patid, icdv)
  ]
mdcr_wide <-
  data.table::dcast(mdcr_wide, patid + icdv ~ variable, value.var = "code")
```

```{r}
#| label: pccc_results
tic <- Sys.time()

pccc_results <-
  rbind(
    pccc::ccc(
      data = mdcr_wide[icdv == 9L],
      id = patid,
      dx_cols = grep("^DX\\d+", names(mdcr_wide), value = TRUE),
      pc_cols = grep("^PR\\d+", names(mdcr_wide), value = TRUE),
      icdv = 9
    )
    ,
    pccc::ccc(
      data = mdcr_wide[icdv == 10L],
      id = patid,
      dx_cols = grep("^DX\\d+", names(mdcr_wide), value = TRUE),
      pc_cols = grep("^PR\\d+", names(mdcr_wide), value = TRUE),
      icdv = 10
    )
  )

pccc_results <-
  aggregate(
    . ~ patid,
    data = pccc_results,
    FUN = max,
    na.rm = TRUE
  )

toc <- Sys.time()

attr(pccc_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))
```

### Compare Results

The time required to apply PCCC version 2 [@feudtner2014pediatric] to the data
set is considerably different between `pccc::ccc()` and `medicalcoder::comorbidities()`.
medicalcoder was
`r tictoc_ratio(pccc_results, medicalcoder_pccc_v2.0)`
times faster.

```{r}
#| label: medicalcoder-vs-pccc-timing
attr(medicalcoder_pccc_v2.0, "tictoc") # seconds
attr(pccc_results, "tictoc") # seconds
```

The speed difference is impressive, but it only matters if the results are the
same or better. We start by checking the flagging for the conditions.

```{r}
#| label: medicalcoder-vs-pccc-results
mdcr_v_pccc <-
  merge(
    x = medicalcoder_pccc_v2.0,
    y = pccc_results,
    all = TRUE,
    by = "patid",
    suffixes = c("_mdcr", "_pccc")
  )
```

Between the two methods, the flag for any condition are the same
```{r}
#| label: mdcr-v-pccc-flag
stopifnot(
  identical(
    mdcr_v_pccc[["cmrb_flag"]],  # medicalcoder::comorbidities()
    mdcr_v_pccc[["ccc_flag"]]    # pccc::ccc()
  )
)

# remove these columns from mdcr_v_pccc.  We'll remove the columns after
# comparisons as a way to show that we've covered all the information in
# mdcr_v_pccc.
mdcr_v_pccc[, cmrb_flag := NULL]
mdcr_v_pccc[, ccc_flag  := NULL]
```

For the conditions we also have agreement between the two packages:
```{r}
#| label: mdcr-v-pccc-conditions
common_names <-
  setdiff(
    intersect(names(medicalcoder_pccc_v2.0), names(pccc_results)),
    "patid"
  )
for (j in common_names) {
  xj <- sprintf("%s_mdcr", j)
  yj <- sprintf("%s_pccc", j)
  if (identical(mdcr_v_pccc[[xj]], mdcr_v_pccc[[yj]])) {
    print(sprintf("medicalcoder and pccc flagged %s the same.", j))
    data.table::set(x = mdcr_v_pccc, j = xj, value = NULL)
    data.table::set(x = mdcr_v_pccc, j = yj, value = NULL)
  } else {
    print(sprintf("medicalcoder and pccc did not flagged %s the same.", j))
  }
}
```

What remains are columns with different names or which were unique to
medicalcoder.  Let's check the technology dependence flag.

```{r}
#| label: medicalcoder-vs-pccc-tech-dep
identical(
  mdcr_v_pccc[["any_tech_dep"]], # medicalcoder::comorbidities()
  mdcr_v_pccc[["tech_dep"]]      # pccc::ccc()
)

all(mdcr_v_pccc[["any_tech_dep"]] >= mdcr_v_pccc[["tech_dep"]])
```

There are differences between the two methods. Since the
`medicalcoder::comorbidities()` flag is
always greater or equal to the `pccc::ccc()` flag it means that there are
technology dependence flags that `medicalcoder::comorbidities()` is flagging
conditions when `pccc::ccc()` is not.

```{r}
#| label: medicalcoder-vs-pccc-tech-dep-codes
mdcrDT[, code2 := paste(icdv, dx, code, sep = "_")]

medicalcoder::comorbidities(
  data = mdcrDT[patid %in% mdcr_v_pccc[any_tech_dep != tech_dep, patid]],
  id.vars = c("patid", "code2"),
  dx.var = "dx",
  icd.code = "code",
  icdv.var = "icdv",
  method = "pccc_v2.0"
)[any_tech_dep == 1, sort(unique(code2))]
```
There are four codes which do not match.  Each of these have been reported on
GitHub as issues with pull requests to fix the issues.

```{r}
#| label: medicalcoder-vs-pccc-tech-dep-codes-lookup
#| include: false
subset(
  medicalcoder::get_pccc_codes(),
  full_code %in% c("V45.85", "V56.2", "86.06", "Z94.2")
)

subset(
  medicalcoder::get_icd_codes(with.descriptions = TRUE),
  full_code %in% c("V45.85", "V56.2", "86.06", "Z94.2")
)
```

1. ICD-9-CM V45.85
   * Documented for version 2 as metabolic (devices) but is only listed under
     metabolic in the software.  Because of how `medicalcoder::comorbidities()`
     is implemented the subconditon (devices) had to be provided in the tables
     for `method = "pccc_v2.0"`.  This results in the (improved) mapping to
     technology dependence.  Perhaps more concerning is that the mapping for
     V45.85 in `pccc::ccc()` maps to a transplant, which is incorrect.  GitHub
     issue and pull request have been submitted to resolve this issue[^1].
2. ICD-9-CM V56.2
   * Not in the v2 document, but in the software for neuromusc only. This raises
     a problem that the subcondition is not defined.  To allow for the methods
     in medicalcoder to work, a non-missing subcondition is needed. Device and
     technology use seems appropriate[^2].
3. ICD-9-PCS 86.06
   * Same issue as ICD-9-CM V48.85[^3]
4. ICD-10-CM Z49.01
   * Same issue as ICD-9-CM V56.2[^4]

[^1]: https://github.com/CUD2V/pccc/issues/49
[^2]: https://github.com/CUD2V/pccc/issues/46
[^3]: https://github.com/CUD2V/pccc/issues/48
[^4]: https://github.com/CUD2V/pccc/issues/55

```{r}
#| label: remove-tech-columns-from-mdcr_v_pccc
mdcr_v_pccc[, any_tech_dep := NULL]
mdcr_v_pccc[, tech_dep := NULL]
```

For transplant we have a similar mismatch between
`medicalcoder::comorbidities()` and `pccc::ccc()`.

```{r}
#| label: medicalcoder-vs-pccc-transplant
identical(
  mdcr_v_pccc[["any_transplant"]],  # medicalcoder::comorbidities()
  mdcr_v_pccc[["transplant"]]       # pccc::ccc()
)

all(mdcr_v_pccc[["any_transplant"]] >= mdcr_v_pccc[["transplant"]])

medicalcoder::comorbidities(
  data = mdcrDT[patid %in% mdcr_v_pccc[any_transplant != transplant, patid]],
  id.vars = c("patid", "code2"),
  dx.var = "dx",
  icd.code = "code",
  icdv.var = "icdv",
  method = "pccc_v2.0"
)[any_transplant == 1, sort(unique(code2))]
```

* ICD-9-CM V42.0[^5]
  * Documented as renal (transplant); missing from the transplant set in the
    pccc_1.0.7/src/pccc.cpp

* ICD-10-CM Z94.x[^6]
  * Specifically Z94.1, Z94.2, Z94.4, Z94.81, Z94.82, Z94.83, Z94.84 All
    documented as subcondition transplant but are missing from the transplant
    set in the pccc_1.0.7/src/pccc.cpp

[^5]: https://github.com/CUD2V/pccc/issues/54
[^6]: https://github.com/CUD2V/pccc/issues/56

```{r}
#| label: remove-transplant-columns-from-mdcr_v_pccc
mdcr_v_pccc[, any_transplant := NULL]
mdcr_v_pccc[, transplant := NULL]
```

Additionally, `medicalcoder::comorbidities()` returns a two columns which are
not returned from `pccc::ccc()`. The
`misc` column flags
miscellaneous conditions.  These conditions are only technology dependence or
transplant subconditions and reported as such in `pccc::ccc()`.  The `num_cmrb`
column reports the "number of comorbidities" flagged for the patid.
```{r}
#| label: mdcr-v-pccc-misc-and-num_cmrb
mdcr_v_pccc
medicalcoder_pccc_v2.0[, .N, keyby = .(misc, any_tech_dep, any_transplant)]
summary(mdcr_v_pccc$num_cmrb)
```

# Nice-to-Haves in `r pkg_with_version("medicalcoder")`

As shown above, medicalcoder applies these common comorbidity algorithms
consistent with other implementations, or with increased precision and
considerably faster as well.  Additionally, medicalcoder has some nice-to-haves
for working with the results from calling `medicalcoder::comorbidities()`.

## summary() methods for `medicalcoder::comorbidities()` results

An S3 `summary()` method is available for the `medicalcoder_comorbidities` class
objects returned from `medicalcoder::comorbidities()`.

Below is the structure of the summary for the Charlson, Elixhauser, and PCCC
returns.  Each is slightly different based on the method.

```{r}
#| label: medicalcoder-summary
str( summary(medicalcoder_charlson_quan2005) )
str( summary(medicalcoder_elixhauser_quan2005) )
str( summary(medicalcoder_pccc_v2.0) )
```

We highlight the `conditions` element from `medicalcoder_charlson_quan2005`,
`medicalcoder_elixhauser_quan2005` and the summary from the
`medicalcoder_pccc_v2.0` object.
These data.frames can be the basis for useful summary tables which can be nicely
formatted via packages like kableExtra [@r-kableExtra].


```{r}
#| label: summary-table-charlson
#| echo: true
#| results: "asis"
x <- summary(medicalcoder_charlson_quan2005)[["conditions"]]
kableExtra::kbl(
  x = x[, c("condition_description", "count", "percent")],
  format = "latex",
  booktabs = TRUE
) |>
kableExtra::pack_rows("Conditions", 1, 17) |>
kableExtra::pack_rows("Total Number of Conditions", 18, nrow(x))
```

## ICD Tools

There are four functions in the medicalcoder package specifically for working
with International Classification of Diseases (ICD) codes.

1. `medicalcoder::get_icd_codes()`: returns a lookup table of ICD codes as a `data.frame`.
2. `medicalcoder::lookup_icd_codes()`: returns details on specific ICD codes.
3. `medicalcoder::is_icd()`: returns `TRUE` or `FALSE` for a vector of codes while considering
   ICD version, type, and billable status.
4. `medicalcoder::icd_compact_to_full()`: insert a decimal point into a string to be
   consistent with ICD-9 diagnostic, ICD-9 procedure, or ICD-10 diagnostic
   codes.  (ICD-10 procedure codes do not have decimal places.)  NOTE: this will
   not validate the return as a valid ICD code, just format the input string
   appropriately.

### `medicalcoder::get_icd_codes()`

A lookup table for the ICD codes has been built as internal data sets within the
_medicalcoder_ package.  The sources for these lookup tables come from the Centers
for Disease Control (CDC) and from the Centers for Medicare & Medicaid Services
(CMS) and World Health Organization (WHO).  The specific links to the source
data sets can be found in the source code for the medicalcoder package on
[GitHub](https://github.com/dewittpe/medicalcoder).

End users can get a `data.frame` with ICD-9 diagnostic, ICD-9 procedure, ICD-10
diagnostic, and ICD-10 procedure codes.

```{r}
#| label: get-icd-codes
icd_codes <- medicalcoder::get_icd_codes()
str(icd_codes)
```
The columns of this data.frame are:

* `icdv`: integer value 9 or 10 indicating ICD-9 or ICD-10
* `dx`: 1 if the code is a diagnostic code, i.e., from the ICD-9-CM or ICD-10-CM
  standard.  This also covers codes from the World Health Organization (WHO) and
  the Center for Disease Control and Prevention (CDC) Mortality codes.
  `dx` will be 0 if the code is a procedure code, i.e., from the ICD-9-PCS or
  ICD-10-PCS standard.
* `full_code`: the full ICD code with the decimal point if applicable.
* `code`: compact code, any applicable decimal point has been omitted.
* `src`: a character vector denoting the source of the information.
* `known_start`: The first year that the medicalcoder package has data for this
  code.
  * For codes based on the ICD-9-CM, ICD-9-PCS, ICD-10-CM, ICD-10-PCS standards
    the year is the _fiscal year_ for the United States Federal Government;
    October 1 - September 30.
    For example, fiscal year 2018 started October 1, 2017 and ended on September
    30, 2018.

  * For codes from the World Health Organization (WHO) and the Centers for
    Disease Control and Prevention (CDC) Mortality coding, the year is
    _calendar year_.
  * For ICD-9, CDC extracts in medicalcoder span fiscal years 1997--2012 and
    CMS extracts span fiscal years 2006--2015.

* `known_end`: the last year the code was part of the standard, or that the
  medicalcoder package has data for.
  * ICD-9 CMS extracts run through FY 2015; CDC ICD-9 extracts end at FY 2012.
  * ICD-10 is active.  The current version of medicalcoder has details on ICD-10
    codes through FY `r max(icd_codes[["known_end"]][icd_codes[["icdv"]] == 10L])`.

```{r}
#| label: example-for-assignable
#| include: false
d4 <- medicalcoder::lookup_icd_codes(x = "^C84\\.6", regex = TRUE, compact.codes = FALSE)
d4 <- subset(d4, src %in% c("cms", "who"), select = c("src", "full_code"))
d4 <- unique(d4)
d4
```

* `assignable_start`: The first year (fiscal or calendar based on src) a code
  was assignable. `NA` indicates the code was never assignable.  Assignable
  status can vary from source to source. For example the code "C84.6" is an
  assignable code using the WHO ICD-10 codes because there is no code with more
  granularity.  The same code is not assignable under ICD-10-CM because codes
  `r paste(d4$full_code[nchar(d4$full_code) > 5], collapse = ", ")`
  exist.  Codes that are not assignable are called header codes.  Ideally codes
  are reported with the greatest level of granularity, but that is not always the
  case.

* `assignable_end`: The last year the code was assignable.

#### with.descriptions

To get the descriptions of the ICD codes call `medicalcoder::get_icd_codes()` with
`with.descriptions = TRUE`.
```{r}
#| label: icd-codes-with-descriptions
str(medicalcoder::get_icd_codes(with.descriptions = TRUE))
```
The return has the additional columns:

* `desc`: the description of the code
* `desc_start`: the first fiscal_year that the description was documented
* `desc_end`: the last fiscal_year that the description was documented

Over time the descriptions for some ICD codes were modified within sources.
There are also many differences between sources.  The tables below have several
examples.

```{r}
#| label: deltas-in-desc
#| results: "hide"
delta_in_desc <-
  subset(
    medicalcoder::get_icd_codes(with.descriptions = TRUE),
    subset = full_code %in% c("Z88.7", "010.93", "V76.49"),
    select = c("full_code", "src", "desc", "desc_start", "desc_end")
  )
```

The only difference in the description for 010.93 is a comma.

```{r}
#| label: deltas-in-desc-show-010.93
#| echo: false
#| results: "asis"
knitr::kable(
  subset(delta_in_desc, subset = full_code == "010.93"),
  row.names = FALSE
)
```

ICD-10-CM Z88.7 has differences in the description over time within `cms` source
and between `cms` and `who`.

```{r}
#| label: deltas-in-desc-show-Z88.7
#| echo: false
#| results: "asis"
knitr::kable(
  subset(delta_in_desc, subset = full_code == "Z88.7"),
  row.names = FALSE
)
```

ICD-9-CM V79.49 had the description of 'other' which would require exploration
of the header codes to understand. Even the most verbose description may
still require consideration of the header codes to fully understand.

```{r}
#| label: deltas-in-desc-show-v79.49
#| echo: false
#| results: "asis"
knitr::kable(
  subset(delta_in_desc, subset = full_code == "V76.49"),
  row.names = FALSE
)
```

#### with.hierarchy

Lastly, the `medicalcoder::get_icd_codes()` function includes the argument
`with.hierarchy` which will provide additional details for the codes.

```{r}
#| label: get-icd-descs-with-hierarchy
str(medicalcoder::get_icd_codes(with.hierarchy = TRUE))
```
The additional columns, in order of hierarchy, are:

* chapter
* subchapter
* category
* subcategory
* subclassification
* subsubclassification
* extension

To keep the install size of medicalcoder under the size
limits for CRAN, the stored data is structured in a way that several joins and
other operations are needed to have a data set that is end user friendly.
Several data sets are generated and cached when the namespace is loaded.

### `medicalcoder::lookup_icd_codes()`

A related function, `medicalcoder::lookup_icd_codes()`, allows the user to look
up specific ICD codes. The return is a `data.frame`.  The columns report the
input code, if it was matched as a full code (with an applicable decimal point)
or a compact code (applicable decimal point omitted) along with the ICD version,
type, and when the code was assignable.

```{r}
#| label: lookup-icd-code-example
#| results: "asis"
codes <- c("0011", "7329", "732", "73291", "not a code", "001.1", "A9248", "A924", "Z00")
knitr::kable(
  medicalcoder::lookup_icd_codes(codes),
  row.names = FALSE
)
```

It is possible to restrict the lookup to just full or compact codes.  The
default, as shown above, is to consider full and compact codes.  Set `full.codes
= FALSE` so only compact codes are considered.
```{r}
#| label: lookup-compact-icd-codes
#| results: "asis"
knitr::kable(
  medicalcoder::lookup_icd_codes(codes, full.codes = FALSE),
  row.names = FALSE
)
```
And set `compact.codes = FALSE` to only consider full codes.
```{r}
#| label: lookup-full-icd-codes
#| results: "asis"
knitr::kable(
  medicalcoder::lookup_icd_codes(codes, compact.codes = FALSE),
  row.names = FALSE
)
```

By default, `medicalcoder::lookup_icd_codes()` considers the input to be a
string and a direct match to the internal lookup table is made.

`medicalcoder::lookup_icd_codes()` can also accept regular expressions.  By
providing a vector of regular expression patterns for the codes (passed to
`grep()`)

```{r}
#| label: lookup-icd-code-by-regex
#| results: "asis"
knitr::kable(
  medicalcoder::lookup_icd_codes(x = "^C84\\.6[0-1A-Z]", regex = TRUE),
  row.names = FALSE
)
```

### `medicalcoder::is_icd()`

By convention, ICD codes are generally reported without decimal points. Under
this convention discriminating between ICD-9 and ICD-10, and between diagnostic
and procedure codes can be difficult.

Is "7993" a valid code?  It is not a valid ICD-10 code as a four digit code
could not be an ICD-10 procedure code, and all ICD-10 diagnostic codes start
with a letter, not a number.  So this string could only be a ICD-9 code.  It is
a valid ICD-9 diagnostic code, and a valid ICD-9 procedure code.

```{r}
#| label: icd-7993
medicalcoder::is_icd(x = "7993")
medicalcoder::is_icd(x = "7993", icdv =  9, dx = 1)
medicalcoder::is_icd(x = "7993", icdv =  9, dx = 0)
medicalcoder::is_icd(x = "7993", icdv = 10, dx = 1)
medicalcoder::is_icd(x = "7993", icdv = 10, dx = 0)
medicalcoder::lookup_icd_codes("7993")
```

A vector of possible codes:
```{r}
#| label: icd-is-icd-vector
x <-
  c("7993", "A924", "7993", "A924", "no", "A92", "516", "5163", "51631", "A00")
medicalcoder::is_icd(x)
```

If you have codes with decimal points then discriminating between ICD-9
diagnostic and procedure codes can be done.

```{r}
#| label: icd-decimal-disambiguation
x <- c("7993",  # valid dx and pr code
       ".7993", # not a valid code
       "7.993", # not a valid code
       "79.93", # invalid dx code; valid pr code
       "799.3", # valid dx code; invalid pr code
       "7993.") # not a valid code
data.frame(
  x = x,
  icd9_dx = medicalcoder::is_icd(x, icdv = 9, dx = 1, warn.ambiguous = FALSE),
  icd9_pr = medicalcoder::is_icd(x, icdv = 9, dx = 0, warn.ambiguous = FALSE)
)
```

#### Assignable codes

Ideally, codes are reported with the greatest level of detail.  While there is
always a chance for incomplete coding, it is possible that an assignable code in
one year becomes a header code in a subsequent year.  Let's look at the
ICD-9 DX code 516.3 and five digit codes 516.30 through 516.39 (not all of these
are valid, as we'll see in the examples.)

Given the default settings, we have the following results for testing if these
strings are valid ICD-9 diagnostic codes.

By default, if no year is provided in the `medicalcoder::is_icd()` call then
return will be `TRUE` if the code was ever assignable.

```{r}
#| label: icd-assignable-table
#| results: "asis"
x <- paste0("516.3", c("", 0:9))
tab <-
  data.frame(
    code       = x,
    default    = medicalcoder::is_icd(x, icdv = 9, dx = 1),
    assignable_1997_cdc = medicalcoder::is_icd(x, src = "cdc", icdv = 9, dx = 1, year = 1997),
    assignable_2010_cms = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1, year = 2010),
    assignable_2011_cms = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1, year = 2011),
    assignable_2012_cdc = medicalcoder::is_icd(x, src = "cdc", icdv = 9, dx = 1, year = 2012),
    assignable_2012_cms = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1, year = 2012),
    assignable_2015_cms = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1, year = 2015),
    assignable_ever_cdc = medicalcoder::is_icd(x, src = "cdc", icdv = 9, dx = 1, ever.assignable = TRUE)
  )
knitr::kable(tab)
```

Similar information can be quickly and easily retrieved via
`medicalcoder::lookup_icd_codes()`.
```{r}
#| label: icd-lookup-assignable
#| results: "asis"
knitr::kable(medicalcoder::lookup_icd_codes(x))
```

For fiscal years `r medicalcoder::lookup_icd_codes("516.3")$assignable_start` through
`r medicalcoder::lookup_icd_codes("516.3")$assignable_end` the code 516.3 was assignable.  In
`r medicalcoder::lookup_icd_codes("516.30")$assignable_start` 516.3 was not assignable due to the
introduction of the five digit codes 516.30, 516.31, 516.32, 516.33, 516.34,
516.35, 516.36, and 516.37.  Codes 526.38 and 516.39 were never in the ICD-9-CM
standard.  When looking at retrospective data over several years the use of the
`ever.assignable` argument will simplify the testing for valid codes.

#### Header codes

There is also an option to considering header codes to be valid.  As seen below,
the code "516" is a header, it was never assignable in ICD-9-CM.
By setting `headerok = TRUE` "516" will be flagged as a valid code.  An ICD-10
header "A00" will be FALSE in the following checks of ICD-9 codes.

```{r}
#| label: icd-headerok-example
x <- c("516", "5163", "51631", "A00")
tab <-
  data.frame(
    code     = x,
    default  = medicalcoder::is_icd(x, icdv = 9, dx = 1, src = "cms", headerok = FALSE, ever.assignable = FALSE, warn.ambiguous = FALSE),
    ever     = medicalcoder::is_icd(x, icdv = 9, dx = 1, src = "cms", headerok = FALSE, ever.assignable = TRUE,  warn.ambiguous = FALSE),
    headerok = medicalcoder::is_icd(x, icdv = 9, dx = 1, src = "cms", headerok = TRUE,                           warn.ambiguous = FALSE)
  )
knitr::kable(tab)
```

A more complex situation is ICD-9-CM code 719.7 and the five digit codes
719.70, 719.75, 719.76, 719.77, 719.78, and 719.79.  The five digit codes were
assignable codes through FY 2004.  Starting in FY 2004 the five digit codes were
removed from the standard and the four digit code became assignable.  This is a
rare example of a header code becoming assignable.
```{r}
#| label: ICD-9-7197
x <- paste0("719.7", c("", "0", 5:9))
tab <-
  data.frame(
    code            = x,
    default         = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1),
    assignable_2002 = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1, year = 2002),
    assignable_2003 = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1, year = 2003),
    assignable_2004 = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1, year = 2004),
    assignable_2005 = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1, year = 2005),
    assignable_ever = medicalcoder::is_icd(x, src = "cms", icdv = 9, dx = 1, ever.assignable = TRUE)
  )
knitr::kable(tab)
```

### `medicalcoder::icd_compact_to_full()`

To go from a full code to a compact code is simple, omit any decimal point in
the string.

To go from a compact code to a full code requires knowing if the code is from
version 9 or 10, and if it is a diagnostic or a procedure code.
`medicalcoder::icd_compact_to_full()` will format a string appropriately, within
reason.  This method only formats the strings and will not validate the return.

For example, the compact code "E1234" is in the format expected for a ICD-9
diagnostic code or ICD-10 diagnostic code.  It could not be a procedure code as
ICD-9 procedure codes are all numeric values and ICD-10 procedure codes are
seven characters long.  The actual code E1234 is not a valid ICD code.  We use
this string as an example.

```{r}
#| label: icd-compact-to-full-dx
medicalcoder::icd_compact_to_full("E1234", icdv =  9, dx = 1)
medicalcoder::icd_compact_to_full("E1234", icdv = 10, dx = 1)

medicalcoder::lookup_icd_codes(c("E1234", "E123.4", "E12.34"))[, c("input_code", "match_type")]
```

Notice that no change to the string is made when trying to convert to a full
procedure code.

```{r}
#| label: icd-compact-to-full-pr
medicalcoder::icd_compact_to_full("E1234", icdv =  9, dx = 0)
medicalcoder::icd_compact_to_full("E1234", icdv = 10, dx = 0)
```

# Session Info

```{r}
#| label: sessioninfo
sessionInfo()
```

```{r}
#| label: same-image
#| include: false
# saving the image to use some of the results in the manuscript body
save.image(file = "medicalcoder-vs.Rdata")
```

# References
