---
title: "MIMIC-IV Data Analysis via medicalcoder"
subtitle: "Supplement to JAMIA Open Manuscript"

format:
  pdf:
    toc: true
    number-sections: true
    include-in-header:
      text: |
        \usepackage[table]{xcolor}

bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  cache = TRUE
)

loadNamespace('medicalcoder')
loadNamespace('data.table')
```

# Objective and Setup

In this supplement we will use the MIMIC-IV v3.1 Data
[@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000]
to illustrate how to flag comorbidities via `medicalcoder::comorbidities()`.

To reproduced this work end users will need to have access to the MIMIC-IV v3.1
data.  We have stored the path to the MIMIC-IV data as a system variable on our
machines.  If you are going to reproduce this work either set the same
environment variable or overwrite the `MIMICIVDATA` variable defined below.
Additionally, we will use some of the SQL code for applying the Charlson
comorbidity algorithm to the MIMIC-IV data provided by the MIT Laboratory for
Computational Physilogy (MIT\_LCP) on
[GitHub](https://github.com/MIT-LCP/mimic-code.git).  We stored the
path to a local copy of the respository in the system environmental variable
`MIMICIVCODE`.

```{r}
#| label: getenv
# System variable stores path to MIMIC-IV data
MIMICIVDATA <- Sys.getenv("MIMICIVDATA")
MIMICIVCODE <- Sys.getenv("MIMICIVCODE")
```

```{r}
#| label: verify mimiciv version
# verify that you are looking at version 3.1 of the MIMIC-IV data
stopifnot(
  identical(
    scan(
      file = file.path(MIMICIVDATA, "CHANGELOG.txt"),
      what = "character", sep = "\n", quiet = TRUE)[1],
    "This is the change log for MIMIC-IV v3.1."
  )
)
```

```{r}
#| label: display checksums
#| echo: false
#| results: "asis"
data.table::fread(
  file = file.path(MIMICIVDATA, "SHA256SUMS.txt"),
  header = FALSE
) |>
kableExtra::kbl(
  x = _,
  col.names = c("sha256sum", "file"),
  format = "latex",
  caption = "SHA256 Check sums for files in the MIMIC-IV v3.1 data.",
  booktabs = TRUE
  ) |>
kableExtra::kable_styling(
  latex_options = c("striped")
)
```

```{r}
#| label: verify-checksums
#| cache: true
system(
  sprintf("cd %s; sha256sum -c SHA256SUMS.txt", MIMICIVDATA),
  intern = TRUE
)
```

The SQL code for applying the Charlson comorbidities:
```{r}
mimiciv_charlson_sql <-
  file.path(MIMICIVCODE, "concepts", "comorbidity", "charlson.sql")

# checksum
stopifnot(
  substr(
    system(
      sprintf("sha256sum %s", mimiciv_charlson_sql),
      intern = TRUE
    ),
  start = 1L,
  stop = 64L
  ) == "5b797b673ace4dcbc6f686848d24371382f49ac61fcc39101f09c3b42969d801"
)
```

# Data Preperation

We will read in the MIMIC-IV hospital data and do some light formating to before
using `medicalcoder::comorbidities()` to flag comorbidities.  Start by importing
the `admissions`, `patients`, `diagnoses_icd`, and `procedures_icd`, datasets.

```{r}
#| label: import-mimiciv-data
mimicivdata <-
  list.files(path = file.path(MIMICIVDATA, "hosp"), full.names = TRUE)
names(mimicivdata) <- sub("\\.csv\\.gz$", "", basename(mimicivdata))

mimicivdata <-
  lapply(
    mimicivdata[c("admissions", "patients", "diagnoses_icd", "procedures_icd")],
    data.table::fread
  )
```

We will build an additional data set for age (in years).

```{r}
mimicivdata$ages <-
  merge(
    x = mimicivdata$admissions[, .(subject_id, hadm_id, admittime)],
    y = mimicivdata$patients[,   .(subject_id, anchor_age, anchor_year)],
    all = FALSE,
    by = "subject_id"
  )
mimicivdata$ages[, anchor_year := as.POSIXct(paste0(anchor_year, "-01-01"), format = "%Y-%m-%d")]
mimicivdata$ages[, age := as.numeric(difftime(admittime, anchor_year, units = "days")) / 365.25 + anchor_age]
```

For the calls to `medicalcoder::comorbidities()` we can have all the ICD codes
in one data.frame (data.table will be used here).
```{r}
mimicivdata$icd <-
  data.table::rbindlist(
    list(dx = mimicivdata$diagnoses_icd, pr = mimicivdata$procedures_icd),
    idcol = "dx",
    use.names = TRUE,
    fill = TRUE
  )

# set the diagnoses indicator to an integer as expected for input to
# medicalcoder::comorbidities()
mimicivdata$icd[, dx := as.integer(dx == "dx")]
```

We put all the parts together into one data.frame (data.table)

```{r}
#| label: one-data-set
mimicivDT <-
  merge(
    x =
      merge(
        x = mimicivdata$icd,
        y =
          merge(
            x = mimicivdata$patients[, .(subject_id)],
            y = mimicivdata$ages[, .(subject_id, hadm_id, age)],
            all = TRUE,
            by = c("subject_id")
          ),
        all = TRUE,
        by = c("subject_id", "hadm_id")
      ),
    y = mimicivdata$admissions[, .(subject_id, hadm_id, admittime)],
    all = TRUE,
    by = c("subject_id", "hadm_id")
  )

# set keys and order the data by subject id, admittime, and the hadm_id
data.table::setkey(mimicivDT, subject_id, admittime, hadm_id)

# set an encounter sequence variable. This works because the data has been
# sorted by subject_id, admittime, hadm_id
mimicivDT[, enc_seq := .GRP, by = .(subject_id, hadm_id)]
```

Data summary:

```{r}
#| label: data-summary
str(mimicivDT)
```

There are `r qwraps2::frmt(nrow(mimicivDT))` total rows of data for 
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique hospital encounters.

# Applying the MIMIC-IV Charlson SQL 

SQL code for applying a Charlson comorbidity algorithm to the MIMIC-IV data is
available from MIT Laboratory for Computational Physilogy (MIT\_LCP) on
GitHub[^a].  The code is expected to run on Google Big Query.  We make a few
small modifications to the code so we can evaluate the SQL locally via RSQLite.

[^a]: https://github.com/MIT-LCP/mimic-code.git

```{r}
#| label: mimic-code-edits
mimiciv_charlson_sql <-
  scan(
    file = mimiciv_charlson_sql,
    what = "character",
    sep = "\n",
    quiet = TRUE
  )

# modify the query to work in SQLite
mimiciv_charlson_sql <-
  gsub(pattern = "physionet-data.mimiciv_hosp.admissions",
       replacement = "admissions",
       x = mimiciv_charlson_sql,
       fixed = TRUE)
mimiciv_charlson_sql <-
  gsub(pattern = "physionet-data.mimiciv_hosp.diagnoses_icd",
       replacement = "diagnoses",
       x = mimiciv_charlson_sql,
       fixed = TRUE)
mimiciv_charlson_sql <-
  gsub(pattern = "physionet-data.mimiciv_derived.age",
       replacement = "ages",
       x = mimiciv_charlson_sql,
       fixed = TRUE)
mimiciv_charlson_sql <-
  gsub(pattern = "GREATEST",
       replacement = "MAX",
       x = mimiciv_charlson_sql,
       fixed = TRUE)

mimiciv_charlson_sql <- paste(mimiciv_charlson_sql, collapse = "\n")
```

We set up an SQL database in memory and evaluate the code.

```{r}
#| label: mimic-iv-charlson-results
#| echo: true
#| results: "hide"
#| cache: true
library(odbc)
library(DBI)
library(RSQLite)
con <- dbConnect(drv = RSQLite::SQLite(), dbname = ":memory:")

# add data to the data base
dbWriteTable(conn = con, name = "diagnoses",  value = mimicivdata$diagnoses)
dbWriteTable(conn = con, name = "admissions", value = mimicivdata$admissions)
dbWriteTable(conn = con, name = "patients",   value = mimicivdata$patients)
dbWriteTable(conn = con, name = "ages",       value = mimicivdata$ages)

# get the charlson results via MIMIC-IV
tic <- Sys.time()
mimiciv_charlson_results <- dbGetQuery(con, mimiciv_charlson_sql)
toc <- Sys.time()
attr(mimiciv_charlson_results, "tictoc") <-
  difftime(toc, tic, units = "secs")

# close DB conection
dbDisconnect(conn = con)

# simplify the column names for comparison with medicalcoder results
data.table::setDT(mimiciv_charlson_results)

nms <- c(
  "age_score"                   = "age_score_mimic",
  "myocardial_infarct"          = "mi_mimic",
  "congestive_heart_failure"    = "chf_mimic",
  "peripheral_vascular_disease" = "pvd_mimic",
  "cerebrovascular_disease"     = "cvd_mimic",
  "dementia"                    = "dem_mimic",
  "chronic_pulmonary_disease"   = "copd_mimic",
  "rheumatic_disease"           = "rd_mimic",
  "peptic_ulcer_disease"        = "pud_mimic",
  "mild_liver_disease"          = "mld_mimic",
  "diabetes_without_cc"         = "dm_mimic",
  "diabetes_with_cc"            = "dmc_mimic",
  "paraplegia"                  = "hp_mimic",
  "renal_disease"               = "rnd_mimic",
  "malignant_cancer"            = "ca_mimic",
  "severe_liver_disease"        = "sld_mimic",
  "metastatic_solid_tumor"      = "mst_mimic",
  "aids"                        = "aids_mimic"
)

data.table::setnames(
    mimiciv_charlson_results,
    old = names(nms),
    new = nms
)
```

# Charlson Comorbidities via `medicalcoder::comorbidities()`

```{r}
#| label: medicalcoder-warmup
#| include: false
# to give the best possible time for medicalcoder, it will be helpful to run the
# code once just to make sure all the needed internal data sets have been built
medicalcoder::comorbidities(
  data = mimicivDT[1:10],
  id.vars = c("subject_id", "hadm_id"),
  icd.codes = "icd_code",
  icdv.var = "icd_version",
  dx.var = "dx",
  age.var = "age",
  method = "charlson_quan2005",
  full.codes = FALSE,
  flag.method = "current",
  primarydx = 0L,
  poa = 1L
)
```

```{r}
#| label: medicalcoder-charlson-results
tic <- Sys.time()
medicalcoder_charlson_results <-
  medicalcoder::comorbidities(
    data = mimicivDT,
    id.vars = c("subject_id", "hadm_id"),
    icd.codes = "icd_code",
    icdv.var = "icd_version",
    dx.var = "dx",
    age.var = "age",
    method = "charlson_quan2005",
    full.codes = FALSE,
    flag.method = "current",
    primarydx = 0L,
    poa = 1L
  )
toc <- Sys.time()

attr(medicalcoder_charlson_results, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

# Charlson Comorbidities via `comorbidity::comorbidity()`

The R package _comorbidity_[@r-comorbidity] is a commonly referenced package
used for flagging Charlson and Elixhauser comorbidities.

```{r}
#| label: comorbidity-charlson
#| cache: true

```

# MIMIC-IV Code Results vs `medicalcoder::comorbidities()` Results

The processing the SQL took considerably longer than
`medicalcoder::comorbidities()`

```{r}
#| label: medicalcoder-vs-mimiciv-time-delta
attr(mimiciv_charlson_results, "tictoc") # seconds
attr(medicalcoder_charlson_results, "tictoc") # seconds
```




# Session Info

```{r}
#| label: sessionInfo
#| echo: true
#| results: "markup"
sessionInfo()
```

# References
