---
title: "MIMIC-IV Data Analysis via medicalcoder"
subtitle: "Supplement to JAMIA Open Manuscript"

format:
  pdf:
    toc: true
    toc-depth: 4
    lof: true
    lot: true
    number-sections: true
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage[table]{xcolor}
        \usepackage{threeparttable}
        \usepackage{longtable}

bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  cache = TRUE
)

options(knitr.kable.NA = "")

# Go Avs Go!
# {
#   "burgundy" : "#6F263D",
#   "blue" : "#236192",
#   "light-blue" : "#4F8FCB"
#   "silver" : "#A2AAAD",
#   "black" : "#000000"
# }

loadNamespace('medicalcoder')
loadNamespace('data.table')
```

# Objective and Setup

This supplement provides reproducible validation and benchmarking of
`medicalcoder::comorbidities()` using MIMIC-IV v3.1 data, including algorithmic
agreement with established implementations and demonstration of longitudinal
flagging behavior.

# Materials and Methods

## Data

We use the MIMIC-IV v3.1 Data
[@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000] as the foundation for
these examples.

The MIMIC-IV v3.1 dataset is a large, publicly available clinical database
containing de-identified health data from patients admitted to intensive care
units at Beth Israel Deaconess Medical Center. It includes detailed information
on diagnoses, procedures, laboratory results, medications, and hospital
encounters, with ICD-9-CM and ICD-10-CM/PCS codes available for billing and
comorbidity research. MIMIC-IV is widely used for method development,
validation, and reproducible research in clinical informatics and outcomes
research.

The MIMIC-IV v3.1 data set can be obtained so long as you meet the requirements
set by PhysioNet[^1].

[^1]: https://physionet.org/content/mimiciv/3.1/

## Comorbidity Algorithms

Within the _medicalcoder_ package are implementations for several variants of
the Charlson, Elixhauser, and PCCC comorbidity algorithms,
(@tbl-comorbidity-methods) [@beyrer2021validation;@quan2005;@quan2011;@glasheen2019;@ahrq2025;@feudtner2014pediatric;@feinstein2024pediatric;@dewitt2026pccc;@feinstein2018pccc;@pcccv3sas].

__NOTE:__
The MIMIC-IV data consists of only adult patients.
Although PCCC was designed for pediatric populations, adult MIMIC-IV data are
used here solely to demonstrate algorithmic implementation.

```{r}
#| label: tbl-comorbidity-methods
#| tbl-cap: "Comorbidity methods implemented in _medicalcoder_ and sorted by comorbidity family and method in alphabetical order."
#| echo: false
cm <- data.table::fread(text = "
Comorbidity Method        | Notes
charlson_beyrer2021       | U.S. extension of Quan et al. (2005) by Beyrer et al. (2021)
charlson_cdmf2019         | ICD-9 and ICD-10 defined in [@glasheen2019]
charlson_deyo1992         | ICD-9 codes defined in Table 1 of Quan et al. (2005)
charlson_mimicivcode      | MIMIC-IV Charlson SQL from [`mimic-code`](https://github.com/MIT-LCP/mimic-code)
charlson_quan2005         | ICD-9 and ICD-10 defined in Table 1 of Quan et al. (2005); index scoring as reported in Table 2 of Quan et al. (2011)
charlson_quan2011         | ICD-9 and ICD-10 defined in Table 1 of Quan et al. (2005); index scoring as reported in Table 2 of Quan et al. (2011)
elixhauser_elixhauser1988 | ICD-9 codes defined in Table 2 of Quan et al. (2005)
elixhauser_ahrq_web       | ICD-9 codes defined in Table 2 of Quan et al. (2005)
elixhauser_quan2005       | ICD-9 and ICD-10 defined in Table 2 of Quan et al. (2005)
elixhauser_ahrq2022       | ICD-10 codes from AHRQ for fiscal year 2022
elixhauser_ahrq2023       | ICD-10 codes from AHRQ for fiscal year 2023
elixhauser_ahrq2024       | ICD-10 codes from AHRQ for fiscal year 2024
elixhauser_ahrq2025       | ICD-10 codes from AHRQ for fiscal year 2025
elixhauser_ahrq2026       | ICD-10 codes from AHRQ for fiscal year 2026
elixhauser_ahrq_icd10     | Any ICD-10 code from all elixhauser_ahrqYYYY
pccc_v2.0                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with pccc::ccc()
pccc_v2.1                 | Extended set of pccc_v2.0 codes based on pccc::ccc() and supplements to Feudtner et al. (2014)
pccc_v3.0                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with SAS code from Children's Hospital Assoication
pccc_v3.1                 | Extended set of pccc_v3.0 codes based on the supplements to Feinstein et al. (2024)"
)
cm <- cm[, lapply(.SD, gsub, pattern = "_", replacement = "\\\\_")]

kableExtra::kbl(
  x = cm,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down", "HOLD_position")
) |>
kableExtra::pack_rows(group_label = "Charlson",   start_row =  1, end_row =  4) |>
kableExtra::pack_rows(group_label = "Elixhauser", start_row =  5, end_row = 13) |>
kableExtra::pack_rows(group_label = "PCCC",       start_row = 14, end_row = 17)
```

## Alternative Utilities

We compared the results from `medicalcoder::comorbidities()` to the Charlson
comorbidities flagged by the MIMIC-IV code, and the R package _comorbidity_
[@r-comorbidity].  For Elixhauser we will compare results between
`medicalcoder::comorbidities()` and `comorbidity::comorbidity()`.  Lastly, we
will compare PCCC version 2 results between `medicalcoder::comorbidities()` and
the R package _pccc_ [@dewitt2026pccc] call `pccc::ccc()`.

# Prepare MIMIC-IV Data For Analysis

We have stored the path to the MIMIC-IV data as a system variable on our
machines as `MIMICIVDATA`.  The following code is provide to help readers verify
that they are using the same version of the MIMIC-IV v3.1 data as used in this
document.

```{r}
#| label: verify mimiciv version
# verify that you are looking at version 3.1 of the MIMIC-IV data
stopifnot(
  identical(
    scan(
      file = file.path(Sys.getenv("MIMICIVDATA"), "CHANGELOG.txt"),
      what = "character", sep = "\n", quiet = TRUE)[1],
    "This is the change log for MIMIC-IV v3.1."
  )
)
```

We only need four of the MIMIC-IV hospital datasets:
`admissions`, `patients`, `diagnoses_icd`, and `procedures_icd`.

We will do some light formatting before using `medicalcoder::comorbidities()` to
flag comorbidities.  Start by importing the `admissions`, `patients`,
`diagnoses_icd`, and `procedures_icd`, datasets.

@tbl-mimicivdata-checksums gives the sha256 checksums for these files
Each file checksum has been verified.

```{r}
#| label: tbl-mimicivdata-checksums
#| tbl-cap: "SHA256 Check sums for files in the MIMIC-IV v3.1 data."
#| echo: false
#| results: "asis"
data.table::fread(
  file = file.path(Sys.getenv("MIMICIVDATA"), "SHA256SUMS.txt"),
  header = FALSE
)[grepl("^hosp.(admissions|patients|.+_icd)\\.csv\\.gz$", V2)] |>
kableExtra::kbl(
  x = _,
  col.names = c("sha256sum", "file"),
  format = "latex",
  booktabs = TRUE
  ) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down", "HOLD_position")
)
```

```{r}
#| label: sha256-verify
#| include: false
verify_sha256sums <- function(dir, sums_file = "SHA256SUMS.txt") {
  old <- setwd(dir)
  on.exit(setwd(old), add = TRUE)

  sys <- Sys.info()[["sysname"]]

  if (identical(sys, "Windows")) {
    # Use CertUtil (built-in)
    lines <- readLines(sums_file, warn = FALSE)
    ok <- logical(length(lines))
    for (i in seq_along(lines)) {
      line <- trimws(lines[[i]])
      if (!nzchar(line)) { ok[[i]] <- TRUE; next }
      # Expected format: "<hash>  <filename>" or "<hash> *<filename>"
      parts <- strsplit(line, "\\s+")[[1]]
      hash <- parts[[1]]
      file <- sub("^\\*", "", parts[[2]])
      out <- suppressWarnings(system2("certutil", c("-hashfile", shQuote(file), "SHA256"), stdout = TRUE))
      got <- tolower(gsub("\\s+", "", out[grepl("^[0-9A-Fa-f]+$", out)]))
      ok[[i]] <- length(got) == 1 && tolower(hash) == got
    }
    return(all(ok))
  }

  # macOS often has `shasum -a 256`, Linux typically has `sha256sum`
  if (nzchar(Sys.which("sha256sum"))) {
    out <- system2("sha256sum", c("-c", sums_file), stdout = TRUE, stderr = TRUE)
    return(all(grepl(": OK$", out)))
  }

  if (nzchar(Sys.which("shasum"))) {
    out <- system2("shasum", c("-a", "256", "-c", sums_file), stdout = TRUE, stderr = TRUE)
    return(all(grepl(": OK$", out)))
  }

  stop("No SHA-256 tool found (need sha256sum, shasum, or certutil).")
}
stopifnot(verify_sha256sums(Sys.getenv("MIMICIVDATA")))
```

```{r}
#| label: import-mimiciv-data
mimicivdata <-
  list.files(
    path = file.path(Sys.getenv("MIMICIVDATA"), "hosp"),
    pattern = "(admissions|patients|.+_icd)\\.csv\\.gz",
    full.names = TRUE
  )
names(mimicivdata) <- sub("\\.csv\\.gz$", "", basename(mimicivdata))

mimicivdata <- lapply(mimicivdata, data.table::fread)
```

We will build an additional data set for age (in years).

```{r}
#| label: build-mimicvidata-ages
mimicivdata$ages <-
  merge(
    x = mimicivdata$admissions[, .(subject_id, hadm_id, admittime)],
    y = mimicivdata$patients[,   .(subject_id, anchor_age, anchor_year)],
    all = FALSE,
    by = "subject_id"
  )
mimicivdata$ages[
  ,
  anchor_year := as.POSIXct(paste0(anchor_year, "-01-01"), format = "%Y-%m-%d")
]
mimicivdata$ages[
  ,
  age :=
    anchor_age +
    as.numeric(difftime(admittime, anchor_year, units = "days")) / 365.25
]
```

For the calls to `medicalcoder::comorbidities()` we can have all the ICD codes
in one `data.frame` (data.table [@r-data.table] will be used here).
```{r}
#| label: build-mimicivdata-icd
mimicivdata$icd <-
  data.table::rbindlist(
    list(dx = mimicivdata$diagnoses_icd, pr = mimicivdata$procedures_icd),
    idcol = "dx",
    use.names = TRUE,
    fill = TRUE
  )

# set the diagnoses indicator to an integer as expected for input to
# medicalcoder::comorbidities()
mimicivdata$icd[, dx := as.integer(dx == "dx")]
```

We put all the parts together into one data.frame (data.table)

```{r}
#| label: one-data-set
mimicivDT <-
  merge(
    x = mimicivdata$patients[, .(subject_id)],
    y = mimicivdata$ages[, .(subject_id, hadm_id, age)],
    all = TRUE,
    by = c("subject_id")
  )

mimicivDT <-
  merge(
    x = mimicivdata$icd,
    y = mimicivDT,
    all = TRUE,
    by = c("subject_id", "hadm_id")
  )

mimicivDT <-
  merge(
    x = mimicivDT,
    y = mimicivdata$admissions[, .(subject_id, hadm_id, admittime)],
    all = TRUE,
    by = c("subject_id", "hadm_id")
  )
```
We need an encounter sequence variable, the `hadm_id` are not sequential.

NOTE: `admittime` can be used in the calls to
`medicalcoder::comorbidities()` but it there is additional overhead and
computational expense when sorting and joining by POSIXct variable.  The
`enc_seq` achieves the same utility with lower computational overhead.

```{r}
#| label: build-enc-seq
data.table::setorder(mimicivDT, subject_id, admittime)
mimicivDT[, enc_seq := cumsum(!duplicated(hadm_id)), by = .(subject_id)]
data.table::setkey(mimicivDT, subject_id, enc_seq, hadm_id)

# Sanity checks:
# expect age to always be increasing as enc_seq increase
mimicivDT[
  ,
  unique(.SD),
  .SDcols = c("subject_id", "hadm_id", "admittime", "age", "enc_seq")
  ][
  ,
  c(NA_real_, diff(age)),
  by = .(subject_id)
  ][
  ,
  all(V1 >= 0, na.rm = TRUE)
  ] |>
print() |>
stopifnot()
```

_medicalcoder_ can use multiple id variables.  Other utilities such as
_comorbidity_ and _pccc_ can only use one id variable.  As such we create `cid`
to use in those examples.
```{r}
#| label: build-cid
mimicivDT[, cid := paste(subject_id, hadm_id, sep = "__")]
```
Our final `mimicivDT` object:
```{r}
#| label: mimicivDT-data-structure
str(mimicivDT)
summary(mimicivDT)
```
There are `r qwraps2::frmt(nrow(mimicivDT))` total rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
hospital encounters.

# Investigating ICD Codes

The _medicalcoder_ package provides utilities for investigating ICD codes.
For the MIMIC-IV data, let's just check if the reported ICD codes are in the
_medicalcoder_ ICD data.

```{r}
#| label: mimicivdata-icd-codes
unique_icdcodes <- mimicivDT[, unique(.SD), .SDcols = c("icd_code", "icd_version", "dx")]

unique_icdcodes[!is.na(icd_code) & icd_version ==  9 & dx == 0, iscms := medicalcoder::is_icd(x = icd_code, icdv =  9, dx = 0, src = "cms")]
unique_icdcodes[!is.na(icd_code) & icd_version ==  9 & dx == 1, iscms := medicalcoder::is_icd(x = icd_code, icdv =  9, dx = 1, src = "cms")]
unique_icdcodes[!is.na(icd_code) & icd_version == 10 & dx == 0, iscms := medicalcoder::is_icd(x = icd_code, icdv = 10, dx = 0, src = "cms")]
unique_icdcodes[!is.na(icd_code) & icd_version == 10 & dx == 1, iscms := medicalcoder::is_icd(x = icd_code, icdv = 10, dx = 1, src = "cms")]

unique_icdcodes[!is.na(icd_code) & icd_version ==  9 & dx == 0, iscmshdrok := medicalcoder::is_icd(x = icd_code, icdv =  9, dx = 0, src = "cms", headerok = TRUE)]
unique_icdcodes[!is.na(icd_code) & icd_version ==  9 & dx == 1, iscmshdrok := medicalcoder::is_icd(x = icd_code, icdv =  9, dx = 1, src = "cms", headerok = TRUE)]
unique_icdcodes[!is.na(icd_code) & icd_version == 10 & dx == 0, iscmshdrok := medicalcoder::is_icd(x = icd_code, icdv = 10, dx = 0, src = "cms", headerok = TRUE)]
unique_icdcodes[!is.na(icd_code) & icd_version == 10 & dx == 1, iscmshdrok := medicalcoder::is_icd(x = icd_code, icdv = 10, dx = 1, src = "cms", headerok = TRUE)]

unique_icdcodes[, .N, keyby = .(iscms, iscmshdrok)]

subset(medicalcoder::get_icd_codes(with.description = TRUE), startsWith(code, "S73109"))
subset(medicalcoder::get_icd_codes(with.description = TRUE), code == "Z31")
```

# Flagging Encounter Level Comorbidities

In the following we will outline how to apply the comorbidities methods to the
`mimicivDT` dataset, focusing on encounter level flagging of comorbidities.  We
start by showing how to apply the
Charlson [@quan2005],
Elixhauser [@quan2005], and
PCCC v2 [@feudtner2014pediatric] methods to the `mimicivDT` dataset using the
`medicalcoder::comorbidities()` call.  Following that we will show how to do the
same work using SQL code provided by MIMIC-code [@johnson2018mimic], the R
package _comorbidity_ [@r-comorbidity], and the R package _pccc_ [@r-pccc].
Lastly, we compare and contrast the differences in the flagged comorbidities
between the methods.

## `medicalcoder::comorbidities()`

### Charlson Comorbidities
There are several variants of Charlson like comorbidity methods implemented in
_medicalcoder_.  One way to see the current list of implemented methods is to
look at the names of the lookup table mapping ICD codes to conditions:
```{r}
#| label: charlson-variants
grep(
  pattern = "^charlson",
  x = names(medicalcoder::get_charlson_codes()),
  value = TRUE
)
```
The `charlson_quan2011` method will be applied to the `mimicivDT` dataset and
the results will be compared to the results from the
`comorbidity::comorbidity()` method.
```{r}
#| label: medicalcoder-charlson-quan2011
medicalcoder_charlson_quan2011 <-
  medicalcoder::comorbidities(
    data        = mimicivDT,  # object inheriting data.frame
    icd.codes   = "icd_code", # character string name of icd codes in data
    id.vars     = c("subject_id", "hadm_id"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L, # consider all codes present on admission
    primarydx   = 0L, # consider all diagnosis codes secondary diagnoses.
    method      = "charlson_quan2011",
    flag.method = "current" # default
  )
data.table::setkey(medicalcoder_charlson_quan2011, subject_id, hadm_id)
```
The SQL code for applying Charlson comorbidities to the MIMIC-IV data provided
by [MIT-LCP](https://github.com/MIT-LCP/mimic-code) has a slightly different set
of ICD-10 codes mapping to conditions.  Quan et al. (2005, 2011) is based on
ICD-10 codes, not ICD-10-CM.  The MIMIC-IV SQL code includes the C4A category
which is a ICD-10-CM specific category (see [this github
issue](https://github.com/MIT-LCP/mimic-code/issues/2017)).  To be consistent
with the expected set of ICD codes identified in the Quan papers and the
MIMIC-IV SQL code, _medicalcoder_ provides distinct methods.
```{r}
#| label: medicalcoder-charlson-mimiciv
medicalcoder_charlson_mimiciv <-
  medicalcoder::comorbidities(
    data        = mimicivDT,  # object inheriting data.frame
    icd.codes   = "icd_code", # character string name of icd codes in data
    id.vars     = c("subject_id", "hadm_id"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    age.var     = "age",
    poa         = 1L,    # consider all codes present on admission
    primarydx   = 0L,    # consider all diagnosis codes secondary diagnoses.
    method      = "charlson_mimicivcode",
    flag.method = "current" # default
  )
data.table::setkey(medicalcoder_charlson_mimiciv, subject_id, hadm_id)
```

### Elixhauser Comorbidities
As with the Charlson comorbidities there are multiple variants of Elixhauser
like comorbidity methods implemented in _medicalcoder_.
```{r}
#| label: elixhauser-variants
grep(
  pattern = "^elixhauser",
  x = names(medicalcoder::get_elixhauser_codes()),
  value = TRUE
)
```
To apply the `elixhauser_quan2005` version, which is comparable to the R package
_comorbidity_, do the following:
```{r}
#| label: medicalcoder-elixhauser-quan2005
medicalcoder_elixhauser_quan2005 <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "hadm_id"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    primarydx   = 0L,
    method      = "elixhauser_quan2005"
  )
data.table::setkey(medicalcoder_elixhauser_quan2005, subject_id, hadm_id)
```

### PCCC v2.0
PCCC v2.0 has been implemented to be consistent with the R package _pccc_.
The v2.1 variants extends the ICD to conditions mapping while using the same
underlying algorithm. v3.0 and v3.1 use a different set of ICD codes and
algorithm for flagging conditions.  For this work we focus on v2.0 so we can
compare it to the results from the R package _pccc_.

```{r}
#| label: medicalcoder-pccc-v2.0
medicalcoder_pccc_v2.0 <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "hadm_id"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    method      = "pccc_v2.0"
  )
data.table::setkey(medicalcoder_pccc_v2.0, subject_id, hadm_id)
```

## MIMIC-IV Code

We use the SQL code for applying the Charlson
comorbidity algorithm to the MIMIC-IV data provided by the MIT Laboratory for
Computational Physiology (MIT\_LCP) on
[GitHub](https://github.com/MIT-LCP/mimic-code.git).  We stored the
path to a local copy of the repository in the system environmental variable
`MIMICIVCODE`.  The specific version of the SQL code being used is from commit
[278df75ec30991ff3a6f5ceb6d2221635a085e9f](https://github.com/MIT-LCP/mimic-code/blob/278df75ec30991ff3a6f5ceb6d2221635a085e9f/mimic-iv/concepts/comorbidity/charlson.sql)

```{r}
#| label: checksum-mimiciv-code
mimiciv_charlson_sql <-
  file.path(Sys.getenv("MIMICIVCODE"), "concepts", "comorbidity", "charlson.sql")

# checksum
stopifnot(
  substr(
    system(
      sprintf("sha256sum %s", mimiciv_charlson_sql),
      intern = TRUE
    ),
  start = 1L,
  stop = 64L
  ) == "5b797b673ace4dcbc6f686848d24371382f49ac61fcc39101f09c3b42969d801"
)
```

The SQL code is expected to run on Google Big Query.  We make a few
small modifications to the code so we can evaluate the SQL locally via RSQLite.

```{r}
#| label: mimic-code-edits
# scan in the sql code
mimiciv_charlson_sql <-
  scan(
    file = mimiciv_charlson_sql,
    what = "character",
    sep = "\n",
    quiet = TRUE
  )

# modify the query to work in SQLite
# replace Google Big Query table names with table names to be using in the local
# RSQLite in memory database.
mimiciv_charlson_sql <-
  gsub(
    pattern = "physionet-data.mimiciv_hosp.admissions",
    replacement = "admissions",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <-
  gsub(
    pattern = "physionet-data.mimiciv_hosp.diagnoses_icd",
    replacement = "diagnoses",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <-
  gsub(
    pattern = "physionet-data.mimiciv_derived.age",
    replacement = "ages",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

# Replace the Google Big Query SQL function GREATEST with MAX
mimiciv_charlson_sql <-
  gsub(
    pattern = "GREATEST",
    replacement = "MAX",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <- paste(mimiciv_charlson_sql, collapse = "\n")
```

We set up an SQL database in memory and evaluate the code.

```{r}
#| label: mimiciv-charlson-results
#| echo: true
#| results: "hide"
con <- odbc::dbConnect(drv = RSQLite::SQLite(), dbname = ":memory:")

# add data to the data base
odbc::dbWriteTable(conn = con, name = "diagnoses",  value = mimicivdata$diagnoses)
odbc::dbWriteTable(conn = con, name = "admissions", value = mimicivdata$admissions)
odbc::dbWriteTable(conn = con, name = "patients",   value = mimicivdata$patients)
odbc::dbWriteTable(conn = con, name = "ages",       value = mimicivdata$ages)

# Run the query and record the time required to do so
mimiciv_charlson <- odbc::dbGetQuery(con, mimiciv_charlson_sql)
# close DB connection
odbc::dbDisconnect(conn = con)

data.table::setDT(mimiciv_charlson)
data.table::setkey(mimiciv_charlson, subject_id, hadm_id)
```

## `comorbidity::comorbidity()`

The R package _comorbidity_ [@r-comorbidity] is a widely used and referenced
tool for applying the Charlson and Elixhauser comorbidities.
One consideration for using `comorbidity::comorbidity()` is that ICD-9 or ICD-10
codes are used independently. There are
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id)][V1 > 1, .N]`
subjects with both ICD-9 and ICD-10 codes in their records, and
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id, hadm_id)][V1 > 1, .N]`
encounters with both ICD versions.

To flag the conditions appropriately we must split the data by ICD version,
apply `comorbidity::comorbidity()` to each split, and then aggregate the
results.  Also, `comorbidity::comorbidity()` only takes one column for the id
and thus we need to use `cid` for the id instead of the `c("subject_id",
"hadm_id")` used in the calls to `medicalcoder::comorbidities()`.  In the
following work we will copy the data built for the encounter level flagging and
extend for cumulative flagging as well.

The following is done using _comorbidity_ version
`r packageVersion("comorbidity")`.

### Charlson

```{r}
#| label: comorbidity-charlson
comorbidity_charlson_icd9 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 9 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "charlson_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe
                   # comorbidities is also flagged
  )

comorbidity_charlson_icd10 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 10 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "charlson_icd10_quan",
    assign0 = TRUE
  )

comorbidity_charlson <-
  rbind(
    comorbidity_charlson_icd9,
    comorbidity_charlson_icd10
  )

# aggregate between the icd versions
comorbidity_charlson <-
  aggregate(. ~ cid, data = comorbidity_charlson, FUN = max)

# apply the scoring, to use comorbidity::score the object needs to have certain
# attributes which were lost when aggregating over the icd versions.  Replace
# the attributes
attributes(comorbidity_charlson)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_charlson_icd9)[c("class", "variable.labels", "map")]

comorbidity_charlson[["score"]] <-
  comorbidity::score(
    x = comorbidity_charlson,
    weights = "quan",
    assign0 = TRUE
  )

# expand the id columns
data.table::setDT(comorbidity_charlson)
comorbidity_charlson[
  ,
  c("subject_id", "hadm_id") :=
    lapply(data.table::tstrsplit(cid, "__"), as.integer)
  ]
data.table::set(comorbidity_charlson, j = "cid", value = NULL)
data.table::setkey(comorbidity_charlson, subject_id, hadm_id)
```

### Elixhauser

Compared to the work done for flagging Charlson comorbidities, Elixhauser
flagging requires changing the `map` argument in the calls to
`comorbidity::comorbidity()` and there is no scoring to be done.

```{r}
#| label: comorbidity-elixhauser
comorbidity_elixhauser_icd9 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 9 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "elixhauser_icd9_quan",
    assign0 = TRUE
  )

comorbidity_elixhauser_icd10 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 10 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "elixhauser_icd10_quan",
    assign0 = TRUE
  )

comorbidity_elixhauser <-
  rbind(
    comorbidity_elixhauser_icd9,
    comorbidity_elixhauser_icd10
  )

# aggregate between the icd versions
comorbidity_elixhauser <-
  aggregate(. ~ cid, data = comorbidity_elixhauser, FUN = max)

# expand the id columns
data.table::setDT(comorbidity_elixhauser)
comorbidity_elixhauser[
  ,
  c("subject_id", "hadm_id") :=
    lapply(data.table::tstrsplit(cid, "__"), as.integer)
  ]
data.table::set(comorbidity_elixhauser, j = "cid", value = NULL)
data.table::setkey(comorbidity_elixhauser, subject_id, hadm_id)
```

## `pccc::ccc()`

The R package _pccc_ [@dewitt2026pccc] implements PCCC v2
[@feudtner2014pediatric].  As with `comorbidity::comorbidity()` and many other
implementations of comorbidities, only one version of ICD is assessed at a time.

The expected input structure of the data for `pccc::ccc()` is a wide format, one
row per encounter, one column for each ICD code.
The following uses _pccc_ version `r packageVersion("pccc")`.

```{r}
#| label: pccc-pcccv2.0
mimicivDT_for_pccc <-
  split(mimicivDT, by = "icd_version") |>
  lapply(
    data.table::dcast,
    formula = cid ~ ifelse(dx == 1, "DX", "PR") + seq_num,
    value.var = "icd_code"
  )

pccc_pcccv2.0 <-
  rbind(
    pccc::ccc(
      data = mimicivDT_for_pccc[["9"]],
      id = cid,
      dx_cols = grep("^DX", names(mimicivDT_for_pccc[["9"]]), value = TRUE),
      pc_cols = grep("^PR", names(mimicivDT_for_pccc[["9"]]), value = TRUE),
      icdv = 9
    )
    ,
    pccc::ccc(
      data = mimicivDT_for_pccc[["10"]],
      id = cid,
      dx_cols = grep("^DX", names(mimicivDT_for_pccc[["10"]]), value = TRUE),
      pc_cols = grep("^PR", names(mimicivDT_for_pccc[["10"]]), value = TRUE),
      icdv = 10
    )
  )

# aggregate over ICD version
pccc_pcccv2.0 <- aggregate(. ~ cid, data = pccc_pcccv2.0, FUN = max)

# expand the id columns
data.table::setDT(pccc_pcccv2.0)
pccc_pcccv2.0[
  ,
  c("subject_id", "hadm_id") :=
    lapply(data.table::tstrsplit(cid, "__"), as.integer)
  ]
data.table::set(pccc_pcccv2.0, j = "cid", value = NULL)
data.table::setkey(pccc_pcccv2.0, subject_id, hadm_id)
```

# Comparing Encounter-level Flags Between Utilities

## Charlson Comorbidities

### `medicalcoder::comorbidities()` vs `comorbidity::comorbidity()`

The first thing to not is that the number of row returned by the two
implementations are different.
```{r}
#| label: nrows-medicalcoder-vs-comorbidity
nrow(medicalcoder_charlson_quan2011)
nrow(comorbidity_charlson)
```
The reason for this is that there are encounters with no diagnostic ICD codes.
`medicalcoder::comorbidities()` returns a row for these encounters and
`comorbidity::comorbidity()` does not.  For these encounters,
`medicalcoder::comorbidities()` returns a row with zero comorbidities.

```{r}
#| label: no-icd-no-comorbibity-result
# subject/hadm_id not in comorbibity::comorbibity() results all have no
# diagnostic codes.
mimicivDT[!comorbidity_charlson][, any(dx == 1 & !is.na(icd_code))]
medicalcoder_charlson_quan2011[!comorbidity_charlson][, all(cmrb_flag == 0L)]
```

We build the `mdcr_v_cmrb` object to make comparing the results between the
`medicalcoder::comorbidities()` call and the `comorbidity::comorbidity()` call
easier.  We will only include the `subject_id`/`hadm_id` reported on by both.

```{r}
#| label: medicalcoder-vs-comorbidity-charlson
mdcr_v_cmrb <-
  merge(
    x = medicalcoder_charlson_quan2011,
    y = comorbidity_charlson,
    all = FALSE,
    by = c("subject_id", "hadm_id"),
    suffixes = c("_mdcr", "_cmrb")
  )
```
The relevant columns to compare are as follows:
```{r}
#| label: mdcr-v-cmrb-charlson-column-map
mvc_columns <-
  data.table::fread(text = "
    medicalcoder   | comorbidity
    aidshiv        | aids
    cebvd          | cevd
    chf_mdcr       | chf_cmrb
    hp_mdcr        | hp_cmrb
    copd           | cpd
    dem            | dementia
    dm             | diab
    dmc            | diabwc
    mal            | canc
    mi_mdcr        | mi_cmrb
    mld_mdcr       | mld_cmrb
    msld_mdcr      | msld_cmrb
    mst            | metacanc
    pud_mdcr       | pud_cmrb
    pvd_mdcr       | pvd_cmrb
    rhd            | rheumd
    rnd            | rend
    cci            | score
  ")
```
If the columns are identical, we remove them from the `mdcr_v_cmrb` object.
This test includes the class of the column.  `medicalcoder::comorbidities()`
returns integer values as does `comorbidity::comorbidity()` save the `score`
columns which is numberic.
```{r}
#| label: column-by-column-mdcr-v-cmrb
data.table::set(
  x = mdcr_v_cmrb,
  j = "score",
  value = as.integer(mdcr_v_cmrb[["score"]])
)

for (i in seq_len(nrow(mvc_columns))) {
  x <- mvc_columns[["medicalcoder"]][i]
  y <- mvc_columns[["comorbidity"]][i]
  z <- identical(mdcr_v_cmrb[[x]], mdcr_v_cmrb[[y]])
  if (z) {
    message(sprintf("`%s` and `%s` are identical.", x, y))
    data.table::set(mdcr_v_cmrb, j = x, value = NULL)
    data.table::set(mdcr_v_cmrb, j = y, value = NULL)
  } else {
    stop(sprintf("`%s` and `%s` are not identical.", x, y))
  }
}
```
```{r}
#| label: verify-only-remaining-mdcr-v-cmrb-columns
#| include: false
stopifnot(
  identical(
    names(mdcr_v_cmrb),
    c("subject_id", "hadm_id", "num_cmrb", "cmrb_flag", "age_score")
  )
)
```
The only columns remaining in the `mdcr_v_cmrb` object are id columns and three
colums unique to the return from `medicalcoder::comorbidities()`:
* `num_cmrb`: the number of flagged comorbidities,
* `cmrb_flag`: a comorbidity flag (at least one comorbidity), and
* `age_score`: an age score (all `NA`) in this case.
```{r}
#| label: remaining-mdcr_v_cmrb
str(mdcr_v_cmrb)
```

### `medicalcoder::comorbidities()` vs MIMIC-IV Code

There are some noteable structural differnces between the results from
`medicalcoder::comorbidities()` and the MIMIC-IV Code.  First, note that the
nubmer of rows differ.  This time for different reasons than when we saw
different results between `medicalcoder::comorbidities()` and
`comorbidity::comorbidity()`.

```{r}
#| label: medicalcoder-vs-mimiciv-code-nrows
nrow(medicalcoder_charlson_mimiciv)
nrow(mimiciv_charlson)
```
Thes records with no comorbidities are not returned by the MIMIC-IV
Code. More percisely, all subject/hadm_id pairs missing from the MIMIC-IV Code
have a missing `hadm_id`.
```{r}
#| label: no-comorb-no-row-in-mimic-code
mimicivDT[!mimiciv_charlson, on = c("hadm_id", "hadm_id")][, all(is.na(hadm_id))]
```
Build the `mdcr_v_mmcc` object to make comparing the results between
the `medicalcoder::comorbidities()` call and the MIMIC-IV SQL easier.
```{r}
#| label: define-mdcr_v_mmcc
mdcr_v_mmcc <-
  merge(
    x = medicalcoder_charlson_mimiciv,
    y = mimiciv_charlson,
    all = FALSE,
    by = c("subject_id", "hadm_id"),
    suffixes = c("_mdcr", "_mimiciv")
  )
```

Another difference in the output form the MIMIC-IV SQL and
`medicalcoder::comorbidities()` is that when an encounter has ICD codes for the
less severe and more severe version of a condition the MIMIC-IV code reports the
flag for both whereas `medicalcoder::comorbidities()` only returns the flag for
the more severe case.   We set the less severe case for the MIMIC-IV code
results to zero when the more severe case is present before comparing all the
restuls.

```{r}
#| label: set-zeros-by-severity
# Set the MIMIC-IV code DM without cc to 0 when cc exist
mdcr_v_mmcc[
  diabetes_without_cc == 1L & diabetes_with_cc == 1L,
  diabetes_without_cc := 0L
]

# Set mild liver disease to 0 when sever liver disease exits for MIMIC-IV code
# results
mdcr_v_mmcc[
  mild_liver_disease == 1L & severe_liver_disease == 1L,
  mild_liver_disease := 0L
]

# set malignant_cancer to zero when metastatic_solid_tumor exits for MIMIC-IV
# code results
mdcr_v_mmcc[
  malignant_cancer == 1L & metastatic_solid_tumor == 1L,
  malignant_cancer := 0L
]
```
Now that the flags have been adjusted, the relevent columns to compare between
`medicalcoder::comorbidities()` and the MIMIC-IV Code are:
```{r}
#| label: mdcr-v-mimic-column-mapping
mvm_columns <-
  data.table::fread(text = "
    medicalcoder   | mimicivcode
    aidshiv        | aids
    cebvd          | cerebrovascular_disease
    chf            | congestive_heart_failure
    copd           | chronic_pulmonary_disease
    dem            | dementia
    dm             | diabetes_without_cc
    dmc            | diabetes_with_cc
    hp             | paraplegia
    mal            | malignant_cancer
    mi             | myocardial_infarct
    mld            | mild_liver_disease
    msld           | severe_liver_disease
    mst            | metastatic_solid_tumor
    pud            | peptic_ulcer_disease
    pvd            | peripheral_vascular_disease
    rhd            | rheumatic_disease
    rnd            | renal_disease
    age_score_mdcr | age_score_mimiciv
    cci            | charlson_comorbidity_index
  ")
```
If the columns are identical, they are removed from the `mdcr_v_mmcc` object.
```{r}
for (i in seq_len(nrow(mvm_columns))) {
  x <- mvm_columns[["medicalcoder"]][i]
  y <- mvm_columns[["mimicivcode"]][i]
  z <- identical(mdcr_v_mmcc[[x]], mdcr_v_mmcc[[y]])
  if (z) {
    message(sprintf("`%s` and `%s` are identical.", x, y))
    data.table::set(mdcr_v_mmcc, j = x, value = NULL)
    data.table::set(mdcr_v_mmcc, j = y, value = NULL)
  } else {
    stop(sprintf("`%s` and `%s` are not identical.", x, y))
  }
}
```
The only remaining columns are id columns and the `num_cmrb` column, which is
unique to the return from `medicalcoder::comorbidities()`.
```{r}
#| label: sanity-check-mdcr-v-mmcc-remaining-columns
#| include: false
stopifnot(
  identical(
    names(mdcr_v_mmcc),
    c("subject_id", "hadm_id", "num_cmrb", "cmrb_flag")
  )
)
```

```{r}
#| label: str-remaining-mdcr-v-mmcc
str(mdcr_v_mmcc)
```

## Elixhauser Comorbidities

### `medicalcoder::comorbidities()` vs `comorbidity::comorbidity()`

As with the Charlson comparison, the numbers of rows between the two results
differ. And for the same reason: no diagnostic ICD codes.
```{r}
#| label: nrow-medicalcoder-vs-comorbidity-elixhauser
nrow(medicalcoder_elixhauser_quan2005)
nrow(comorbidity_elixhauser)
# subject/hadm_id not in comorbibity::comorbibity() results all have no
# diagnostic codes.
mimicivDT[!comorbidity_elixhauser][, any(dx == 1 & !is.na(icd_code))]
medicalcoder_elixhauser_quan2005[!comorbidity_elixhauser][, all(cmrb_flag == 0L)]
```
We build the `mdcr_v_cmrb_elix` object to make comparing the results between the
`medicalcoder::comorbidities()` call and the `comorbidity::comorbidity()` call
easier.  We will only include the `subject_id`/`hadm_id` reported on by both.

```{r}
#| label: medicalcoder-vs-comorbidity-elixhauser
mdcr_v_cmrb_elix <-
  merge(
    x = medicalcoder_elixhauser_quan2005,
    y = comorbidity_elixhauser,
    all = FALSE,
    by = c("subject_id", "hadm_id"),
    suffixes = c("_mdcr", "_cmrb")
  )
```
The relevant columns to compare are as follows:
```{r}
#| label: mdcr-v-cmrb-elixhauser-column-map
mvc_elix_columns <-
  data.table::fread(text = "
    medicalcoder        | comorbidity
    AIDS                | aids
    ALCOHOL             | alcohol
    ANEMDEF             | dane
    ARTH                | rheumd
    BLDLOSS             | blane
    CARDIAC_ARRHYTHMIAS | carit
    CHF                 | chf
    CHRNLUNG            | cpd
    COAG                | coag
    DEPRESS             | depre
    DM                  | diabunc
    DMCX                | diabc
    DRUG                | drug
    HTN_CX              | hypc
    HTN_UNCX            | hypunc
    HYPOTHY             | hypothy
    LIVER               | ld
    LYMPH               | lymph
    LYTES               | fed
    METS                | metacanc
    NEURO               | ond
    OBESE               | obes
    PARA                | para
    PERIVASC            | pvd
    PSYCH               | psycho
    PULMCIRC            | pcd
    RENLFAIL            | rf
    TUMOR               | solidtum
    ULCER               | pud
    VALVE               | valv
    WGHTLOSS            | wloss
  ")
```
If the columns are identical, we remove them from the `mdcr_v_cmrb_elix` object.
```{r}
#| label: column-by-column-mdcr-v-cmrb-elix
for (i in seq_len(nrow(mvc_elix_columns))) {
  x <- mvc_elix_columns[["medicalcoder"]][i]
  y <- mvc_elix_columns[["comorbidity"]][i]
  z <- identical(mdcr_v_cmrb_elix[[x]], mdcr_v_cmrb_elix[[y]])
  if (z) {
    message(sprintf("`%s` and `%s` are identical.", x, y))
    data.table::set(mdcr_v_cmrb_elix, j = x, value = NULL)
    data.table::set(mdcr_v_cmrb_elix, j = y, value = NULL)
  } else {
    stop(sprintf("`%s` and `%s` are not identical.", x, y))
  }
}
```
```{r}
#| label: verify-only-remaining-mdcr-v-cmrb-columns-elixhauser
#| include: false
stopifnot(
  identical(
    names(mdcr_v_cmrb_elix),
    c("subject_id", "hadm_id", "HTN_C", "num_cmrb", "cmrb_flag", "mortality_index", "readmission_index")
  )
)
```
The only columns remaining in the `mdcr_v_cmrb_elix` object are id columns and
colums unique to the return from `medicalcoder::comorbidities()`:
* `num_cmrb`: the number of flagged comorbidities,
* `cmrb_flag`: a comorbidity flag (at least one comorbidity),
* `HTN_C`: any hypertension,
* `mortality_index` and `readmission_index`: based on the weights from AHRQ.
```{r}
#| label: remaining-mdcr_v_cmrb_elix
str(mdcr_v_cmrb_elix)
```

## PCCC

### `medicalcoder::comorbidities()` vs `pccc:::ccc()`
As we have seen in other cases, the absence of ICD codes for a `hadm_id` results
in the omision of the record from the return from `pccc::ccc()`.
```{r}
#| label: nrow-medicalcoder-vs-pccc
nrow(medicalcoder_pccc_v2.0)
nrow(pccc_pcccv2.0)
```
For the common `subject_id`/`hadm_id` pairs we compare the flags.
```{r}
mdcr_v_pccc <-
  merge(
    x = medicalcoder_pccc_v2.0,
    y = pccc_pcccv2.0,
    all = FALSE,
    by = c("subject_id", "hadm_id"),
    suffixes = c("_mdcr", "_pccc")
  )
```

```{r}
#| label: mvp-columns
mvp_columns <- data.table::fread(text = "
  medicalcoder | pccc
  congeni_genetic_mdcr | congeni_genetic_pccc
  cvd_mdcr | cvd_pccc
  gi_mdcr | gi_pccc
  hemato_immu_mdcr | hemato_immu_pccc
  malignancy_mdcr | malignancy_pccc
  metabolic_mdcr | metabolic_pccc
  neonatal_mdcr | neonatal_pccc
  neuromusc_mdcr | neuromusc_pccc
  renal_mdcr | renal_pccc
  respiratory_mdcr | respiratory_pccc
  cmrb_flag | ccc_flag"
)
```
```{r}
#| label: check-identical-columns-mdcr-v-pccc
for (i in seq_len(nrow(mvp_columns))) {
  x <- mvp_columns[["medicalcoder"]][i]
  y <- mvp_columns[["pccc"]][i]
  z <- identical(mdcr_v_pccc[[x]], mdcr_v_pccc[[y]])
  if (z) {
    message(sprintf("`%s` and `%s` are identical.", x, y))
    data.table::set(mdcr_v_pccc, j = x, value = NULL)
    data.table::set(mdcr_v_pccc, j = y, value = NULL)
  } else {
    stop(sprintf("`%s` and `%s` are not identical.", x, y))
  }
}
```
There are some differences in the results between the two approaches.
```{r}
#| label: structre-remaining-mdcr-v-pccc
str(mdcr_v_pccc)
```

The `misc` column is the "miscellaneous" category reported by
`medicalcoder::comorbidities()` and is not reported by `pccc::ccc()`.  The
existence of the `misc` column and some differences in the returned results
between `pccc::ccc()` version 1.0.7, and `medicalcoder::comorbidities()` is due
to how medicalcoder is implemented.

The differences in any technology dependence and transplant are due to design
difference between _medicalcoder_ and _pccc_.  _medicalcoder_ uses database
style joins to map precomputed ICD-codes to conditions.  This allows for a
simple extension of mapping ICD-codes to both the primary condition and
subcondition as the latter require only an additional column in a lookup table.
The _pccc_ uses partical string matching to map ICD codes to condtions for
each of the primary conditions and only the two subconditions of technology
dependence and transplant.  When building _medicalcoder_, the lookup table with
condition and subconditons was extended to match the publication documentation.

```{r}
#| label: any-tech-dep-and-transplant
mdcr_v_pccc[, .N, keyby = .(medicalcoder = any_tech_dep, pccc = tech_dep)]
mdcr_v_pccc[, .N, keyby = .(medicalcoder = any_transplant, pccc = transplant)]
```

Several GitHub issues have be created for _pccc_ to correct these issues for
specific ICD codes:

* [ICD-9 349.1](https://github.com/CUD2V/pccc/issues/45)
* [ICD-9 V56](https://github.com/CUD2V/pccc/issues/46)
* [ICD-10 Z49](https://github.com/CUD2V/pccc/issues/47)
* [ICD-9 86.06](https://github.com/CUD2V/pccc/issues/48)
* [ICD-9 V45.85](https://github.com/CUD2V/pccc/issues/49)
* [ICD-9 V53.3](https://github.com/CUD2V/pccc/issues/50)
* [ICD-9 V53.91](https://github.com/CUD2V/pccc/issues/51)
* [ICD-9 V65.46](https://github.com/CUD2V/pccc/issues/52)
* [ICD-9 37.52](https://github.com/CUD2V/pccc/issues/53)
* [ICD-9 V42.0](https://github.com/CUD2V/pccc/issues/54)
* [ICD-10 Z94](https://github.com/CUD2V/pccc/issues/55)

# Longitudinal Flagging

Under the assumption that all relevent ICD codes are (re)reported on each
encounter then there is no need to carry-forward comorbidity flags.  However,
that assumption is rarely valid. _medicalcoder_ provides a utility for
indefinite carry-forwared of a condition. As reported in the primary manuscript
this includes accounting for condtions which should not be reported on a
specific encounter but are reported on the following encounters. In the
following we will explore some of the pros and cons of the indefinite
carry-forward of comorbidities.

To simplify the work we will reapply the Quan 2011 method using `enc_seq`
instead of `hadm_id` so that the encounters are considered in chronologic order.

```{r}
#| label: current-and-cumulative
medicalcoder_charlson_quan2011_current <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    primarydx   = 0L,
    method      = "charlson_quan2011",
    flag.method = "current"
  )

medicalcoder_charlson_quan2011_cumulative <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    primarydx   = 0L,
    method      = "charlson_quan2011",
    flag.method = "cumulative"
  )
```

## Diabetes

Non-gestational diabetes is generally considered a chronic condition.  Even when
well managed, it is still a relevent diagnosis.  The ICD-10 code defining
diabetes (with and without complications) for Charlson comorbidities as
defined by Quan et al. (2005, 2011), are all in the E10, E11, E12, E13, and E14
categories.

```{r}
#| label: ICD-10-categories-dm-quan
subset(
  x = medicalcoder::get_charlson_codes(),
  subset = charlson_quan2005 == 1L & condition %in% c("dm", "dmc") & icdv == 10,
)[["code"]] |>
substring(1, 3) |>
unique() |>
sort()
```
Let's look at the records for those who have at least two encounters and at
least one of the encounters was flagged with diabetes (with or without
complications).
```{r}
#| level: minimum-two-encounters-at-least-one-with-dm
sids <-
  medicalcoder_charlson_quan2011_current[
    ,
    .(encounters = length(unique(enc_seq)), any_dm = sum(dm + dmc)),
    by = .(subject_id)
  ][
    encounters > 1 & any_dm > 0,
    subject_id
  ]

dms <-
  merge(
    x = medicalcoder_charlson_quan2011_current[
          subject_id %in% sids,
          .(subject_id, enc_seq, dm_current = dm, dmc_current = dmc)
        ],
    y = medicalcoder_charlson_quan2011_cumulative[
          subject_id %in% sids,
          .(subject_id, enc_seq, dm_cumulative = dm, dmc_cumulative = dmc)
        ],
    all = TRUE,
    by = c("subject_id", "enc_seq")
  )
data.table::setDT(dms)
dms[, current := (dm_current + dmc_current)]
dms[, cumulative := (dm_cumulative + dmc_cumulative)]
dms[, total_encs := max(enc_seq), by = .(subject_id)]
dms
```
How many cumulative flags are on encounters were DM i(with or without
complications) is flagged on both a preceding and a proceding encounter, or only
after a preceding encounter.
```{r}
#| label: cumulative-flags-between-and-after-current-flag
dms[, current_flags_before := as.integer((cumsum(current) - current) > 0), by = .(subject_id)]
dms[, current_flags_after  := as.integer(rev(cumsum(rev(current)) - current) > 0), by = .(subject_id)]

dms[, cumulative_between_current_flags   := as.integer((current != cumulative) & current_flags_before == 1L & current_flags_after == 1L)]
dms[, cumulative_only_after_current_flag := as.integer((current != cumulative) & current_flags_before == 1L & current_flags_after == 0L)]

# Number of Encounters:
dms[, .(Encounters = .N), keyby = .(cumulative_between_current_flags, cumulative_only_after_current_flag)]
```
At a subject-level, the
```{r}
#| label: cumulative-flags-between-and-after-current-flag-subject-level
# counts of subjects with encounters flagged
dms[
    ,
    .(
      between = any(cumulative_between_current_flags),
      after = any(cumulative_only_after_current_flag)
    ),
    by = .(subject_id)
  ][
    ,
    .(Subjects = .N),
    keyby = .(between, after)
  ]
```

```{r}
#| label: more-dm-aggregations
#| eval: false
#| include: false
dms[
  ,
  .(
     dm    = all(dm_current == dm_cumulative),
     dmc   = all(dmc_current == dmc_cumulative),
     anydm = all(current == cumulative)
  ),
  by = .(subject_id)
  ][, .N, keyby = .(dm, dmc, anydm)]

dms[
  ,
  .(
     dm = all(dm_current == dm_cumulative),
     dmc = all(dmc_current == dmc_cumulative)
  ),
  by = .(subject_id, cumulative_flag_current_no_flag)
  ][, .N, keyby = .(dm, dmc, cumulative_flag_current_no_flag)]

4732+3993+697 # 9422
17634 / (4732+3993+697+17634) # patients with no difference

dms[
  ,
  .(
     dm = all(dm_current == dm_cumulative),
     dmc = all(dmc_current == dmc_cumulative)
  ),
  by = .(subject_id, enc_seq)
  ][, .N, keyby = .(dm, dmc)]

12905+11649+3345 # 27899
118844 / (12905+11649+3345+118844) # encounters with no difference

# How many encounters was dm/dmc status missed using only encounter level data?
dms[, .(encounters = .N, subjects = data.table::uniqueN(subject_id))]
dms[, qwraps2::n_perc(cumulative > current, show_denom = "always")]
dms[, any(cumulative > current), by = .(subject_id)][, qwraps2::n_perc(V1, show_denom = "always")]
```

Longitudinal flagging of a chronic condition could be beneficial, especially in
the cases were we see the condition flagged on encounters preceding and
proceding a specific encounter.

End users are cuationed to understand the specific conditions and comorbidity
methods being used before applying the indefinite carry-forward.  For example,
gestational diabetes is part of the comorbidity mappings for the AHRQ Elixhauser
variants based on ICD-10 codes.  Note that ICD-10 category E08, E09, E10, E11,
E13, _and_ O24 are all diabetes codes with O24 codes mapping to uncomplicated
diabetes.
```{r}
#| label: ahrq-icd-10-dm-codes
ahrq_icd10_dm_codes <-
  subset(
    x = medicalcoder::get_elixhauser_codes(),
    subset = elixhauser_ahrq_icd10 == 1 & condition %in% c("DIAB_CX", "DIAB_UNCX"),
    select = c("code", "condition")
  )
data.table::setDT(ahrq_icd10_dm_codes)
ahrq_icd10_dm_codes[, category := substring(code, 1, 3)]

ahrq_icd10_dm_codes[, .N, keyby = .(category, condition)] |>
  data.table::dcast(category ~ condition, value.var = "N", fill = 0L)
```
In the MIMIC-IV data, there are subjects with both gestational and
non-gestational diabetes, subjects with only gestational diabetes, and subjects
with only non-gestational diabetes.
```{r}
#| label: gestational-and-non-gestational-dm
DT <- mimicivDT[icd_code %in% ahrq_icd10_dm_codes[condition == "DIAB_UNCX", code]]
DT[, Gestational := any(startsWith(icd_code, "O")), by = .(subject_id)]
DT[, `Non-Gestational` := any(startsWith(icd_code, "E")), by = .(subject_id)]
data.table::setkey(DT, subject_id, enc_seq)
DT[
  , unique(.SD), .SDcols = c("subject_id", "Gestational", "Non-Gestational")
  ][
  , .(Subjects = .N), keyby = .(Gestational, `Non-Gestational`)
  ]
```
Since the AHRQ variant of Elixhauser comorbidities include both chronic and
non-chronic versions of diabetes `flag.method = "cumulative"` may results false
positives of diabetes.

## Metastatic Tumors

Metastatic tumors could be consider non-chronic conditions as the tumors could
be removed via surgery or treated with other therapies.  As such, the flags
reported via `flag.method = "cumulative"` for malignancy (Charlson and PCCC),
or solid tumors (Elixhauser) should not be interpreted as "present disease" but
as "a history of _observed_ disease."

We empathize _observed disease_ because there are ICD codes specifically for
personal history of malignancies.  We collect those ICD codes and the codes
mapping to malignancies per Quan et al. (2005, 2011) and subset the `mimicivDT`
to subjects with these codes.

```{r}
#| label: history-of-malignant
personal_history_of_malignant <-
  subset(
    x = medicalcoder::get_icd_codes(with.description = TRUE),
    subset = src == "cms" & grepl("personal history of malignant", desc, ignore.case = TRUE),
    select = c("code", "icdv", "dx")
  )
personal_history_of_malignant[["history_of"]] <- 1L

charlson_quan2011_mal <-
  subset(
    x = medicalcoder::get_charlson_codes(),
    subset = condition == "mal" & charlson_quan2011 == 1L,
    select = c("code", "icdv", "dx")
  )
charlson_quan2011_mal[["observed_mal"]] <- 1L

mal <-
  merge(
    x = mimicivDT[!is.na(icd_code), .(subject_id, enc_seq, icd_code, icd_version, dx)],
    y = personal_history_of_malignant,
    all.x = TRUE,
    by.x = c("icd_code", "icd_version", "dx"),
    by.y = c("code", "icdv", "dx")
  )

mal <-
  merge(
    x = mal,
    y = charlson_quan2011_mal,
    all.x = TRUE,
    by.x = c("icd_code", "icd_version", "dx"),
    by.y = c("code", "icdv", "dx")
  )
data.table::setkey(mal, subject_id, enc_seq)

# set flag for any history or observation of a malignancy
mal <-
  mal[
    ,
    .(
      enc_history_of   = as.integer(any(history_of, na.rm = TRUE)),
      enc_observed_mal = as.integer(any(observed_mal, na.rm = TRUE))
    ),
    by = .(subject_id, enc_seq)
  ]
mal[, subject_history_of   := as.integer(sum(enc_history_of) > 0), by = .(subject_id)]
mal[, subject_observed_mal := as.integer(sum(enc_observed_mal) > 0), by = .(subject_id)]

mal <- mal[subject_history_of + subject_observed_mal > 0]

# join on the cumulative flags from medicalcoder for both malignancy and
# metastatic solid tumors.  An encounter with both a malignancy and a metastatic
# solid tumor will have the malignancy flag set to zero and retain only the more
# severe condition.
mal <-
  merge(
    x = mal,
    y = medicalcoder_charlson_quan2011_cumulative[, .(subject_id, enc_seq, medicalcoder_cummulative_mal = mal, medicalcoder_cummulative_mst = mst)],
    all.x = TRUE,
    by = c("subject_id", "enc_seq")
  )

mal[, .(Encounters = .N), keyby = .(enc_observed_mal, enc_history_of)]
```
The next chunk will count the number of subjects with reported history of a
malignancy and those with an observed malignancy
```{r}
#| label: mal-subject-level-variables
subject_history <-
  mal[
    ,
    lapply(lapply(.SD, any), as.integer),
    .SDcols = c("subject_history_of", "subject_observed_mal"),
    by = .(subject_id)
  ]
subject_history[
    ,
    .(subjects = .N), keyby = .(subject_history_of, subject_observed_mal)
  ]
```
What is notable in the above is that there are
`r with(subject_history, qwraps2::n_perc(subject_history_of == 1 & subject_observed_mal == 0))`
patients who have only a personal history of malignancy in the data set.  If the
cumulative flag was to be considered a "history of" these would patients would
be considered a false negative.  The, the `flag.method = 'cumulative'` can only
be interpreted as "current or history of observed disease."

# References
