---
title: "MIMIC-IV Data Analysis via medicalcoder"
subtitle: "Supplement to JAMIA Open Manuscript"

format:
  pdf:
    toc: true
    toc-depth: 4
    lof: true
    lot: true
    number-sections: true
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage[table]{xcolor}
        \usepackage{threeparttable}
        \usepackage{longtable}

bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE
)

options(knitr.kable.NA = "")

# Go Avs Go!
# {
#   "burgundy" : "#6F263D",
#   "blue" : "#236192",
#   "light-blue" : "#4F8FCB"
#   "silver" : "#A2AAAD",
#   "black" : "#000000"
# }

loadNamespace('medicalcoder')
library(data.table)

knitr::read_chunk("mimiciv-data-analysis.R")
```

# Objective and Setup

This supplement provides reproducible validation and benchmarking of
`medicalcoder::comorbidities()` using MIMIC-IV v3.1 data, including algorithmic
agreement with established implementations and demonstration of longitudinal
flagging behavior.

# Materials and Methods

## Data

We use the MIMIC-IV v3.1 Data
[@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000] as the foundation for
these examples.

The MIMIC-IV v3.1 dataset is a large, publicly available clinical database
containing de-identified health data from patients admitted to intensive care
units at Beth Israel Deaconess Medical Center. It includes detailed information
on diagnoses, procedures, laboratory results, medications, and hospital
encounters, with ICD-9-CM and ICD-10-CM/PCS codes available for billing and
comorbidity research. MIMIC-IV is widely used for method development,
validation, and reproducible research in clinical informatics and outcomes
research.

The MIMIC-IV v3.1 data set can be obtained so long as you meet the requirements
set by PhysioNet[^1].

[^1]: https://physionet.org/content/mimiciv/3.1/

## Comorbidity Algorithms

Within the _medicalcoder_ package are implementations for several variants of
the Charlson, Elixhauser, and PCCC comorbidity algorithms,
(@tbl-comorbidity-methods) [@beyrer2021validation;@quan2005;@quan2011;@glasheen2019;@ahrq2025;@feudtner2014pediatric;@feinstein2024pediatric;@dewitt2026pccc;@feinstein2018pccc;@pcccv3sas].

__NOTE:__
The MIMIC-IV data consists of only adult patients.
Although PCCC was designed for pediatric populations, adult MIMIC-IV data are
used here solely to demonstrate algorithmic implementation.

```{r}
#| label: tbl-comorbidity-methods
#| tbl-cap: "Comorbidity methods implemented in _medicalcoder_ and sorted by comorbidity family and method in alphabetical order."
#| echo: false
```

## Alternative Utilities

We compared the results from `medicalcoder::comorbidities()` to the Charlson
comorbidities flagged by the MIMIC-IV code, and the R package _comorbidity_
[@r-comorbidity].  For Elixhauser we will compare results between
`medicalcoder::comorbidities()` and `comorbidity::comorbidity()`.  Lastly, we
will compare PCCC version 2 results between `medicalcoder::comorbidities()` and
the R package _pccc_ [@dewitt2026pccc] call `pccc::ccc()`.

# Prepare MIMIC-IV Data For Analysis

We have stored the path to the MIMIC-IV data as a system variable on our
machines as `MIMICIVDATA`.  The following code is provide to help readers verify
that they are using the same version of the MIMIC-IV v3.1 data as used in this
document.

```{r}
#| label: verify-mimiciv-version
```

We only need four of the MIMIC-IV hospital datasets:
`admissions`, `patients`, `diagnoses_icd`, and `procedures_icd`.

We will do some light formatting before using `medicalcoder::comorbidities()` to
flag comorbidities.  Start by importing the `admissions`, `patients`,
`diagnoses_icd`, and `procedures_icd`, datasets.

@tbl-mimicivdata-checksums gives the sha256 checksums for these files
Each file checksum has been verified.

```{r}
#| label: tbl-mimicivdata-checksums
#| tbl-cap: "SHA256 Check sums for files in the MIMIC-IV v3.1 data."
#| echo: false
#| results: "asis"
```
```{r}
#| label: sha256-verify
#| include: false
#| cache: true
```
```{r}
#| label: import-mimiciv-data
#| cache: true
#| dependson: ["sha256-verify"]
```

We will build an additional data set for age (in years).

```{r}
#| label: build-mimicivdata-ages
```

For the calls to `medicalcoder::comorbidities()` we can have all the ICD codes
in one `data.frame` (data.table [@r-data.table] will be used here).
```{r}
#| label: build-mimicivdata-icd
```

We put all the parts together into one data.frame (data.table)

```{r}
#| label: one-data-set
```
We need an encounter sequence variable, the `hadm_id` are not sequential.

NOTE: `admittime` can be used in the calls to
`medicalcoder::comorbidities()`, but there is additional overhead and
computational expense when sorting and joining by POSIXct variable.  The
`enc_seq` achieves the same utility with lower computational overhead.

```{r}
#| label: build-enc-seq
```

_medicalcoder_ can use multiple id variables.  Other utilities such as
_comorbidity_ and _pccc_ can only use one id variable.  As such we create `cid`
to use in those examples.
```{r}
#| label: build-cid
```
Our final `mimicivDT` object:
```{r}
#| label: mimicivDT-data-structure
```

There are `r qwraps2::frmt(nrow(mimicivDT))` total rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
hospital encounters.

# Investigating ICD Codes

The _medicalcoder_ package provides utilities for investigating ICD codes.
For the MIMIC-IV data, let's just check if the reported ICD codes are in the
_medicalcoder_ ICD data.

## Utilities

There are several functions in the _medicalcoder_ package for working with ICD
codes:

* `get_icd_codes()`: returns a lookup table of ICD codes as a `data.frame`.
* `lookup_icd_codes()`: returns details on specific ICD codes.
* `is_icd()`: returns `TRUE` or `FALSE` for a vector of codes while considering
  ICD version, type, and billable status.
* `icd_compact_to_full()`: insert a decimal point into a string to be
  consistent with ICD-9 diagnostic, ICD-9 procedure, or ICD-10 diagnostic
  codes.  (ICD-10 procedure codes do not have decimal places.)  NOTE: this will
  not validate the return as a valid ICD code, just format the input string
  appropriately.

We encourage readers to review the vignette in _medicalcoder_ for details on
these utilities.
```{r}
#| label: vignette
#| eval: false
vignette(topic = "icd", package = "medicalcoder")
```

## ICD-Codes in MIMIC-IV Data

Let's quickly verify that all the reported ICD codes in the MIMIC-IV data are
known and ICD-9-CM, ICD-9-PCS, ICD-10-CM, or ICD-10-PCS codes.

```{r}
#| label: mimicivdata-unique-icd-codes
```
The call `medicalcoder::is_icd()` will return `TRUE` or `FALSE` for a vector of
strings if the input is a known ICD code based on version (`icdv = 9` or `icdv =
10`), diagnostic (`dx = 1`) or procedure (`dx = 0`) codes, and source.
```{r}
#| label: mimicivdata-is-icd-is-cms
```
There are some codes which appear to not be valid ICD codes.
```{r}
#| label: mimicivdata-n-by-iscms
```
Let's investigate these codes.  All of them are header codes, never assignable.
```{r}
#| label: mimicivdata-not-iscms
```
If `headerok = TRUE` in the call `medicalcoder::is_icd()`, then these header
codes will be considered valid.
```{r}
#| label: mimicivdata-is-icd-is-cms-headers
```
Lastly, to get more detail on any of the codes you can use
`medicalcoder::lookup_icd_codes()` or `medicalcoder::get_icd_codes()`.
```{r}
#| label: icd-code-look-up-examples
```

# Flagging Encounter Level Comorbidities

In the following we will outline how to apply the comorbidities methods to the
`mimicivDT` dataset, focusing on encounter level flagging of comorbidities.  We
start by showing how to apply the
Charlson [@quan2005],
Elixhauser [@quan2005], and
PCCC v2 [@feudtner2014pediatric] methods to the `mimicivDT` dataset using the
`medicalcoder::comorbidities()` call.  Following that we will show how to do the
same work using SQL code provided by MIMIC-code [@johnson2018mimic], the R
package _comorbidity_ [@r-comorbidity], and the R package _pccc_ [@dewitt2026pccc].
Lastly, we compare and contrast the differences in the flagged comorbidities
between the methods.

## `medicalcoder::comorbidities()`

### Charlson Comorbidities
There are several variants of Charlson like comorbidity methods implemented in
_medicalcoder_.  One way to see the current list of implemented methods is to
look at the names of the lookup table mapping ICD codes to conditions:
```{r}
#| label: charlson-variants
```
The `charlson_quan2011` method will be applied to the `mimicivDT` dataset and
the results will be compared to the results from the
`comorbidity::comorbidity()` method.
```{r}
#| label: medicalcoder-charlson-quan2011
```
The SQL code for applying Charlson comorbidities to the MIMIC-IV data provided
by [MIT-LCP](https://github.com/MIT-LCP/mimic-code) has a slightly different set
of ICD-10 codes mapping to conditions.  Quan et al. (2005, 2011) is based on
ICD-10 codes, not ICD-10-CM.  The MIMIC-IV SQL code includes the C4A category
which is an ICD-10-CM specific category (see [this github
issue](https://github.com/MIT-LCP/mimic-code/issues/2017)).  To be consistent
with the expected set of ICD codes identified in the Quan papers and the
MIMIC-IV SQL code, _medicalcoder_ provides distinct methods.
```{r}
#| label: medicalcoder-charlson-mimiciv
```

### Elixhauser Comorbidities
As with the Charlson comorbidities there are multiple variants of Elixhauser
like comorbidity methods implemented in _medicalcoder_.
```{r}
#| label: elixhauser-variants
```
To apply the `elixhauser_quan2005` version, which is comparable to the R package
_comorbidity_, do the following:
```{r}
#| label: medicalcoder-elixhauser-quan2005
```

### PCCC v2.0
PCCC v2.0 has been implemented to be consistent with the R package _pccc_.
The v2.1 variants extends the ICD to conditions mapping while using the same
underlying algorithm. v3.0 and v3.1 use a different set of ICD codes and
algorithm for flagging conditions.  For this work we focus on v2.0 so we can
compare it to the results from the R package _pccc_.

```{r}
#| label: medicalcoder-pccc-v2.0
```

## MIMIC-IV Code

We use the SQL code for applying the Charlson
comorbidity algorithm to the MIMIC-IV data provided by the MIT Laboratory for
Computational Physiology (MIT\_LCP) on
[GitHub](https://github.com/MIT-LCP/mimic-code.git).  We stored the
path to a local copy of the repository in the system environmental variable
`MIMICIVCODE`.  The specific version of the SQL code being used is from commit
[278df75ec30991ff3a6f5ceb6d2221635a085e9f](https://github.com/MIT-LCP/mimic-code/blob/278df75ec30991ff3a6f5ceb6d2221635a085e9f/mimic-iv/concepts/comorbidity/charlson.sql)

```{r}
#| label: checksum-mimiciv-code
```

The SQL code is expected to run on Google Big Query.  We make a few
small modifications to the code so we can evaluate the SQL locally via RSQLite.

```{r}
#| label: mimic-code-edits
```

We set up an SQL database in memory and evaluate the code.

```{r}
#| label: mimiciv-charlson-results
#| cache: true
```

## `comorbidity::comorbidity()`

The R package _comorbidity_ [@r-comorbidity] is a widely used and referenced
tool for applying the Charlson and Elixhauser comorbidities.
One consideration for using `comorbidity::comorbidity()` is that ICD-9 or ICD-10
codes are used independently. There are
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id)][V1 > 1, .N]`
subjects with both ICD-9 and ICD-10 codes in their records, and
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id, hadm_id)][V1 > 1, .N]`
encounters with both ICD versions.

To flag the conditions appropriately we must split the data by ICD version,
apply `comorbidity::comorbidity()` to each split, and then aggregate the
results.  Also, `comorbidity::comorbidity()` only takes one column for the id
and thus we need to use `cid` for the id instead of the `c("subject_id",
"hadm_id")` used in the calls to `medicalcoder::comorbidities()`.  In the
following work we will copy the data built for the encounter level flagging and
extend for cumulative flagging as well.

The following is done using _comorbidity_ version
`r packageVersion("comorbidity")`.

### Charlson

```{r}
#| label: comorbidity-charlson
#| cache: true
```

### Elixhauser

Compared to the work done for flagging Charlson comorbidities, Elixhauser
flagging requires changing the `map` argument in the calls to
`comorbidity::comorbidity()` and there is no scoring to be done.

```{r}
#| label: comorbidity-elixhauser
#| cache: true
```

## `pccc::ccc()`

The R package _pccc_ [@dewitt2026pccc] implements PCCC v2
[@feudtner2014pediatric].  As with `comorbidity::comorbidity()` and many other
implementations of comorbidities, only one version of ICD is assessed at a time.

The expected input structure of the data for `pccc::ccc()` is a wide format, one
row per encounter, one column for each ICD code.
The following uses _pccc_ version `r packageVersion("pccc")`.

```{r}
#| label: pccc-pcccv2.0
#| cache: true
```

# Comparing Encounter-level Flags Between Utilities

## Charlson Comorbidities

### `medicalcoder::comorbidities()` vs `comorbidity::comorbidity()`

The first thing to not is that the number of row returned by the two
implementations are different.
```{r}
#| label: nrows-medicalcoder-vs-comorbidity
```
The reason for this is that there are encounters with no diagnostic ICD codes.
`medicalcoder::comorbidities()` returns a row for these encounters and
`comorbidity::comorbidity()` does not.  For these encounters,
`medicalcoder::comorbidities()` returns a row with zero comorbidities.

```{r}
#| label: no-icd-no-comorbibity-result
```

We build the `mdcr_v_cmrb` object to make comparing the results between the
`medicalcoder::comorbidities()` call and the `comorbidity::comorbidity()` call
easier.  We will only include the `subject_id`/`hadm_id` reported on by both.

```{r}
#| label: medicalcoder-vs-comorbidity-charlson
```
The relevant columns to compare are as follows:
```{r}
#| label: mdcr-v-cmrb-charlson-column-map
```
If the columns are identical, we remove them from the `mdcr_v_cmrb` object.
This test includes the class of the column.  `medicalcoder::comorbidities()`
returns integer values as does `comorbidity::comorbidity()` save the `score`
columns which is numberic.
```{r}
#| label: column-by-column-mdcr-v-cmrb
```
```{r}
#| label: verify-only-remaining-mdcr-v-cmrb-columns
#| include: false
```
The only columns remaining in the `mdcr_v_cmrb` object are id columns and three
columns unique to the return from `medicalcoder::comorbidities()`:
* `num_cmrb`: the number of flagged comorbidities,
* `cmrb_flag`: a comorbidity flag (at least one comorbidity), and
* `age_score`: an age score (all `NA`) in this case.
```{r}
#| label: remaining-mdcr_v_cmrb
```

### `medicalcoder::comorbidities()` vs MIMIC-IV Code

There are some noteable structural differnces between the results from
`medicalcoder::comorbidities()` and the MIMIC-IV Code.  First, note that the
nubmer of rows differ.  This time for different reasons than when we saw
different results between `medicalcoder::comorbidities()` and
`comorbidity::comorbidity()`.

```{r}
#| label: medicalcoder-vs-mimiciv-code-nrows
```
These records with no comorbidities are not returned by the MIMIC-IV
Code. More precisely, all subject/hadm_id pairs missing from the MIMIC-IV Code
have a missing `hadm_id`.
```{r}
#| label: no-comorb-no-row-in-mimic-code
```
Build the `mdcr_v_mmcc` object to make comparing the results between
the `medicalcoder::comorbidities()` call and the MIMIC-IV SQL easier.
```{r}
#| label: define-mdcr_v_mmcc
```

Another difference in the output form the MIMIC-IV SQL and
`medicalcoder::comorbidities()` is that when an encounter has ICD codes for the
less severe and more severe version of a condition the MIMIC-IV code reports the
flag for both whereas `medicalcoder::comorbidities()` only returns the flag for
the more severe case.   We set the less severe case for the MIMIC-IV code
results to zero when the more severe case is present before comparing all the
results.

```{r}
#| label: set-zeros-by-severity
```
Now that the flags have been adjusted, the relevant columns to compare between
`medicalcoder::comorbidities()` and the MIMIC-IV Code are:
```{r}
#| label: mdcr-v-mimic-column-mapping
```
If the columns are identical, they are removed from the `mdcr_v_mmcc` object.
```{r}
#| label: check-columns-mdcr-v-mmcc
```
The only remaining columns are id columns and the `num_cmrb` column, which is
unique to the return from `medicalcoder::comorbidities()`.
```{r}
#| label: sanity-check-mdcr-v-mmcc-remaining-columns
#| include: false
```
```{r}
#| label: str-remaining-mdcr-v-mmcc
```

## Elixhauser Comorbidities

### `medicalcoder::comorbidities()` vs `comorbidity::comorbidity()`

As with the Charlson comparison, the numbers of rows between the two results
differ. And for the same reason: no diagnostic ICD codes.
```{r}
#| label: nrow-medicalcoder-vs-comorbidity-elixhauser
nrow(medicalcoder_elixhauser_quan2005)
nrow(comorbidity_elixhauser)
# subject/hadm_id not in comorbibity::comorbibity() results all have no
# diagnostic codes.
mimicivDT[!comorbidity_elixhauser][, any(dx == 1 & !is.na(icd_code))]
medicalcoder_elixhauser_quan2005[!comorbidity_elixhauser][, all(cmrb_flag == 0L)]
```
We build the `mdcr_v_cmrb_elix` object to make comparing the results between the
`medicalcoder::comorbidities()` call and the `comorbidity::comorbidity()` call
easier.  We will only include the `subject_id`/`hadm_id` reported on by both.

```{r}
#| label: medicalcoder-vs-comorbidity-elixhauser
```
The relevant columns to compare are as follows:
```{r}
#| label: mdcr-v-cmrb-elixhauser-column-map
```
If the columns are identical, we remove them from the `mdcr_v_cmrb_elix` object.
```{r}
#| label: column-by-column-mdcr-v-cmrb-elix
```
```{r}
#| label: verify-only-remaining-mdcr-v-cmrb-columns-elixhauser
#| include: false
```
The only columns remaining in the `mdcr_v_cmrb_elix` object are id columns and
columns unique to the return from `medicalcoder::comorbidities()`:
* `num_cmrb`: the number of flagged comorbidities,
* `cmrb_flag`: a comorbidity flag (at least one comorbidity),
* `HTN_C`: any hypertension,
* `mortality_index` and `readmission_index`: based on the weights from AHRQ.
```{r}
#| label: remaining-mdcr_v_cmrb_elix
```

## PCCC

### `medicalcoder::comorbidities()` vs `pccc:::ccc()`
As we have seen in other cases, the absence of ICD codes for a `hadm_id` results
in the omission of the record from the return from `pccc::ccc()`.
```{r}
#| label: nrow-medicalcoder-vs-pccc
```
For the common `subject_id`/`hadm_id` pairs we compare the flags.
```{r}
#| label: define-mdcr_v_pccc
```
```{r}
#| label: mvp-columns
```
```{r}
#| label: check-identical-columns-mdcr-v-pccc
```
There are some differences in the results between the two approaches.
```{r}
#| label: structre-remaining-mdcr-v-pccc
```

The `misc` column is the "miscellaneous" category reported by
`medicalcoder::comorbidities()` and is not reported by `pccc::ccc()`.  The
existence of the `misc` column and some differences in the returned results
between `pccc::ccc()` version 1.0.7, and `medicalcoder::comorbidities()` is due
to how medicalcoder is implemented.

The differences in any technology dependence and transplant are due to design
difference between _medicalcoder_ and _pccc_.  _medicalcoder_ uses database
style joins to map precomputed ICD-codes to conditions.  This allows for a
simple extension of mapping ICD-codes to both the primary condition and
subcondition as the latter require only an additional column in a lookup table.
The _pccc_ uses partial string matching to map ICD codes to conditions for
each of the primary conditions and only the two subconditions of technology
dependence and transplant.  When building _medicalcoder_, the lookup table with
condition and subconditions was extended to match the publication documentation.

```{r}
#| label: any-tech-dep-and-transplant
```

Several GitHub issues have be created for _pccc_ to correct these issues for
specific ICD codes:

* [ICD-9 349.1](https://github.com/CUD2V/pccc/issues/45)
* [ICD-9 V56](https://github.com/CUD2V/pccc/issues/46)
* [ICD-10 Z49](https://github.com/CUD2V/pccc/issues/47)
* [ICD-9 86.06](https://github.com/CUD2V/pccc/issues/48)
* [ICD-9 V45.85](https://github.com/CUD2V/pccc/issues/49)
* [ICD-9 V53.3](https://github.com/CUD2V/pccc/issues/50)
* [ICD-9 V53.91](https://github.com/CUD2V/pccc/issues/51)
* [ICD-9 V65.46](https://github.com/CUD2V/pccc/issues/52)
* [ICD-9 37.52](https://github.com/CUD2V/pccc/issues/53)
* [ICD-9 V42.0](https://github.com/CUD2V/pccc/issues/54)
* [ICD-10 Z94](https://github.com/CUD2V/pccc/issues/55)

## Summary of _medicalcoder_ vs Other Implementations

Using the MIMIC-IV data set, we compared encounter-level _medicalcoder_ results
with comparator implementations for shared algorithms and compatible inputs.
The comparison focuses on whether outputs agree after harmonizing structural
differences in returned objects and severity conventions.

| _medicalcoder_ method   | Comparator    | Agreement summary                               | Main reason for discordance                |
|:------------------------|:--------------|:------------------------------------------------|:-------------------------------------------|
| `charlson_mimicivcode`  | MIMIC-IV Code | Exact match after harmonizing severity handling | Structural output/severity handling        |
| `charlson_quan2011`     | _comorbidity_ | Exact match for shared flags                    | Regex/prefix matching vs valid-code lookup |
| `elixhauser_quan2005`   | _comorbidity_ | Exact match for shared flags                    | Extra medicalcoder summary/index fields    |
| `pccc_v2.0`             | _pccc_        | High agreement, with documented differences     | Technology dependence/transplant mapping   |
: Summary of encounter-level agreement between _medicalcoder_ and comparator implementations using MIMIC-IV v3.1 data. {#tbl-validation-summary}


# Precomputed vs Regular Expressions

The default method in _medicalcoder_ for mapping ICD codes to comorbidities is
to use a precomputed sets of known ICD codes and use database style joins to and
lookup tables.   _comorbidity_, and many other utilities, use regular
expressions or partial string matching to make the maps. This can lead to
problems.  For example, the strings "C189N" and "C189NOTACODE" are not a valid
ICD code, and will be flagged as a cancer comorbidity via
`comorbidity::comorbidity()` but will not be flagged by
`medicalcoder::comorbidities().`
```{r}
#| label: precomputed-v-regex-canc
```

There are cases where a regular expressions could be useful.  Consider the
ICD-10-GM (German version of ICD-10).  As of _medicalcoder_ version
`r packageVersion("medicalcoder")`, the ICD-10-GM codes are not part of the
internal ICD database.

There are codes in ICD-10-GM which are not in the CDC, CMS, WHO, IHACPA, or
Socialstyrelsen systems which are part of the _medicalcoder_ ICD database.  For
example, ICD-10-GM [E11.74](https://gesund.bund.de/en/icd-code-suche/e11-74):
Type 2 diabetes mellitus With multiple complications With diabetic foot
syndrome, controlled, and
[E11.75](https://gesund.bund.de/en/icd-code-suche/e11-75): Type 2 diabetes
mellitus With multiple complications With diabetic foot syndrome, uncontrolled.
These two codes are extensions under the E11 category; subcategory E11.7 which is
listed in @quan2005 as diabetes with complications for Charlson and Elixhauser.

```{r}
#| label: precomputed-v-regex-e117
```

Thus, if you have a data-set with ICD-10-GM codes, you should expect false
negative flags for diabetes with complications when using the default settings
in `medicalcoder::comorbidities()`.  In the following example we have E11.72,
E11.73, E11.74, and E11.75.  E11.72 and E11.73 should map to complex diabetes
since those codes overlap with the IHACPA codes.

```{r}
#| label: precomputed-v-regex-false-negative
```

If, however, you use `mapping = "regex"` in the call to
`medicalcoder::comorbidities()` (new in v0.9.0), then these conditions will be
flagged, at least for Charlson, the Elixhauser regex has not yet been
implemented.

```{r}
#| label: precomputed-v-regex-with-regex
#| error: true
```

The current regular expressions are avaiable to end users via
```{r}
#| label: precomputed-v-regex-patterns
```
The regular expression patterns have been written to match both full and compact
ICD codes.

Revisiting the first example in this section, we can see that using `mapping =
"regex"` in `medicalcoder::comorbidities()` will give the same false-positives
as `comorbidity::comorbidity()` does.   The utility exists, but it should be
used with care.

```{r}
#| label: precomputed-v-regex-canc-part2
```


# Longitudinal Flagging

Under the assumption that all relevant ICD codes are (re)reported on each
encounter then there is no need to carry-forward comorbidity flags.  However,
that assumption is rarely valid. _medicalcoder_ provides a utility for
indefinite carry-forward of a condition. As reported in the primary manuscript
this includes accounting for conditions which should not be reported on a
specific encounter but are reported on the following encounters. In the
following we will explore some of the pros and cons of the indefinite
carry-forward of comorbidities.

To simplify the work we will reapply the comorbidity algorithms using `enc_seq`
instead of `hadm_id` so that the encounters are considered in chronologic order.
We present here the code for Charlson, Elixhauser, and PCCC, which will be
referenced several times within the remainder of this document.

```{r}
#| label: medicalcoder-charlson-current
#| cache: true
```
```{r}
#| label: medicalcoder-charlson-cumulative
#| cache: true
```
```{r}
#| label: medicalcoder-elixhauser-current
#| cache: true
```
```{r}
#| label: medicalcoder-elixhauser-cumulative
#| cache: true
```
```{r}
#| label: medicalcoder-pcccv2.0-current
#| cache: true
```
```{r}
#| label: medicalcoder-pcccv2.0-cumulative
#| cache: true
```
```{r}
#| label: medicalcoder-pcccv3.1-current
#| cache: true
```
```{r}
#| label: medicalcoder-pcccv3.1-cumulative
#| cache: true
```

## Severity of Disease

Let's look at the impact of longitudinal flagging vs encounter level flagging of
Charlson comorbidities.

```{r}
#| label: build-flagmethod-delta
#| include: false
```

@fig-flagmethod-delta-charlson shows the number of encounters flagged with a
given condition for `flag.method = 'current'` and `flag.method = 'cumulative'`
for the Charlson condition flagged by `medicalcoder::comorbidities()`.  For most
conditions the number of encounters flagged is greater when using the cumulative
flag method compared to the the current flag method.  However, there is a
notable counter example: dm (diabetes without complications) and dmc (diabetes
with complications).  _medicalcoder_ will carry forward the severity of disease
an report the more severe version after it has been reported.

```{r}
#| label: define-flagmethod-plot
#| include: false
```
```{r}
#| label: fig-flagmethod-delta-charlson
#| fig-scap: Charlson comorbidities by flag.method.
#| fig-cap: |
#|   Number of encounters flagged with a Charlson comorbidities based on
#|   encounter only (`flag.method = "current"`) verses the whole patient history
#|   (`flag.method = "cumulative"`).
#| fig-width: 7
#| fig-height: 7
#| echo: false
```

```{r}
#| label: build-dmseverity
#| include: false
```
```{r}
#| label: s10009326
#| include: false
```

An example of a patient record where on an encounter level flagging approach
diabetes will be missed entirely and the severity is inaccurate is in
@tbl-dm-s10009326.  Subject 10009326 had `r nrow(s10009326)` encounters.  At
least one ICD code mapping to uncomplicated diabetes under the Charlson _a la_
@quan2005 appeared in `r sum(s10009326$dm_current)` encounters, and
`r sum(s10009326$dmc_current)` encounter for complicated diabetes.  Under an
encounter-level only flagging of comorbidities, this patient would erroneously
not have diabetes of any kind on
`r nrow(s10009326) - sum(s10009326$dm_current) - sum(s10009326$dmc_current)`
encounters.  Using the longitudinal flagging in `medicalcoder::comorbidities()`
via `flag.method = 'cumulative'`, we have diabetes flagged on all encounters,
and the severity correctly marked for encounters later in the patient record.

```{r}
#| label: tbl-dm-s10009326
#| tbl-cap: |
#|   Flagging Diabetes under Charlson comorbidities for MIMIC-IV subject_id
#|   10009326 using `medicalcoder::comorbidities()` `flag.method = "current"`
#|   and `flag.method = "cumulative"`.
#| results: "asis"
#| echo: false
```

### Elixhauser
For Elixhauser, @fig-flagmethod-delta-elixhauser, shows similar behavior for
diabetes as Charlson.
```{r}
#| label: fig-flagmethod-delta-elixhauser
#| fig-scap: Elixhauser comorbidities by flag.method.
#| fig-cap: |
#|    Number of encounters flagged with a Elixhauser comorbidities based on
#|    encounter only (`flag.method = current`) verses the whole patient history
#|    (`flag.method = "cumulative"`).
#| fig-width: 7
#| fig-height: 7
#| echo: false
```

### PCCC

@fig-flagmethod-delta-pcccv2.0 and @fig-flagmethod-delta-pcccv3.1 illustrate
the impact of the flagging method when identifying PCCC comorbidities under
PCCC v2 and PCCC v3 respectively.  @fig-flagmethod-delta-pcccv3.1 is of
particular interest as the implications of technology-dependent codes in PCCC v3
are handled have considerable impact on the flagged comorbidities.

```{r}
#| label: fig-flagmethod-delta-pcccv2.0
#| fig-scap: PCCC v2 comorbidities by flag.method.
#| fig-cap: |
#|    Number of encounters flagged with a PCCC v2 comorbidities based on
#|    encounter only (`flag.method = current`) verses the whole patient history
#|    (`flag.method = "cumulative"`).
#| fig-width: 7
#| fig-height: 7
#| echo: false
```

```{r}
#| label: fig-flagmethod-delta-pcccv3.1
#| fig-scap: PCCC v3 comorbidities by flag.method.
#| fig-cap: |
#|    Number of encounters flagged with a PCCC v3 comorbidities based on
#|    encounter only (`flag.method = current`) verses the whole patient history
#|    (`flag.method = "cumulative"`).
#| fig-width: 7
#| fig-height: 9
#| echo: false
```

```{r}
#| label: s10728333
#| include: false
```

Let's look at the records for `subject_id = 10728333`.  @tbl-s10728333 shows the
flagging of metabolic, miscellaneous, and respiratory conditions for PCCC v3
using `flag.method = 'current'` or `flag.method = 'cumulative'`.  For this
patient, on encounter 1, a technology-dependent ICD code related to a
respiratory comorbidity was observed in the patient record.  No other ICD codes
related to a PCCC v3 comorbidity were observed for encounter 1.  As such, no
comorbidities are flagged for encounter 1 as at least one
non-technology-dependent ICD code is required for any comorbidity to be flagged.
Encounter 2 had no PCCC v3 related ICD codes reported.  Encounter 3 had a
technology-dependent ICD code for a miscellaneous comorbidity reported and
again, because no non-technology-dependent ICD was reported no comorbidity is
flagged.  Starting on encounter 4, a non-technology-dependent ICD code for a
metabolic comorbidity was reported.  Under the `flag.method = 'current'` that is
the only code that is reported and thus only one comorbidity is reported.  Under
`flag.method = 'cumulative'` the technology-dependent ICD codes from encounters
1 and 3 have been carried forward in the record to encounter 4 and are thus
flagged.  If the end user were to try to build the cumulative comorbidities from
the results of the `flag.method = 'current'` results, the end user would never
capture the miscellaneous comorbidity, and would fail to capture the respiratory
comorbidity until encounter 6, and would incorrectly classify as
non-technology-dependent until the final encounter.

```{r}
#| label: tbl-s10728333
#| echo: false
#| results: "asis"
```

## Diabetes - Chronic vs Acute Disease

Non-gestational diabetes is generally considered a chronic condition.  Even when
well managed, it is still a relevant diagnosis.  The ICD-10 code defining
diabetes (with and without complications) for Charlson comorbidities as
defined by Quan et al. (2005, 2011), are all in the E10, E11, E12, E13, and E14
categories.

```{r}
#| label: ICD-10-categories-dm-quan
```
Let's look at the records for those who have at least two encounters and at
least one of the encounters was flagged with diabetes (with or without
complications).
```{r}
#| label: minimum-two-encounters-at-least-one-with-dm
```
How many cumulative flags are on encounters were DM i(with or without
complications) is flagged on both a preceding and a proceding encounter, or only
after a preceding encounter.
```{r}
#| label: cumulative-flags-between-and-after-current-flag
```
At a subject-level, the
```{r}
#| label: cumulative-flags-between-and-after-current-flag-subject-level
```

Longitudinal flagging of a chronic condition could be beneficial, especially in
the cases were we see the condition flagged on encounters preceding and
proceding a specific encounter.

End users are cuationed to understand the specific conditions and comorbidity
methods being used before applying the indefinite carry-forward.  For example,
gestational diabetes is part of the comorbidity mappings for the AHRQ Elixhauser
variants based on ICD-10 codes.  Note that ICD-10 category E08, E09, E10, E11,
E13, _and_ O24 are all diabetes codes with O24 codes mapping to uncomplicated
diabetes.
```{r}
#| label: ahrq-icd-10-dm-codes
```
In the MIMIC-IV data, there are subjects with both gestational and
non-gestational diabetes, subjects with only gestational diabetes, and subjects
with only non-gestational diabetes.
```{r}
#| label: gestational-and-non-gestational-dm
```
Since the AHRQ variant of Elixhauser comorbidities include both chronic and
non-chronic versions of diabetes `flag.method = "cumulative"` may results false
positives of diabetes.

## Metastatic Tumors

Metastatic tumors could be consider non-chronic conditions as the tumors could
be removed via surgery or treated with other therapies.  As such, the flags
reported via `flag.method = "cumulative"` for malignancy (Charlson and PCCC),
or solid tumors (Elixhauser) should not be interpreted as "present disease" but
as "a history of _observed_ disease."

We empathize _observed disease_ because there are ICD codes specifically for
personal history of malignancies.  We collect those ICD codes and the codes
mapping to malignancies per Quan et al. (2005, 2011) and subset the `mimicivDT`
to subjects with these codes.

```{r}
#| label: history-of-malignant
```
The next chunk will count the number of subjects with reported history of a
malignancy and those with an observed malignancy
```{r}
#| label: mal-subject-level-variables
```
What is notable in the above is that there are
`r with(subject_history, qwraps2::n_perc(subject_history_of == 1 & subject_observed_mal == 0))`
patients who have only a personal history of malignancy in the data set.  If the
cumulative flag was to be considered a "history of" these would patients would
be considered a false negative.  The, the `flag.method = 'cumulative'` can only
be interpreted as "current or history of observed disease."

## AHRQ (2022+) Example - POA

AHRQ Elixhauser for years 2022 and beyond use ICD-10 diagnostic codes and
require some ICD-codes to be present-on-admission for the comorbidity to be
flag.  The MIMIC data does not have a present-on-admission flag.  For this
example let's assume that all codes are not POA.

```{r}
#| label: ahrq-longitudinial-example
```

Let's look at subject_id 19997538.  This subject has ICD code K76.0 reported on
encounter 1.

```{r}
#| label: s19997538
```
K76.0 maps to mild liver disease and is required to be POA for the
condition to be flagged.
```{r}
#| label: K760
```
When `flag.method = "current"` we see what we expect, mild liver disease is not
flagged on any encounter.  When `flag.method = "cumulative"` we see that on
encounter 1 the condition is not flagged, as expected since it is not POA, but
it is flagged on encounters 2 and 3 as the patient history is accounted for.
```{r}
#| label: s19997538-current-v-cumulative
```

# Time Required To Compute

The times required (median and IQR) to apply the different comorbidity
algorithms to the MIMIC-IV data are shown in @tbl-time-to-compute.  To process
the `r qwraps2::frmt(nrow(mimicivDT))` total rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of `r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))`
unique hospital encounters, `medicalcoder::comorbidities()` is considerably
faster than all the alternatives.

For this work we will include several additional calls to
`medicalcoder::comorbidities()`.
```{r}
#| label: figure2
#| include: false
```
```{r}
#| label: save-figure2
#| include: false
```


The benchmarking is done using _microbenchmark_ [@r-microbenchmark] and calling
each expression ten times.
```{r}
#| label: benchmarking
#| cache: true
#| include: false
```
@fig-benchmark-summary shows the median time in seconds along with error bars
for the IQR for the ten iterations.  The reason for the drastic speed
improvement between _medicalcoder_ and other utilities is that _medicalcoder_
uses precomputed maps between known ICD codes and conditions whereas
_comorbidity_ uses regular expressions, and _pccc_ uses partial string
matching.  The MIMIC-IV code also uses partial string matching and lexicographical
between statements.  The numeric values represented in @fig-benchmark-summary
are reported in @tbl-time-to-compute.
```{r}
#| label: fig-benchmark-summary
#| echo: false
#| fig-cap: |
#|   Time (bar height = median, error bars = IQR) in seconds to apply several
#|   comorbidity  algorithms to the MIMIC-IV dataset.  Each method was evaluated
#|   10 times.
#| fig-width: 7
#| fig-height: 7
```
```{r}
#| label: tbl-time-to-compute
#| results: asis
#| echo: false
#| tbl-cap: |
#|   Time, in seconds, median and IQR, to flag comorbidities via Charlson,
#|   Elixhauser, PCCC v2, and PCCC v3 and several different utilities in the
#|   MIMIC-IV v3.1 dataset. The flagging method, current or cumulative is
#|   reported.  Each method was evaluated 10 times.
```

# Summarizing `medicalcoder_comorbidities` Objects

The _medicalcoder_ package provides an S3 summary method for the objects
returned by `medicalcoder::comorbidities()`.  The summaries differ slightly
depending of the comorbidity algorithm (`method`). Charlson and Elixhauser
return lists with multiple summaries and the PCCC return is a data.frame of
summaries.

When the `flag.method = 'current'`, the `conditions` element of the summary for
the Charlson and Elixhauser comorbidities is a data.frame with counts and
percentages of the distinct `id.vars` with a given comorbidity, and counts of
comorbidities.  For PCCC v2 and v3, the return is only this data.frame.   When
the flag method is cumulative, a warning is given stating that the summary may
not be meaningful as it is an encounter level summary.

```{r}
#| label: structure-of-summaries
```

The `data.frame`s are provided so that end users can easily build their own
summary tables using tools such as _knitr_
[@r-knitr,@r-knitr-collection,@r-knitr-book] and _kableExtra_
[@r-kableExtra]

```{r}
#| label: tbl-medicalcoder-pcccv2.0-summary
#| tbl-cap: |
#|   Code and resulting table for a summary of the PCCC v2 conditions flagged in
#|   the MIMIC-IV dataset.
#| echo: true
#| results: asis
```

# Session Info

@tbl-session-info-platform and @tbl-session-info-packages reports details on the
machine, R, and the R packages used to produce this document.

```{r}
#| label: mem-total-helper
#| include: false
```
```{r}
#| label: get-cpu-brand
#| include: false
```
```{r}
#| label: sessioninfo-with-memory-cpu
```
```{r}
#| label: tbl-session-info-platform
#| tbl-cap: "Details on the machine to generate this document."
#| echo: false
#| results: "asis"
```
```{r}
#| label: tbl-session-info-packages
#| tbl-cap: "R packages and versions used to generate this document."
#| echo: false
#| results: asis
```

# References
