---
title: "MIMIC-IV Data Analysis via medicalcoder"
subtitle: "Supplement to JAMIA Open Manuscript"

format:
  pdf:
    toc: true
    number-sections: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[table]{xcolor}

bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  cache = TRUE
)

options(knitr.kable.NA = "")

# Go Avs Go!
# {
#   "burgundy" : "#6F263D",
#   "blue" : "#236192",
#   "light-blue" : "#4F8FCB"
#   "silver" : "#A2AAAD",
#   "black" : "#000000"
# }

loadNamespace('medicalcoder')
loadNamespace('data.table')
```

# Objective and Setup

This supplement is used to showcase `medicalcoder::comorbidities()` on its own
and against alternative tools for applying Charlson, Elixhauser, and Pediatric
Complex Chronic Conditions (PCCC) to a data set.

# Materials and Methods

## Data

We use the MIMIC-IV v3.1 Data
[@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000] as the foundation for
these examples. To reproduced this work end users will need to have access to
the MIMIC-IV v3.1 data.

__NOTE:__
The MIMIC-IV data consists of only
adult patients.  The PCCC is intended for pediatrics and may not be of interest
in adult data.  However,
In this work we are focused on the use of the _medicalcoder_ package and so
we use adult data here to illustrate the use the package, not to make clinical
inference.

## Alternative Utilities

We compared the results from `medicalcoder::comorbidities()` to the Charlson
comorbidities flagged by the MIMIC-IV code, and the R package _comorbidity_
[@r-comorbidity].  For Elixhauser we will compare results between
`medicalcoder::comorbidities()` and `comorbidity::comorbidity()`.  Lastly, we
will compare PCCC version 2 results between `medicalcoder::comorbidities()` and
the R package _pccc_ [@r-pccc] call `pccc::ccc()`.

# Results

## MIMIC-IV Data Prep

We have stored the path to the MIMIC-IV data as a
system variable on our machines.  If you are going to reproduce this work either
set the same environment variable or overwrite the `MIMICIVDATA` variable
defined below.

```{r}
#| label: MIMICIVDATA
# System variable stores path to MIMIC-IV data
MIMICIVDATA <- Sys.getenv("MIMICIVDATA")
```

```{r}
#| label: verify mimiciv version
# verify that you are looking at version 3.1 of the MIMIC-IV data
stopifnot(
  identical(
    scan(
      file = file.path(MIMICIVDATA, "CHANGELOG.txt"),
      what = "character", sep = "\n", quiet = TRUE)[1],
    "This is the change log for MIMIC-IV v3.1."
  )
)
```

@tbl-mimicivdata-checksums gives the sha256 checksums for the files in the
MIMIC-IV data set.

```{r}
#| label: tbl-mimicivdata-checksums
#| tab-cap: "SHA256 Check sums for files in the MIMIC-IV v3.1 data."
#| echo: false
#| results: "asis"
data.table::fread(
  file = file.path(MIMICIVDATA, "SHA256SUMS.txt"),
  header = FALSE
) |>
kableExtra::kbl(
  x = _,
  col.names = c("sha256sum", "file"),
  format = "latex",
  #caption = "SHA256 Check sums for files in the MIMIC-IV v3.1 data.",
  booktabs = TRUE
  ) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down")
)
```

Each file checksum has been verified.

```{r}
#| label: verify-checksums
#| cache: true
system(
  sprintf("cd %s; sha256sum -c SHA256SUMS.txt", MIMICIVDATA),
  intern = TRUE
)
```

We will read in the MIMIC-IV hospital data and do some light formating to before
using `medicalcoder::comorbidities()` to flag comorbidities.  Start by importing
the `admissions`, `patients`, `diagnoses_icd`, and `procedures_icd`, datasets.

```{r}
#| label: import-mimiciv-data
mimicivdata <-
  list.files(path = file.path(MIMICIVDATA, "hosp"), full.names = TRUE)
names(mimicivdata) <- sub("\\.csv\\.gz$", "", basename(mimicivdata))

mimicivdata <-
  lapply(
    mimicivdata[c("admissions", "patients", "diagnoses_icd", "procedures_icd")],
    data.table::fread
  )
```

We will build an additional data set for age (in years).

```{r}
#| label: build-mimicvidata-ages
mimicivdata$ages <-
  merge(
    x = mimicivdata$admissions[, .(subject_id, hadm_id, admittime)],
    y = mimicivdata$patients[,   .(subject_id, anchor_age, anchor_year)],
    all = FALSE,
    by = "subject_id"
  )
mimicivdata$ages[
  ,
  anchor_year := as.POSIXct(paste0(anchor_year, "-01-01"), format = "%Y-%m-%d")
]
mimicivdata$ages[
  ,
  age :=
    anchor_age +
    as.numeric(difftime(admittime, anchor_year, units = "days")) / 365.25
]
```

For the calls to `medicalcoder::comorbidities()` we can have all the ICD codes
in one data.frame (data.table will be used here).
```{r}
#| label: build-mimicivdata-icd
mimicivdata$icd <-
  data.table::rbindlist(
    list(dx = mimicivdata$diagnoses_icd, pr = mimicivdata$procedures_icd),
    idcol = "dx",
    use.names = TRUE,
    fill = TRUE
  )

# set the diagnoses indicator to an integer as expected for input to
# medicalcoder::comorbidities()
mimicivdata$icd[, dx := as.integer(dx == "dx")]
```

We put all the parts together into one data.frame (data.table)

```{r}
#| label: one-data-set
mimicivDT <-
  merge(
    x = mimicivdata$patients[, .(subject_id)],
    y = mimicivdata$ages[, .(subject_id, hadm_id, age)],
    all = TRUE,
    by = c("subject_id")
  )

mimicivDT <-
  merge(
    x = mimicivdata$icd,
    y = mimicivDT,
    all = TRUE,
    by = c("subject_id", "hadm_id")
  )

mimicivDT <-
  merge(
    x = mimicivDT,
    y = mimicivdata$admissions[, .(subject_id, hadm_id, admittime)],
    all = TRUE,
    by = c("subject_id", "hadm_id")
  )

# set keys and order the data by subject id, admittime, hadm_id
data.table::setkey(mimicivDT, subject_id, admittime, hadm_id)

# set an encounter sequence variable. This works because the data has been
# sorted by subject_id, admittime
mimicivDT[, enc_seq := cumsum(!duplicated(hadm_id)), by = .(subject_id)]

# For some of the methods, only one id variable is allowed so we create a
# composite id:
mimicivDT[, cid := paste(subject_id, enc_seq, sep = "__")]
```

Data summary:

```{r}
#| label: data-summary
str(mimicivDT)
mimicivDT
```

There are `r qwraps2::frmt(nrow(mimicivDT))` total rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique hospital encounters.

## Flagging comorbidities

In the following we will outline how to apply the comorbidities methods to the
`mimicivDT` data set.  We will define some wrapper functions to simpify calls.
We will return the results and time how long it takes to apply the methods.  We
will flag the following comorbidities:

* Charlson _a la_ @quan2005
* Elixhauser _a la_ @quan2005
* PCCC v2 _a la_ @feudtner2014pediatric
* PCCC v3 _a la_ @feinstein2024pediatric

For each of these comorbidity methods we will flag on a per-encounter basis and
a longitudinal basis within a patient record.

### `medicalcoder::comorbidities()`

```{r}
#| label: medicalcoder-warmup
#| include: false
# to give the best possible time for medicalcoder, it will be helpful to run the
# code once just to make sure all the needed internal data sets have been built
medicalcoder::comorbidities(
  data = mimicivDT[1:10],
  id.vars = c("subject_id", "hadm_id"),
  icd.codes = "icd_code",
  icdv.var = "icd_version",
  dx.var = "dx",
  age.var = "age",
  method = "charlson_quan2005",
  full.codes = FALSE,
  flag.method = "current",
  primarydx = 0L,
  poa = 1L
)
```

The call to `medicalcoder::comorbidities()` has been designed to make calling
multiple comorbidity methods and multiple flagging methods easy.  For this use
case it would be programatically preferable to use a simple wrapper about
`medicalcoder::comorbidities()` to simplify the forthcoming calls.  However, we
opt to repeat ourselves in the following calls for clarity and ease of selective
reproduction by the reader.

#### Charlson

```{r}
#| label: medicalcoder-charlson-current
tic <- Sys.time()

medicalcoder_charlson_current <-
  medicalcoder::comorbidities(
    data        = mimicivDT,  # object inheriting data.frame
    icd.codes   = "icd_code", # character string name of icd codes in data
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,    # consider all codes present on admission
    age.var     = "age", # in years, only relevant to charlson
    primarydx   = 0L,    # consider all diagnosis codes secondary diagnoses.
    method      = "charlson_quan2005",
    flag.method = "current" # default
  )

toc <- Sys.time()
attr(medicalcoder_charlson_current, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

The one difference to get longitudinal results within a subject is to change the
`flag.method = 'current'` to `flag.method = 'cumulative'`.


```{r}
#| label: medicalcoder-charlson-cumulative
tic <- Sys.time()

medicalcoder_charlson_cumulative <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    age.var     = "age",
    primarydx   = 0L,
    method      = "charlson_quan2005",
    flag.method = "cumulative"
  )

toc <- Sys.time()
attr(medicalcoder_charlson_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

#### Elixhauser

Applying Elixhauser requires a only a change to the `method` argument.

```{r}
#| label: medicalcoder-elixhauser-current
tic <- Sys.time()

medicalcoder_elixhauser_current <-
  medicalcoder::comorbidities(
    data        = mimicivDT,  # object inheriting data.frame
    icd.codes   = "icd_code", # character string name of icd codes in data
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,    # consider all codes present on admission
    age.var     = "age", # in years, only relevant to elixhauser
    primarydx   = 0L,    # consider all diagnosis codes secondary diagnoses.
    method      = "elixhauser_quan2005",
    flag.method = "current" # default
  )

toc <- Sys.time()
attr(medicalcoder_elixhauser_current, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

```{r}
#| label: medicalcoder-elixhauser-cumulative
tic <- Sys.time()

medicalcoder_elixhauser_cumulative <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    age.var     = "age",
    primarydx   = 0L,
    method      = "elixhauser_quan2005",
    flag.method = "cumulative"
  )

toc <- Sys.time()
attr(medicalcoder_elixhauser_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

#### PCCC v2.0

```{r}
#| label: medicalcoder-pcccv2-current
tic <- Sys.time()

medicalcoder_pcccv2.0_current <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    method      = "pccc_v2.0",
    flag.method = "current"
  )

toc <- Sys.time()

attr(medicalcoder_pcccv2.0_current, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

```{r}
#| label: medicalcoder-pcccv2-cumulative
tic <- Sys.time()

medicalcoder_pcccv2.0_cumulative <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    age.var     = "age",
    method      = "pccc_v2.0",
    flag.method = "cumulative"
  )

toc <- Sys.time()

attr(medicalcoder_pcccv2.0_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

#### PCCC v3

```{r}
#| label: medicalcoder-pcccv3-current
tic <- Sys.time()

medicalcoder_pcccv3.1_current <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    method      = "pccc_v3.1",
    flag.method = "current"
  )

toc <- Sys.time()

attr(medicalcoder_pcccv3.1_current, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

```{r}
#| label: medicalcoder-pcccv3-cumulative
tic <- Sys.time()

medicalcoder_pcccv3.1_cumulative <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    age.var     = "age",
    method      = "pccc_v3.1",
    flag.method = "cumulative"
  )

toc <- Sys.time()

attr(medicalcoder_pcccv3.1_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs")
```


### MIMIC-IV Code

We use some of the SQL code for applying the Charlson
comorbidity algorithm to the MIMIC-IV data provided by the MIT Laboratory for
Computational Physilogy (MIT\_LCP) on
[GitHub](https://github.com/MIT-LCP/mimic-code.git).  We stored the
path to a local copy of the respository in the system environmental variable
`MIMICIVCODE`.


```{r}
#| label: MIMICIVCODE
# System variable stores path to MIMIC-IV code repo
MIMICIVCODE <- Sys.getenv("MIMICIVCODE")
```

The SQL code for applying the Charlson comorbidities and checksum:
```{r}
mimiciv_charlson_sql <-
  file.path(MIMICIVCODE, "concepts", "comorbidity", "charlson.sql")

# checksum
stopifnot(
  substr(
    system(
      sprintf("sha256sum %s", mimiciv_charlson_sql),
      intern = TRUE
    ),
  start = 1L,
  stop = 64L
  ) == "5b797b673ace4dcbc6f686848d24371382f49ac61fcc39101f09c3b42969d801"
)
```

The SQL code is expected to run on Google Big Query.  We make a few
small modifications to the code so we can evaluate the SQL locally via RSQLite.

```{r}
#| label: mimic-code-edits
mimiciv_charlson_sql <-
  scan(
    file = mimiciv_charlson_sql,
    what = "character",
    sep = "\n",
    quiet = TRUE
  )

# modify the query to work in SQLite
mimiciv_charlson_sql <-
  gsub(
    pattern = "physionet-data.mimiciv_hosp.admissions",
    replacement = "admissions",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <-
  gsub(
    pattern = "physionet-data.mimiciv_hosp.diagnoses_icd",
    replacement = "diagnoses",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <-
  gsub(
    pattern = "physionet-data.mimiciv_derived.age",
    replacement = "ages",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <-
  gsub(
    pattern = "GREATEST",
    replacement = "MAX",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <- paste(mimiciv_charlson_sql, collapse = "\n")
```

We set up an SQL database in memory and evaluate the code.

```{r}
#| label: mimic-iv-charlson-results
#| echo: true
#| results: "hide"
#| cache: true
con <- odbc::dbConnect(drv = RSQLite::SQLite(), dbname = ":memory:")

# add data to the data base
odbc::dbWriteTable(conn = con, name = "diagnoses",  value = mimicivdata$diagnoses)
odbc::dbWriteTable(conn = con, name = "admissions", value = mimicivdata$admissions)
odbc::dbWriteTable(conn = con, name = "patients",   value = mimicivdata$patients)
odbc::dbWriteTable(conn = con, name = "ages",       value = mimicivdata$ages)

# Run the query and record the time required to do so
tic <- Sys.time()

mimiciv_charlson_current <- odbc::dbGetQuery(con, mimiciv_charlson_sql)

toc <- Sys.time()

attr(mimiciv_charlson_current, "tictoc") <- difftime(toc, tic, units = "secs")

# close DB connection
odbc::dbDisconnect(conn = con)
```

### `comorbidity::comorbidity()`

The R package _comorbidity_ [@r-comorbidity] is a widely used and referenced
tool for applying the Charlson and Elixhauser comorbidities.
One consideration for
using `comorbidity::comorbidity()` is that ICD-9 or ICD-10 codes are used
independently. There are
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id)][V1 > 1, .N]`
subjects with both ICD-9 and ICD-10 codes in their records, and
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id, hadm_id)][V1 > 1, .N]`
encounters with both ICD versions.

To flag the conditions appropriately we must split the data by ICD version,
apply `comorbidity::comorbidity()` to each split, and then aggregate the
results.  Also, `comorbidity::comorbidity()` only takes one column for the id
and thus we need to build an composite id.  In the following work we will copy
the data built for the encounter level flagging and extend for cumulative
flagging as well.

```{r}
#| label: comorbidity-charlson
tic <- Sys.time()

comorbidity_charlson_icd9 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 9 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "charlson_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe
                   # comorbidities is also flagged
  )

comorbidity_charlson_icd10 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 10 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "charlson_icd10_quan",
    assign0 = TRUE
  )

comorbidity_charlson_current <-
  rbind(
    comorbidity_charlson_icd9,
    comorbidity_charlson_icd10
  )

# aggregate between the icd versions
comorbidity_charlson_current <-
  aggregate(. ~ cid, data = comorbidity_charlson_current, FUN = max)

# Note, Not all the cid are in comorbidity_charlson_current.  Especially for
# building the cumulative flags we need all the subjects and encounters
comorbidity_charlson_current <-
  merge(
    x = comorbidity_charlson_current,
    y = unique(mimicivDT[, .(cid)]),
    all = TRUE,
    by = "cid"
  )

# set NA values to zero
for (j in names(comorbidity_charlson_current)[-1]) {
  idx <- which(is.na(comorbidity_charlson_current[[j]]))
  comorbidity_charlson_current[[j]][idx] <- 0L
}

# apply the scoring, to use comorbidity::score the object needs to have certain
# attributes which were lost when aggregating over the icd versions.  Replace
# the attributes
attributes(comorbidity_charlson_current)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_charlson_icd9)[c("class", "variable.labels", "map")]

comorbidity_charlson_current[["score"]] <-
  comorbidity::score(
    x = comorbidity_charlson_current,
    weights = "quan",
    assign0 = TRUE
  )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
comorbidity_charlson_current[["score"]] <-
  as.integer(comorbidity_charlson_current[["score"]])

# expand the id columns and sort
data.table::setDT(comorbidity_charlson_current)
comorbidity_charlson_current[
  ,
  c("subject_id", "enc_seq") := data.table::tstrsplit(cid, "__")
  ][
  ,
  `:=`(subject_id = as.integer(subject_id), enc_seq = as.integer(enc_seq))
  ]
data.table::setkey(comorbidity_charlson_current, subject_id, enc_seq)

# end timing for encounter level flagging
toc <- Sys.time()

# For the cumulative flagging
tic1 <- Sys.time()

comorbidity_charlson_cumulative <- data.table::copy(comorbidity_charlson_current)

# next we use cumsum to flag conditions moving forward in time
cumflag <- function(x) as.integer(cumsum(x) > 0)

for (j in names(comorbidity_charlson_cumulative)) {
  if (!(j %in% c("subject_id", "enc_seq", "cid"))) {
    e <- substitute(
          comorbidity_charlson_cumulative[, J := cumflag(J), by = .(subject_id)],
          list(J = as.name(j))
          )
    eval(e)
  }
}

# clean up flags by severity
comorbidity_charlson_cumulative[diab == 1 & diabwc == 1, diab := 0L]
comorbidity_charlson_cumulative[mld == 1 & msld == 1, mld := 0L]
comorbidity_charlson_cumulative[canc == 1 & metacanc == 1, canc := 0L]

# apply the scoring, to use comorbidity::score the object needs to have certain
# attributes which were lost when aggregating over the icd versions.  Replace
# the attributes
data.table::setDF(comorbidity_charlson_cumulative)
attributes(comorbidity_charlson_cumulative)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_charlson_icd9)[c("class", "variable.labels", "map")]

comorbidity_charlson_cumulative[["score"]] <-
  comorbidity::score(
    x = comorbidity_charlson_cumulative,
    weights = "quan",
    assign0 = TRUE
  )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
comorbidity_charlson_cumulative[["score"]] <-
  as.integer(comorbidity_charlson_cumulative[["score"]])

data.table::setDT(comorbidity_charlson_cumulative)

toc1 <- Sys.time()

attr(comorbidity_charlson_current, "tictoc") <-
  difftime(toc, tic, units = "secs")

attr(comorbidity_charlson_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs") +
  difftime(toc1, tic1, units = "secs")
```

### Flagging Elixhauser via `comorbidity::comorbidity()`
```{r}
#| label: comorbidity-elixhauser
tic <- Sys.time()

comorbidity_elixhauser_icd9 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 9 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "elixhauser_icd9_quan",
    assign0 = TRUE
  )

comorbidity_elixhauser_icd10 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 10 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "elixhauser_icd10_quan",
    assign0 = TRUE
  )

comorbidity_elixhauser_current <-
  rbind(
    comorbidity_elixhauser_icd9,
    comorbidity_elixhauser_icd10
  )

# aggregate between the icd versions
comorbidity_elixhauser_current <-
  aggregate(. ~ cid, data = comorbidity_elixhauser_current, FUN = max)

# Note, Not all the cid are in comorbidity_elixhauser_current.  Especially for
# building the cumulative flags we need all the subjects and encounters
comorbidity_elixhauser_current <-
  merge(
    x = comorbidity_elixhauser_current,
    y = unique(mimicivDT[, .(cid)]),
    all = TRUE,
    by = "cid"
  )

# set NA values to zero
for (j in names(comorbidity_elixhauser_current)[-1]) {
  idx <- which(is.na(comorbidity_elixhauser_current[[j]]))
  comorbidity_elixhauser_current[[j]][idx] <- 0L
}

# expand the id columns and sort
data.table::setDT(comorbidity_elixhauser_current)
comorbidity_elixhauser_current[
  ,
  c("subject_id", "enc_seq") := data.table::tstrsplit(cid, "__")
  ][
  ,
  `:=`(subject_id = as.integer(subject_id), enc_seq = as.integer(enc_seq))
  ]
data.table::setkey(comorbidity_elixhauser_current, subject_id, enc_seq)

# end timing for encounter level flagging
toc <- Sys.time()

# For the cumulative flagging
tic1 <- Sys.time()

comorbidity_elixhauser_cumulative <- data.table::copy(comorbidity_elixhauser_current)

# next we use cumsum to flag conditions moving forward in time
cumflag <- function(x) as.integer(cumsum(x) > 0)

for (j in names(comorbidity_elixhauser_cumulative)) {
  if (!(j %in% c("subject_id", "enc_seq", "cid"))) {
    e <- substitute(
          comorbidity_elixhauser_cumulative[, J := cumflag(J), by = .(subject_id)],
          list(J = as.name(j))
          )
    eval(e)
  }
}

# clean up flags by severity?
comorbidity_elixhauser_cumulative[diabunc  == 1L & diabc    == 1L, diabunc  := 0L]
comorbidity_elixhauser_cumulative[hypunc   == 1L & hypc     == 1L, hypunc   := 0L]
comorbidity_elixhauser_cumulative[solidtum == 1L & metacanc == 1L, solidtum := 0L]

toc1 <- Sys.time()

attr(comorbidity_elixhauser_current, "tictoc") <-
  difftime(toc, tic, units = "secs")

attr(comorbidity_elixhauser_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs") +
  difftime(toc1, tic1, units = "secs")
```


### `pccc::ccc()`

The R package _pccc_ [@r-pccc] implements PCCC v2 [@feudtner2014pediatric].  As
with `comorbidity::comorbidity()` and many other implementations of
comorbidities, only one version of ICD is assessed at a time.

The expected input structure of the data for `pccc::ccc()` is a wide format, one
row per encounter, one column for each ICD code.

```{r}
#| label: pccc

mimicivDT[, dx_char := data.table::fifelse(dx == 1, "DX", "PR")]

mimicivDT_for_pccc <-
  split(mimicivDT, by = "icd_version") |>
  lapply(
    data.table::dcast,
    formula = cid ~ dx_char + seq_num,
    value.var = "icd_code"
    )

tic <- Sys.time()
pccc_icd9 <-
  pccc::ccc(
    data = mimicivDT_for_pccc[["9"]],
    id = cid,
    dx_cols = grep("^DX", names(mimicivDT_for_pccc[["9"]]), value = TRUE),
    pc_cols = grep("^PR", names(mimicivDT_for_pccc[["9"]]), value = TRUE),
    icdv = 9
  )
pccc_icd10 <-
  pccc::ccc(
    data = mimicivDT_for_pccc[["10"]],
    id = cid,
    dx_cols = grep("^DX", names(mimicivDT_for_pccc[["10"]]), value = TRUE),
    pc_cols = grep("^PR", names(mimicivDT_for_pccc[["10"]]), value = TRUE),
    icdv = 10
  )

pccc_pcccv2.0_current <- rbind(pccc_icd9, pccc_icd10)

# aggregate over ICD version
pccc_pcccv2.0_current <-
  aggregate(. ~ cid, data = pccc_pcccv2.0_current, FUN = max)

# Not all the cid are in this data set.  add them and set the missing values to
# zeros.
pccc_pcccv2.0_current <-
  merge(
    x = pccc_pcccv2.0_current,
    y = unique(mimicivDT[, .(cid)]),
    all = TRUE,
    by = "cid"
  )

for (j in names(pccc_pcccv2.0_current)[-1]) {
  idx <- which(is.na(pccc_pcccv2.0_current[[j]]))
  pccc_pcccv2.0_current[[j]][idx] <- 0L
}

toc <- Sys.time()

# cumflag
tic1 <- Sys.time()

data.table::setDT(pccc_pcccv2.0_current)
pccc_pcccv2.0_current[,
  c("subject_id", "enc_seq") := data.table::tstrsplit(cid, "__")
  ][
  ,
  `:=`(
    subject_id = as.integer(subject_id),
    enc_seq = as.integer(enc_seq)
  )]

data.table::setkey(pccc_pcccv2.0_current, subject_id, enc_seq)

pccc_pcccv2.0_cumulative <- data.table::copy(pccc_pcccv2.0_current)

for (j in names(pccc_pcccv2.0_cumulative)) {
  if (!(j %in% c("cid", "subject_id", "enc_seq"))) {
    e <- substitute(pccc_pcccv2.0_cumulative[, J := cumflag(J), by = .(subject_id)],
      list(J = as.name(j)))
    eval(e)
  }
}
toc1 <- Sys.time()

attr(pccc_pcccv2.0_current, "tictoc") <- difftime(toc, tic, units = "secs")
attr(pccc_pcccv2.0_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs") +
  difftime(toc1, tic1, units = "secs")
```

## Same Flags?

### `medicallcoder::comorbidities()` vs MIMIC-IV Code
Comparing the results between
_medicalcoder_ and MIMIC-IV code for Charlson comorbidities at the encounter
level should start by looking at the number or rows in the returned objects.

```{r}
#| lable: output-dim
unique(mimicivDT[, .(subject_id, enc_seq)])
dim(medicalcoder_charlson_current)
dim(mimiciv_charlson_current)
```

`medicalcoder::comorbidities()` returned a row for each encounter whereas the
MIMIC-IV code only returned a row for encounters with at least one flag.

```{r}
#| label: build-mdcr-v-mimiciv
mdcr_v_mimiciv <-
  merge(
    x = merge(
          x = medicalcoder_charlson_current,
          y = unique(mimicivDT[, .(subject_id, enc_seq, hadm_id)]),
          all.x = TRUE,
          by = c("subject_id", "enc_seq")
        ),
    y = mimiciv_charlson_current,
    all = TRUE,
    by = c("subject_id", "hadm_id"),
    suffixes = c("_mdcr", "_mimiciv")
  )

mdcr_v_mimiciv[
  cmrb_flag == 1L, # at least one flagged comorbidity via medicalcoder
  !any(is.na(charlson_comorbidity_index))
]
```

For the rest of the comparisons we will ignore the encounters not flagged by the
MIMIC-IV code.

```{r}
#| label: remove-missing-values-from-mdcr-v-mimiciv
mdcr_v_mimiciv <- mdcr_v_mimiciv[!is.na(charlson_comorbidity_index)]
```

There are three comorbidities where there are different levels of severity.
`medicalcoder::comorbidities()` will set the less severe condition indicator to
0 when the more severe condition is flagged.  The MIMIC-IV code returns a flag
for both levels.  Both `medicalcoder::comorbidities()` and MIMIC-IV Code only
consider the more severe case in the index scoring.  For the comorbidities
diabetes, cancer, and liver disease we will set the less sever flag to 0 if the
more sever flag is present for the results from the MIMIC-IV Code.

```{r}
#| label: set-zeros-by-severity
mdcr_v_mimiciv[dm == 1L & dmc == 1L, .N == 0L] # medicalcoder
mdcr_v_mimiciv[diabetes_without_cc == 1L & diabetes_with_cc == 1L, .N > 0L] # MIMIC
mdcr_v_mimiciv[diabetes_without_cc == 1L & diabetes_with_cc == 1L, diabetes_without_cc := 0L]

mdcr_v_mimiciv[mal == 1L & mst == 1L, .N == 0L] # medicalcoder
mdcr_v_mimiciv[malignant_cancer == 1L & metastatic_solid_tumor == 1L, .N > 0L] # MIMIC
mdcr_v_mimiciv[malignant_cancer == 1L & metastatic_solid_tumor == 1L, malignant_cancer := 0L]

mdcr_v_mimiciv[mld == 1L & msld == 1L, .N == 0L] # medicalcoder
mdcr_v_mimiciv[mild_liver_disease == 1L & severe_liver_disease == 1L, .N > 0L] # MIMIC
mdcr_v_mimiciv[mild_liver_disease == 1L & severe_liver_disease == 1L, mild_liver_disease := 0L]
```

```{r}
#| label: mdcr-v-mimic-check-for-identical-columns
mvm_columns <-
  data.table::fread(text = "
  medicalcoder   | mimicivcode
  aidshiv        | aids
  cebvd          | cerebrovascular_disease
  chf            | congestive_heart_failure
  copd           | chronic_pulmonary_disease
  dem            | dementia
  dm             | diabetes_without_cc
  dmc            | diabetes_with_cc
  hp             | paraplegia
  mal            | malignant_cancer
  mi             | myocardial_infarct
  mld            | mild_liver_disease
  msld           | severe_liver_disease
  mst            | metastatic_solid_tumor
  pud            | peptic_ulcer_disease
  pvd            | peripheral_vascular_disease
  rhd            | rheumatic_disease
  rnd            | renal_disease
  age_score_mdcr | age_score_mimiciv
  cci            | charlson_comorbidity_index
  ")

for (i in seq_len(nrow(mvm_columns))) {
  j1 <- mvm_columns[["medicalcoder"]][i]
  j2 <- mvm_columns[["mimicivcode"]][i]
  e <- identical(mdcr_v_mimiciv[[j1]], mdcr_v_mimiciv[[j2]])
  if (e) {
    mdcr_v_mimiciv[[j1]] <- NULL
    mdcr_v_mimiciv[[j2]] <- NULL
  } else {
    print(sprintf("%s and %s are not identical", j1, j2))
  }
}
```

All that is left in `mdcr_v_mimiciv` are the id.vars and the
`num_cmrb`, and `cmrb_flag`.  These columns are from
`medicalcoder::comorbidities()` and report the number of comorbidities flagged
and indicator for any comorbidity.
```{r}
#| label: show-remainder-of-mdcr-v-mimicivcode
head(mdcr_v_mimiciv)
```

### `medicalcoder::comorbidities()` vs `comorbidity::comorbidity()`

#### Charlson
```{r}
#| label: build-mdcr-v-cmrb-charlson
mdcr_v_cmrb_charlson_current <-
  merge(
    x = medicalcoder_charlson_current,
    y = comorbidity_charlson_current,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_cmrb")
  )

mdcr_v_cmrb_charlson_cumulative <-
  merge(
    x = medicalcoder_charlson_cumulative,
    y = comorbidity_charlson_cumulative,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_cmrb")
  )
```

```{r}
#| label: mdcr-v-cmrb-charlson-column-tests
mvc_columns <-
  data.table::fread(text = "
  medicalcoder   | comorbidity
  aidshiv        | aids
  cebvd          | cevd
  chf_mdcr       | chf_cmrb
  hp_mdcr        | hp_cmrb
  copd           | cpd
  dem            | dementia
  dm             | diab
  dmc            | diabwc
  mal            | canc
  mi_mdcr        | mi_cmrb
  mld_mdcr       | mld_cmrb
  msld_mdcr      | msld_cmrb
  mst            | metacanc
  pud_mdcr       | pud_cmrb
  pvd_mdcr       | pvd_cmrb
  rhd            | rheumd
  rnd            | rend
  cci            | score
  ")

for (i in seq_len(nrow(mvc_columns))) {
  #jm <- "mal"; jc <- "canc"
  jm <- mvc_columns[["medicalcoder"]][i]
  jc <- mvc_columns[["comorbidity"]][i]
  e <- identical(mdcr_v_cmrb_charlson_current[[jm]], mdcr_v_cmrb_charlson_current[[jc]])
  if (e) {
    mdcr_v_cmrb_charlson_current[[jm]] <- NULL
    mdcr_v_cmrb_charlson_current[[jc]] <- NULL
  } else {
    print(sprintf("mdcr_v_cmrb_charlson_current: %s is not identical to %s", jm, jc))
  }
  e <- identical(mdcr_v_cmrb_charlson_cumulative[[jm]], mdcr_v_cmrb_charlson_cumulative[[jc]])
  if (e) {
    mdcr_v_cmrb_charlson_cumulative[[jm]] <- NULL
    mdcr_v_cmrb_charlson_cumulative[[jc]] <- NULL
  } else {
    print(sprintf("mdcr_v_cmrb_charlson_cumulative: %s is not identical to %s", jm, jc))
  }
}
```

There is a missmatch for malignant cancer.  In all cases
`medicalcoder::comorbidities()` flagged malignant cancer while
`comorbidity::comorbidity()` did not flag.
```{r}
#| label: mal-delta-mdcr-v-cmrb
mdcr_v_cmrb_charlson_current[mal != canc, all(mal > canc)]

DT <-
  mimicivDT[
    mdcr_v_cmrb_charlson_current[mal != canc, .(cid)],
    on = "cid"
  ][
    ,
    code_id := as.numeric(factor(icd_code))
  ]

medicalcoder::comorbidities(
  data = DT,
  id.var = c("icd_code", "code_id"),
  icd.codes = "icd_code",
  icdv.var = "icd_version",
  dx.var = "dx",
  poa = 1L,
  primarydx = 0L,
  method = "charlson_quan2005"
)[mal == 1, .(icd_code, code_id, mal)]

comorbidity::comorbidity(
  x = DT,
  id = "code_id",
  code = "icd_code",
  map = "charlson_icd10_quan",
  assign0 = TRUE
  ) |>
subset(subset = canc == 1, select = c("code_id", "canc"))

subset(
  medicalcoder::get_icd_codes(with.description = TRUE),
  code %in% c("C4A111", "C4A39", "C4A52", "C4A62", "C4A71", "C4A72", "C4A9")
  )
```

Table 2 of @quan2005 lists C45.x - C58.x as mapping to any malignancy.  We argue
that C4A is included and that these are flase negatives from
`comorbidity::comorbidity()`.

#### Elixhauser

```{r}
#| label: build-mdcr-v-cmrb-elixhauser
mdcr_v_cmrb_elixhauser_current <-
  merge(
    x = medicalcoder_elixhauser_current,
    y = comorbidity_elixhauser_current,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_cmrb")
  )

mdcr_v_cmrb_elixhauser_cumulative <-
  merge(
    x = medicalcoder_elixhauser_cumulative,
    y = comorbidity_elixhauser_cumulative,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_cmrb")
  )
```

```{r}
#| label: mdcr-v-cmrb-column-tests
mvc_columns <-
  data.table::fread(text = "
  medicalcoder        | comorbidity
  AIDS                | aids
  ALCOHOL             | alcohol
  ANEMDEF             | dane
  ARTH                | rheumd
  BLDLOSS             | blane
  CARDIAC_ARRHYTHMIAS | carit
  CHF                 | chf
  CHRNLUNG            | cpd
  COAG                | coag
  DEPRESS             | depre
  DM                  | diabunc
  DMCX                | diabc
  DRUG                | drug
  HTN_CX              | hypc
  HTN_UNCX            | hypunc
  HYPOTHY             | hypothy
  LIVER               | ld
  LYMPH               | lymph
  LYTES               | fed
  METS                | metacanc
  NEURO               | ond
  OBESE               | obes
  PARA                | para
  PERIVASC            | pvd
  PSYCH               | psycho
  PULMCIRC            | pcd
  RENLFAIL            | rf
  TUMOR               | solidtum
  ULCER               | pud
  VALVE               | valv
  WGHTLOSS            | wloss
  ")

for (i in seq_len(nrow(mvc_columns))) {
  jm <- mvc_columns[["medicalcoder"]][i]
  jc <- mvc_columns[["comorbidity"]][i]
  e <- identical(
    mdcr_v_cmrb_elixhauser_current[[jm]],
    mdcr_v_cmrb_elixhauser_current[[jc]]
  )
  if (e) {
    mdcr_v_cmrb_elixhauser_current[[jm]] <- NULL
    mdcr_v_cmrb_elixhauser_current[[jc]] <- NULL
  } else {
    print(sprintf("mdcr_v_cmrb_elixhauser_current: %s is not identical to %s", jm, jc))
  }
  e <- identical(
    mdcr_v_cmrb_elixhauser_cumulative[[jm]],
    mdcr_v_cmrb_elixhauser_cumulative[[jc]]
  )
  if (e) {
    mdcr_v_cmrb_elixhauser_cumulative[[jm]] <- NULL
    mdcr_v_cmrb_elixhauser_cumulative[[jc]] <- NULL
  } else {
    print(sprintf("mdcr_v_cmrb_elixhauser_cumulative: %s is not identical to %s", jm, jc))
  }
}
```

```{r}
#| label: whats-left-mdcr_v_cmrb_elixhauser
mdcr_v_cmrb_elixhauser_current
mdcr_v_cmrb_elixhauser_cumulative
```

The remaining columns in `mdcr_v_cmrb_elixhauser_current` and
`mdcr_v_cmrb_elixhauser_cumulative` are unique to
`medicalcoder::comorbidities()`. `HTN_C` is any hypertension.


```{r}
#| label: verify-htn-c
stopifnot(
  identical(
    medicalcoder_elixhauser_current[["HTN_C"]],
    as.integer(
      medicalcoder_elixhauser_current[["HTN_UNCX"]] |
        medicalcoder_elixhauser_current[["HTN_CX"]]
    )
  ),
  identical(
    medicalcoder_elixhauser_cumulative[["HTN_C"]],
    as.integer(
      medicalcoder_elixhauser_cumulative[["HTN_UNCX"]] |
        medicalcoder_elixhauser_cumulative[["HTN_CX"]]
    )
  )
)
```

The
`mortality_index` and `readmission_index` are based on the weights from
@ahrq_web in this case were `method = "elixhauser_quan2005"`.  For `method =
"elixhauser_ahrqXXXX"` other weights may be applied.  To see the specific
weights used by method call `medicalcoder::get_elixhauser_index_scores()`

```{r}
#| label: medicalcoder-get-elixhauser-index-scores
str(
  medicalcoder::get_elixhauser_index_scores()
)
```

### `medicalcoder::comorbidities()` vs `pccc::ccc()`

```{r}
#| label: build-mdcr-v-pccc
mdcr_v_pccc_current <-
  merge(
    x = medicalcoder_pcccv2.0_current,
    y = pccc_pcccv2.0_current,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_pccc")
  )

mdcr_v_pccc_cumulative <-
  merge(
    x = medicalcoder_pcccv2.0_cumulative,
    y = pccc_pcccv2.0_cumulative,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_pccc")
  )
```

The return from `pccc::ccc()` does not inlcude all the `subject_id` and
`enc_seq`.  The missing ones are all flagged as no comorbidities by
`medicalcoder::comorbidities()`.  Implicitly this is the same result.

```{r}
#| label: mdcr-v-pccc-na-zero
# verify that medicalcoder's flag for any comorbidity (cmrb_flag) is zero when
# the ccc_flag from pccc::ccc()
mdcr_v_pccc_current[is.na(ccc_flag), all(cmrb_flag == 0)]
mdcr_v_pccc_cumulative[is.na(ccc_flag), all(cmrb_flag == 0)]
```

```{r}
#| label: mdcr-v-pccc-colwise-checks
for (j in c(unique(medicalcoder::get_pccc_conditions()$condition), "tech_dep", "transplant", "ccc_flag")) {
  if (j %in% c("tech_dep", "transplant")) {
    jm <- sprintf("any_%s", j)
    jp <- j
  } else if (j == "ccc_flag") {
    jm <- "cmrb_flag"
    jp <- j
  } else {
    jm <- sprintf("%s_mdcr", j)
    jp <- sprintf("%s_pccc", j)
  }
  e1 <- identical(mdcr_v_pccc_current[[jm]], mdcr_v_pccc_current[[jp]])
  e2 <- identical(mdcr_v_pccc_cumulative[[jm]], mdcr_v_pccc_cumulative[[jp]])
  if (e1) {
    mdcr_v_pccc_current[[jm]] <- NULL
    mdcr_v_pccc_current[[jp]] <- NULL
  } else {
    print(sprintf("for mdcr_v_pccc_current: %s is not identical to %s", jm, jp))
  }
  if (e1) {
    mdcr_v_pccc_cumulative[[jm]] <- NULL
    mdcr_v_pccc_cumulative[[jp]] <- NULL
  } else {
    print(sprintf("for mdcr_v_pccc_cumulative: %s is not identical to %s", jm, jp))
  }
}
```

There are some differences between the results for technology dependence
(`tech_dep`) and transplant.  The underlying reason for these differences is
that the implementation of `medicalcoder::comorbidities()` allows for flagging
of subconditions.  Under this paradigm both technology dependence and transplant
are subconditions.  There are several ICD codes which had/have errors in _pccc_
version 1.0.7.

```{r}
#| label: mdcr-v-pccc-tech-and-trans
DT <-
  mimicivDT[
    mdcr_v_pccc_current[any_tech_dep != tech_dep | any_transplant != transplant],
    on = c("cid", "subject_id", "enc_seq")
    ][, .(icd_code, icd_version, dx)
    ][,
    code_id := paste0("ICD-", icd_version, "-", ifelse(dx == 1, "CM", "PCS"), " ", icd_code)
    ][,
    dx_char := data.table::fifelse(dx == 1, "DX", "PR")
    ]

DT <- unique(DT)

mdcr <-
  medicalcoder::comorbidities(
    data = DT,
    id.var = "code_id",
    icd.codes = "icd_code",
    icdv.var = "icd_version",
    dx.var = "dx",
    poa = 1L,
    method = 'pccc_v2.0'
  )

DT_for_pccc <-
  split(DT, by = "icd_version") |>
  lapply(
    data.table::dcast,
    formula = code_id ~ dx_char,
    value.var = "icd_code"
    )

pccc <-
  rbind(
    pccc::ccc(
      data = DT_for_pccc[["9"]],
      id = code_id,
      dx_cols = grep("^DX", names(DT_for_pccc[["9"]]), value = TRUE),
      pc_cols = grep("^PR", names(DT_for_pccc[["9"]]), value = TRUE),
      icdv = 9
    ),
    pccc::ccc(
      data = DT_for_pccc[["10"]],
      id = code_id,
      dx_cols = grep("^DX", names(DT_for_pccc[["10"]]), value = TRUE),
      pc_cols = grep("^PR", names(DT_for_pccc[["10"]]), value = TRUE),
      icdv = 10
    )
  )

delta <-
  merge(mdcr, pccc, all = TRUE, by = "code_id")

delta <-
  delta[
    any_tech_dep != tech_dep | any_transplant != transplant,
    .(code_id, any_tech_dep, any_transplant, tech_dep, transplant)
  ]

delta[, .N, keyby = .(any_tech_dep, tech_dep, any_transplant, transplant)]

data.table::setkey(delta, code_id)

print(delta, nrow = Inf)
```

These are known and reported issues:

* ICD-10-CM Z94x
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/56
  * Specifically Z94.1, Z94.2, Z94.4, Z94.81, Z94.82, Z94.83, Z94.84 All
    documented as subcondition transplant but are missing from the transplant
    set in the pccc_1.0.7/src/pccc.cpp

* ICD-9-CM 3491
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/45
  * Documented and implemented for neuromusc in `pccc::ccc()` tech_dep is
    missing

* ICD-9-CM V420
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/54
  * Documented as renal (transplant); missing from the transplant set in the
    pccc_1.0.7/src/pccc.cpp

* ICD-9-CM V4585
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/49
  * Documented for version 2 as metabolic (devices) but is only listed under
    metabolic in the software.  Because of how `medicalcoder::comorbidities()`
    is implemented the subconditon (devices) had to be provided in the tables
    for `method = "pccc_v2.0"`.  This results in the (improved) mapping to
    technology dependence.  Perhaps more concerning is that the mapping for
    V45.85 in `pccc::ccc()` maps to a transplant, which is incorrect.  GitHub
    issue and pull request have been submitted to resolve this issue.

* ICD-9-CM V5391
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/51
  * In `pccc::ccc()` this codes maps to transplant when it should map to
    technology dependence.

* ICD-9-CM V56.x
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/46
  * Same as ICD-9-CM 349.1

* ICD-9-CM V6546
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/52
  * Missing tech dependence flag from `pccc::ccc()`

* ICD-9-PCS 8606
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/48
  * Missing tech dependence flag from `pccc::ccc()`

We remove the transplant and technology dependence columns from the
`mdcr_v_pccc_current` and `mdcr_v_pccc_cumulative` objects.
```{r}
#| label: remove-tech-trans-from-mdcr-v-pccc
mdcr_v_pccc_current[, any_transplant := NULL]
mdcr_v_pccc_current[, any_tech_dep := NULL]
mdcr_v_pccc_current[, transplant := NULL]
mdcr_v_pccc_current[, tech_dep := NULL]
mdcr_v_pccc_cumulative[, any_transplant := NULL]
mdcr_v_pccc_cumulative[, any_tech_dep := NULL]
mdcr_v_pccc_cumulative[, transplant := NULL]
mdcr_v_pccc_cumulative[, tech_dep := NULL]
```

The only remaining columns are the id columns and two columns from
`medicalcoder::comorbidities()`: `misc`, indicating a technology dependence or
transplant not classified as one of the other comorbidities, and `num_cmrb`, the
number of comorbidities flagged.

```{r}
#| label: whats-left-mdcr-v-pccc
mdcr_v_pccc_current
mdcr_v_pccc_cumulative
```

## Current vs Cumulative Flagging

```{r}
#| label: build-list-of-objects
#| include: false
objs <-
  c(
    "mimiciv_charlson_current",
    "comorbidity_charlson_current",
    "comorbidity_charlson_cumulative",
    "comorbidity_elixhauser_current",
    "comorbidity_elixhauser_cumulative",
    "medicalcoder_charlson_current",
    "medicalcoder_charlson_cumulative",
    "medicalcoder_elixhauser_current",
    "medicalcoder_elixhauser_cumulative",
    "medicalcoder_pcccv2.0_current",
    "medicalcoder_pcccv2.0_cumulative",
    "medicalcoder_pcccv3.1_current",
    "medicalcoder_pcccv3.1_cumulative",
    "pccc_pcccv2.0_current",
    "pccc_pcccv2.0_cumulative"
  ) |>
  sapply(get)
```

Let's look at the impact of longitudinal flagging vs encounter level flagging of
Charlson comorbidities.

```{r}
#| label: build-flagmethod-delta
#| include: false
flagmethod_delta <-
  data.table::rbindlist(
    objs[grepl("^medicalcoder", names(objs))],
    use.names = TRUE,
    fill = TRUE,
    idcol = "obj"
  )

flagmethod_delta[
  ,
  c("algo", "flag.method") := data.table::tstrsplit(obj, "_", keep = 2:3)
]

flagmethod_delta[, obj := NULL]

flagmethod_delta <-
  data.table::melt(
    data = flagmethod_delta,
    id.vars = c("subject_id", "enc_seq", "algo", "flag.method"),
    variable.factor = FALSE,
    variable.name = "condition",
    na.rm = TRUE
  )

# subset to the conditions and add labels
#flagmethod_delta <-
#  merge(
#    x = flagmethod_delta,
#    y = medicalcoder::get_charlson_index_scores()[c("condition", "condition_description")],
#    by = "condition"
#  )

flagmethod_delta <-
  flagmethod_delta[
   , .(N = sum(value), p = mean(value)),
   by = .(algo, flag.method, condition)
 ]

flagmethod_delta[,
  flag.method := factor(flag.method, levels = c("current", "cumulative"))
]

flagmethod_delta <-
  flagmethod_delta[
    !(condition %in% c("readmission_index", "mortality_index",
        "cci", "num_cmrb", "age_score"))
  ]

flagmethod_delta[condition == "cmrb_flag"]
```

@fig-flagmethod-delta shows the number of encounters flagged with a given
condition for `flag.method = 'current'` and `flag.method = 'cumulative'` for the
Charlson condition flagged by `medicalcoder::comorbidities()`.

```{r}
#| label: fig-flagmethod-delta
#| echo: false
g <- function(data) {
  ggplot2::ggplot(data = data) +
    ggplot2::theme_bw() +
    ggplot2::aes(x = condition, y = N, fill = flag.method) +
    ggplot2::geom_col(position = ggplot2::position_dodge()) +
    ggplot2::scale_fill_manual(
      name = "Flag Method",
      values = c("current" = "#6F263D", "cumulative" = "#4F8FCB")
    ) +
    ggplot2::scale_y_continuous(
      name = "Encounters",
      label = scales::label_comma()
    ) +
    ggplot2::theme(
      axis.title.y = ggplot2::element_blank(),
      legend.position = "bottom",
    ) +
    ggplot2::facet_wrap( ~ algo, scales = "free_y") +
    ggplot2::coord_flip()
}
g(flagmethod_delta[algo == "charlson"])
g(flagmethod_delta[algo == "exlihauser"])
g(flagmethod_delta[algo == "pcccv2.0"])
g(flagmethod_delta[algo == "pcccv3.1"])
```

Diabetes is the most interesting, as we account for a patient history, the
number of encounters flagged as having uncomplicated diabetes decreases and the
number of encounters flagged as having complicated diabetes increases.

```{r}
#| label: build-dm
#| include: false
dm <-
  merge(
    x = medicalcoder_charlson_current,
    y = medicalcoder_charlson_cumulative,
    by = c("subject_id", "enc_seq"),
    suffix = c("_current", "_cumulative")
  )
# find a record with a change from uncomplicated to complicated dm

# sanity checks
dm[dm_current  == 1, all(dmc_current == 0)]
dm[dmc_current == 1, all(dm_current == 0)]
```

```{r}
#| label: find-dm-patient-example
#| include: false
dm[,
  .(
    V1 = mean(dm_current + dmc_current),
    V2 = mean(dm_cumulative + dmc_cumulative),
    .N
   ),
 by = .(subject_id)
 ][, d := V2 - V1][d > 0 & N > 5]

s10009326 <-
  dm[
    subject_id == 10009326,
    .SD,
    .SDcols = patterns("enc_seq|^dm")
  ]
```

An example of a patient record where on an encounter level flagging approach
diabetes will be missed entirely and the severity is inaccurate is in
@tbl-dm-10009326.  Subject 10009326 had `r nrow(s10009326)` encounters.  At
least one ICD code mapping to uncomplicated diabetes under the Charlson _a la_
@quan2005 appeared in `r sum(s10009326$dm_current)` encounters, and
`r sum(s10009326$dmc_current)` encounter for complicated diabetes.  Under an
encounter-level only flagging of comorbidities, this patient would erroneously
not have diabetes of any kind on
`r nrow(s10009326) - sum(s10009326$dm_current) - sum(s10009326$dmc_current)`
encounters.  Using the longitudinal flagging in `medicalcoder::comorbidities()`
via `flag.method = 'cumulative'`, we have diabetes flagged on all encounters,
and the severity correctly marked for encounters later in the patient record.

```{r}
#| label: tbl-dm-10009326
#| tbl-cap: "Diabetes without complications (dm) and diabetes with complications (dmc) by flagging method."
#| results: "asis"
#| echo: false
s10009326[
  ,
  lapply(.SD, function(x) data.table::fifelse(x == 1, "X", "")),
  .SDcols = patterns("^dm"),
  by = .(enc_seq)
  ] |>
kableExtra::kbl(
  x = _,
  format = "latex",
  booktabs = TRUE,
  align = "ccccc",
  row.names = FALSE
  ) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down")
)
```


# Time Required To Compute

The amount of time required to apply the several different comorbidity
algorithms to the MIMIC-IV data are shown in @tbl-time-to-compute.

```{r}
#| label: tbl-time-to-compute
#| tbl-cap: "Time, in seconds, to process the MIMIC-IV data."
#| results: "asis"
#| echo: false

ttc <-
  lapply(objs, attr, "tictoc") |>
  lapply(as.numeric) |>
  data.table::as.data.table() |>
  data.table::melt(measure.vars = names(objs), variable.factor = FALSE)

ttc[, c("Tool", "Algorithm", "Flag Method") := data.table::tstrsplit(variable, "_")]
ttc[, variable := NULL]
ttc[, value := sprintf("%.2f", value)]

ttc <-
  data.table::dcast(ttc, Algorithm + `Flag Method` ~ Tool, value.var = "value")
ttc[, `Flag Method` := factor(`Flag Method`, levels = c("current", "cumulative"))]

data.table::setcolorder(
  ttc,
  c("Algorithm", "Flag Method", "medicalcoder", "comorbidity", "mimiciv", "pccc")
)

data.table::setnames(
  ttc,
  old = c("medicalcoder", "comorbidity", "mimiciv", "pccc"),
  new = c("medicalcoder::comorbidities()", "comorbidity::comorbidity()", "MIMIC-IV Code", "pccc::ccc()")
  )

data.table::setkey(ttc, Algorithm, `Flag Method`)

kableExtra::kbl(
  x = ttc[, 2:6],
  format = "latex",
  booktabs = TRUE,
  row.names = FALSE,
  col.names = c("", names(ttc)[3:6]),
  align = "lrrrr"
  ) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down")
) |>
kableExtra::pack_rows(
  index = table(ttc$Algorithm)
)
```

# Session Info

```{r}
#| label: sessionInfo
#| echo: true
#| results: "markup"
sessionInfo()
```

# References
