---
title: "MIMIC-IV Data Analysis via medicalcoder"
subtitle: "Supplement to JAMIA Open Manuscript"

format:
  pdf:
    toc: true
    toc-depth: 4
    lof: true
    lot: true
    number-sections: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage[table]{xcolor}
        \usepackage{threeparttable}
        \usepackage{longtable}

bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  cache = TRUE
)

options(knitr.kable.NA = "")

# Go Avs Go!
# {
#   "burgundy" : "#6F263D",
#   "blue" : "#236192",
#   "light-blue" : "#4F8FCB"
#   "silver" : "#A2AAAD",
#   "black" : "#000000"
# }

loadNamespace('medicalcoder')
loadNamespace('data.table')
```

# Objective and Setup

This supplement provides reproducible validation and benchmarking of
`medicalcoder::comorbidities()` using MIMIC-IV v3.1 data, including algorithmic
agreement with established implementations and demonstration of longitudinal
flagging behavior.

# Materials and Methods

## Data

We use the MIMIC-IV v3.1 Data
[@PhysioNet-mimiciv-3.1;@johnson2023mimic;@PhysioNet2000] as the foundation for
these examples.

The MIMIC-IV v3.1 dataset is a large, publicly available clinical database
containing de-identified health data from patients admitted to intensive care
units at Beth Israel Deaconess Medical Center. It includes detailed information
on diagnoses, procedures, laboratory results, medications, and hospital
encounters, with ICD-9-CM and ICD-10-CM/PCS codes available for billing and
comorbidity research. MIMIC-IV is widely used for method development,
validation, and reproducible research in clinical informatics and outcomes
research.

The MIMIC-IV v3.1 data set can be obtained so long at you meet the requirements
set by PhysioNet[^1].

[^1]: https://physionet.org/content/mimiciv/3.1/

## Comorbidity Algorithms

Within the _medicalcoder_ package are implimentations for several variants of
the Charlson, Elixhauser, and PCCC comorbidity algorithms,
(@tbl-comorbidity-methods) [@quan2005;@quan2011;@glasheen2019;@ahrq2025;@feudtner2014pediatric;@feinstein2024pediatric;@r-pccc;@feinstein2018pccc;@pcccv3sas].

For this supplement we focus on `charlson_quan2005`,
`elixhauser_quan2005`, `pccc_v2.0`, and `pccc_3.1`.

__NOTE:__
The MIMIC-IV data consists of only adult patients.
Although PCCC was designed for pediatric populations, adult MIMIC-IV data are
used here solely to demonstrate algorithmic implementation.


```{r}
#| label: tbl-comorbidity-methods
#| tbl-cap: "Comorbidity methods implemented in _medicalcoder_."
#| echo: false
cm <- data.table::fread(text = "
Comorbidity Method        | Notes
charlson_deyo1992         | ICD-9 codes defined in Table 1 of Quan et al. (2005)
charlson_quan2005         | ICD-9 and ICD-10 defined in Table 1 of Quan et al. (2005); index scoring as reported in Table 2 of Quan et al. (2011)
charlson_quan2011         | ICD-9 and ICD-10 defined in Table 1 of Quan et al. (2005); index scoring as reported in Table 2 of Quan et al. (2011)
charlson_cdmf2019         | ICD-9 and ICD-10 defined in [@glasheen2019]
elixhauser_elixhauser1988 | ICD-9 codes defined in Table 2 of Quan et al. (2005)
elixhauser_ahrq_web       | ICD-9 codes defined in Table 2 of Quan et al. (2005)
elixhauser_quan2005       | ICD-9 and ICD-10 defined in Table 2 of Quan et al. (2005)
elixhauser_ahrq2022       | ICD-10 codes from AHRQ for fiscal year 2022
elixhauser_ahrq2023       | ICD-10 codes from AHRQ for fiscal year 2023
elixhauser_ahrq2024       | ICD-10 codes from AHRQ for fiscal year 2024
elixhauser_ahrq2025       | ICD-10 codes from AHRQ for fiscal year 2025
elixhauser_ahrq2026       | ICD-10 codes from AHRQ for fiscal year 2026
elixhauser_ahrq_icd10     | Any ICD-10 code from all elixhauser_ahrqYYYY
pccc_v2.0                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with pccc::ccc()
pccc_v2.1                 | Extended set of pccc_v2.0 codes based on pccc::ccc() and supplements to Feudtner et al. (2014)
pccc_v3.0                 | ICD-9 and ICD-10 diagnostic and procedure codes consistent with SAS code from Children's Hospital Assoication
pccc_v3.1                 | Extended set of pccc_v3.0 codes based on the supplements to Feinstein et al. (2024)"
)
cm <- cm[, lapply(.SD, gsub, pattern = "_", replacement = "\\\\_")]

kableExtra::kbl(
  x = cm,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down", "HOLD_position")
) |>
kableExtra::pack_rows(group_label = "Charlson",   start_row =  1, end_row =  4) |>
kableExtra::pack_rows(group_label = "Elixhauser", start_row =  5, end_row = 13) |>
kableExtra::pack_rows(group_label = "PCCC",       start_row = 14, end_row = 17)
```

## Alternative Utilities

We compared the results from `medicalcoder::comorbidities()` to the Charlson
comorbidities flagged by the MIMIC-IV code, and the R package _comorbidity_
[@r-comorbidity].  For Elixhauser we will compare results between
`medicalcoder::comorbidities()` and `comorbidity::comorbidity()`.  Lastly, we
will compare PCCC version 2 results between `medicalcoder::comorbidities()` and
the R package _pccc_ [@r-pccc] call `pccc::ccc()`.

# Results

## MIMIC-IV Data Prep

We have stored the path to the MIMIC-IV data as a system variable on our
machines.

```{r}
#| label: verify mimiciv version
# verify that you are looking at version 3.1 of the MIMIC-IV data
stopifnot(
  identical(
    scan(
      file = file.path(Sys.getenv("MIMICIVDATA"), "CHANGELOG.txt"),
      what = "character", sep = "\n", quiet = TRUE)[1],
    "This is the change log for MIMIC-IV v3.1."
  )
)
```

We only need four of the MIMIC-IV hospital datasets:
`admissions`, `patients`, `diagnoses_icd`, and `procedures_icd`.



and do some light formating to before
using `medicalcoder::comorbidities()` to flag comorbidities.  Start by importing
the `admissions`, `patients`, `diagnoses_icd`, and `procedures_icd`, datasets.

@tbl-mimicivdata-checksums gives the sha256 checksums for these files
Each file checksum has been verified.

```{r}
#| label: tbl-mimicivdata-checksums
#| tbl-cap: "SHA256 Check sums for files in the MIMIC-IV v3.1 data."
#| echo: false
#| results: "asis"
data.table::fread(
  file = file.path(Sys.getenv("MIMICIVDATA"), "SHA256SUMS.txt"),
  header = FALSE
)[grepl("^hosp.(admissions|patients|.+_icd)\\.csv\\.gz", V2)] |>
kableExtra::kbl(
  x = _,
  col.names = c("sha256sum", "file"),
  format = "latex",
  booktabs = TRUE
  ) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down", "HOLD_position")
)
```

```{r}
#| label: sha256-verify
#| include: false
verify_sha256sums <- function(dir, sums_file = "SHA256SUMS.txt") {
  old <- setwd(dir)
  on.exit(setwd(old), add = TRUE)

  sys <- Sys.info()[["sysname"]]

  if (identical(sys, "Windows")) {
    # Use CertUtil (built-in)
    lines <- readLines(sums_file, warn = FALSE)
    ok <- logical(length(lines))
    for (i in seq_along(lines)) {
      line <- trimws(lines[[i]])
      if (!nzchar(line)) { ok[[i]] <- TRUE; next }
      # Expected format: "<hash>  <filename>" or "<hash> *<filename>"
      parts <- strsplit(line, "\\s+")[[1]]
      hash <- parts[[1]]
      file <- sub("^\\*", "", parts[[2]])
      out <- suppressWarnings(system2("certutil", c("-hashfile", shQuote(file), "SHA256"), stdout = TRUE))
      got <- tolower(gsub("\\s+", "", out[grepl("^[0-9A-Fa-f]+$", out)]))
      ok[[i]] <- length(got) == 1 && tolower(hash) == got
    }
    return(all(ok))
  }

  # macOS often has `shasum -a 256`, Linux typically has `sha256sum`
  if (nzchar(Sys.which("sha256sum"))) {
    out <- system2("sha256sum", c("-c", sums_file), stdout = TRUE, stderr = TRUE)
    return(any(grepl(": OK$", out)))
  }

  if (nzchar(Sys.which("shasum"))) {
    out <- system2("shasum", c("-a", "256", "-c", sums_file), stdout = TRUE, stderr = TRUE)
    return(any(grepl(": OK$", out)))
  }

  stop("No SHA-256 tool found (need sha256sum, shasum, or certutil).")
}
stopifnot(verify_sha256sums(Sys.getenv("MIMICIVDATA")))
```

```{r}
#| label: import-mimiciv-data
mimicivdata <-
  list.files(
    path = file.path(Sys.getenv("MIMICIVDATA"), "hosp"),
    pattern = "(admissions|patients|.+_icd)\\.csv\\.gz",
    full.names = TRUE
  )
names(mimicivdata) <- sub("\\.csv\\.gz$", "", basename(mimicivdata))

mimicivdata <- lapply(mimicivdata, data.table::fread)
```

We will build an additional data set for age (in years).

```{r}
#| label: build-mimicvidata-ages
mimicivdata$ages <-
  merge(
    x = mimicivdata$admissions[, .(subject_id, hadm_id, admittime)],
    y = mimicivdata$patients[,   .(subject_id, anchor_age, anchor_year)],
    all = FALSE,
    by = "subject_id"
  )
mimicivdata$ages[
  ,
  anchor_year := as.POSIXct(paste0(anchor_year, "-01-01"), format = "%Y-%m-%d")
]
mimicivdata$ages[
  ,
  age :=
    anchor_age +
    as.numeric(difftime(admittime, anchor_year, units = "days")) / 365.25
]
```

For the calls to `medicalcoder::comorbidities()` we can have all the ICD codes
in one data.frame (data.table will be used here).
```{r}
#| label: build-mimicivdata-icd
mimicivdata$icd <-
  data.table::rbindlist(
    list(dx = mimicivdata$diagnoses_icd, pr = mimicivdata$procedures_icd),
    idcol = "dx",
    use.names = TRUE,
    fill = TRUE
  )

# set the diagnoses indicator to an integer as expected for input to
# medicalcoder::comorbidities()
mimicivdata$icd[, dx := as.integer(dx == "dx")]
```

We put all the parts together into one data.frame (data.table)

```{r}
#| label: one-data-set
mimicivDT <-
  merge(
    x = mimicivdata$patients[, .(subject_id)],
    y = mimicivdata$ages[, .(subject_id, hadm_id, age)],
    all = TRUE,
    by = c("subject_id")
  )

mimicivDT <-
  merge(
    x = mimicivdata$icd,
    y = mimicivDT,
    all = TRUE,
    by = c("subject_id", "hadm_id")
  )

mimicivDT <-
  merge(
    x = mimicivDT,
    y = mimicivdata$admissions[, .(subject_id, hadm_id, admittime)],
    all = TRUE,
    by = c("subject_id", "hadm_id")
  )

# set keys and order the data by subject id, admittime, hadm_id
data.table::setkey(mimicivDT, subject_id, admittime, hadm_id)

# set an encounter sequence variable. This works because the data has been
# sorted by subject_id, admittime
mimicivDT[, enc_seq := cumsum(!duplicated(hadm_id)), by = .(subject_id)]

# For some of the methods, only one id variable is allowed so we create a
# composite id:
mimicivDT[, cid := paste(subject_id, enc_seq, sep = "__")]
```

```{r}
#| label: data-structure
str(mimicivDT)
```

```{r}
#| label: save-mimicivDT
#| include: false
# save this data set to disk so it can be used when writting the manuscript
saveRDS(mimicivDT, file = "mimicivDT.rds")
```

There are `r qwraps2::frmt(nrow(mimicivDT))` total rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of
`r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))` unique
hospital encounters.

## Flagging comorbidities

In the following we will outline how to apply the comorbidities methods to the
`mimicivDT` data set.  We will define some wrapper functions to simpify calls.
We will return the results and time how long it takes to apply the methods.  We
will flag the following comorbidities:

* Charlson [@quan2005]: `method = 'charlson_quan2005'`
* Elixhauser [@quan2005]: `method = 'elixhauser_quan2005'`
* PCCC v2 [@feudtner2014pediatric]: `method = 'pccc_v2.0'`
* PCCC v3 [@feinstein2024pediatric] with additional codes: `method = 'pccc_v3.1'`

For each of these comorbidity methods we will flag on a per-encounter basis and
a longitudinal basis within a patient record.

### `medicalcoder::comorbidities()`

```{r}
#| label: medicalcoder-warmup
#| include: false
# to give the best possible time for medicalcoder, it will be helpful to run the
# code once just to make sure all the needed internal data sets have been built
medicalcoder::comorbidities(
  data = mimicivDT[1:10],
  id.vars = c("subject_id", "hadm_id"),
  icd.codes = "icd_code",
  icdv.var = "icd_version",
  dx.var = "dx",
  age.var = "age",
  method = "charlson_quan2005",
  full.codes = FALSE,
  flag.method = "current",
  primarydx = 0L,
  poa = 1L
)
```

The call to `medicalcoder::comorbidities()` has been designed to make calling
multiple comorbidity methods and multiple flagging methods easy.  For this use
case it would be programatically preferable to use a simple wrapper about
`medicalcoder::comorbidities()` to simplify the forthcoming calls.  However, we
opt to repeat ourselves in the following calls for clarity and ease of selective
reproduction by the reader.

#### Charlson

The following code chunk will apply the Charlson [@quan2005] comorbidities
algorithms to each encounter in the data set.

```{r}
#| label: medicalcoder-charlson-current
tic <- Sys.time()

medicalcoder_charlson_current <-
  medicalcoder::comorbidities(
    data        = mimicivDT,  # object inheriting data.frame
    icd.codes   = "icd_code", # character string name of icd codes in data
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,    # consider all codes present on admission
    age.var     = "age", # in years, only relevant to charlson
    primarydx   = 0L,    # consider all diagnosis codes secondary diagnoses.
    method      = "charlson_quan2005",
    flag.method = "current" # default
  )

toc <- Sys.time()

attr(medicalcoder_charlson_current, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

The next call will apply the Charlson [@quan2005] comorbidities to the MIMIC-IV
v3.1 data within a subject's record.  That is, a comorbidities flagged on on
encounter $i$ will also be flagged on encounters $i+1$, $i+2$, ...

Two important things to note about the following call.

1. Longitudinal results within a subject is returned when `flag.method = 'cumulative'`.

2. There is an expectation that there are at least two identification variables.
   The last one needs to be sortable variable.  That is  `id.vars =
   c("subject_id", "eqn_seq")` means that the data will be sorted by `subject_id`
   and `enc_seq` with `enc_seq = 1` considered to be occur before `enc_seq = 2`,
   etc.  In the MIMIC-IV v3.1 data, the `hadm_id` are not sequential within a
   patient record, hence the construction of the `enc_seq` variable.

```{r}
#| label: medicalcoder-charlson-cumulative
tic <- Sys.time()

medicalcoder_charlson_cumulative <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    age.var     = "age",
    primarydx   = 0L,
    method      = "charlson_quan2005",
    flag.method = "cumulative"
  )

toc <- Sys.time()

attr(medicalcoder_charlson_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

#### Elixhauser

Applying Elixhauser requires a only a change to the `method` argument.

```{r}
#| label: medicalcoder-elixhauser-current
tic <- Sys.time()

medicalcoder_elixhauser_current <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    primarydx   = 0L,
    method      = "elixhauser_quan2005",
    flag.method = "current"
  )

toc <- Sys.time()

attr(medicalcoder_elixhauser_current, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

```{r}
#| label: medicalcoder-elixhauser-cumulative
tic <- Sys.time()

medicalcoder_elixhauser_cumulative <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    primarydx   = 0L,
    method      = "elixhauser_quan2005",
    flag.method = "cumulative"
  )

toc <- Sys.time()

attr(medicalcoder_elixhauser_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

#### PCCC v2.0

PCCC v2.0 has been built to be consistent with the output from `pccc::ccc()` and
is called used `method = "pccc_v2.0"`.  Another option is `pccc_v2.1` which has
additional ICD codes based on our interpretation of the intention of the
mappings based on the suplements to @feudtner2014pediatric.

```{r}
#| label: medicalcoder-pcccv2-current
tic <- Sys.time()

medicalcoder_pcccv2.0_current <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    method      = "pccc_v2.0",
    flag.method = "current"
  )

toc <- Sys.time()

attr(medicalcoder_pcccv2.0_current, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

```{r}
#| label: medicalcoder-pcccv2-cumulative
tic <- Sys.time()

medicalcoder_pcccv2.0_cumulative <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    age.var     = "age",
    method      = "pccc_v2.0",
    flag.method = "cumulative"
  )

toc <- Sys.time()

attr(medicalcoder_pcccv2.0_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

#### PCCC v3

PCCC v3 [@feinstein2024pediatric] differs from PCCC v2 [@feudtner2014pediatric]
in two major ways: (1) Updated mappings between ICD codes and conditions. (2)
How conditions flagged by technology dependence are handled.  If the only reason
for a conditions to be flagged is due to technology dependence then under PCCC
v3 the condition _will not_ be flagged.  At least one non-technology dependent
code needs to be in the patient record for the technology-dependent conditions
to also be flagged.  This makes longitudinal flagging particularly difficult.
Examples to follow.

```{r}
#| label: medicalcoder-pcccv3-current
tic <- Sys.time()

medicalcoder_pcccv3.1_current <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    method      = "pccc_v3.1",
    flag.method = "current"
  )

toc <- Sys.time()

attr(medicalcoder_pcccv3.1_current, "tictoc") <-
  difftime(toc, tic, units = "secs")
```

```{r}
#| label: medicalcoder-pcccv3-cumulative
tic <- Sys.time()

medicalcoder_pcccv3.1_cumulative <-
  medicalcoder::comorbidities(
    data        = mimicivDT,
    icd.codes   = "icd_code",
    id.vars     = c("subject_id", "enc_seq"),
    icdv.var    = "icd_version",
    dx.var      = "dx",
    poa         = 1L,
    age.var     = "age",
    method      = "pccc_v3.1",
    flag.method = "cumulative"
  )

toc <- Sys.time()

attr(medicalcoder_pcccv3.1_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs")
```


### MIMIC-IV Code

We use some of the SQL code for applying the Charlson
comorbidity algorithm to the MIMIC-IV data provided by the MIT Laboratory for
Computational Physilogy (MIT\_LCP) on
[GitHub](https://github.com/MIT-LCP/mimic-code.git).  We stored the
path to a local copy of the respository in the system environmental variable
`MIMICIVCODE`.

The SQL code for applying the Charlson comorbidities and checksum:
```{r}
mimiciv_charlson_sql <-
  file.path(Sys.getenv("MIMICIVCODE"), "concepts", "comorbidity", "charlson.sql")

# checksum
stopifnot(
  substr(
    system(
      sprintf("sha256sum %s", mimiciv_charlson_sql),
      intern = TRUE
    ),
  start = 1L,
  stop = 64L
  ) == "5b797b673ace4dcbc6f686848d24371382f49ac61fcc39101f09c3b42969d801"
)
```

The SQL code is expected to run on Google Big Query.  We make a few
small modifications to the code so we can evaluate the SQL locally via RSQLite.

```{r}
#| label: mimic-code-edits
mimiciv_charlson_sql <-
  scan(
    file = mimiciv_charlson_sql,
    what = "character",
    sep = "\n",
    quiet = TRUE
  )

# modify the query to work in SQLite
mimiciv_charlson_sql <-
  gsub(
    pattern = "physionet-data.mimiciv_hosp.admissions",
    replacement = "admissions",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <-
  gsub(
    pattern = "physionet-data.mimiciv_hosp.diagnoses_icd",
    replacement = "diagnoses",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <-
  gsub(
    pattern = "physionet-data.mimiciv_derived.age",
    replacement = "ages",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <-
  gsub(
    pattern = "GREATEST",
    replacement = "MAX",
    x = mimiciv_charlson_sql,
    fixed = TRUE
  )

mimiciv_charlson_sql <- paste(mimiciv_charlson_sql, collapse = "\n")
```

We set up an SQL database in memory and evaluate the code.

```{r}
#| label: mimic-iv-charlson-results
#| echo: true
#| results: "hide"
#| cache: true
con <- odbc::dbConnect(drv = RSQLite::SQLite(), dbname = ":memory:")

# add data to the data base
odbc::dbWriteTable(conn = con, name = "diagnoses",  value = mimicivdata$diagnoses)
odbc::dbWriteTable(conn = con, name = "admissions", value = mimicivdata$admissions)
odbc::dbWriteTable(conn = con, name = "patients",   value = mimicivdata$patients)
odbc::dbWriteTable(conn = con, name = "ages",       value = mimicivdata$ages)

# Run the query and record the time required to do so
tic <- Sys.time()

mimiciv_charlson_current <- odbc::dbGetQuery(con, mimiciv_charlson_sql)
data.table::setDT(mimiciv_charlson_current)

# Note: not all subject, enc_seq (hadm_id) values are in the return.  Add the
# needed rows # and set all the flags to zero
nrow(unique(mimicivDT[, .(subject_id, hadm_id, enc_seq)])) >
  nrow(mimiciv_charlson_current)

mimiciv_charlson_current <-
  merge(
    x = unique(mimicivDT[, .(subject_id, hadm_id, enc_seq)]),
    y = mimiciv_charlson_current,
    all = TRUE,
    by = c("subject_id", "hadm_id")
  )

for (j in names(mimiciv_charlson_current)) {
  if (!(j %in% c("subject_id", "hadm_id", "enc_seq"))) {
    idx <- which(is.na(mimiciv_charlson_current[[j]]))
    mimiciv_charlson_current[[j]][idx] <- 0L
  }
}

toc <- Sys.time()

attr(mimiciv_charlson_current, "tictoc") <- difftime(toc, tic, units = "secs")

# close DB connection
odbc::dbDisconnect(conn = con)
```

### `comorbidity::comorbidity()`

The R package _comorbidity_ [@r-comorbidity] is a widely used and referenced
tool for applying the Charlson and Elixhauser comorbidities.
One consideration for using `comorbidity::comorbidity()` is that ICD-9 or ICD-10
codes are used independently. There are
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id)][V1 > 1, .N]`
subjects with both ICD-9 and ICD-10 codes in their records, and
`r mimicivDT[, length(unique(icd_version)), by = .(subject_id, hadm_id)][V1 > 1, .N]`
encounters with both ICD versions.

To flag the conditions appropriately we must split the data by ICD version,
apply `comorbidity::comorbidity()` to each split, and then aggregate the
results.  Also, `comorbidity::comorbidity()` only takes one column for the id
and thus we need to use `cid` for the id instead of the `c("subject_id",
"enc_seq")` used in the calls to `medicalcoder::comorbidities()`.  In the
following work we will copy the data built for the encounter level flagging and
extend for cumulative flagging as well.

The following is done using _comorbidity_ version
`r packageVersion("comorbidity")`.

#### Charlson

```{r}
#| label: comorbidity-charlson
tic <- Sys.time()

comorbidity_charlson_icd9 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 9 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "charlson_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe
                   # comorbidities is also flagged
  )

comorbidity_charlson_icd10 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 10 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "charlson_icd10_quan",
    assign0 = TRUE
  )

comorbidity_charlson_current <-
  rbind(
    comorbidity_charlson_icd9,
    comorbidity_charlson_icd10
  )

# aggregate between the icd versions
comorbidity_charlson_current <-
  aggregate(. ~ cid, data = comorbidity_charlson_current, FUN = max)

# Note, Not all the cid are in comorbidity_charlson_current.  Especially for
# building the cumulative flags we need all the subjects and encounters
comorbidity_charlson_current <-
  merge(
    x = comorbidity_charlson_current,
    y = unique(mimicivDT[, .(cid)]),
    all = TRUE,
    by = "cid"
  )

# set NA values to zero
for (j in names(comorbidity_charlson_current)[-1]) {
  idx <- which(is.na(comorbidity_charlson_current[[j]]))
  comorbidity_charlson_current[[j]][idx] <- 0L
}

# apply the scoring, to use comorbidity::score the object needs to have certain
# attributes which were lost when aggregating over the icd versions.  Replace
# the attributes
attributes(comorbidity_charlson_current)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_charlson_icd9)[c("class", "variable.labels", "map")]

comorbidity_charlson_current[["score"]] <-
  comorbidity::score(
    x = comorbidity_charlson_current,
    weights = "quan",
    assign0 = TRUE
  )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
comorbidity_charlson_current[["score"]] <-
  as.integer(comorbidity_charlson_current[["score"]])

# expand the id columns and sort
data.table::setDT(comorbidity_charlson_current)
comorbidity_charlson_current[
  ,
  c("subject_id", "enc_seq") := data.table::tstrsplit(cid, "__")
  ][
  ,
  `:=`(subject_id = as.integer(subject_id), enc_seq = as.integer(enc_seq))
  ]
data.table::setkey(comorbidity_charlson_current, subject_id, enc_seq)

# end timing for encounter level flagging
toc <- Sys.time()

# For the cumulative flagging
tic1 <- Sys.time()

comorbidity_charlson_cumulative <- data.table::copy(comorbidity_charlson_current)

# next we use cumsum to flag conditions moving forward in time
cumflag <- function(x) as.integer(cumsum(x) > 0)

for (j in names(comorbidity_charlson_cumulative)) {
  if (!(j %in% c("subject_id", "enc_seq", "cid"))) {
    e <- substitute(
          comorbidity_charlson_cumulative[, J := cumflag(J), by = .(subject_id)],
          list(J = as.name(j))
          )
    eval(e)
  }
}

# clean up flags by severity
comorbidity_charlson_cumulative[diab == 1 & diabwc == 1, diab := 0L]
comorbidity_charlson_cumulative[mld == 1 & msld == 1, mld := 0L]
comorbidity_charlson_cumulative[canc == 1 & metacanc == 1, canc := 0L]

# apply the scoring, to use comorbidity::score the object needs to have certain
# attributes which were lost when aggregating over the icd versions.  Replace
# the attributes
data.table::setDF(comorbidity_charlson_cumulative)
attributes(comorbidity_charlson_cumulative)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_charlson_icd9)[c("class", "variable.labels", "map")]

comorbidity_charlson_cumulative[["score"]] <-
  comorbidity::score(
    x = comorbidity_charlson_cumulative,
    weights = "quan",
    assign0 = TRUE
  )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
comorbidity_charlson_cumulative[["score"]] <-
  as.integer(comorbidity_charlson_cumulative[["score"]])

data.table::setDT(comorbidity_charlson_cumulative)

toc1 <- Sys.time()

attr(comorbidity_charlson_current, "tictoc") <-
  difftime(toc, tic, units = "secs")

attr(comorbidity_charlson_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs") +
  difftime(toc1, tic1, units = "secs")
```

#### Elixhauser

Compared to the work done for flagging Charlson comorbidities, Elixhauser
flagging requires changing the `map` argument in the calls to
`comorbidity::comorbidity()` and there is no scoring to be done.

```{r}
#| label: comorbidity-elixhauser
tic <- Sys.time()

comorbidity_elixhauser_icd9 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 9 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "elixhauser_icd9_quan",
    assign0 = TRUE
  )

comorbidity_elixhauser_icd10 <-
  comorbidity::comorbidity(
    x       = subset(mimicivDT, icd_version == 10 & dx == 1L),
    id      = "cid",
    code    = "icd_code",
    map     = "elixhauser_icd10_quan",
    assign0 = TRUE
  )

comorbidity_elixhauser_current <-
  rbind(
    comorbidity_elixhauser_icd9,
    comorbidity_elixhauser_icd10
  )

# aggregate between the icd versions
comorbidity_elixhauser_current <-
  aggregate(. ~ cid, data = comorbidity_elixhauser_current, FUN = max)

# Note, Not all the cid are in comorbidity_elixhauser_current.  Especially for
# building the cumulative flags we need all the subjects and encounters
comorbidity_elixhauser_current <-
  merge(
    x = comorbidity_elixhauser_current,
    y = unique(mimicivDT[, .(cid)]),
    all = TRUE,
    by = "cid"
  )

# set NA values to zero
for (j in names(comorbidity_elixhauser_current)[-1]) {
  idx <- which(is.na(comorbidity_elixhauser_current[[j]]))
  comorbidity_elixhauser_current[[j]][idx] <- 0L
}

# expand the id columns and sort
data.table::setDT(comorbidity_elixhauser_current)
comorbidity_elixhauser_current[
  ,
  c("subject_id", "enc_seq") := data.table::tstrsplit(cid, "__")
  ][
  ,
  `:=`(subject_id = as.integer(subject_id), enc_seq = as.integer(enc_seq))
  ]
data.table::setkey(comorbidity_elixhauser_current, subject_id, enc_seq)

# end timing for encounter level flagging
toc <- Sys.time()

# For the cumulative flagging
tic1 <- Sys.time()

comorbidity_elixhauser_cumulative <- data.table::copy(comorbidity_elixhauser_current)

# next we use cumsum to flag conditions moving forward in time
cumflag <- function(x) as.integer(cumsum(x) > 0)

for (j in names(comorbidity_elixhauser_cumulative)) {
  if (!(j %in% c("subject_id", "enc_seq", "cid"))) {
    e <- substitute(
          comorbidity_elixhauser_cumulative[, J := cumflag(J), by = .(subject_id)],
          list(J = as.name(j))
          )
    eval(e)
  }
}

# clean up flags by severity?
comorbidity_elixhauser_cumulative[diabunc  == 1L & diabc    == 1L, diabunc  := 0L]
comorbidity_elixhauser_cumulative[hypunc   == 1L & hypc     == 1L, hypunc   := 0L]
comorbidity_elixhauser_cumulative[solidtum == 1L & metacanc == 1L, solidtum := 0L]

toc1 <- Sys.time()

attr(comorbidity_elixhauser_current, "tictoc") <-
  difftime(toc, tic, units = "secs")

attr(comorbidity_elixhauser_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs") +
  difftime(toc1, tic1, units = "secs")
```


### `pccc::ccc()`

The R package _pccc_ [@r-pccc] implements PCCC v2 [@feudtner2014pediatric].  As
with `comorbidity::comorbidity()` and many other implementations of
comorbidities, only one version of ICD is assessed at a time.

The expected input structure of the data for `pccc::ccc()` is a wide format, one
row per encounter, one column for each ICD code.

The following uses _pccc_ version `r packageVersion("pccc")`.

```{r}
#| label: pccc
mimicivDT_for_pccc <-
  split(mimicivDT, by = "icd_version") |>
  lapply(
    data.table::dcast,
    formula = cid ~ ifelse(dx == 1, "DX", "PR") + seq_num,
    value.var = "icd_code"
    )

tic <- Sys.time()

pccc_pcccv2.0_current <-
  rbind(
    pccc::ccc(
      data = mimicivDT_for_pccc[["9"]],
      id = cid,
      dx_cols = grep("^DX", names(mimicivDT_for_pccc[["9"]]), value = TRUE),
      pc_cols = grep("^PR", names(mimicivDT_for_pccc[["9"]]), value = TRUE),
      icdv = 9
    )
    ,
    pccc::ccc(
      data = mimicivDT_for_pccc[["10"]],
      id = cid,
      dx_cols = grep("^DX", names(mimicivDT_for_pccc[["10"]]), value = TRUE),
      pc_cols = grep("^PR", names(mimicivDT_for_pccc[["10"]]), value = TRUE),
      icdv = 10
    )
  )


# aggregate over ICD version
pccc_pcccv2.0_current <-
  aggregate(. ~ cid, data = pccc_pcccv2.0_current, FUN = max)

# Not all the cid are in this data set.  add them and set the missing values to
# zeros.
pccc_pcccv2.0_current <-
  merge(
    x = pccc_pcccv2.0_current,
    y = unique(mimicivDT[, .(cid)]),
    all = TRUE,
    by = "cid"
  )

for (j in names(pccc_pcccv2.0_current)[-1]) {
  idx <- which(is.na(pccc_pcccv2.0_current[[j]]))
  pccc_pcccv2.0_current[[j]][idx] <- 0L
}

toc <- Sys.time()

# cumflag
tic1 <- Sys.time()

data.table::setDT(pccc_pcccv2.0_current)
pccc_pcccv2.0_current[,
  c("subject_id", "enc_seq") := data.table::tstrsplit(cid, "__")
  ][
  ,
  `:=`(
    subject_id = as.integer(subject_id),
    enc_seq = as.integer(enc_seq)
  )]

data.table::setkey(pccc_pcccv2.0_current, subject_id, enc_seq)

pccc_pcccv2.0_cumulative <- data.table::copy(pccc_pcccv2.0_current)

for (j in names(pccc_pcccv2.0_cumulative)) {
  if (!(j %in% c("cid", "subject_id", "enc_seq"))) {
    e <- substitute(pccc_pcccv2.0_cumulative[, J := cumflag(J), by = .(subject_id)],
      list(J = as.name(j)))
    eval(e)
  }
}
toc1 <- Sys.time()

attr(pccc_pcccv2.0_current, "tictoc") <- difftime(toc, tic, units = "secs")
attr(pccc_pcccv2.0_cumulative, "tictoc") <-
  difftime(toc, tic, units = "secs") +
  difftime(toc1, tic1, units = "secs")
```

## Consistency in Flags between Utilties

### `medicalcoder::comorbidities()` vs MIMIC-IV Code

Comparing the results between `medicalcoder::comorbidities()` and MIMIC-IV code
for Charlson comorbidities at the encounter level should note that the number of
rows returned differed.  In the code above that was rectified.
`medicalcoder::comorbidities()` returned a row for each subject_id, enc_seq
(hadm_id) pair, whereas the MIMIC-IV code did not return a row for each.  This
was resolved above and zeros imputed for the subject_id, hadm_id which were
omitted from the `mimiciv_charlson_current` object.

We build one data.frame to make the comparisons between the two methods easier.
```{r}
#| label: build-mdcr-v-mimiciv
mdcr_v_mimiciv <-
  merge(
    x = medicalcoder_charlson_current,
    y = mimiciv_charlson_current,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_mimiciv")
  )
```

There are three comorbidities where there are different levels of severity.
`medicalcoder::comorbidities()` will set the less severe condition indicator to
0 when the more severe condition is flagged.  The MIMIC-IV code returns a flag
for both levels.  Both `medicalcoder::comorbidities()` and MIMIC-IV Code only
consider the more severe case in the index scoring.  For the comorbidities
diabetes, cancer, and liver disease we will set the less sever flag to 0 if the
more sever flag is present for the results from the MIMIC-IV Code.

```{r}
#| label: set-zeros-by-severity
mdcr_v_mimiciv[
  ,
  .(
    mdcr_dm    = sum(dm == 1L & dmc == 1L),
    mimiciv_dm = sum(diabetes_without_cc == 1L & diabetes_with_cc == 1L),
    mdcr_canc    = sum(mal == 1L & mst == 1L),
    mimiciv_canc = sum(malignant_cancer == 1L & metastatic_solid_tumor == 1L),
    mdcr_ld    = sum(mld == 1L & msld == 1L),
    mimiciv_ld = sum(mild_liver_disease == 1L & severe_liver_disease == 1L)
   )
]

mdcr_v_mimiciv[
  diabetes_without_cc == 1L & diabetes_with_cc == 1L,
  diabetes_without_cc := 0L
]

mdcr_v_mimiciv[
  malignant_cancer == 1L & metastatic_solid_tumor == 1L,
  malignant_cancer := 0L
]

mdcr_v_mimiciv[
  mild_liver_disease == 1L & severe_liver_disease == 1L,
  mild_liver_disease := 0L
]
```

We define the mapping between the columns returned from
`medicalcoder::comorbidities()` and the MIMIC-IV Code:
```{r}
#| label: mdcr-v-mimic-column-mapping
mvm_columns <-
  data.table::fread(text = "
    medicalcoder   | mimicivcode
    aidshiv        | aids
    cebvd          | cerebrovascular_disease
    chf            | congestive_heart_failure
    copd           | chronic_pulmonary_disease
    dem            | dementia
    dm             | diabetes_without_cc
    dmc            | diabetes_with_cc
    hp             | paraplegia
    mal            | malignant_cancer
    mi             | myocardial_infarct
    mld            | mild_liver_disease
    msld           | severe_liver_disease
    mst            | metastatic_solid_tumor
    pud            | peptic_ulcer_disease
    pvd            | peripheral_vascular_disease
    rhd            | rheumatic_disease
    rnd            | renal_disease
    age_score_mdcr | age_score_mimiciv
    cci            | charlson_comorbidity_index
  ")
```

The following `for` loop will check that the columns between
`medicalcoder::comorbidities()` and MIMIC-IV Code are identical.  If so, the
columns are removed from the `mdcr_v_mimiciv` object, otherwise a message is
printed.

```{r}
#| label: mdcr-v-mimic-check-for-identical-columns
for (i in seq_len(nrow(mvm_columns))) {
  j1 <- mvm_columns[["medicalcoder"]][i]
  j2 <- mvm_columns[["mimicivcode"]][i]
  e <- identical(mdcr_v_mimiciv[[j1]], mdcr_v_mimiciv[[j2]])
  if (e) {
    mdcr_v_mimiciv[[j1]] <- NULL
    mdcr_v_mimiciv[[j2]] <- NULL
  } else {
    print(sprintf("%s and %s are not identical", j1, j2))
  }
}
```

The differences in the age score is simple.  In the MIMIC-IV Code, the records
without an age were dropped, `medicalcoder::comorbidities()` returned rows for
these subjects. In all cases there where no comorbidities flagged.

```{r}
#| label: mdcr-v-mimiciv-age-score
mdcr_v_mimiciv[, .N, keyby = .(age_score_mdcr, age_score_mimiciv)]

medicalcoder_charlson_current[
  is.na(age_score), all(cci == 0) & all(cmrb_flag == 0)
]

mdcr_v_mimiciv[, age_score_mdcr := NULL]
mdcr_v_mimiciv[, age_score_mimiciv := NULL]
```

All that is left in `mdcr_v_mimiciv` are the id.vars and the
`num_cmrb`, and `cmrb_flag`.  These columns are from
`medicalcoder::comorbidities()` and report the number of comorbidities flagged
and indicator for any comorbidity.

```{r}
#| label: show-remainder-of-mdcr-v-mimicivcode
head(mdcr_v_mimiciv)
```

### `medicalcoder::comorbidities()` vs `comorbidity::comorbidity()`

We will compare the returns from the two utilities for both Charlson and
Elixhauser comorbidities and the flagging of the comorbidities on an encounter
by encounter basis (`flag.method = 'current'`) and longitudinally (`flag.method
= 'cumulative'`).

#### Charlson
```{r}
#| label: build-mdcr-v-cmrb-charlson
mdcr_v_cmrb_charlson_current <-
  merge(
    x = medicalcoder_charlson_current,
    y = comorbidity_charlson_current,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_cmrb")
  )

mdcr_v_cmrb_charlson_cumulative <-
  merge(
    x = medicalcoder_charlson_cumulative,
    y = comorbidity_charlson_cumulative,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_cmrb")
  )
```

The following maps the columns in the returned objects.
```{r}
#| label: mdcr-v-cmrb-charlson-column-map
mvc_columns <-
  data.table::fread(text = "
    medicalcoder   | comorbidity
    aidshiv        | aids
    cebvd          | cevd
    chf_mdcr       | chf_cmrb
    hp_mdcr        | hp_cmrb
    copd           | cpd
    dem            | dementia
    dm             | diab
    dmc            | diabwc
    mal            | canc
    mi_mdcr        | mi_cmrb
    mld_mdcr       | mld_cmrb
    msld_mdcr      | msld_cmrb
    mst            | metacanc
    pud_mdcr       | pud_cmrb
    pvd_mdcr       | pvd_cmrb
    rhd            | rheumd
    rnd            | rend
    cci            | score
  ")
```

The following `for` loop will check if the columns are identical, removing from
the object is so, and printing a message if not.
```{r}
#| label: mdcr-v-cmrb-charlson-column-tests
for (i in seq_len(nrow(mvc_columns))) {
  jm <- mvc_columns[["medicalcoder"]][i]
  jc <- mvc_columns[["comorbidity"]][i]
  e <- identical(
    mdcr_v_cmrb_charlson_current[[jm]],
    mdcr_v_cmrb_charlson_current[[jc]]
  )
  if (e) {
    mdcr_v_cmrb_charlson_current[[jm]] <- NULL
    mdcr_v_cmrb_charlson_current[[jc]] <- NULL
  } else {
    print(sprintf("mdcr_v_cmrb_charlson_current: %s is not identical to %s", jm, jc))
  }
  e <- identical(mdcr_v_cmrb_charlson_cumulative[[jm]], mdcr_v_cmrb_charlson_cumulative[[jc]])
  if (e) {
    mdcr_v_cmrb_charlson_cumulative[[jm]] <- NULL
    mdcr_v_cmrb_charlson_cumulative[[jc]] <- NULL
  } else {
    print(sprintf("mdcr_v_cmrb_charlson_cumulative: %s is not identical to %s", jm, jc))
  }
}
```

There is a mismatch for malignant cancer.  In all cases
`medicalcoder::comorbidities()` flagged malignant cancer while
`comorbidity::comorbidity()` did not flag.

```{r}
#| label: mal-delta-mdcr-v-cmrb
mdcr_v_cmrb_charlson_current[mal != canc, all(mal > canc)]
```

Let's look for the ICD-codes that `medicalcoder::comorbidities()` flagged and
`comorbidity::comorbidity()` did not.
```{r}
#| label: mal-delta-codes
DT <-
  mimicivDT[
    mdcr_v_cmrb_charlson_current[mal != canc, .(cid)],
    on = "cid"
  ][
    ,
    code_id := as.numeric(factor(icd_code))
  ]

medicalcoder::comorbidities(
  data = DT,
  id.var = c("icd_code", "code_id"),
  icd.codes = "icd_code",
  icdv.var = "icd_version",
  dx.var = "dx",
  poa = 1L,
  primarydx = 0L,
  method = "charlson_quan2005"
)[mal == 1, .(icd_code, code_id, mal)]

comorbidity::comorbidity(
  x = DT,
  id = "code_id",
  code = "icd_code",
  map = "charlson_icd10_quan",
  assign0 = TRUE
  ) |>
subset(subset = canc == 1, select = c("code_id", "canc"))
```

The ICD-10-CM codes C4A.x are flagged as a malignancy by
`medicalcoder::comorbidities()` but not flagged by `comorbidity::comorbidity()`.
Table 2 of @quan2005 lists C45.x - C58.x as mapping to any malignancy.  We argue
that C4A.x is included in that range of codes and that these are false negatives
from `comorbidity::comorbidity()`.

**NOTE:** _medicalcoder_ has a robust internal database of ICD codes.  End users
can access the codes via `medicalcoder::get_icd_codes()`.
```{r}
#| label: lookup-C4A.x
subset(
  medicalcoder::get_icd_codes(with.description = TRUE),
  code %in% c("C4A111", "C4A39", "C4A52", "C4A62", "C4A71", "C4A72", "C4A9")
  )
```


#### Elixhauser

The checks between `medicalcoder::comorbidities()` and
`comorbidity::comorbidity()` for Elixhauser comorbidities is done the same way
the prior checks have been done.

```{r}
#| label: build-mdcr-v-cmrb-elixhauser
mdcr_v_cmrb_elixhauser_current <-
  merge(
    x = medicalcoder_elixhauser_current,
    y = comorbidity_elixhauser_current,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_cmrb")
  )

mdcr_v_cmrb_elixhauser_cumulative <-
  merge(
    x = medicalcoder_elixhauser_cumulative,
    y = comorbidity_elixhauser_cumulative,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_cmrb")
  )
```

```{r}
#| label: mdcr-v-cmrb-column-map
mvc_columns <-
  data.table::fread(text = "
    medicalcoder        | comorbidity
    AIDS                | aids
    ALCOHOL             | alcohol
    ANEMDEF             | dane
    ARTH                | rheumd
    BLDLOSS             | blane
    CARDIAC_ARRHYTHMIAS | carit
    CHF                 | chf
    CHRNLUNG            | cpd
    COAG                | coag
    DEPRESS             | depre
    DM                  | diabunc
    DMCX                | diabc
    DRUG                | drug
    HTN_CX              | hypc
    HTN_UNCX            | hypunc
    HYPOTHY             | hypothy
    LIVER               | ld
    LYMPH               | lymph
    LYTES               | fed
    METS                | metacanc
    NEURO               | ond
    OBESE               | obes
    PARA                | para
    PERIVASC            | pvd
    PSYCH               | psycho
    PULMCIRC            | pcd
    RENLFAIL            | rf
    TUMOR               | solidtum
    ULCER               | pud
    VALVE               | valv
    WGHTLOSS            | wloss
  ")
```

```{r}
#| label: mdcr-v-cmrb-column-tests
for (i in seq_len(nrow(mvc_columns))) {
  jm <- mvc_columns[["medicalcoder"]][i]
  jc <- mvc_columns[["comorbidity"]][i]
  e <- identical(
    mdcr_v_cmrb_elixhauser_current[[jm]],
    mdcr_v_cmrb_elixhauser_current[[jc]]
  )
  if (e) {
    mdcr_v_cmrb_elixhauser_current[[jm]] <- NULL
    mdcr_v_cmrb_elixhauser_current[[jc]] <- NULL
  } else {
    print(sprintf("mdcr_v_cmrb_elixhauser_current: %s is not identical to %s", jm, jc))
  }
  e <- identical(
    mdcr_v_cmrb_elixhauser_cumulative[[jm]],
    mdcr_v_cmrb_elixhauser_cumulative[[jc]]
  )
  if (e) {
    mdcr_v_cmrb_elixhauser_cumulative[[jm]] <- NULL
    mdcr_v_cmrb_elixhauser_cumulative[[jc]] <- NULL
  } else {
    print(sprintf("mdcr_v_cmrb_elixhauser_cumulative: %s is not identical to %s", jm, jc))
  }
}
```

The remaining columns in `mdcr_v_cmrb_elixhauser_current` and
`mdcr_v_cmrb_elixhauser_cumulative` are unique to
`medicalcoder::comorbidities()`. `HTN_C` is any hypertension.

```{r}
#| label: whats-left-mdcr_v_cmrb_elixhauser
mdcr_v_cmrb_elixhauser_current
mdcr_v_cmrb_elixhauser_cumulative
```

```{r}
#| label: verify-htn-c
stopifnot(
  identical(
    medicalcoder_elixhauser_current[["HTN_C"]],
    as.integer(
      medicalcoder_elixhauser_current[["HTN_UNCX"]] |
        medicalcoder_elixhauser_current[["HTN_CX"]]
    )
  ),
  identical(
    medicalcoder_elixhauser_cumulative[["HTN_C"]],
    as.integer(
      medicalcoder_elixhauser_cumulative[["HTN_UNCX"]] |
        medicalcoder_elixhauser_cumulative[["HTN_CX"]]
    )
  )
)
```

The `mortality_index` and `readmission_index` are based on the weights from AHRQ
in this case were `method = "elixhauser_quan2005"`.  For `method =
"elixhauser_ahrqXXXX"` other weights may be applied.  To see the specific
weights used by method call `medicalcoder::get_elixhauser_index_scores()`

```{r}
#| label: medicalcoder-get-elixhauser-index-scores
str(
  medicalcoder::get_elixhauser_index_scores()
)
```

### `medicalcoder::comorbidities()` vs `pccc::ccc()`

```{r}
#| label: build-mdcr-v-pccc
mdcr_v_pccc_current <-
  merge(
    x = medicalcoder_pcccv2.0_current,
    y = pccc_pcccv2.0_current,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_pccc")
  )

mdcr_v_pccc_cumulative <-
  merge(
    x = medicalcoder_pcccv2.0_cumulative,
    y = pccc_pcccv2.0_cumulative,
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_mdcr", "_pccc")
  )
```

The return from `pccc::ccc()` does not inlcude all the `subject_id` and
`enc_seq`.  The missing ones are all flagged as no comorbidities by
`medicalcoder::comorbidities()`.  Implicitly this is the same result.

```{r}
#| label: mdcr-v-pccc-na-zero
# verify that medicalcoder's flag for any comorbidity (cmrb_flag) is zero when
# the ccc_flag from pccc::ccc()
mdcr_v_pccc_current[is.na(ccc_flag), all(cmrb_flag == 0)]
mdcr_v_pccc_cumulative[is.na(ccc_flag), all(cmrb_flag == 0)]
```

```{r}
#| label: mdcr-v-pccc-colwise-checks
for (j in c(unique(medicalcoder::get_pccc_conditions()$condition), "tech_dep", "transplant", "ccc_flag")) {
  if (j %in% c("tech_dep", "transplant")) {
    jm <- sprintf("any_%s", j)
    jp <- j
  } else if (j == "ccc_flag") {
    jm <- "cmrb_flag"
    jp <- j
  } else {
    jm <- sprintf("%s_mdcr", j)
    jp <- sprintf("%s_pccc", j)
  }
  e1 <- identical(mdcr_v_pccc_current[[jm]], mdcr_v_pccc_current[[jp]])
  e2 <- identical(mdcr_v_pccc_cumulative[[jm]], mdcr_v_pccc_cumulative[[jp]])
  if (e1) {
    mdcr_v_pccc_current[[jm]] <- NULL
    mdcr_v_pccc_current[[jp]] <- NULL
  } else {
    print(sprintf("for mdcr_v_pccc_current: %s is not identical to %s", jm, jp))
  }
  if (e1) {
    mdcr_v_pccc_cumulative[[jm]] <- NULL
    mdcr_v_pccc_cumulative[[jp]] <- NULL
  } else {
    print(sprintf("for mdcr_v_pccc_cumulative: %s is not identical to %s", jm, jp))
  }
}
```

There are some differences between the results for technology dependence
(`tech_dep`) and transplant.  The underlying reason for these differences is
that the implementation of `medicalcoder::comorbidities()` allows for flagging
of subconditions.  Under this paradigm both technology dependence and transplant
are subconditions.  There are several ICD codes which had/have errors in _pccc_
version 1.0.7.

```{r}
#| label: mdcr-v-pccc-tech-and-trans
DT <-
  mimicivDT[
      mdcr_v_pccc_current[
        any_tech_dep != tech_dep | any_transplant != transplant
      ],
      on = c("cid", "subject_id", "enc_seq")
    ][
      , .(icd_code, icd_version, dx)
    ][
      ,
      code_id := paste0(
        "ICD-", icd_version, "-", ifelse(dx == 1, "CM", "PCS"), " ", icd_code
      )
    ] |>
  unique()

mdcr <-
  medicalcoder::comorbidities(
    data = DT,
    id.var = "code_id",
    icd.codes = "icd_code",
    icdv.var = "icd_version",
    dx.var = "dx",
    poa = 1L,
    method = 'pccc_v2.0'
  )

DT_for_pccc <-
  split(DT, by = "icd_version") |>
  lapply(
    data.table::dcast,
    formula = code_id ~ ifelse(dx == 1, "DX", "PR"),
    value.var = "icd_code"
    )

pccc <-
  rbind(
    pccc::ccc(
      data = DT_for_pccc[["9"]],
      id = code_id,
      dx_cols = grep("^DX", names(DT_for_pccc[["9"]]), value = TRUE),
      pc_cols = grep("^PR", names(DT_for_pccc[["9"]]), value = TRUE),
      icdv = 9
    ),
    pccc::ccc(
      data = DT_for_pccc[["10"]],
      id = code_id,
      dx_cols = grep("^DX", names(DT_for_pccc[["10"]]), value = TRUE),
      pc_cols = grep("^PR", names(DT_for_pccc[["10"]]), value = TRUE),
      icdv = 10
    )
  )

delta <- merge(mdcr, pccc, all = TRUE, by = "code_id")

delta <-
  delta[
    any_tech_dep != tech_dep | any_transplant != transplant,
    .(code_id, any_tech_dep, any_transplant, tech_dep, transplant)
  ]

delta[, .N, keyby = .(any_tech_dep, tech_dep, any_transplant, transplant)]

data.table::setkey(delta, code_id)

print(delta, nrow = Inf)
```

These are known and reported issues:

* ICD-10-CM Z94x
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/56
  * Specifically Z94.1, Z94.2, Z94.4, Z94.81, Z94.82, Z94.83, Z94.84 All
    documented as subcondition transplant but are missing from the transplant
    set in the pccc_1.0.7/src/pccc.cpp

* ICD-9-CM 3491
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/45
  * Documented and implemented for neuromusc in `pccc::ccc()` tech_dep is
    missing

* ICD-9-CM V420
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/54
  * Documented as renal (transplant); missing from the transplant set in the
    pccc_1.0.7/src/pccc.cpp

* ICD-9-CM V4585
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/49
  * Documented for version 2 as metabolic (devices) but is only listed under
    metabolic in the software.  Because of how `medicalcoder::comorbidities()`
    is implemented the subconditon (devices) had to be provided in the tables
    for `method = "pccc_v2.0"`.  This results in the (improved) mapping to
    technology dependence.  Perhaps more concerning is that the mapping for
    V45.85 in `pccc::ccc()` maps to a transplant, which is incorrect.  GitHub
    issue and pull request have been submitted to resolve this issue.

* ICD-9-CM V5391
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/51
  * In `pccc::ccc()` this codes maps to transplant when it should map to
    technology dependence.

* ICD-9-CM V56.x
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/46
  * Same as ICD-9-CM 349.1

* ICD-9-CM V6546
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/52
  * Missing tech dependence flag from `pccc::ccc()`

* ICD-9-PCS 8606
  * GitHub Issue: https://github.com/CUD2V/pccc/issues/48
  * Missing tech dependence flag from `pccc::ccc()`

We remove the transplant and technology dependence columns from the
`mdcr_v_pccc_current` and `mdcr_v_pccc_cumulative` objects.
```{r}
#| label: remove-tech-trans-from-mdcr-v-pccc
mdcr_v_pccc_current[, any_transplant := NULL]
mdcr_v_pccc_current[, any_tech_dep := NULL]
mdcr_v_pccc_current[, transplant := NULL]
mdcr_v_pccc_current[, tech_dep := NULL]
mdcr_v_pccc_cumulative[, any_transplant := NULL]
mdcr_v_pccc_cumulative[, any_tech_dep := NULL]
mdcr_v_pccc_cumulative[, transplant := NULL]
mdcr_v_pccc_cumulative[, tech_dep := NULL]
```

The only remaining columns are the id columns and two columns from
`medicalcoder::comorbidities()`: `misc`, indicating a technology dependence or
transplant not classified as one of the other comorbidities, and `num_cmrb`, the
number of comorbidities flagged.

```{r}
#| label: whats-left-mdcr-v-pccc
str(mdcr_v_pccc_current)
str(mdcr_v_pccc_cumulative)
```

## Current vs Cumulative Flagging

```{r}
#| label: build-list-of-objects
#| include: false
objs <-
  c(
    "mimiciv_charlson_current",
    "comorbidity_charlson_current",
    "comorbidity_charlson_cumulative",
    "comorbidity_elixhauser_current",
    "comorbidity_elixhauser_cumulative",
    "medicalcoder_charlson_current",
    "medicalcoder_charlson_cumulative",
    "medicalcoder_elixhauser_current",
    "medicalcoder_elixhauser_cumulative",
    "medicalcoder_pcccv2.0_current",
    "medicalcoder_pcccv2.0_cumulative",
    "medicalcoder_pcccv3.1_current",
    "medicalcoder_pcccv3.1_cumulative",
    "pccc_pcccv2.0_current",
    "pccc_pcccv2.0_cumulative"
  ) |>
  sapply(get)

saveRDS(object = objs, file = "objs.rds")
```

Let's look at the impact of longitudinal flagging vs encounter level flagging of
Charlson comorbidities.

```{r}
#| label: build-flagmethod-delta
#| include: false
flagmethod_delta <-
  data.table::rbindlist(
    objs[grepl("^medicalcoder", names(objs))],
    use.names = TRUE,
    fill = TRUE,
    idcol = "obj"
  )

flagmethod_delta[
  ,
  c("algo", "flag.method") := data.table::tstrsplit(obj, "_", keep = 2:3)
]

flagmethod_delta[, obj := NULL]

flagmethod_delta <-
  data.table::melt(
    data = flagmethod_delta,
    id.vars = c("subject_id", "enc_seq", "algo", "flag.method"),
    variable.factor = FALSE,
    variable.name = "condition",
    na.rm = TRUE
  )

# subset to the conditions and add labels
#flagmethod_delta <-
#  merge(
#    x = flagmethod_delta,
#    y = medicalcoder::get_charlson_index_scores()[c("condition", "condition_description")],
#    by = "condition"
#  )

flagmethod_delta <-
  flagmethod_delta[
   , .(N = sum(value), p = mean(value)),
   by = .(algo, flag.method, condition)
 ]

flagmethod_delta[,
  flag.method := factor(flag.method, levels = c("current", "cumulative"))
]

flagmethod_delta <-
  flagmethod_delta[
    !(condition %in% c("readmission_index", "mortality_index",
        "cci", "num_cmrb", "age_score"))
  ]

flagmethod_delta[condition == "cmrb_flag"]
```

@fig-flagmethod-delta-charlson shows the number of encounters flagged with a
given condition for `flag.method = 'current'` and `flag.method = 'cumulative'`
for the Charlson condition flagged by `medicalcoder::comorbidities()`.

```{r}
#| label: define-flagmethod-plot
#| include: false
flagmethod_plot <- function(data) {
  ggplot2::ggplot(data = data) +
    ggplot2::theme_bw() +
    ggplot2::aes(x = condition, y = N, fill = flag.method) +
    ggplot2::geom_col(position = ggplot2::position_dodge()) +
    ggplot2::scale_fill_manual(
      name = "Flag Method",
      values = c("current" = "#6F263D", "cumulative" = "#4F8FCB")
    ) +
    ggplot2::scale_y_continuous(
      name = "Encounters",
      label = scales::label_comma()
    ) +
    ggplot2::theme(
      axis.title.y = ggplot2::element_blank(),
      legend.position = "bottom",
    ) +
    ggplot2::coord_flip()
}
```
```{r}
#| label: fig-flagmethod-delta-charlson
#| fig-scap: Charlson comorbidities by flag.method.
#| fig-cap: |
#|   Number of encounters flagged with a Charlson comorbidities based on
#|   encounter only (`flag.method = "current"`) verses the whole patient history
#|   (`flag.method = "cumulative"`).
#| fig-width: 7
#| fig-height: 7
#| echo: false
flagmethod_plot(flagmethod_delta[algo == "charlson"])
```

Diabetes is a great example for the benefits of `flag.method = "cumulative"`.
As we account for a patient history, the number of encounters flagged as having
uncomplicated diabetes decreases and the number of encounters flagged as having
complicated diabetes increases.

```{r}
#| label: build-dm
#| include: false
dm <-
  merge(
    x = medicalcoder_charlson_current,
    y = medicalcoder_charlson_cumulative,
    by = c("subject_id", "enc_seq"),
    suffix = c("_current", "_cumulative")
  )
# find a record with a change from uncomplicated to complicated dm

# sanity checks
dm[dm_current  == 1, all(dmc_current == 0)]
dm[dmc_current == 1, all(dm_current == 0)]
```

```{r}
#| label: find-dm-patient-example
#| include: false
dm[,
  .(
    V1 = mean(dm_current + dmc_current),
    V2 = mean(dm_cumulative + dmc_cumulative),
    .N
   ),
 by = .(subject_id)
 ][, d := V2 - V1][d > 0 & N > 5]

s10009326 <-
  dm[
    subject_id == 10009326,
    .SD,
    .SDcols = patterns("enc_seq|^dm")
  ]
```

An example of a patient record where on an encounter level flagging approach
diabetes will be missed entirely and the severity is inaccurate is in
@tbl-dm-10009326.  Subject 10009326 had `r nrow(s10009326)` encounters.  At
least one ICD code mapping to uncomplicated diabetes under the Charlson _a la_
@quan2005 appeared in `r sum(s10009326$dm_current)` encounters, and
`r sum(s10009326$dmc_current)` encounter for complicated diabetes.  Under an
encounter-level only flagging of comorbidities, this patient would erroneously
not have diabetes of any kind on
`r nrow(s10009326) - sum(s10009326$dm_current) - sum(s10009326$dmc_current)`
encounters.  Using the longitudinal flagging in `medicalcoder::comorbidities()`
via `flag.method = 'cumulative'`, we have diabetes flagged on all encounters,
and the severity correctly marked for encounters later in the patient record.

```{r}
#| label: tbl-dm-10009326
#| tbl-cap: |
#|   Flagging Diabetes under Charlson comorbidities for MIMIC-IV subject_id
#|   10009326 using `medicalcoder::comorbidities()` `flag.method = "current"`
#|   and `flag.method = "cumulative"`.
#| results: "asis"
#| echo: false
s10009326[
  ,
  lapply(.SD, function(x) data.table::fifelse(x == 1, "X", "")),
  .SDcols = patterns("^dm"),
  by = .(enc_seq)
  ] |>
kableExtra::kbl(
  x = _,
  format = "latex",
  booktabs = TRUE,
  align = "ccccc",
  row.names = FALSE,
  col.names = c("Encounter", "Complicated", "Uncomplicated", "Complicated", "Uncomplicated")
  ) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down", "HOLD_position")
) |>
kableExtra::add_header_above(
  header = c("", "Current" = 2, "Cumulative" = 2)
)
```

For Elixhauser, @fig-flagmethod-delta-elixhauser, shows similar behavior for
diabetes as Charlson.
```{r}
#| label: fig-flagmethod-delta-elixhauser
#| fig-scap: Elixhauser comorbidities by flag.method.
#| fig-cap: |
#|    Number of encounters flagged with a Elixhauser comorbidities based on
#|    encounter only (`flag.method = current`) verses the whole patient history
#|    (`flag.method = "cumulative"`).
#| fig-width: 7
#| fig-height: 7
#| echo: false
flagmethod_plot(flagmethod_delta[algo == "elixhauser"])
```

@fig-flagmethod-delta-pcccv2.0 and @fig-flagmethod-delta-pcccv3.1 illustrate
the impact of the flagging method when identifying PCCC comorbidities under
PCCC v2 and PCCC v3 respectively.  @fig-flagmethod-delta-pcccv3.1 is of
particular interest as the implications of technology-dependent codes in PCCC v3
are handled have considerable impact on the flagged comorbidities.

```{r}
#| label: fig-flagmethod-delta-pcccv2.0
#| fig-scap: PCCC v2 comorbidities by flag.method.
#| fig-cap: |
#|    Number of encounters flagged with a PCCC v2 comorbidities based on
#|    encounter only (`flag.method = current`) verses the whole patient history
#|    (`flag.method = "cumulative"`).
#| fig-width: 7
#| fig-height: 7
#| echo: false
flagmethod_plot(flagmethod_delta[algo == "pcccv2.0"])
```

```{r}
#| label: fig-flagmethod-delta-pcccv3.1
#| fig-scap: PCCC v3 comorbidities by flag.method.
#| fig-cap: |
#|    Number of encounters flagged with a PCCC v3 comorbidities based on
#|    encounter only (`flag.method = current`) verses the whole patient history
#|    (`flag.method = "cumulative"`).
#| fig-width: 7
#| fig-height: 9
#| echo: false
flagmethod_plot(flagmethod_delta[algo == "pcccv3.1"])
```

```{r}
#| label: respository-tech
#| eval: false
#| include: false
resp <-
  merge(
    x = medicalcoder_pcccv3.1_current[, .SD, .SDcols = patterns("subject_id|enc_seq|^resp")],
    y = medicalcoder_pcccv3.1_cumulative[, .SD, .SDcols = patterns("subject_id|enc_seq|^resp")],
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_current", "_cumulative")
 )

resp[respiratory_tech_only_current < respiratory_tech_only_cumulative]

# a missed flag on enc 3 of 4
resp[subject_id == 10027381]

# i want to find a tech only cumulative flag as the first flag in the record

resp[, cs_current    := cumsum(respiratory_tech_only_current),    by = .(subject_id)]
resp[, cs_cumulative := cumsum(respiratory_tech_only_cumulative), by = .(subject_id)]

resp[cs_current < cs_cumulative & cs_current == 0 & cs_cumulative > 1, unique(subject_id)]  |> cat(sep = "\n")

resp[subject_id == 10728333]

Filter(f = function(x) sum(x) > 0, s19855614)
```


```{r}
#| include: false
s19855614 <-
  merge(
    x = medicalcoder_pcccv3.1_current[subject_id == 19855614],
    y = medicalcoder_pcccv3.1_cumulative[subject_id == 19855614],
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_current", "_cumulative")
 )
Filter(f = function(x) sum(x) > 0, s19855614)
# need gi, malignancy, metabolic, neuromusc, respriatory,

s10728333 <-
  merge(
    x = medicalcoder_pcccv3.1_current[subject_id == 10728333],
    y = medicalcoder_pcccv3.1_cumulative[subject_id == 10728333],
    all = TRUE,
    by = c("subject_id", "enc_seq"),
    suffixes = c("_current", "_cumulative")
 )
Filter(f = function(x) sum(x) > 0, s10728333)
# metabolic, respriatory, any_tech, num_cmrb, misc,
```

Let's look at the records for `subject_id = 10728333`.  \ref{tbl:s10728333} shows the
flagging of metabolic, miscellaneous, and respiratory conditions for PCCC v3
using `flag.method = 'current'` or `flag.method = 'cumulative'`.  For this
patient, on encounter 1, a technology-dependent ICD code related to a
respiratory comorbidity was observed in the patient record.  No other ICD codes
related to a PCCC v3 comorbidity were observed for encounter 1.  As such, no
comorbidities are flagged for encounter 1 as at least one
non-technology-dependent ICD code is required for any comorbidity to be flagged.
Encounter 2 had no PCCC v3 related ICD codes reported.  Encounter 3 had a
technology-dependent ICD code for a miscellaneous comorbidity reported and
again, because no non-technology-dependent ICD was reported no comorbidity is
flagged.  Starting on encounter 4, a non-technology-dependent ICD code for a
metabolic comorbidity was reported.  Under the `flag.method = 'current'` that is
the only code that is reported and thus only one comorbidity is reported.  Under
`flag.method = 'cumulative'` the technology-dependent ICD codes from encounters
1 and 3 have been carried forward in the record to encounter 4 and are thus
flagged.  If the end user were to try to build the cumulative comorbidities from
the results of the `flag.method = 'current'` results, the end user would never
capture the miscellaneous comorbidity, and would fail to capture the respiratory
comorbidity until encounter 6, and would incorrectly classify as
non-technology-dependent until the final encounter.

```{r}
#| label: s10728333
#| echo: false
#| results: "asis"
# NOTE: not using the quarto yaml for the table crossref so I can use short
# captions
DT0 <- s10728333[, .SD, .SDcols = patterns("enc_seq|^(met|resp|misc|num_|any_tech)")]
DT1 <-
  DT0[
    ,
    .(enc_seq,
      metabolic__current = data.table::fcase(
        metabolic_dxpr_only_current == 1, "DxPr",
        metabolic_tech_only_current == 1, "Tech",
        metabolic_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      metabolic__cumulative = data.table::fcase(
        metabolic_dxpr_only_cumulative == 1, "DxPr",
        metabolic_tech_only_cumulative == 1, "Tech",
        metabolic_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      misc__current = data.table::fcase(
        misc_dxpr_only_current == 1, "DxPr",
        misc_tech_only_current == 1, "Tech",
        misc_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      misc__cumulative = data.table::fcase(
        misc_dxpr_only_cumulative == 1, "DxPr",
        misc_tech_only_cumulative == 1, "Tech",
        misc_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      respiratory__current = data.table::fcase(
        respiratory_dxpr_only_current == 1, "DxPr",
        respiratory_tech_only_current == 1, "Tech",
        respiratory_dxpr_and_tech_current == 1, "DxPrTech",
        default = ""),
      respiratory__cumulative = data.table::fcase(
        respiratory_dxpr_only_cumulative == 1, "DxPr",
        respiratory_tech_only_cumulative == 1, "Tech",
        respiratory_dxpr_and_tech_cumulative == 1, "DxPrTech",
        default = ""),
      any_tech_dep_current = data.table::fifelse(any_tech_dep_current == 1, "1", ""),
      any_tech_dep_cumulative = data.table::fifelse(any_tech_dep_cumulative == 1, "1", ""),
      num_cmrb_current = as.character(num_cmrb_current),
      num_cmrb_cumulative = as.character(num_cmrb_cumulative)
    )
  ]

ftm1 <- "*"#kableExtra::footnote_marker_symbol(1, format = "latex")
ftm2 <- "\\dag"#kableExtra::footnote_marker_symbol(2, format = "latex")
DT1[enc_seq == 1, `:=`(respiratory__current = ftm1, respiratory__cumulative = ftm1)]
DT1[enc_seq %in% 2:3,`:=`(respiratory__cumulative = ftm2)]
DT1[enc_seq == 3, `:=`(misc__current = ftm1, misc__cumulative = ftm1)]

kableExtra::kbl(
  x = DT1,
  col.names = c("Encounter", rep(c("Current", "Cumulative"), 5)),
  format = "latex",
  booktabs = TRUE,
  row.names = FALSE,
  align = rep("c", ncol(DT)),
  escape = FALSE,
  caption.short = "Flagging of comorbidities under PCCC v3 for MIMIC-IV subject 10728333.",
  caption = "Flagging of comorbidities under PCCC v3 for MIMIC-IV subject 10728333. Cells marked with DxPr denote that the comorbidity was flagged due to a ICD code that is not technology-dependent.  Tech denotes flagging of a comorbidity base on the presence of at least one technology-dependent ICD code and the presence of at least one non-technology-dependent code for another comorbidity. DxPrTech denotes flagging due to both technology-dependent and non-technology-dependent ICD codes.  Under the Any Tech columns we mark the encounters where some tech dependence is identified. For this patient, on encounter 1, a technology-dependent ICD code for a respiratory comorbidity was found in the subject's record.  Because no non-technology-dependent code was observed on encounter 1, under PCCC v3 this subject has no comorbidities for encounter 1.  This persists for encounters 2 and 3 with respect to respiratory and on encounter 3 for miscellaneous as well.  On encounter 4 a non-technology-dependent code for a metabolic comorbidity was reported.  Under a cumulative flagging paradigm the miscellaneous and respiratory comorbidities are also flagged as the reported codes from encounters 3 and 1 have been carried forward respectively.  Under the 'current' flagging method miscellaneous is never flagged as the technology-dependent code never occurs on the same encounter as the non-technology-dependent code."
  ) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down", "HOLD_position")
) |>
kableExtra::add_header_above(
  header = c("", "Metabolic" = 2, "Miscellaneous" = 2, "Respiratory" = 2, "Any Tech" = 2, "Number of Comorbiditys" = 2)
) |>
kableExtra::add_footnote(
  label = c("A technology-dependent code was reported on this encounter.  Since no non-technology-dependent was reported no comorbidities are flagged.",
            "A technology-dependent code has been carried forward in the record.  Since no non-technology-dependent was reported on this, or a prior, enconter, no comorbidities are flagged."),
  notation = "symbol",
  threeparttable = TRUE
)

```

```{r}
#| label: s10728333-ICD-codes
#| include: false
#| eval: false
merge(
  x = mimicivDT[subject_id == 10728333],
  y = subset(medicalcoder::get_pccc_codes(), pccc_v3.1 == 1),
  all = FALSE,
  by.x = c("icd_version", "dx", "icd_code"),
  by.y = c("icdv", "dx", "code")
) |>
dplyr::select(icd_version, dx, icd_code, enc_seq, condition, subcondition) |>
dplyr::distinct() |>
dplyr::arrange(enc_seq, condition, subcondition) |>
print(nrow = Inf)

subset(
  medicalcoder::get_icd_codes(with.description = TRUE),
  code %in% c("J9509", "Z9981")
)
```


## Time Required To Compute

The amount of time required to apply the several different comorbidity
algorithms to the MIMIC-IV data are shown in @tbl-time-to-compute.  To process
the `r qwraps2::frmt(nrow(mimicivDT))` total rows of data for
`r qwraps2::frmt(length(unique(mimicivDT$subject_id)))` unique subjects and a
total of `r qwraps2::frmt(nrow(unique(mimicivDT[, .(subject_id, hadm_id)])))`
unique hospital encounters, `medicalcoder::comorbidities()` is considerably
faster than all the alternatives.

```{r}
#| label: tbl-time-to-compute
#| tbl-cap: |
#|   Time, in seconds, to flag comorbidities via Charlson, Elixhauser, PCCC v2,
#|   and PCCC v3 and several different utilities in the MIMIC-IV v3.1 dataset.
#|   The flagging method, current or cumulative is reported.
#| results: asis
#| echo: false

ttc <-
  lapply(objs, attr, "tictoc") |>
  lapply(as.numeric) |>
  data.table::as.data.table() |>
  data.table::melt(measure.vars = names(objs), variable.factor = FALSE)

ttc[, c("Tool", "Algorithm", "Flag Method") := data.table::tstrsplit(variable, "_")]
ttc[, variable := NULL]
ttc[, value := sprintf("%.2f", value)]

ttc <-
  data.table::dcast(ttc, Algorithm + `Flag Method` ~ Tool, value.var = "value")
ttc[, `Flag Method` := factor(`Flag Method`, levels = c("current", "cumulative"))]

data.table::setcolorder(
  ttc,
  c("Algorithm", "Flag Method", "medicalcoder", "comorbidity", "mimiciv", "pccc")
)

data.table::setnames(
  ttc,
  old = c("medicalcoder", "comorbidity", "mimiciv", "pccc"),
  new = c("medicalcoder::comorbidities()", "comorbidity::comorbidity()", "MIMIC-IV Code", "pccc::ccc()")
  )

data.table::setkey(ttc, Algorithm, `Flag Method`)

kableExtra::kbl(
  x = ttc[, 2:6],
  format = "latex",
  booktabs = TRUE,
  row.names = FALSE,
  col.names = c("", names(ttc)[3:6]),
  align = "lrrrr"
  ) |>
kableExtra::kable_styling(
  latex_options = c("striped", "scale_down", "HOLD_position")
) |>
kableExtra::pack_rows(
  index = table(ttc$Algorithm)
)
```

# Session Info

@tbl-session-info-platform and @tbl-session-info-packages reports details on the
machine, R, and the R packages used to produce this document.

```{r}
#| label: mem-total-helper
#| include: false
mem_total_bytes <- function() {
  sys <- Sys.info()[["sysname"]]
  if (is.na(sys)) sys <- .Platform$OS.type

  if (identical(sys, "Windows")) {
    # memory.size() returns MB
    return(utils::memory.size(max = TRUE) * 1024^2)
  }

  if (identical(sys, "Darwin")) {
    out <- suppressWarnings(system("sysctl -n hw.memsize", intern = TRUE))
    if (length(out) == 1 && nchar(out)) return(as.numeric(out))
  }

  if (identical(sys, "Linux")) {
    if (file.exists("/proc/meminfo")) {
      lines <- readLines("/proc/meminfo", warn = FALSE)
      mem <- sub("^MemTotal:\\s+([0-9]+)\\s+kB.*$", "\\1",
                 lines[grepl("^MemTotal:", lines)])
      if (length(mem) == 1 && nchar(mem)) return(as.numeric(mem) * 1024)
    }
  }

  NA_real_
}

format_gb <- function(bytes) {
  if (is.na(bytes)) return(NA_character_)
  sprintf("%.1f GB", bytes / 1024^3)
}

mem_total_gb <- format_gb(mem_total_bytes())
```

```{r}
#| label: tbl-session-info-platform
#| tbl-cap: "Details on the machine to generate this document."
#| echo: false
#| results: "asis"
si <- sessioninfo::session_info()
si$platform$RAM <- mem_total_gb

si$platform <-
  data.table::data.table(
    V1 = names(si$platform),
    V2 = do.call(c, si$platform)
  )

si$platform <-
  si$platform[
    V1 %in% c("version", "os", "system", "RAM", "date", "pandoc", "quarto")
  ][c(1:3, 7, 4:6)]

kableExtra::kbl(
  x = si$platform,
  format = "latex",
  row.names = FALSE,
  col.names = c("Setting", "Value"),
  booktabs = TRUE
) |>
kableExtra::kable_styling(
  latex_options = c("striped", "HOLD_position")
)
```

```{r}
#| label: tbl-session-info-packages
#| tbl-cap: "R packages and versions used to generate this document."
#| echo: false
#| results: asis
pkgs <- as.data.frame(si$packages)[c("package", "loadedversion")]
blks <- 4
n <- ceiling(nrow(pkgs) / blks)
mat <- matrix("", nrow = n, ncol = 2 * blks)
for (i in seq_len(nrow(pkgs))) {
  rw  <- (i - 1) %% n + 1
  blk <- (i - 1) %/% n
  cl  <-  blk * 2 + 1
  mat[rw, cl:(cl+1)] <- unlist(pkgs[i, ])
}

kableExtra::kbl(
  x = mat,
  col.names = NA, #rep("", 8),
  format = "latex",
  booktabs = TRUE
) |>
kableExtra::kable_styling(
  latex_options = c("striped", "HOLD_position")
) |>
kableExtra::add_header_above(rep(c("Package Version" = 2), 4))
```

# References
