---
title: "Supplementary Materials"
subtitle: "medicalcoder: An R Package for Working with ICD Codes and Comorbidity Algorithms with Considerations for Longitudinal Data"

code-fold: false
echo: true
number-sections: false

format:
  pdf:
    toc: true

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
stopifnot(packageVersion("medicalcoder") >= "0.7.0")
```

# Example Data in medicalcoder

There are two example data sets provided in the medicalcoder R package, `mdcr`
and `mdcr_longitudinal`.

## mdcr

```{r}
#| label: mdcr
#| collapse: true
data(mdcr, package = "medicalcoder")
str(mdcr)
```

`mdcr` is a `r class(mdcr)` with `r nrow(mdcr)` rows and `r ncol(mdcr)` columns.
The data is a synthetic set of ICD-9 and ICD-10 codes for
`r length(unique(mdcr$patid))` patients.  Each row of the data set represents
one ICD-code in the patient record.

The columns:

* `patid`: patient identifier, integer values
* `icdv`: ICD version; integer values, 9 or 10
* `dx`: indicator column for ICD diagnostic (1) or procedure (0) codes
* `code`: compact ICD code; character values

## mdcr_longitudinal

```{r}
#| label: mdcr_longitudinal
#| collapse: true
data(mdcr_longitudinal, package = "medicalcoder")
str(mdcr)
```

`mdcr_longitudinal` is a `r class(mdcr_longitudinal)` with `r
nrow(mdcr_longitudinal)` rows and `r ncol(mdcr_longitudinal)` columns.  The data
is a synthetic set of ICD codes collected on specific dates for
`r length(unique(mdcr_longitudinal$patid))` patients.  The ICD provided are
expected to be treated as diagnostic codes but the end user is warned that some
of the codes can match to procedure codes as well.

The columns:
* `patid`: patient identifier, integer values
* `date`: date the diagnostic code was recorded
* `icdv`: ICD version 9 or 10, integer valued
* `code`: ICD codes; character values

# Basic Use of medicalcoder

## ICD Utilities

## Comorbidities

# medicalcoder versus other packages

## vs comorbidity

The R package comorbidity [@r-comorbidity]

## vs icdcomorbid

## vs multimorbidity

# SessionInfo

```{r}
#| label: sessioninfo
sessionInfo()
```

# References
